---
title: "N-mixture abundance with tulpaObs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{N-mixture abundance with tulpaObs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## From presence to counts

Occupancy models ask one question per site: is the species there or not.
A repeated-count survey asks a harder one. A surveyor visits a site five
times and records not whether a marbled salamander was seen, but how
many: 3, then 1, then 4, then 0, then 2. Those numbers carry more than
presence. They carry the size of the local population, blurred by the
fact that nobody counts every individual on every visit.

Royle's (2004) N-mixture model is the tool for that data. It keeps the
two-level structure occupancy taught us, a true state we cannot see and a
noisy observation we can, but the latent state is now a count rather than
a coin flip. Each site holds some unknown number of animals $N_i$, fixed
across the short survey window. On each visit the surveyor detects each
animal independently with probability $p_{ij}$, so the recorded count
$y_{ij}$ is a thinned version of $N_i$. Five visits to a site with eight
animals and detection 0.4 might read 4, 2, 3, 4, 1. The counts never reach
eight, and the model's job is to see eight through the thinning.

This buys you something occupancy cannot give. Where occupancy estimates
the probability a site is used, N-mixture estimates how many individuals
use it, and how that number responds to a covariate, all from counts that
systematically undercount. The price is the same identifiability tension
occupancy has, sharpened: abundance and detection both push the counts
down or up, and the data have to separate them. A site that averages two
animals per visit is consistent with twenty animals seen one in ten times,
or with three animals seen two in three times. What breaks the tie is the
spread of the counts across visits. Twenty animals at detection 0.1 produce
counts that jump around a lot relative to their mean; three animals at
detection 0.67 produce counts that barely move. The model reads that spread
and infers the abundance behind it. More visits sharpen the read, and
counts that vary across them are what carry the information at all.

The model leans on one assumption worth naming up front: the population is
closed across the survey window. The $N_i$ animals at a site stay put and
stay the same animals from the first visit to the last, so every count is a
thinning of the same latent number. A survey spread over a season that lets
animals move in and out, breed, or die violates this, and the model will
read the movement as detection noise. The closure window is usually a few
days to a few weeks, short enough that the standing population is fixed but
long enough to gather replicate counts. Designing the survey to respect
that window matters as much as any modelling choice that follows.

This vignette builds an N-mixture fit from a simulation with known truth,
reads the estimates back, handles the count-ceiling parameter, switches to
a negative-binomial state for overdispersed counts, predicts abundance at
new covariate values, adds an areal spatial field, and closes with the
rules of thumb that decide when the model is the right one.

## The model

Write the two levels for site $i$ and visit $j$. The latent abundance is
Poisson with a log-linear mean, and the counts are binomial thinnings of
that abundance with a logit-linear detection:

$$
\begin{aligned}
N_i        &\sim \mathrm{Poisson}(\lambda_i), &
  \log \lambda_i &= x_i^\top \beta_\lambda, \\
y_{ij} \mid N_i &\sim \mathrm{Binomial}(N_i, p_{ij}), &
  \mathrm{logit}\, p_{ij} &= w_{ij}^\top \beta_p.
\end{aligned}
$$

Here $\lambda_i$ is the expected abundance at site $i$, driven by site
covariates $x_i$ through a log link, and $p_{ij}$ is the per-individual
detection probability on visit $j$, driven by detection covariates
$w_{ij}$ through a logit link. The abundance arm uses a log link because
$\lambda$ is a rate that must stay positive; that detail comes back the
first time you read a coefficient.

The latent $N_i$ never appears in the data. What occupancy did with a
single sum over $\{0, 1\}$, N-mixture does with a sum over every possible
abundance:

$$
p(y_{i\cdot} \mid \lambda_i, p_{i\cdot}) \;=\;
  \sum_{N = \max_j y_{ij}}^{K} \mathrm{Poisson}(N \mid \lambda_i)
  \prod_j \mathrm{Binomial}(y_{ij} \mid N, p_{ij}).
$$

The sum runs from the largest count seen at the site (you cannot have
fewer animals than you counted on your best visit) up to a ceiling $K$.
This marginal has a closed form, so tulpa evaluates it directly with
analytical gradients and reads the curvature at the mode for standard
errors. There is no EM loop and no data augmentation over $N$: the latent
count integrates out exactly, term by term, which is why the fit returns
in a fraction of a second. The ceiling $K$ is the one approximation, and
the next sections treat it as such.

The exact marginalisation is what separates this from occupancy mechanically.
An occupancy model sums over two states and the arithmetic is trivial. An
N-mixture model sums over a few hundred, and a naive implementation would
recompute the binomial product for every candidate $N$ at every site on
every gradient step. tulpa caches the parts of each term that do not depend
on the coefficients, so the repeated evaluations across the optimisation
reuse work rather than redo it. The practical effect is that a fit on a few
hundred sites with a ceiling in the low hundreds is a sub-second operation,
not a sampler you wait on. The same closed form also means the standard
errors are the genuine observed-information curvature of the marginal, not a
Monte-Carlo approximation, so the credible intervals are deterministic and
repeatable across runs.

One consequence of the log link deserves a flag before the first fit. A
coefficient on the abundance arm is a multiplicative effect on expected
abundance, not an additive one. A slope of 0.5 means a one-unit increase in
the covariate multiplies $\lambda$ by $e^{0.5} \approx 1.65$, a 65 percent
rise, and the same slope adds more animals where abundance is already high
than where it is low. The intercept is the log of baseline abundance, so it
reads as a number of animals only after `exp()`. This is the same arithmetic
a Poisson GLM uses, and the coefficients carry the same interpretation; the
only twist is that the counts feeding them are thinned, and the model
corrects for the thinning before it reports the slope.

## Simulating and fitting a Poisson model

`simulate_abun()` draws a dataset with the truth you set, so the fit has a
target to recover. Abundance depends on one site covariate with intercept
$\log 3$ (so baseline expected abundance is 3 animals) and slope 0.5;
detection depends on one covariate with intercept 0.4 (detection near
0.6) and slope 0.3. Five visits per site give the thinning enough
replication to separate the two arms.

```{r sim}
sa <- simulate_abun(
  N = 200, J = 5,
  n_abund_covs = 1, n_det_covs = 1,
  beta_lambda = c(log(3), 0.5),   # log link: intercept is log(baseline lambda)
  beta_p      = c(0.4, 0.3),      # logit link on detection
  seed = 2
)
dat <- sa$data       # site covariates, one row per site
y   <- sa$y          # 200 x 5 count matrix
y[1:4, ]
```

The counts undercount the latent abundance. Averaging the per-site maximum
count against the simulated truth shows the gap that detection opens.

```{r naive}
c(mean_max_count = mean(apply(y, 1, max)),
  mean_true_N    = mean(sa$truth$N))
```

The fit follows the same call shape as occupancy. The abundance formula
is the first argument, detection is its own formula, and `abun()` selects
the N-mixture family. `method = "laplace"` is the default and the only
backend wired for counts; it finds the mode of the marginal likelihood and
the curvature there.

```{r fit}
fa <- tobs(
  ~ abund_cov1,
  data      = dat,
  family    = abun(),
  detection = ~ det_cov1,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)   # the fitter prints iterations otherwise
)
```

`coef()` returns the estimates as a list with one element per arm.
Read the `lambda` arm with the log link in mind: the intercept is on the
log scale, so back-transforming with `exp()` gives baseline expected
abundance.

```{r coef}
coef(fa)
```

```{r coef-exp}
c(baseline_lambda = exp(coef(fa)$lambda[["(Intercept)"]]),
  true_baseline   = 3)
```

The abundance intercept back-transforms to roughly the three animals the
data were built with, and the slope on `abund_cov1` lands near the
simulated 0.5. `summary()` adds the posterior standard deviation and the
credible bounds for every coefficient; the `lambda_` rows are the
abundance arm and the `p_` rows the detection arm.

```{r summary}
summary(fa)
```

`fitted()` returns the per-site quantities the fit implies: expected
abundance `lambda`, detection probability `p`, and the posterior mean
abundance `N` given the counts. The last one is the estimate that matters
for a population total, since it conditions on what was actually seen at
each site.

```{r fitted}
fv <- fitted(fa)
str(lapply(fv, head, 4))
```

Plotting the posterior mean abundance against the simulated truth shows
the model recovers the site-level signal through the thinning, not just
the population average.

```{r fitted-plot, fig.alt = "Posterior mean abundance per site against the true simulated abundance, scattered around the identity line"}
plot(sa$truth$N, fv$N, pch = 16, col = rgb(0, 0, 0, 0.35),
     xlab = "true N", ylab = "fitted E[N | y]")
abline(0, 1, col = "steelblue", lwd = 2)
```

The points track the identity line with scatter that grows at higher
abundance, which is the right behaviour: a site with more animals leaves
more room for the binomial thinning to vary the counts, so its posterior
abundance is less certain.

## The count ceiling K_max

The marginal likelihood sums over abundance from each site's maximum count
up to a ceiling. That ceiling is `K_max`, and the fit chooses a default
from the data when you do not set it. The sum has to reach high enough
that the Poisson mass above `K_max` is negligible; if `K_max` sits too
close to the true abundance, the model truncates probability that belongs
in the likelihood and biases $\lambda$ downward.

The default works for ordinary counts. You raise `K_max` when abundance is
large relative to detection, so the latent $N$ can sit far above any
single count. A site with $\lambda = 40$ and detection 0.2 might never
record more than a dozen animals on a visit, yet its abundance routinely
exceeds thirty, and the sum has to climb past that. The symptom of a
ceiling set too low is a boundary warning or an abundance estimate that
drifts when you raise it. The check is to refit with a higher `K_max` and
confirm the estimate is stable.

```{r kmax}
fa_k <- tobs(~ abund_cov1, data = dat, family = abun(),
             detection = ~ det_cov1, y = y, method = "laplace",
             control = list(verbose = FALSE), K_max = 200)
rbind(default_K = coef(fa)$lambda,
      K_max_200 = coef(fa_k)$lambda)
```

The two fits agree to several decimals, so the default ceiling was already
high enough here. On data where they disagree, take the higher-`K_max`
fit and keep raising it until the estimate settles. The cost of a generous
ceiling is compute time, since the inner sum runs over more terms, and
that cost is linear in `K_max`, so a ceiling a few times the largest
plausible abundance is a safe default rather than an expensive one.

It helps to see why a low ceiling pulls the estimate down. The likelihood
asks how probable the observed counts are under a given $\lambda$, and it
answers by summing over every abundance that could have produced them. If
the true $\lambda$ wants to place real probability on abundances above the
ceiling, cutting the sum off there throws that probability away. The
optimiser then finds a smaller $\lambda$ that fits its mass entirely below
the ceiling, and the abundance estimate comes out biased low. The bias is
invisible in any single fit; it shows only when you raise the ceiling and
watch the estimate move. That is why the stability check is the diagnostic
rather than any one number. A practical starting point is to set `K_max`
to several times the largest count you would believe at the most abundant
site, then confirm a doubling of it leaves the coefficients unchanged. The
negative-binomial state, with its heavier tail, needs more headroom than
Poisson for the same mean, so a ceiling that suffices under Poisson can be
tight under overdispersion.

## Overdispersion and the negative binomial

The Poisson state assumes the variance of abundance equals its mean. Real
populations rarely oblige. Animals cluster, habitat varies in ways no
covariate captures, and the spread of true abundance across sites runs
wider than Poisson allows. When it does, a Poisson N-mixture absorbs the
extra spread into $\lambda$ and detection, and the standard errors come
out too small.

The biology behind overdispersion is worth holding in mind, because it
tells you when to expect it. A Poisson abundance is what you get when
individuals settle independently and at the same rate everywhere a covariate
says they should. Real animals do neither. They aggregate around resources,
they respond to gradients no survey measured, and a good patch holds far
more than its covariates predict while a poor one holds far fewer. Each of
those produces sites that are more variable than their fitted mean, which is
overdispersion exactly. The clustering need not be spatial; it can be any
unmeasured driver that makes some sites richer and others poorer than the
covariates explain. Whenever you suspect such a driver, and in count data
you usually should, the negative binomial is the safer default state.

`abun(mixture = "negbin")` replaces the Poisson state with a negative
binomial, $N_i \sim \mathrm{NegBin}(\lambda_i, r)$, where the size $r$
sets the overdispersion through $\mathrm{Var}(N) = \lambda +
\lambda^2 / r$. A small $r$ means heavy clustering; as $r$ grows the
negative binomial approaches the Poisson, and the extra parameter costs
nothing. Simulate overdispersed counts with $r = 2$, so a site with mean
abundance 4 has variance $4 + 16/2 = 12$, three times the Poisson value.

```{r sim-nb}
sb <- simulate_abun(
  N = 200, J = 5, n_abund_covs = 1, n_det_covs = 1,
  mixture = "negbin", size = 2,
  beta_lambda = c(log(4), 0.4), beta_p = c(0.5, 0.2),
  seed = 5
)
```

Fit the negative-binomial model by passing the same `mixture` flag to the
family. The size $r$ is estimated jointly with the coefficients on the log
scale and reported on the object.

```{r fit-nb}
fb <- tobs(~ abund_cov1, data = sb$data, family = abun(mixture = "negbin"),
           detection = ~ det_cov1, y = sb$y, method = "laplace",
           control = list(verbose = FALSE))
coef(fb)
```

The dispersion lives in `fb$nmix_dispersion`: the natural-scale size `$r`,
its log-scale value `$log_r`, and the delta-method standard error `$r_sd`.

```{r nb-dispersion}
fb$nmix_dispersion
```

The estimated $r$ sits near the simulated 2, with a standard error wide
enough to tell you the size is the least certain quantity in the fit. That
is normal. The size is identified only through the tail of the count
distribution, and a single dataset of 200 sites pins it loosely.

Fitting a negative binomial when the data are actually Poisson is close to
free, since the size simply drifts large. Fitting a Poisson when the data
are overdispersed is the costly mistake, because the missing variance has
nowhere to go but the standard errors. A quick dispersion check tells you
which regime you are in. Compare the variance of the per-site maximum count
against its mean: a ratio well above one is the signature of
overdispersion that a Poisson state cannot represent.

```{r dispersion-check}
maxcount <- apply(sb$y, 1, max)
c(mean = mean(maxcount), var = var(maxcount),
  ratio = var(maxcount) / mean(maxcount))
```

A ratio near one says Poisson will do; a ratio of two or more says reach
for the negative binomial. This is a rough diagnostic on the observed
counts, not the latent state, so treat it as a flag rather than a test.
The cleaner comparison is to fit both and read the marginal log-likelihood,
which the next-to-last section does.

## Predicting abundance

`predict()` with a design matrix `X.0` returns the abundance arm's
prediction at new covariate values, on the response scale, with the
posterior spread propagated through the joint coefficient covariance. Pass
one column per abundance coefficient.

```{r predict-design}
X0 <- cbind(`(Intercept)` = 1, abund_cov1 = c(-1.5, 0, 1.5))
predict(fa, X.0 = X0)
```

Expected abundance climbs across the three covariate values, and the
credible bounds widen with the mean, the multiplicative spread a log link
produces. That widening is what the log link means. A fixed uncertainty on
the log scale becomes a proportional
uncertainty on the count scale, so a prediction of 8 animals carries a wider
absolute band than a prediction of 2, even when the coefficient is equally
well known. Reading the bounds on the response scale, as `predict()` returns
them, keeps that proportionality visible rather than hiding it on the log
scale where it looks constant.

The draws behind these summaries come from the joint coefficient
covariance, so the abundance and detection arms vary together rather than
independently. This matters for any quantity that combines the two, a
detection-corrected total or a ratio of abundances at two covariate values,
because the two arms are negatively correlated through the counts: a fit
that nudges detection up has to nudge abundance down to match the same
counts. Drawing them jointly propagates that correlation into the derived
quantity, where drawing them separately would overstate its uncertainty. To
trace the response across the whole covariate range rather than at three
points, pass `terms` and the predictor name; the result carries an estimate
and a band at each grid point, ready to plot.

```{r predict-terms, fig.alt = "Predicted expected abundance rising with abund_cov1, with a shaded 95 percent credible band"}
pr <- predict(fa, terms = "abund_cov1")
plot(pr$x, pr$estimate, type = "n",
     xlab = "abund_cov1", ylab = "expected abundance",
     ylim = range(pr$lower, pr$upper))
polygon(c(pr$x, rev(pr$x)), c(pr$lower, rev(pr$upper)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(pr$x, pr$estimate, lwd = 2)
```

The band is the abundance arm's uncertainty alone. The detection arm has
its own prediction surface, reached with `type = "detection"`, which is
the right call when the question is about survey efficiency rather than
population size.

```{r predict-detection}
predict(fa, X.0 = cbind(`(Intercept)` = 1, det_cov1 = c(-1, 0, 1)),
        type = "detection")
```

## Random effects on abundance

Counts arrive grouped. Sites cluster into watersheds, plots nest inside
forest stands, transects fall under the field crew that walked them. The
group carries its own baseline abundance that no measured covariate
captures, and treating that grouping as one fixed intercept per level
burns a parameter on every group and leaves the sparse ones with wild
estimates. A random intercept ties the levels together instead: each
group's abundance baseline is a draw from $N(0, \sigma^2)$, and the model
estimates the single spread $\sigma$ rather than one free intercept per
group. Groups with many sites keep their own signal; groups with few are
pulled toward the population mean, the shrinkage that is the point of the
random effect rather than a side effect of it.

`abun()` carries the `lme4` bar syntax on either arm. A term in the
abundance formula varies log-abundance by group; a term in the
`detection` formula varies the logit of detection. The single-species
N-mixture takes `(1 | g)` for a random intercept and the slope variants
`(x | g)`, `(x || g)`, and `(1 + x | g)` the same way the occupancy model
does, with one restriction: a term cannot sit on both arms of one fit, and
a random effect cannot share a fit with an areal spatial field. Either of
those needs the joint sampler. Put the bar on one arm at a time and the
deterministic path handles it.

Simulate grouped abundance with eight watersheds, each holding its own
log-abundance offset drawn from $N(0, 0.6^2)$, on top of a site covariate
`x1`. Detection is constant near 0.5, and four visits give the thinning
enough replication.

```{r sim-re}
sim_re <- function(N = 120, J = 4, ngrp = 8, sigma_b = 0.6, seed = 11) {
  set.seed(seed)
  grp <- rep(seq_len(ngrp), length.out = N)
  b   <- rnorm(ngrp, sd = sigma_b)
  dat <- data.frame(x1 = rnorm(N), g = factor(grp))
  eta <- as.numeric(model.matrix(~ x1, dat) %*% c(0.6, 0.4)) + b[grp]
  Nlat <- rpois(N, exp(eta))
  y <- matrix(NA_integer_, N, J)
  for (i in seq_len(N)) y[i, ] <- rbinom(J, Nlat[i], plogis(0))
  list(y = y, data = dat, b = b)
}
sr <- sim_re()
```

The fit reads like the others, with the bar inside the abundance formula.
`control = list(n.quad = 5)` sets the adaptive Gauss-Hermite resolution
that debiases the small-cluster variance attenuation; `n.quad = 1` is the
plain joint Laplace, and higher values recover the spread the single
Gaussian per group misses.

```{r fit-re}
fr <- tobs(
  ~ x1 + (1 | g),
  data      = sr$data,
  family    = abun(),
  detection = ~ 1,
  y         = sr$y,
  method    = "laplace",
  verbose   = FALSE,
  control   = list(n.quad = 5)
)
coef(fr)$lambda
```

The abundance intercept and slope recover near the simulated 0.6 and 0.4
with the group structure absorbed into the random effect rather than
blended into the coefficients. The variance component lives in `fr$means`
under a `sigma_g`-prefixed name, the spread across watershed baselines.

```{r re-sigma}
sig_name <- grep("^sigma_g", names(fr$means), value = TRUE)
c(estimate = fr$means[[sig_name]], truth = 0.6)
```

`vcov()` returns the joint fixed-effect covariance over the abundance and
detection coefficients alone, the same block the non-grouped fit carries.
The variance-component point estimate sits outside it, since the
marginal-Hessian path does not surface its joint posterior covariance with
the fixed effects.

```{r re-vcov}
round(vcov(fr), 4)
```

`ranef()` returns the per-group effects, the BLUPs, one row per group level
and coefficient: the grouping factor, the level, the term, the estimate,
and its standard error. The standard error is the genuine AGHQ marginal
posterior spread, wider for groups with fewer sites.

```{r re-ranef}
re <- ranef(fr)
re
```

Plot the recovered group effects against the simulated truth with the BLUP
standard error as an interval. The points track the identity line, with
the shrinkage that pulls sparse groups toward zero and the uncertainty band
that widens where a group carries less data.

```{r re-plot, fig.alt = "Recovered per-group random intercepts against the simulated truth, with standard-error bars, scattered along the identity line"}
library(ggplot2)
re$truth <- sr$b[as.integer(re$level)]
ggplot(re, aes(truth, estimate)) +
  geom_abline(slope = 1, intercept = 0, colour = "steelblue", linewidth = 1) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error),
                width = 0, colour = "grey50") +
  geom_point(size = 2) +
  labs(x = "true group effect", y = "estimated BLUP") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The same machinery moves to detection by writing the bar in the
`detection` formula, where it varies the per-group logit of detection and
the variance component is named `sigma_p` for the detection arm. Under
`mixture = "negbin"` the random effect and the overdispersion size are
estimated together on the same AGHQ likelihood, with the size as the
trailing log-scale coordinate. The recovery here used eight groups; the
variance starts to mean something around five to eight levels, and below
that a fixed effect that names each group is the honest choice.

## Counts on a map

Sites near each other often hold similar numbers of animals for reasons no
covariate records: shared microclimate, a corridor of good habitat, a
patch the survey design happened to cluster. Left unmodelled, that
structure does two kinds of damage. It inflates confidence in the slopes,
because neighbouring sites carry correlated leftovers and the model counts
each as fresh information when they are partly the same observation seen
twice. And it biases the slopes themselves whenever a covariate is
spatially patterned, since the covariate and the unmeasured regional driver
vary together and a model with nowhere to put the driver loads it onto the
covariate. An areal spatial field gives the regional signal a place to
live, one random effect per spatial unit, tied to neighbours so the surface
stays smooth. The slope is then read against a background that already
accounts for location, and its standard error widens to its honest size
because the field, not the covariate, now carries the autocorrelation.

The same logic that motivates a spatial field for a binomial or Gaussian
response carries over to counts unchanged. What differs is only where the
field attaches: it enters the log-abundance predictor, alongside the
covariate slopes, so it soaks up the regional variation in how many animals
a site holds. The detection arm is left non-spatial here, which is the usual
choice, since detection is a property of the survey rather than the place.
If survey effort is itself spatially clustered, a field on detection is
possible too, but the common case puts the field on abundance and reads it
as the smoothed map of population density the covariates do not explain.

For N-mixture the field attaches to the abundance arm through a spatial
term inside the formula, with `method = "nested_laplace"`. The areal terms
are `icar()`, `bym2()`, and `car_proper()`; each takes a dense 0/1
adjacency matrix with one row and column per site. Build a small grid and
its rook adjacency, where each cell neighbours the cells directly above,
below, left, and right.

```{r grid-adj}
grid_adj <- function(side) {
  ng <- side * side
  co <- expand.grid(x = seq_len(side), y = seq_len(side))
  adj <- matrix(0L, ng, ng)
  for (i in seq_len(ng)) for (j in seq_len(ng)) {
    if (i != j && abs(co$x[i] - co$x[j]) + abs(co$y[i] - co$y[j]) == 1L)
      adj[i, j] <- 1L
  }
  adj
}
adj <- grid_adj(6)        # 36 cells, one site per cell
dim(adj)
```

Simulate counts on the grid with a smooth abundance field added to the
log-linear mean, then thin them to visits. The field is the regional
signal the fit has to recover alongside the covariate slope.

```{r sim-spatial}
sim_spatial <- function(adj, b_lambda, b_p, J = 5, sd_phi = 0.6, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  ng <- nrow(adj)
  phi <- as.numeric(scale(rnorm(ng)))
  for (rep in 1:3) {                       # smooth toward neighbours
    pn <- phi
    for (i in seq_len(ng)) {
      nb <- which(adj[i, ] == 1L)
      pn[i] <- 0.5 * phi[i] + 0.5 * mean(phi[nb])
    }
    phi <- pn
  }
  phi <- sd_phi * as.numeric(scale(phi)); phi <- phi - mean(phi)
  x_ab <- rnorm(ng); x_det <- rnorm(ng)
  lambda <- exp(b_lambda[1] + b_lambda[2] * x_ab + phi)
  p <- plogis(b_p[1] + b_p[2] * x_det)
  N <- rpois(ng, lambda)
  y <- matrix(NA_integer_, ng, J)
  for (i in seq_len(ng)) y[i, ] <- rbinom(J, N[i], p[i])
  list(y = y, data = data.frame(abund_cov1 = x_ab, det_cov1 = x_det),
       phi = phi)
}
ss <- sim_spatial(adj, b_lambda = c(log(5), 0.5), b_p = c(0.3, 0.4), seed = 42)
```

Fit the spatial model with a `bym2()` term on the abundance formula. The
field's smoothing strength is integrated over a grid rather than fixed, so
the slope standard errors carry the field's own uncertainty. The fit
returns the same `tobs_fit` object as the non-spatial path, with
`fs$method` reading `"nested_laplace"`.

```{r fit-spatial}
fs <- suppressWarnings(tobs(
  ~ abund_cov1 + bym2(graph = adj),
  data      = ss$data,
  family    = abun(),
  detection = ~ det_cov1,
  y         = ss$y,
  method    = "nested_laplace",
  control   = list(verbose = FALSE, max.iter = 50)
))
coef(fs)
```

The covariate slopes recover with the regional signal absorbed into the
field rather than blended into the coefficient: `coef(fs)` reads back the
abundance slope near the simulated 0.5 and the detection slope near 0.4.

Read the slopes, not the abundance intercept, under an intrinsic field.
The intrinsic prior pins the field only up to an overall level, so the
intercept and the field mean trade off along a flat direction; the BYM2
field's independent component tames that confounding but the slopes are
the inferential payload regardless. The recovered field comes back on the
object as `fs$spatial_field`, one value per cell, and
`cor(fs$spatial_field, ss$phi)` is positive: the fit cannot reconstruct one
draw of a noisy field exactly, but it recovers its shape, and that shape is
what removes the bias from the slopes.

```{r spatial-field}
c(field_cor = cor(fs$spatial_field, ss$phi),
  method    = fs$method)
```

Under `mixture = "negbin"` the size $r$ is integrated over the
outer grid alongside the field's smoothing parameters and reported in
`fs$nmix_hyper$r` rather than as a coefficient, so the spatial path
handles overdispersion without adding a fixed parameter to the covariance.

## Comparing models and practical guidance

Does the negative binomial earn its extra parameter? Fit both states on
the overdispersed data and read the marginal log-likelihood through
`logLik()`. The marginal already integrates the latent abundance out, so a
higher value means the state genuinely fits the counts better, not that it
has more freedom.

```{r compare}
fb_pois <- tobs(~ abund_cov1, data = sb$data, family = abun(),
                detection = ~ det_cov1, y = sb$y, method = "laplace",
                control = list(verbose = FALSE))

c(poisson = as.numeric(logLik(fb_pois)),
  negbin  = as.numeric(logLik(fb)))
```

The negative binomial carries the higher log-likelihood on data built with
$r = 2$, which is the answer the simulation should give. On real data the
same comparison decides the state: a meaningfully higher negative-binomial
log-likelihood says the counts are overdispersed and the Poisson was
hiding it in the standard errors. A near-tie says Poisson is adequate and
the simpler state is the honest choice.

A handful of rules of thumb make the rest of the modelling decisions.

- **At least three visits, more when detection is low.** The abundance and
  detection arms separate only through the spread of counts across visits.
  With two visits and detection below about 0.3 they are weakly
  identified, and the abundance estimate can drift toward the count ceiling
  or collapse toward the mean count. Three visits is the practical floor;
  five or more is comfortable, and they buy more than extra sites do
  because replication, not site count, is what pins detection.

- **Counts must vary within a site.** A site that reads 2, 2, 2, 2 tells
  the model almost nothing about detection, since constant counts are
  consistent with high abundance and high detection or low abundance seen
  perfectly. Variation across visits is the signal the binomial thinning
  leaves, and it is what the detection arm reads. If your counts barely
  move across visits, the fit will lean on the prior and the covariates
  rather than the data.

- **Check the dispersion before trusting Poisson standard errors.** A
  variance-to-mean ratio of the per-site counts above roughly two is a
  flag for the negative binomial. Fitting Poisson on overdispersed counts
  does not bias the slopes badly, but it understates their uncertainty, so
  a coefficient can read significant when it is not. When in doubt, fit
  both and compare the marginal log-likelihood.

- **Raise K_max when abundance dwarfs detection.** The default ceiling
  suits ordinary counts. When expected abundance is large and detection is
  small, the latent $N$ sits far above any visit's count, and a ceiling set
  too low truncates the likelihood and biases $\lambda$ down. A workable
  rule is to set `K_max` to roughly three to five times the largest count
  you would believe at the richest site, then double it and confirm the
  coefficients do not move. The negative-binomial state needs the upper end
  of that range for the same mean, since its heavier tail puts mass higher.

- **Group with five to eight levels before trusting the variance.** A
  random effect on the abundance arm estimates the spread across groups, and
  with two or three groups there is almost no information about that spread.
  Five to eight levels is the rough floor; below it, name each group with a
  fixed intercept instead. Keep at least ten sites per group when the term
  carries a random slope rather than only an intercept, since a slope asks
  each group's own data to inform its covariate effect before shrinkage.

- **When N-mixture is the wrong tool.** Three cases call for a different
  model. If the surveys are not closed, animals moving in and out between
  visits so abundance is not fixed, the binomial thinning assumption breaks
  and the model reads movement as detection. If you only have
  detection or non-detection rather than counts, occupancy is the honest
  model and N-mixture adds a latent abundance the data cannot inform. And
  if detection is near one, the counts are nearly the abundance itself, so
  a plain Poisson or negative-binomial regression on the counts is simpler
  and loses nothing.

## Where to go next

- Single-season occupancy, the binary cousin of this model, with the same
  detection structure: `vignette("occupancy")`.
- Many species at once, each with its own N-mixture, tied by community
  hyperpriors: `vignette("community-models")`.
- Grouped random effects in full, with slopes, the `||` forms, and the
  AGHQ debias: `vignette("random-effects")`.
- Areal and continuous spatial fields in depth, with field extraction and
  the smoothing-precision posterior: the `tulpa` spatial vignettes.
- Posterior-predictive checks and residual diagnostics:
  `vignette("diagnostics")`.
