---
title: "Community models in tulpaObs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Community models in tulpaObs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## Why fit many species at once

A survey rarely records one species. You walk a transect, run a camera grid, or
net a pond, and you come back with detections for dozens of taxa across the same
sites and visits. The obvious move is to fit each species on its own: one
occupancy model per species, one abundance model per species, stack the results.
That works for the common species. It falls apart for the rare ones.

The rare species are where most of the data lives, in the sense that they are
the majority of the species list and the minority of the detections. A bird seen
at three sites out of two hundred carries almost no information on its own. Its
single-species fit will have a slope with a credible interval running from deeply
negative to deeply positive, an intercept pinned to the floor, and detection
probability that the optimiser cannot separate from occupancy. You report it
anyway, the interval covers everything, and the species adds nothing.

A community model changes the arithmetic. Instead of treating each species as an
isolated problem, it assumes the species coefficients are draws from a shared
distribution, a community. The mean of that distribution is estimated from every
species jointly, and a species with thin data is pulled toward the community
mean rather than left to drift. The pull is strongest where the data are weakest,
which is exactly where you want it. A common species, well informed by its own
detections, barely moves. A rare species, with nothing to anchor it, borrows the
shape of the assemblage. This is partial pooling, and for assemblage data it is
usually the difference between a species list you can interpret and a table of
noise.

There is a second reason to fit the assemblage jointly, beyond rescuing the rare
species. Many of the questions you ask of survey data are about the community, not
about any one species. How many species occupy a site. How richness changes along
a gradient. Whether the assemblage as a whole responds to disturbance. None of
these has a single-species answer you can stack up afterward, because the
uncertainty in each species' contribution has to combine correctly into the
community summary. A model that fits the species jointly carries that uncertainty
through to the community quantity. A pile of separate fits does not, and the
honest error bars on a richness map assembled from independent fits are hard to
get right.

The same idea spans both response types tulpaObs supports for assemblages.
`ms_occu()` fits community occupancy: presence-absence with imperfect detection,
pooled across species. `ms_abun()` fits community N-mixture: repeated counts with
a latent abundance per site, again pooled across species. The model structure is
identical; only the per-species likelihood differs. This vignette builds both
from simulated data with a known truth, reads the community means against that
truth, pulls out species richness with its uncertainty, and ends with the rules
of thumb for when pooling helps and when a single-species fit is the honest call.

## The model

Write the coefficient vector for species $s$ as $\beta_s$. In a single-species
fit, each $\beta_s$ is a free parameter estimated only from species $s$'s data. A
community model adds one layer above it: the species coefficients are themselves
draws from a community distribution,

$$
\beta_s \;\sim\; \mathcal{N}(\mu,\; \Sigma),
$$

where $\mu$ is the community mean (the average response across the assemblage)
and $\Sigma$ is the covariance of species around that mean. The community mean
$\mu$ is the fixed effect: it is what `coef()` returns, and it is the quantity
you interpret as "the typical response in this community". The per-species
departures $\beta_s - \mu$ are the random effects, the deviations that say how
each species differs from the assemblage average.

For occupancy, $\beta_s$ holds the occupancy-arm coefficients (intercept and
covariate slopes on the logit of $\psi$) and a parallel detection-arm vector with
its own community mean and covariance. For the N-mixture, $\beta_s$ holds the
abundance-arm coefficients on the log of $\lambda$ and the detection-arm
coefficients on the logit of $p$, each with its own community distribution. In
both cases the latent state, occupancy $z$ in one and abundance $N$ in the other,
integrates out, so the fit works directly with the marginal likelihood of the
observed detections.

The detection arm pools just as the state arm does, and this is more than a
symmetry of notation. Detection is the hardest thing to estimate per species,
because a species seen rarely gives you almost no replicate detections to learn
its detectability from. Pooling detection across the assemblage borrows the
community's detection mean for the species that cannot estimate their own, which
is what keeps a rare species' occupancy from being confounded with its
detectability. The two arms have independent community distributions, so a species
can be a detection outlier without being an occupancy outlier and the model keeps
the two departures separate.

The width of $\Sigma$ controls how hard the pooling pulls. A narrow $\Sigma$ says
species are alike, so a rare species is dragged close to $\mu$. A wide $\Sigma$
says species are heterogeneous, so each keeps more of its own estimate. Crucially,
$\Sigma$ is estimated from the data, not fixed by hand. The assemblage decides how
much it resembles itself, and the pooling strength follows from that.

It helps to see the two extremes the community structure interpolates between.
Fix $\Sigma$ at zero and every species collapses onto the community mean: this is
complete pooling, a single coefficient for the whole assemblage that ignores
species identity entirely. Let $\Sigma$ run to infinity and the prior tells each
species nothing, so each $\beta_s$ is estimated from its own data alone: this is
no pooling, the stack of independent single-species fits. The community model sits
between, and because $\Sigma$ is fitted, the data choose where between. An
assemblage of near-identical species lands near complete pooling; a heterogeneous
one lands near no pooling. You do not pick the pooling strength, you let it be
inferred, which is the difference between a defensible community fit and an
arbitrary one.

The estimation works because the latent state never has to be sampled. In a
naive formulation you would carry the occupancy indicators $z_{s,i}$ or the
abundances $N_{s,i}$ as parameters and integrate them by simulation. tulpaObs
instead sums them out in closed form, occupancy over the two states $z \in
\{0,1\}$ and abundance over a truncated count range, so the likelihood the fitter
sees is already marginal in the latent state. What remains are the per-species
coefficients and their community distribution, and those are handled by a Laplace
approximation at the mode with the community covariance updated from how the
fitted species scatter. The result returns in well under a second on an assemblage
of a few dozen species across a few hundred sites.

## Community occupancy

`simulate_ms_occu()` draws an assemblage of occupancy histories with a known
community truth. The community mean occupancy coefficients are `beta_comm_mean`
(intercept and one covariate slope `x`), and species scatter around them with
standard deviations `beta_comm_sd`. Here eight species share an occupancy
intercept of 0 and a slope of 0.5, with species-level spread of 0.5 on the
intercept and 0.3 on the slope.

```{r ms-occu-sim}
sm <- simulate_ms_occu(N = 120, J = 5, n_species = 8, seed = 3)
dim(sm$y)            # 120 sites x 5 visits x 8 species
sm$truth$beta_comm_mean   # community occupancy intercept + slope on x
sm$truth$beta_comm_sd     # species-level SD around each community mean
```

The response `y` is a three-dimensional array: sites by visits by species. Each
species occupies a slice `sm$y[, , s]`, a sites-by-visits detection matrix of the
same shape a single-species `occu()` would take. Stacking those slices is what
turns eight occupancy models into one community fit. The detection histories vary
a lot in how much they carry: the first species below is detected at many sites,
a later one at only a handful, and the community prior is what keeps the sparse
species from drifting.

```{r ms-occu-detected}
detected <- apply(sm$y, 3, function(s) mean(rowSums(s, na.rm = TRUE) > 0))
round(setNames(detected, paste0("sp", 1:8)), 2)
```

Fit with `ms_occu()`. The call shape is the single-species occupancy call with
two additions: the response is the array, and `species` names the third dimension.
The occupancy formula and the `detection` formula apply to every species; the
community structure is what lets their coefficients differ.

```{r ms-occu-fit}
fm <- tobs(
  ~ x,
  data      = sm$data,
  family    = ms_occu(),
  detection = ~ 1,
  y         = sm$y,
  species   = paste0("sp", 1:8),
  method    = "laplace",
  control   = list(verbose = FALSE)
)
```

`coef()` returns the community means, one element per process. The `psi` element
is the community occupancy mean $\mu$; the `p` element is the community detection
mean. These are the assemblage-level fixed effects.

```{r ms-occu-coef}
coef(fm)
```

Read the occupancy means against the truth. The simulated community intercept was
0 and the slope 0.5, and the fit lands close to both.

```{r ms-occu-recovery}
data.frame(
  term      = c("intercept", "slope_x"),
  estimate  = round(coef(fm)$psi, 3),
  truth     = sm$truth$beta_comm_mean,
  row.names = NULL
)
```

`summary()` adds the posterior standard deviation and credible bounds for each
community mean. The `psi_` rows are the occupancy community coefficients, the
`p_` row the detection one.

```{r ms-occu-summary}
summary(fm)
```

The credible bounds here are tight because they describe the *community mean*,
which is informed by all eight species at once. A single species would report a
far wider interval on the same slope. That narrowing is the pooling working: the
assemblage estimates its average response sharply even though no individual
species could.

The contrast is worth making concrete. Fit the sparsest species on its own with a
single-species `occu()` and compare its slope to the community mean. The
single-species interval is wide and may not even pin the sign; the community mean
is sharp. The community fit does not make the sparse species' own estimate
sharper, it supplies a sharp assemblage average against which that species'
departure is small.

```{r ms-occu-vs-single}
sparse_sp <- which.min(detected)        # the least-detected species
fit_one <- tobs(~ x, data = sm$data, family = occu(), detection = ~ 1,
                y = sm$y[, , sparse_sp], method = "laplace",
                control = list(verbose = FALSE))
rbind(
  single_species = round(confint(fit_one)["psi_x", ], 3),
  community_mean  = round(confint(fm)["psi_x", ], 3)
)
```

The single-species slope carries an interval several times wider than the
community mean's. On a real survey, that single-species interval is what you would
report species by species if you fit them separately, and most of them would say
nothing. The community mean says something because it pooled.

## Species richness

The reason to fit occupancy for a whole assemblage is often a single derived
quantity: how many species occupy each site. The naive count, species detected at
a site, understates richness because an occupied species can go undetected on
every visit. The community model fixes this by carrying each species' occupancy
probability $\psi_{s,i}$, so expected richness at site $i$ is the sum of
occupancy probabilities across species, $\sum_s \psi_{s,i}$, and the uncertainty
in those probabilities propagates into the sum.

`tobs_richness()` does this from the fitted object. It returns a per-site table
with the posterior mean richness, its standard deviation, and a 95% credible
interval.

```{r richness}
rich <- tobs_richness(fm)
head(rich)
```

The column `mean` is the expected number of species present, summed over their
occupancy probabilities rather than counted from detections. Because the sum runs
over probabilities, it is a real number, not an integer: a site with five species
each at $\psi = 0.6$ has expected richness 3.0. Plotting richness against site,
ordered, shows the spread across the survey with its uncertainty band.

```{r richness-plot, fig.alt = "Estimated species richness per site, sorted, with a shaded 95 percent credible band"}
ord <- order(rich$mean)
r   <- rich[ord, ]
plot(seq_len(nrow(r)), r$mean, type = "n",
     xlab = "site (ordered by richness)", ylab = "expected richness",
     ylim = range(r$q2.5, r$q97.5))
polygon(c(seq_len(nrow(r)), rev(seq_len(nrow(r)))),
        c(r$q2.5, rev(r$q97.5)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(seq_len(nrow(r)), r$mean, lwd = 2, col = "steelblue")
```

The band carries the occupancy uncertainty for every species into the richness
estimate, so a site whose species are all marginally present reports both a lower
richness and a wider interval than a site of confident occupants. That is the
honest picture a raw detected-species count throws away.

It is worth seeing how far the estimate departs from the naive count. The naive
richness at a site is the number of species detected there at least once; the
modelled richness adds the species that are probably present but went undetected.
The gap between them is the detection correction, and it grows with the number of
visits a survey skipped and the species it would have caught with more effort.

```{r richness-naive}
naive <- rowSums(apply(sm$y, c(1, 3), function(v) any(v > 0, na.rm = TRUE)))
data.frame(
  site        = 1:6,
  detected    = naive[1:6],
  modelled    = round(rich$mean[1:6], 2),
  correction  = round(rich$mean[1:6] - naive[1:6], 2)
)
```

The modelled richness exceeds the detected count at most sites, by the amount the
model attributes to imperfect detection. A survey that visited each site five
times misses less than one that visited twice, so the correction shrinks as
visits accumulate. Reporting richness without this correction systematically
undercounts, and the bias is worst exactly where detection is poor.

One note on what `ms_occu()` exposes. The community structure is handled inside
the fitter, so `ranef()` returns an empty table for an `ms_occu` fit; the
per-species occupancy probabilities reach you through `tobs_richness()` and
`fitted()` rather than through a coefficient-deviation table. The community
N-mixture below does expose per-species deviations through `ranef()`, because its
random-effect structure is reported directly.

## Community N-mixture

The count analogue is `ms_abun()`, the community version of Royle's N-mixture.
Each species has a latent abundance per site drawn from a Poisson with mean
$\lambda_{s,i}$, and the repeated visits are binomial detections of that
abundance. The per-species abundance coefficients are drawn from a community
Normal on the log scale, the detection coefficients from another on the logit
scale, exactly the two-arm structure of the occupancy model with counts in place
of presence.

`simulate_ms_abun()` draws this with a known community truth. The community
abundance mean `mu_lambda` is on the log scale, so an intercept of $\log 3$ means
the typical species averages three individuals per site at the covariate mean.

```{r ms-abun-sim}
sab <- simulate_ms_abun(n_species = 8, N = 80, J = 4,
                        n_abund_covs = 1, n_det_covs = 1, seed = 7)
dim(sab$y)              # 80 sites x 4 visits x 8 species
sab$truth$mu_lambda     # community abundance mean (log scale): log(3), slope 0.4
sab$truth$mu_p          # community detection mean (logit scale)
```

Fit with `ms_abun()`. The call mirrors the single-species `abun()` call, with the
species array and `species` added.

```{r ms-abun-fit}
fab <- tobs(
  ~ abund_cov1,
  data      = sab$data,
  family    = ms_abun(),
  detection = ~ det_cov1,
  y         = sab$y,
  species   = paste0("sp", 1:8),
  method    = "laplace",
  control   = list(verbose = FALSE)
)
```

`coef()` returns the community means: the `lambda` element on the log scale, the
`p` element on the logit scale. The log link on the abundance arm matters for
reading the intercept. A community abundance intercept near $\log 3 \approx 1.1$
back-transforms to about three individuals; do not read it as a count directly.

```{r ms-abun-coef}
coef(fab)
exp(coef(fab)$lambda["(Intercept)"])   # community mean abundance at covariate mean
```

Compare the community means to the truth. The abundance arm is on the log scale,
the detection arm on the logit scale.

```{r ms-abun-recovery}
data.frame(
  arm      = c("lambda", "lambda", "p", "p"),
  term     = c("intercept", "abund_cov1", "intercept", "det_cov1"),
  estimate = round(c(coef(fab)$lambda, coef(fab)$p), 3),
  truth    = round(c(sab$truth$mu_lambda, sab$truth$mu_p), 3),
  row.names = NULL
)
```

### Per-species deviations

Where the community mean is the assemblage average, the per-species deviations
say how each species departs from it. `ranef()` returns those deviations for an
`ms_abun` fit, in long form: one row per species, arm, and term. The estimate is
$\beta_s - \mu$, the species' offset from the community mean on the link scale.

```{r ms-abun-ranef}
re <- ranef(fab)
head(re, 8)
```

A positive abundance-intercept deviation marks a species more abundant than the
community average, a negative one a sparser species. Pull the abundance-intercept
deviations and order them into a caterpillar plot, the standard way to read
random effects: species sorted by their offset, each a point against the
community-mean line at zero.

```{r ms-abun-caterpillar, fig.alt = "Caterpillar plot of per-species abundance-intercept deviations, ordered, against the community mean at zero"}
dev_int <- re[re$arm == "lambda" & re$term == "(Intercept)", ]
dev_int <- dev_int[order(dev_int$estimate), ]
plot(dev_int$estimate, seq_len(nrow(dev_int)),
     yaxt = "n", pch = 19, col = "steelblue",
     xlab = "abundance-intercept deviation (log scale)", ylab = "",
     xlim = range(dev_int$estimate) + c(-0.1, 0.1))
axis(2, at = seq_len(nrow(dev_int)), labels = dev_int$species, las = 1)
abline(v = 0, lty = 2, col = "grey50")
```

The community covariances $\Sigma_\lambda$ and $\Sigma_p$ sit on the fitted object
as `ms_community$sd_lambda` and `ms_community$sd_p`, the per-coefficient community
standard deviations. They report how heterogeneous the assemblage is around its
mean, the estimated width of the community distribution that drove the pooling.

```{r ms-abun-sigma}
fab$ms_community$sd_lambda   # community SD of abundance coefficients
fab$ms_community$sd_p        # community SD of detection coefficients
```

A small community SD says the species behave alike, so the pooling pulled rare
species hard toward the mean. A large one says they are heterogeneous, so each
species kept more of its own estimate. Reading the SD alongside the deviations
tells you how much work the community prior is doing.

The per-species coefficients themselves, community mean plus deviation, give the
per-species abundance you would predict at the covariate mean. Back-transform the
abundance intercept of each species through `exp()` to read it as a count, and the
spread across species shows the assemblage from sparse to abundant.

```{r ms-abun-perspecies}
coef_lambda <- fab$ms_community$coef_lambda          # one row per species
per_species_N <- round(exp(coef_lambda[, "(Intercept)"]), 2)
sort(per_species_N)
```

These are the species-level estimates the community structure produced. A sparse
species' value is close to the community mean because the pooling supplied most of
it; an abundant species with ample counts sits at its own data. The community SD
above is what set how far each was allowed to stray.

`ms_abun()` fits a community N-mixture under `method = "laplace"` -- Poisson by
default, or negative-binomial with `mixture = "negbin"`. The NB arm gives each
species its own overdispersion through a log-normal random effect
(`log r_s ~ N(mu_log_r, sigma_log_r)`), so the dispersion is partially pooled
across the assemblage the same way the coefficients are; `fit$ms_dispersion`
returns the per-species `r_s` alongside the community `mu_log_r` and
`sigma_log_r`. For a single overdispersed species on its own,
`abun(mixture = "negbin")` covers that case directly (see
`vignette("abundance")`).

### Per-species overdispersion

Counts in the wild are rarely Poisson. A site that suits a species holds a
cluster; an unsuitable one holds none, and the variance of the counts outruns
their mean. The Poisson abundance arm has no room for that, so an overdispersed
species inflates the detection estimate to soak up the extra variance and the
abundance mean drifts. `mixture = "negbin"` gives each species its own
negative-binomial size $r_s$, with the per-species log-dispersions drawn from a
community Normal, $\log r_s \sim \mathcal{N}(\mu_{\log r}, \sigma_{\log r}^2)$.
The dispersion pools across the assemblage the same way the coefficients do: a
species with thin counts borrows the community log-dispersion, a species with
many counts keeps its own.

Simulate an assemblage with genuine overdispersion. `size = 5` sets the
community-mean size (so $\mu_{\log r} = \log 5$), and `sigma_logr = 0.5` spreads
the per-species sizes around it. Twenty species over 120 sites with five visits
gives each species enough replicate counts for its own dispersion to register.

```{r ms-nb-sim}
snb <- simulate_ms_abun(n_species = 20, N = 120, J = 5,
                        n_abund_covs = 1, n_det_covs = 1,
                        mu_lambda = c(log(5), 0.4), mu_p = c(0.3, -0.3),
                        sd_lambda = 0.4, sd_p = 0.35,
                        mixture = "negbin", size = 5, sigma_logr = 0.5,
                        seed = 14)
snb$truth$mu_log_r          # community log-dispersion: log(5)
snb$truth$sigma_log_r       # spread of per-species log r
```

Fit with `ms_abun(mixture = "negbin")`. The call is the Poisson community call
with the mixture switched. The per-species RE vector widens by one coordinate
for `log_r_s`, so the AGHQ grid grows; `control$n.quad = 3` keeps it tractable
at this species count without losing the small-cluster correction.

```{r ms-nb-fit}
fnb <- tobs(
  ~ abund_cov1,
  data      = snb$data,
  family    = ms_abun(mixture = "negbin"),
  detection = ~ det_cov1,
  y         = snb$y,
  species   = snb$species,
  method    = "laplace",
  control   = list(verbose = FALSE, n.quad = 3L)
)
fnb$mixture
```

The community log-dispersion sits on `fit$ms_dispersion`, alongside the
per-species sizes `r_s` and the community spread `sigma_log_r`. `mu_log_r` is the
assemblage-mean log size; `r = exp(mu_log_r)` is the size of the typical species.

```{r ms-nb-dispersion}
disp <- fnb$ms_dispersion
c(mu_log_r    = round(disp$mu_log_r, 3),
  sigma_log_r = round(disp$sigma_log_r, 3),
  r_community = round(disp$r, 3))
```

Read the community log-dispersion against its truth. `mu_log_r` enters the
coefficient surface as `log_r`, so its standard error is on `fit$sds`, and the
estimate lands within a couple of standard errors of $\log 5$.

```{r ms-nb-recovery}
data.frame(
  quantity  = c("mu_log_r", "sigma_log_r"),
  estimate  = round(c(disp$mu_log_r, disp$sigma_log_r), 3),
  truth     = round(c(snb$truth$mu_log_r, snb$truth$sigma_log_r), 3),
  row.names = NULL
)
```

A community fit gains more from per-species dispersion than a single-species one
does. The negative binomial carries variance $\lambda + \lambda^2 / r_s$, so a
small $r_s$ lets a species' counts spread well beyond their mean without forcing
the detection probability down to absorb the slack. Under a shared Poisson the
clumped species would each push the community detection mean low, and the rare
species, leaning on that mean, would inherit the bias. Pooling the dispersion
keeps the overdispersion where it belongs, on the abundance arm, species by
species, so the detection community mean stays clean for the species that depend
on it.

The per-species sizes are in `disp$r_s`, one per species. A small $r_s$ marks a
clumped species whose counts vary far more than their mean; a large one
approaches the Poisson limit. The deviations behind them reach `ranef()` as a
`logr` arm, the per-species $\log r_s - \mu_{\log r}$ on the same scale the
coefficient deviations use.

```{r ms-nb-ranef}
re_nb <- ranef(fnb)
head(re_nb[re_nb$arm == "logr", ], 6)
```

Plot the per-species sizes with a band for the community spread. The points are
each species' fitted $r_s$, sorted; the dashed line is the community-mean size
$\exp(\mu_{\log r})$, and the shaded band spans $\exp(\mu_{\log r} \pm
\sigma_{\log r})$, the central interval the community distribution places on a
new species.

```{r ms-nb-plot, fig.alt = "Per-species negative-binomial size r_s, sorted, against the community-mean size and its one-SD band"}
rs   <- sort(disp$r_s)
band <- exp(disp$mu_log_r + c(-1, 1) * disp$sigma_log_r)
plot(seq_along(rs), rs, pch = 19, col = "steelblue", log = "y",
     xlab = "species (ordered by size)", ylab = expression(r[s]~"(log axis)"),
     panel.first = rect(0, band[1], length(rs) + 1, band[2],
                        col = adjustcolor("steelblue", 0.15), border = NA))
abline(h = disp$r, lty = 2, col = "grey40")
```

A species sitting below the band is more overdispersed than the assemblage
typically is; one above it is closer to Poisson. The species that fall inside the
band are the ones whose own counts told the model little, so the community
log-dispersion supplied most of their size. With `sigma_logr = 0` the simulator
would give a single shared $r$ and every point would land on the line; the spread
here is the assemblage genuinely differing in how clumped its species are.

## Reading community against species-level effects

The two scales a community model reports answer different questions, and keeping
them straight is the whole point of the structure.

The community mean from `coef()` answers "what is the typical response across this
assemblage". Its credible interval is narrow because every species informs it.
This is the right number for a statement about the community as a whole: the
assemblage occupies sites more often where `x` is high, the typical species
averages three individuals per site.

The per-species coefficients answer "how does this particular species respond".
For `ms_abun()` they are the community mean plus the `ranef()` deviation; for
`ms_occu()` the per-species occupancy probabilities reach you through `fitted()`
and `tobs_richness()`. A rare species' per-species estimate sits close to the
community mean precisely because the model could not learn much from its own data
and leaned on the assemblage. Read its credible interval, not just its point, and
remember that the point is partly borrowed.

The mistake to avoid is reading a shrunk per-species estimate as if it were an
independent measurement. A rare species pulled to the community mean has not told
you it resembles the assemblage; the model has assumed it does and supplied the
rest. The community SD on the fitted object is the honest measure of how much that
assumption was allowed to matter. When the SD is wide, the pooling is gentle and
the per-species estimates are mostly their own; when it is narrow, the rare
species are mostly the community mean wearing a species label.

This is also why the community mean and the per-species deviations should be read
together rather than in isolation. The deviation `ranef()` reports is a departure
from a mean that the same fit estimated, so a deviation of zero means "this
species behaves like the assemblage", not "this species has no effect". A species
whose deviation interval covers zero is consistent with the community average, and
for a rare species that consistency is largely the prior speaking. The deviations
that carry information are the ones that pull clear of zero despite the shrinkage,
because those species had enough of their own data to resist the pull.

## Practical guidance

A few rules of thumb for community fits, with the numbers that make them
actionable.

- **At least 8 to 10 species before pooling pays.** The community mean and
  covariance are learned from how species scatter, and a handful of species
  leaves $\Sigma$ barely informed. Below roughly five species, the community prior
  is mostly your assumption rather than the data's, and separate single-species
  fits are the more honest report. The pooling repays its machinery when you have
  dozens of species, many of them rare.

- **Sites in the dozens to hundreds.** Each species still needs enough sites for
  its own signal to register before the community can refine it. With fewer than
  about 20 sites, even the common species are weakly estimated and the community
  mean inherits that weakness. The community model borrows strength across
  species; it does not manufacture it from nothing.

- **Pooling helps most for the rare species, least for the common ones.** A
  species detected at 5% of sites gains the most from the community mean; a
  species detected at half its sites barely moves. If your question is only about
  the abundant species, the community structure adds machinery for little gain,
  and a single-species fit on each is simpler to defend.

- **A single-species fit is the honest choice when one species is the question.**
  If you care about one focal species and have ample data on it, fit it alone with
  `occu()` or `abun()`. Borrowing from an assemblage you do not care about only
  muddies a clean estimate, and the community mean is irrelevant to a
  single-species inference.

- **When not to pool: heterogeneous assemblages.** If the species genuinely do not
  share a response, a generalist and a specialist responding to opposite ends of a
  gradient, the community Normal is the wrong model and will fight the data. A wide
  fitted community SD is the warning sign: the assemblage is telling you it does
  not resemble itself, and the pooling is buying little. In that case the community
  mean is an average over species that have no common centre, and it describes none
  of them.

- **Check the community SD before trusting a shrunk estimate.** The fitted
  `ms_community$sd_lambda` and `sd_p` (or the occupancy community SD reported in
  the fit) are the single most useful diagnostic. A community SD near zero on a
  coefficient means the model decided the species are interchangeable on that axis,
  and every per-species estimate there is essentially the community mean. That is
  fine when it is true and misleading when it is not, so confirm it against what
  you know of the natural history before reporting per-species effects on a
  near-zero-SD axis.

The pattern under all five rules is the same. The community model trades
per-species independence for assemblage-level precision, and the trade is worth
making when the species inform each other and a waste when they do not. The fitted
community SD is where the model tells you which case you are in, and reading it is
the habit that separates a community fit you can defend from one that merely
produced numbers.

## Laplace and NUTS

The examples here use `method = "laplace"`, the fast default. It finds the
posterior mode and the curvature there and returns in a fraction of a second,
which is what you want for the first look and for comparing models. For community
occupancy, `method = "nuts"` runs the gradient sampler and returns the full
posterior, including a proper posterior for the community covariance $\Sigma$
rather than its modal estimate. When the assemblage is small or the community SD
is itself a quantity you want to report with an interval, the sampler is the
honest choice; for richness and community means on a well-populated assemblage,
the Laplace fit and the sampler agree to within their uncertainty.

The community N-mixture `ms_abun()` is Laplace-only at present; its
community-mean covariances come from the marginal observed information at the
mode. Both the Poisson and the negative-binomial arm run under that one engine,
the NB through the AGHQ path that integrates the per-species log-dispersion.

## Where to go next

- Single-season occupancy end to end: `vignette("occupancy")`.
- N-mixture abundance, negative binomial, and areal spatial counts:
  `vignette("abundance")`.
- Random effects with `lme4` bar syntax, the single-species analogue of community
  pooling: `vignette("random-effects")`.
- Posterior-predictive checks and WAIC for model comparison:
  `vignette("diagnostics")`.
```
