---
title: "Cover hurdle (independent arms)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle (independent arms)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

Vegetation cover comes with a pile of zeros. A plot either holds a species or
it does not, and where the species is present its cover lands somewhere in
$(0, 1]$. Two questions ride on one column of data: where does the species
occur, and how much of it is there once it occurs. A single Gaussian or beta
regression on cover answers neither cleanly, because the zeros pull the fit one
way and the positive values pull it another. A hurdle model splits the column
into an occurrence arm and a positive-cover arm and lets each answer its own
question.

This vignette fits the simplest version of that model: the two arms treated as
independent. `cover()` under `method = "laplace"` runs a binomial GLM on the
occurrence indicator and a separate Gaussian (or beta) GLM on the positive
subset, with no field shared between them. That is the right tool when a zero
always means true absence and the two processes are not tied through a common
spatial signal. When they are tied, the joint shared-field model in
`vignette("cover-hurdle-motivate")` couples the arms through one BYM2 field and
a fitted copy coefficient; this vignette is the baseline that joint fit is
measured against. The worked example runs end to end on `simulate_cover()` data
where the generating parameters are known, so every estimate can be checked
against a truth.

The fit uses two independent Laplace approximations through `tulpa`:

* a binomial GLM on the occurrence indicator $\mathbf{1}\{y_i > 0\}$;
* a Gaussian GLM on $\log y_i$ for the sites where $y_i > 0$.

The shared-field joint model -- occurrence, detection, and cover arms coupled
through one spatial field -- is the `occu_cover()` family;
`vignette("cover-hurdle-motivate")` builds it end to end, including the beta
positive variant. A separate-fit comparison against `INLA::inla()` on simulated
data lives in `vignette("cover-hurdle-vs-inla")`.

## The model

Write $y_i$ for the cover at site $i$ on the natural $[0, 1]$ scale, with
$y_i = 0$ recording absence. The hurdle splits $y_i$ into an occurrence
indicator $z_i = \mathbf{1}\{y_i > 0\}$ and, where $z_i = 1$, a positive part.
The two arms carry their own coefficient vectors and are fit without any shared
term:

$$
z_i \sim \mathrm{Bernoulli}(p_i), \qquad
\operatorname{logit} p_i = \mathbf{x}_i^\top \boldsymbol\beta^{\mathrm{occ}},
$$
$$
y_i \mid z_i = 1 \sim f\!\left(\eta_i^{\mathrm{pos}}, \theta\right), \qquad
\eta_i^{\mathrm{pos}} = \mathbf{x}_i^\top \boldsymbol\beta^{\mathrm{pos}} .
$$

Here $\mathbf{x}_i$ is the design row from the formula, $\boldsymbol\beta^{\mathrm{occ}}$
and $\boldsymbol\beta^{\mathrm{pos}}$ are the per-arm fixed effects, $p_i$ is the
occurrence probability, and $\eta_i^{\mathrm{pos}}$ is the positive-arm linear
predictor. The positive family $f$ comes in two variants. Under
`cover("lognormal")` the engine fits a Gaussian on $\log y_i$, so
$\log y_i \mid z_i = 1 \sim \mathrm{Normal}(\eta_i^{\mathrm{pos}}, \sigma_{\mathrm{pos}}^2)$,
and the expected cover given presence back-transforms as
$E[y_i \mid z_i = 1] = \exp(\eta_i^{\mathrm{pos}} + \sigma_{\mathrm{pos}}^2 / 2)$.
Under `cover("beta")` the engine fits a logit-link beta directly on $y_i$ with
precision $\phi$, so $\mu_i = \operatorname{logis}(\eta_i^{\mathrm{pos}})$ is the
conditional mean and $\theta = \phi$ is the dispersion. Either way the expected
cover combines the two arms as $E[y_i] = p_i \cdot \mu_i$.

The two arms share nothing in this model. The occurrence coefficients are
estimated from all $N$ sites; the positive coefficients are estimated from the
$n_{\mathrm{pos}}$ sites with $y_i > 0$ only. That independence is exactly what
the joint `occu_cover()` engine relaxes by tying the arms through a copy
coefficient. The dispersion symbol $\theta$ stands in for either
$\sigma_{\mathrm{pos}}$ (lognormal) or $\phi$ (beta), and the choice between the
two families is the main modelling decision the practical guidance returns to.

The two positive variants differ in where they put their probability mass and
in how they treat the cover scale. The lognormal variant models $\log y_i$ as
Gaussian, so it lives on the multiplicative scale: a one-unit change in
$\eta^{\mathrm{pos}}$ multiplies cover by a constant factor regardless of the
baseline level, and the variance of cover grows with its mean. That matches a
species whose cover ranges over orders of magnitude, from a trace patch to near
full ground coverage, which is the common vegetation case. The lognormal puts no
hard ceiling at one, so the engine clips simulated and predicted cover to the
$[0, 1]$ range after the back-transform; in practice the ceiling rarely binds
because the positive mass sits well below one. The beta variant models $y_i$
directly on the natural $(0, 1)$ scale with a logit link on the mean and a
single precision $\phi$ that controls the spread. Its variance
$\mu(1 - \mu) / (1 + \phi)$ shrinks toward both boundaries, so it fits cover that
clusters in the middle of the unit interval and respects the ceiling at one
without clipping. The occurrence arm is identical under both, a plain binomial
on the indicator, so swapping the positive family changes only how the
above-zero values are described.

## Simulating data

`simulate_cover()` generates data from exactly this generative model. It draws
one continuous covariate `x`, builds both linear predictors from the supplied
coefficient vectors, draws the occurrence indicator from a Bernoulli, and draws
log-cover from a Gaussian on the occupied sites before exponentiating back to
the natural scale. The covariate enters both arms with its own slope, so the
simulator can produce a species whose occurrence rises with `x` while its
conditional cover falls, which is the case the hurdle split is built to
separate.

```{r simulate}
library(tulpaObs)
sim <- simulate_cover(
  N         = 400,
  beta_occ  = c(-0.4, 0.7),
  beta_pos  = c(-1.0, 0.3),
  sigma_pos = 0.4,
  seed      = 7
)
head(sim$data)
```

The returned `sim$data` carries the cover values, the covariate `x`, and a pair
of coordinates `lon` / `lat` that the non-spatial fit ignores. The cover vector
itself is `sim$y`, the object passed to `tobs()` through the `y` argument. The
generating parameters sit in `sim$truth`: the occurrence coefficients
`beta_occ`, the positive-cover coefficients `beta_pos`, the lognormal residual
SD `sigma_pos`, plus the per-site occurrence probability `p`, the conditional
mean cover `mu`, and the realized occurrence indicator `occur`.

```{r sim-summary}
c(
  n_sites    = length(sim$y),
  n_positive = sum(sim$y > 0),
  prevalence = mean(sim$y > 0),
  median_pos = median(sim$y[sim$y > 0])
)
```

Prevalence near one half on 400 sites leaves roughly 200 positive observations
for the cover arm, comfortably above the coefficient count. The two arms see
different sample sizes by construction: every site informs occurrence, only the
occupied ones inform cover. A picture of the response makes the zero spike and
the right-skewed positive tail visible, which is the structure that motivates
the hurdle in the first place.

```{r sim-plot, message = FALSE, fig.height = 3.5}
library(ggplot2)
df <- data.frame(cover = sim$y, occupied = sim$y > 0)
ggplot(df, aes(cover, fill = occupied)) +
  geom_histogram(bins = 40, colour = NA) +
  labs(x = "cover", y = "count", fill = "y > 0") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The tall bar at zero is the absence spike; the spread of positive values is the
cover-arm signal. A Gaussian fit on the whole column would try to straddle both
and land badly between them. The hurdle fits the spike with a binomial and the
spread with the positive family, each on its own data.

The simulator exposes the knobs that matter for the fit's behaviour. `beta_occ`
sets the occurrence intercept and slope, so lowering the intercept thins the
positives and raising the slope sharpens the dependence of presence on `x`;
`beta_pos` does the same for conditional cover. `sigma_pos` is the lognormal
residual SD, the width of the positive distribution on the log scale, and
shrinking it tightens the cover values around the fitted curve. The `seed`
argument fixes the random draw so the example reproduces exactly, which is what
lets the recovery table below compare estimates to the named truth rather than
to a moving target. Changing `N` trades sampling noise for compute: more sites
narrow every interval, and the recovery tolerances in the package's own tests
loosen visibly as `N` drops from 800 toward 200. For a first look at a real
dataset, simulating from coefficients in the rough neighbourhood of the data's
prevalence and median cover is a quick way to see how tight the intervals will
be before fitting the real thing.

## Fitting

The fit is one `tobs()` call. The formula carries the fixed effects shared by
both arms (here a single covariate `x`), the data frame supplies the design, the
family is `cover(response = "lognormal")`, and the cover values pass through
`y`. The default `method = "auto"` resolves to `"laplace"` for the cover family,
which is the two-independent-Laplace path; nothing else needs setting for the
non-spatial fit.

```{r fit}
fit <- tobs(
  formula = ~ x,
  data    = sim$data,
  family  = cover(response = "lognormal"),
  y       = sim$y
)
fit
```

The print method lays out both arms. The header reports the positive family,
the total and positive sample sizes, the fitted residual SD `sigma_pos`, and the
convergence flag. The occurrence block is the binomial-logit fit on $z_i$; the
log-cover block is the Gaussian fit on $\log y_i$ over the positive subset. Each
block shows the coefficient estimate and its standard error per term, so the
slope on `x` can be read off directly for each process.

The fitted object is a `cover_fit`, and the two arms are reachable as named
vectors. `beta_occ` and `se_occ` are the occurrence coefficients and their SEs;
`beta_pos` and `se_pos` are the positive-arm coefficients and SEs. The lognormal
residual SD is `sigma_pos` (a scalar). Pulling them into a table puts the
estimates next to the generating truth.

```{r fit-extract}
est <- data.frame(
  parameter = c("beta_occ_(Intercept)", "beta_occ_x",
                "beta_pos_(Intercept)", "beta_pos_x", "sigma_pos"),
  truth     = c(sim$truth$beta_occ, sim$truth$beta_pos, sim$truth$sigma_pos),
  estimate  = c(fit$beta_occ, fit$beta_pos, fit$sigma_pos),
  se        = c(fit$se_occ, fit$se_pos, NA)
)
print(est, row.names = FALSE, digits = 3)
```

The occurrence slope recovers near its generating value of +0.7 and the
positive-cover slope near +0.3, each within a couple of SEs. The residual SD
`sigma_pos` lands near the generating 0.4. One detail to keep in mind:
`tulpa::tulpa_laplace(family = "gaussian")` runs with the dispersion fixed at
one and the cover arm estimates `sigma_pos` post-hoc as the residual standard
error $\sqrt{\mathrm{SS}_{\mathrm{res}} / (n_{\mathrm{pos}} - p)}$. That is why
`sigma_pos` has no SE column here, and why its intercept partner `beta_pos`
absorbs whatever residual scale the engine attributes to the data. The
occurrence arm carries no dispersion parameter at all, being a binomial with the
trial count fixed at one.

Two arms means two convergence states folded into one flag. `fit$converged` is
`TRUE` only when both the binomial and the Gaussian Laplace solves converged, so
a single number reports the health of the whole hurdle. The per-arm marginal
likelihoods sit in `fit$log_marginal` under the names `occ` and `pos`, which the
model-comparison section uses.

The fit also stamps the design it actually used onto `fit$encoding`, which is
worth knowing for two reasons. First, the engine autoscales the numeric columns
of each arm's design matrix so the optimizer sees well-conditioned predictors,
then transforms the coefficients and SEs back to the natural scale before
returning them, so the `beta_occ` and `beta_pos` you read are on the same scale
as the covariate, not the scaled one. Second, the encoding records which rows
fell into the positive subset and how many positive observations each arm saw,
which the convergence flag and the too-few-positives guard both lean on. If the
positive subset has fewer than `ncol(X) + 1` rows the fit stops with an error
rather than returning an unidentified cover arm, because a Gaussian on fewer
points than coefficients has no residual degrees of freedom to estimate
`sigma_pos` from. Swapping the family to `cover("beta")` keeps the same call and
the same extraction, with `phi_pos` replacing `sigma_pos` as the dispersion and
the positive block printing as a logit-link beta.

## Checking the fit

A fit that printed cleanly can still be wrong, so the two arms get checked
separately against the data they were fit on. The occurrence arm is a binomial,
so the natural check is calibration: bin sites by their predicted occupancy
probability and compare the predicted rate in each bin to the observed
occurrence rate. If the binomial link and linear predictor are right, the two
track the diagonal.

```{r check-occ, fig.height = 3.5}
p_hat <- predict(fit, sim$data, type = "occupancy")
bins  <- cut(p_hat, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
cal   <- aggregate(list(observed = as.integer(sim$y > 0)),
                   by = list(bin = bins), FUN = mean)
cal$predicted <- tapply(p_hat, bins, mean)[cal$bin]

ggplot(cal, aes(predicted, observed)) +
  geom_abline(linetype = 2) +
  geom_point(size = 2) +
  labs(x = "predicted occupancy", y = "observed occurrence rate") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The points fall close to the dashed diagonal across the probability range, which
is what a well-calibrated occurrence arm looks like. Departures would show up as
points sitting systematically above or below the line in a region of predicted
probability, signalling a missing covariate or a link mismatch. A single
numeric summary of the same idea is the agreement between mean predicted
occupancy and overall prevalence.

```{r check-occ-num}
c(mean_predicted = mean(p_hat), observed_prevalence = mean(sim$y > 0))
```

The positive arm is a Gaussian on $\log y_i$, so its check is a residual look on
the log scale. Standardized residuals against the fitted log-cover should sit in
a flat band around zero with no funnel and no trend; a funnel would mean the
constant-variance assumption is off, and a trend would mean a missing covariate
on the cover arm.

```{r check-pos, fig.height = 3.5}
pos       <- sim$y > 0
eta_pos   <- as.numeric(model.matrix(~ x, sim$data)[pos, ] %*% fit$beta_pos)
resid_std <- (log(sim$y[pos]) - eta_pos) / fit$sigma_pos

ggplot(data.frame(fitted = eta_pos, resid = resid_std), aes(fitted, resid)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4) +
  labs(x = "fitted log-cover", y = "standardized residual") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The residuals scatter evenly around zero with no fanning, so the lognormal arm's
constant-variance assumption holds on this data, as it should given the
simulator drew log-cover with a fixed `sigma_pos`. On real data a visible funnel
would argue for the beta family instead, whose variance shrinks toward the
$(0, 1)$ boundaries. The standardized residuals should also look roughly
standard-normal, which a quick spread check confirms.

```{r check-pos-num}
c(resid_sd = sd(resid_std), frac_within_2 = mean(abs(resid_std) < 2))
```

A standard deviation near one and about 95% of residuals inside $\pm 2$ are the
marks of a Gaussian arm fitting its data. These checks are deliberately built
from the fit's own coefficients and `predict()` rather than from the
single-season occupancy diagnostics (`tobs_ppc()`, `tobs_pit_residuals()`),
which operate on replicated detection data and do not apply to the plot-level
cover hurdle.

The two checks divide cleanly because the arms are independent here. The
calibration plot exercises the occurrence arm alone, since it compares predicted
occupancy to observed presence and never touches the cover values. The residual
plot exercises the positive arm alone, since it runs over the occupied sites and
never touches the zeros. Nothing couples them in this model, so a failure in one
plot localizes to one arm and one set of coefficients, which is faster to act on
than a single joint diagnostic that mixes both. On a real dataset the most
common finding is a calibration plot that bows away from the diagonal at high
predicted occupancy, which points to a saturating covariate or a missing
interaction on the occurrence formula, and a residual funnel on the positive arm
that argues for the beta family. Both are fixes to the formula or the family
rather than to the engine, and re-running the same two chunks after the change
confirms whether the fix landed.

## Interpreting results

The coefficients are posterior summaries, so each comes with a credible
interval rather than a bare point. With a Laplace fit the interval is the Wald
band $\hat\beta \pm 1.96\,\mathrm{SE}$ on each arm, read straight off `beta_occ`
/ `se_occ` and `beta_pos` / `se_pos`. The occurrence slope is the change in
occurrence logit per unit `x`; the positive slope is the change in conditional
log-cover per unit `x`.

```{r interp-ci}
z   <- qnorm(0.975)
ci  <- data.frame(
  parameter = c("beta_occ_(Intercept)", "beta_occ_x",
                "beta_pos_(Intercept)", "beta_pos_x"),
  estimate  = c(fit$beta_occ, fit$beta_pos),
  lwr       = c(fit$beta_occ, fit$beta_pos) - z * c(fit$se_occ, fit$se_pos),
  upr       = c(fit$beta_occ, fit$beta_pos) + z * c(fit$se_occ, fit$se_pos)
)
print(ci, row.names = FALSE, digits = 3)
```

Both slopes on `x` have intervals clear of zero, so the covariate moves both
occurrence and conditional cover credibly. The two effects are on different
scales (logit on the occurrence arm, log-cover on the positive arm), so the more
readable summary is the expected cover across the covariate range, with a band.
Expected cover is $E[y] = p \cdot \mu$, and propagating the per-arm SEs through
that product gives an approximate band. The cleanest route is to draw
coefficients from each arm's Gaussian posterior, push every draw through the
prediction, and take per-`x` quantiles, which carries the curvature of the
logit and lognormal links rather than a linearization.

```{r interp-draws}
newdata <- data.frame(x = seq(-2, 2, length.out = 60))
Xn      <- model.matrix(~ x, newdata)
S       <- 2000L
draw    <- function(m, se) sweep(matrix(rnorm(S * length(m)), S), 2, se, "*") +
                           rep(m, each = S)
B_occ   <- draw(fit$beta_occ, fit$se_occ)
B_pos   <- draw(fit$beta_pos, fit$se_pos)
E_draws <- plogis(B_occ %*% t(Xn)) * exp(B_pos %*% t(Xn) + fit$sigma_pos^2 / 2)
band    <- data.frame(x = newdata$x, mean = colMeans(E_draws),
                      lwr = apply(E_draws, 2, quantile, 0.025),
                      upr = apply(E_draws, 2, quantile, 0.975))
```

Each draw is a coefficient pair sampled from the per-arm Gaussian posterior,
pushed through the full prediction, so the band carries the link curvature
rather than a delta-method linearization.

```{r interp-curve, fig.height = 3.5}
ggplot(band, aes(x, mean)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line() +
  labs(x = "x", y = "expected cover E[y]") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The curve rises with `x` because the occurrence slope dominates the product, and
the band widens at the extremes where the covariate has fewer supporting sites.
Reading the curve without the band would overstate confidence in the tails. The
band here treats the two arms as independent, which they are in this model, so
the interval is the per-arm uncertainty combined through the product and no
cross-arm covariance is folded in.

## Prediction

`predict()` on a `cover_fit` returns one of three quantities at new covariate
values, selected by `type`. `type = "occupancy"` returns the occurrence
probability $p = \operatorname{logis}(\mathbf{x}^\top \boldsymbol\beta^{\mathrm{occ}})$;
`type = "conditional"` returns the conditional mean cover $\mu$ (the lognormal
back-transform $\exp(\eta^{\mathrm{pos}} + \sigma_{\mathrm{pos}}^2 / 2)$, or
$\operatorname{logis}(\eta^{\mathrm{pos}})$ under beta); and `type = "expected"`
returns their product $E[y] = p \cdot \mu$. The three are linked by an exact
identity, expected equals occupancy times conditional, which holds row by row.

```{r predict}
newdata <- data.frame(x = seq(-2, 2, length.out = 50))
preds <- data.frame(
  x         = newdata$x,
  occupancy = predict(fit, newdata, type = "occupancy"),
  expected  = predict(fit, newdata, type = "expected")
)
head(preds)
```

The `newdata` frame needs the same covariate columns the formula used; a
mismatch on column count or name errors rather than guessing. The prediction is
fixed-effects only, which is all this dispatcher carries: there is no spatial
field to project at new locations in the independent-arm path. A plot of all
three curves at once shows how the expected cover is the occupancy curve scaled
down by the conditional-cover curve.

```{r predict-shape}
qty   <- c("occupancy", "conditional", "expected")
long3 <- do.call(rbind, lapply(qty, function(q) data.frame(
  x = newdata$x, quantity = q, value = predict(fit, newdata, type = q))))
```

Stacking the three prediction types into long form lets one `ggplot()` layer
all three curves with a shared colour scale, which makes the occupancy-times-
conditional structure of expected cover read at a glance.

```{r predict-plot, fig.height = 3.5}
ggplot(long3, aes(x, value, colour = quantity)) +
  geom_line(linewidth = 0.8) +
  labs(x = "x", y = "prediction", colour = NULL) +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The occupancy curve climbs toward one as `x` rises, the conditional cover drifts
up gently with its smaller slope, and the expected cover tracks their product.
At low `x` the expected cover is pulled down by the low occupancy even where
conditional cover is moderate, which is the hurdle structure showing through:
near-zero expected cover comes from near-zero occurrence, not from low cover
where the species is present. Confirming the identity numerically is a one-line
check.

```{r predict-identity}
e <- predict(fit, newdata, type = "expected")
pm <- predict(fit, newdata, type = "occupancy") *
      predict(fit, newdata, type = "conditional")
max(abs(e - pm))
```

The maximum gap is at numerical-tolerance level, confirming that
`type = "expected"` is exactly the product of the other two and not a separately
fit surface.

Which prediction type to report depends on the question. Occupancy answers where
the species is likely to occur and is the right map for a presence survey or a
range estimate. Conditional cover answers how dense the species is where it does
occur and is the right quantity for a biomass or competition argument, but it is
silent about the zeros and so overstates the average cover across a landscape
that includes empty plots. Expected cover folds the two together and is the
honest landscape average, the quantity a monitoring programme usually wants,
because a plot contributes its conditional cover weighted by how likely the
species is to be there at all. The three predictions are fixed-effects only in
this dispatcher, so they extrapolate the fitted covariate relationship to the
`newdata` covariate values without any spatial adjustment; a site far outside
the covariate range seen in training gets a prediction driven entirely by the
extrapolated linear predictor, which the widening band in the interpretation
plot already flagged. For the spatial case, where a fitted field would shift the
prediction at a known location, the joint `occu_cover()` engine carries the
projection machinery this dispatcher does not.

## Model comparison

The positive family is a modelling choice, and the hurdle is fully observed, so
WAIC is available to score it. `tobs_waic()` assembles the pointwise
log-likelihood across both arms (the occurrence Bernoulli plus, where
$y_i > 0$, the positive-part density) and returns the WAIC, the expected
log-pointwise predictive density `elpd`, the effective number of parameters
`p_waic`, and the log-pointwise predictive density `lppd`. Fitting the same data
with the lognormal and beta positive arms and scoring both is the cleanest way
to decide which positive family the data prefers.

```{r compare-fit, message = FALSE, warning = FALSE}
fit_ln   <- tobs(~ x, data = sim$data, family = cover("lognormal"), y = sim$y)
fit_beta <- tobs(~ x, data = sim$data, family = cover("beta"),      y = sim$y)

waic_ln   <- tobs_waic(fit_ln)
waic_beta <- tobs_waic(fit_beta)
```

The two fits share the occurrence arm exactly (the Bernoulli sees the same
zeros), so the WAIC difference is driven entirely by how well each positive
family describes the spread of cover above zero. A lower WAIC is the better fit
on this scoring.

```{r compare-table}
comparison <- data.frame(
  positive = c("lognormal", "beta"),
  waic     = c(waic_ln$waic, waic_beta$waic),
  elpd     = c(waic_ln$elpd, waic_beta$elpd),
  p_waic   = c(waic_ln$p_waic, waic_beta$p_waic)
)
comparison$delta_waic <- comparison$waic - min(comparison$waic)
print(comparison, row.names = FALSE, digits = 4)
```

Because this data was simulated from the lognormal generative model, the
lognormal arm should score at or below the beta arm here, and a `delta_waic`
larger than a few units on the wrong family is the signal to switch. On real
cover data the verdict can go either way: cover concentrated in the middle of
$(0, 1)$ favours beta, while cover with many small-but-positive values and a
long right tail favours lognormal. The WAIC difference is the number to report,
but it should be read alongside the conditional-cover residual check above,
because WAIC rewards the family that fits the bulk of the data and can miss a
boundary misfit that the residual plot makes obvious.

The same `tobs_waic()` call also scores the hurdle against a simpler nested
model, for instance an intercept-only cover formula versus the covariate model,
which tests whether `x` earns its place on the arms. The comparison is valid as
long as both fits are scored on the same response with the same WAIC assembly,
which `tobs_waic()` guarantees by reading each fit's own encoding.

## Practical guidance

A few numbers and rules of thumb that hold up across the recovery tests.

* **At least `ncol(X) + 1` positive observations, and comfortably more.** The
  positive arm only sees the sites with $y_i > 0$. The encoder hard-errors below
  `ncol(X) + 1` positive rows, but that floor is a structural minimum, not a
  working target. Aim for roughly 20 to 30 positive observations per cover-arm
  coefficient; a two-covariate cover formula wants the positive subset well into
  the dozens before the slopes and `sigma_pos` settle. The recovery tests run at
  600 to 800 sites with prevalence near one half, which leaves a few hundred
  positive rows, and even there the per-fit `sigma_pos` carries a tolerance of
  about 15% around its truth.

* **Prevalence away from the extremes.** With prevalence below roughly 0.1 the
  occurrence arm has few positives to estimate the slope from and the cover arm
  has few observations at all; with prevalence above 0.95 the occurrence arm has
  few zeros and the binomial slope is poorly pinned. The simulator's default
  coefficients keep prevalence in a workable middle band, and a real dataset
  with prevalence at an extreme should expect wide intervals on the
  corresponding arm.

* **Sample size scales the `sigma_pos` agreement.** The lognormal residual SD is
  computed post-hoc and converges to the generating value as the positive subset
  grows. At a few hundred positive rows it lands within about 15% of truth; at a
  few dozen it can be off by more, which then feeds through the lognormal
  back-transform and scales every conditional-cover prediction. When the
  positive subset is small, treat `sigma_pos` and the conditional-cover curve as
  the least certain outputs of the fit.

Choosing the positive family comes down to where the cover mass sits. Use
`cover("lognormal")` when cover is right-skewed with many small-but-positive
values and a long upper tail, which is the common vegetation pattern; the
Gaussian-on-log fit handles the skew and the residual check on the log scale is
easy to read. Use `cover("beta")` when cover is bounded strictly inside
$(0, 1)$ and concentrated in the middle, where the beta's shrinking variance
toward the boundaries matches the data and the lognormal would overstate the
spread near one. When the call is genuinely close, fit both and compare WAIC as
above, then confirm the winner with the conditional-cover residual plot, because
the two diagnostics catch different failures.

When to use the joint `occu_cover()` instead: the independent-arm `cover()`
model here assumes the two arms share no structure and that a zero always means
true absence. Reach for the joint shared-field model in
`vignette("cover-hurdle-motivate")` when either assumption fails. If a habitat
gradient that lifts occurrence also lifts conditional cover, the arms are
coupled through a common spatial field and fitting them apart throws that
coupling away along with the cover observations' information about occurrence at
neighbouring plots; the joint fit recovers a copy coefficient `alpha` with a
credible interval that formally tests whether the coupling is real. And if a
zero can mean the species was present but missed, the three-arm `occu_cover()`
adds a detection layer that separates non-detection from absence, which the
plot-level `cover()` cannot do because it has no replicated visits. With fewer
than roughly 30 spatial units the shared field is weakly identified and the
joint machinery buys little, so the independent-arm fit here, with spatial
structure folded into covariates, is the more honest model at small N. The joint
path stays affordable on large resurvey datasets because it collapses both arms
to exact sufficient statistics before each inner solve (`control$aggregate.occ`
and `control$aggregate.pos`, both on by default and byte-identical to the full
per-plot fit); `vignette("cover-hurdle-motivate")` shows those controls.

```{r compare-inla, eval = FALSE}
# Phase 1e: compare against the joint INLA hurdle fit on the MOTIVATE
# resurvey data. Skeleton retained here so the comparison drops in
# without restructuring the vignette.
#
# library(INLA)
# joint <- inla(
#   list(
#     occur ~ x + f(site, model = "bym2", graph = adj),
#     log_cover ~ x + f(site, model = "bym2", graph = adj, copy = "site",
#                       fixed = FALSE)
#   ),
#   family = c("binomial", "gaussian"),
#   data   = inla_stack
# )
```

## Scope

This dispatcher predicts from fixed effects only. Spatial fields fit through
the `spatial` argument, and spatial projection at new coordinates -- together
with the within/between (Mundlak) temporal decomposition -- comes with the
joint `occu_cover()` engine; see `vignette("cover-hurdle-motivate")`.
