---
title: "Cover hurdle (Phase 1a)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle (Phase 1a)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette shows the Phase 1a cover-hurdle dispatcher on simulated data.
The fit uses two independent Laplace approximations through `tulpa`:

* a binomial GLM on the occurrence indicator $\mathbf{1}\{y_i > 0\}$;
* a Gaussian GLM on $\log y_i$ for the sites where $y_i > 0$.

The shared-field joint model (and the beta positive variant) land in
Phase 1c / 1d. Comparison to `INLA::inla()` on the real MOTIVATE data is
Phase 1e; the code chunks below for that comparison are marked
`eval = FALSE`.

```{r simulate}
library(TulpaObs)
sim <- simulate_cover_hurdle(
  N         = 400,
  beta_occ  = c(-0.4, 0.7),
  beta_pos  = c(-1.0, 0.3),
  sigma_pos = 0.4,
  seed      = 7
)
head(sim$data)
```

```{r fit}
fit <- tulpa_obs(
  formula = ~ x,
  data    = sim$data,
  family  = cover_hurdle(positive = "lognormal"),
  y       = sim$y
)
fit
```

```{r predict}
newdata <- data.frame(x = seq(-2, 2, length.out = 50))
preds <- data.frame(
  x         = newdata$x,
  occupancy = predict(fit, newdata, type = "occupancy"),
  expected  = predict(fit, newdata, type = "expected")
)
head(preds)
```

```{r compare-inla, eval = FALSE}
# Phase 1e: compare against the joint INLA hurdle fit on the MOTIVATE
# resurvey data. Skeleton retained here so the comparison drops in
# without restructuring the vignette.
#
# library(INLA)
# joint <- inla(
#   list(
#     occur ~ x + f(site, model = "bym2", graph = adj),
#     log_cover ~ x + f(site, model = "bym2", graph = adj, copy = "site",
#                       fixed = FALSE)
#   ),
#   family = c("binomial", "gaussian"),
#   data   = inla_stack
# )
```

## Phase 1a scope

The Phase 1a fit is fixed-effects-only at prediction time. Spatial fields
fit through the `spatial` argument, but spatial projection at new
coordinates lands together with the shared-field joint engine in Phase 1c.
Temporal helpers (the MOTIVATE within/between Mundlak decomposition) land
in Phase 1d.
