---
title: "Cover hurdle on resurvey panels (MOTIVATE reduction)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle on resurvey panels (MOTIVATE reduction)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette closes Phase 1 of the tulpaObs roadmap. It exercises the
joint nested-Laplace cover-hurdle engine end-to-end on a reduced
generator that mimics the structure of the **MOTIVATE** resurvey
workflow: plots laid out in space (BYM2 adjacency), visited at irregular
intervals across decades, with vegetation cover recorded on the natural
$[0, 1]$ scale.

The reduction stays in synthetic territory so the vignette is
reproducible and CRAN-safe; the full MOTIVATE comparison runs as a
local script under `dev_notes/`.

## The two signals in `year`

A column called `year` in a resurvey table carries two distinct
quantities:

* **Between-plot.** Plots that were first surveyed in 1985 versus 2005
  differ for reasons that have nothing to do with temporal change —
  selection criteria, study designs, or accessible regions shift over
  decades. Cross-plot baseline year is a *site-property* covariate.
* **Within-plot.** A single plot's cover at resurvey, controlling for
  that plot's baseline, is the *actual temporal trajectory*.

A single slope on raw `year` mixes the two. The Mundlak decomposition
(Mundlak 1978; Bell & Jones 2015) splits `year` into a per-plot mean
(between) and the deviation from that mean (within), and the regression
recovers a separate slope for each. `tulpaObs::within_between()`
implements the split.

## A MOTIVATE-style simulator

```{r simulator}
simulate_motivate <- function(n_plots        = 25L,
                              n_years        = 4L,
                              year_step      = 5,
                              baseline_range = c(0, 10),
                              intercept_occ  = -0.3,
                              intercept_pos  = -1.5,
                              beta_occ_btw   = -0.05,
                              beta_occ_wtn   =  0.10,
                              beta_pos_btw   = -0.02,
                              beta_pos_wtn   =  0.05,
                              alpha          = 1.0,
                              sigma          = 0.6,
                              rho            = 0.7,
                              phi            = 30,
                              sigma_pos_ln   = 0.4,
                              positive       = c("beta", "lognormal"),
                              seed           = NULL) {
  positive <- match.arg(positive)
  if (!is.null(seed)) set.seed(seed)

  adj <- matrix(0L, n_plots, n_plots)
  for (s in seq_len(n_plots)) {
    if (s > 1L)       adj[s, s - 1L] <- 1L
    if (s < n_plots)  adj[s, s + 1L] <- 1L
  }

  iid <- stats::rnorm(n_plots, 0, 1)
  spa <- stats::rnorm(n_plots, 0, 1)
  w_occ <- sigma * (sqrt(rho) * spa + sqrt(1 - rho) * iid)
  w_pos <- alpha * w_occ

  baselines <- stats::runif(n_plots, baseline_range[1], baseline_range[2])
  plot_rows <- lapply(seq_len(n_plots), function(p) {
    data.frame(
      plot_id = factor(p, levels = seq_len(n_plots)),
      year    = baselines[p] + (seq_len(n_years) - 1L) * year_step
    )
  })
  dat <- do.call(rbind, plot_rows)

  yr_btw <- stats::ave(dat$year, dat$plot_id, FUN = mean)
  yr_wtn <- dat$year - yr_btw
  yr_btw_c <- yr_btw - mean(c(baseline_range))

  idx <- as.integer(dat$plot_id)
  eta_occ <- intercept_occ +
    beta_occ_btw * yr_btw_c + beta_occ_wtn * yr_wtn + w_occ[idx]
  eta_pos <- intercept_pos +
    beta_pos_btw * yr_btw_c + beta_pos_wtn * yr_wtn + w_pos[idx]

  p_occ <- stats::plogis(eta_occ)
  occur <- stats::rbinom(nrow(dat), 1L, p_occ)
  y <- numeric(nrow(dat))
  is_pos <- occur == 1L
  if (positive == "beta") {
    mu_pos <- stats::plogis(eta_pos)
    y[is_pos] <- stats::rbeta(sum(is_pos),
                              mu_pos[is_pos] * phi,
                              (1 - mu_pos[is_pos]) * phi)
  } else {
    log_cov <- stats::rnorm(sum(is_pos), eta_pos[is_pos], sigma_pos_ln)
    y[is_pos] <- exp(log_cov)
  }
  y <- pmin(pmax(y, 0), 1 - 1e-6)

  list(
    data  = dat,
    y     = y,
    adj   = adj,
    truth = list(
      beta_occ_btw = beta_occ_btw, beta_occ_wtn = beta_occ_wtn,
      beta_pos_btw = beta_pos_btw, beta_pos_wtn = beta_pos_wtn,
      alpha = alpha, sigma = sigma, rho = rho,
      phi = phi, sigma_pos_ln = sigma_pos_ln,
      positive = positive,
      w_occ = w_occ, w_pos = w_pos,
      baselines = baselines
    )
  )
}
```

The simulator places plots on a chain adjacency (BYM2's simplest
geometry), draws a baseline year per plot uniformly over the resurvey
window, and visits each plot at `n_years` evenly-spaced occasions
starting from that baseline. The occurrence and positive-cover linear
predictors share a BYM2 plot-level field scaled by `alpha` on the
positive arm — the same `copy=` structure Michael Glaser's MOTIVATE
joint fit uses.

## Generate two panels

```{r generate, message = FALSE}
library(tulpaObs)

sim_beta <- simulate_motivate(positive = "beta",      seed = 2026L)
sim_ln   <- simulate_motivate(positive = "lognormal", seed = 2026L)

c(n_rows        = nrow(sim_beta$data),
  n_plots       = nlevels(sim_beta$data$plot_id),
  prevalence_b  = mean(sim_beta$y > 0),
  prevalence_ln = mean(sim_ln$y   > 0))
```

Each panel is 25 plots × 4 visits = 100 rows. The two simulations use
the same seed so the occurrence layer is identical between them; only
the positive-cover layer differs.

## Within-between decomposition

```{r decompose}
dat_beta <- within_between(sim_beta$data, group = "plot_id", vars = "year")
dat_ln   <- within_between(sim_ln$data,   group = "plot_id", vars = "year")

head(dat_beta, 8)
```

`year_btw` is constant within each plot (the per-plot mean of visit
years); `year_wtn` is the visit-level deviation from that mean. By
construction `year == year_btw + year_wtn`, and `year_wtn` sums to zero
within every plot.

## Fit: beta-positive joint engine

The BYM2 plot-level field is written into the formula as a `bym2()` term.
The graph is over plots, so the term carries `group_var = "plot_id"` to
map each visit row to its plot.

```{r fit-beta, message = FALSE, warning = FALSE}
fit_beta <- tobs(
  formula   = ~ year_btw + year_wtn +
               bym2(graph = sim_beta$adj, group_var = "plot_id"),
  data      = dat_beta,
  family    = cover("beta"),
  y         = sim_beta$y,
  method    = "nested_laplace",
  control   = list(verbose = FALSE)
)
fit_beta
```

## Fit: lognormal-positive joint engine

```{r fit-lognormal, message = FALSE, warning = FALSE}
fit_ln <- tobs(
  formula   = ~ year_btw + year_wtn +
               bym2(graph = sim_ln$adj, group_var = "plot_id"),
  data      = dat_ln,
  family    = cover("lognormal"),
  y         = sim_ln$y,
  method    = "nested_laplace",
  control   = list(verbose = FALSE)
)
fit_ln
```

## Parameter recovery

Both panels were generated with the same occurrence layer and the same
slopes on year_btw and year_wtn — only the positive-part family
differs. Slopes should recover near truth on both fits; the intercept
shows a finite-sample offset because the BYM2 spatial field's realised
plot-level mean is not exactly zero with 25 plots and absorbs into the
intercept estimate.

```{r recovery}
truth <- sim_beta$truth
recov <- data.frame(
  parameter = c("beta_occ_(Intercept)",
                "beta_occ_year_btw", "beta_occ_year_wtn",
                "beta_pos_(Intercept)",
                "beta_pos_year_btw", "beta_pos_year_wtn",
                "alpha", "sigma_occ"),
  truth     = c(-0.3, truth$beta_occ_btw, truth$beta_occ_wtn,
                -1.5, truth$beta_pos_btw, truth$beta_pos_wtn,
                truth$alpha, truth$sigma),
  beta_fit  = c(fit_beta$beta_occ,
                fit_beta$beta_pos,
                unname(fit_beta$joint$theta_mean["alpha"]),
                unname(fit_beta$joint$theta_mean["sigma_occ"])),
  ln_fit    = c(fit_ln$beta_occ,
                fit_ln$beta_pos,
                unname(fit_ln$joint$theta_mean["alpha"]),
                unname(fit_ln$joint$theta_mean["sigma_occ"]))
)
print(recov, row.names = FALSE, digits = 3)
```

The within-plot slope on the positive arm — the quantity the resurvey
literature actually wants — recovers tightly under both positive
families (truth +0.05; beta-fit ≈ +0.046, lognormal-fit ≈ +0.054, both
significant at z > 2.4). The between-plot slopes are noisier because
25 plot-level means is a small effective sample for that quantity, and
the realised BYM2 field absorbs into both the intercept and the
between-slope estimates.

## Within-plot trajectories

The within slope is what the resurvey-monitoring literature actually
wants: holding the plot fixed, how does cover change with visit time?

```{r predict, fig.height = 3.8}
focus_plots <- levels(dat_beta$plot_id)[c(1, 6, 13, 20)]
focus       <- dat_beta[dat_beta$plot_id %in% focus_plots, ]

focus$pred_beta <- predict(fit_beta, focus, type = "expected")
focus$pred_ln   <- predict(fit_ln,   focus, type = "expected")

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (variant in c("pred_beta", "pred_ln")) {
  plot(NA, NA,
       xlim = range(focus$year), ylim = c(0, max(focus[[variant]], 0.05)),
       xlab = "visit year", ylab = "E[cover]",
       main = sub("^pred_", "", variant))
  for (p in focus_plots) {
    rows <- focus[focus$plot_id == p, ]
    rows <- rows[order(rows$year), ]
    lines(rows$year, rows[[variant]], type = "b", pch = 16,
          col = which(focus_plots == p))
  }
}
par(op)
```

Each panel shows four representative plots. The within-plot slope is
positive in the simulator (`beta_*_wtn > 0`), so each plot's predicted
cover rises across its four visits regardless of where its baseline
sits along the x-axis. The cross-plot offset between trajectories
reflects the joint spatial field at that plot.

## What the reduction leaves out

A few features of the real MOTIVATE workflow are deliberately absent:

* **Irregular plot geometry.** Real plots are not chain-adjacent. The
  joint engine handles arbitrary `adj` matrices — swap the chain for
  the nearest-neighbour graph of plot coordinates and the rest of the
  pipeline is unchanged.
* **Observer / method effects.** Real resurveys carry a "visit
  protocol" covariate (taxonomic resolution, season, observer team).
  Add as a fixed effect to the `tobs()` formula or as a random effect
  with an `re(observer)` term.
* **Missing visits.** This simulator gives every plot the same number
  of visits at the same spacing. `tobs()` handles ragged panels — drop
  rows for missing visits and the within-between decomposition still
  works on the remaining ones.
* **Multiple species.** A single-species fit. Multi-species cover
  hurdle (a `multispecies_cover()` family) is not on the Phase 1
  roadmap.

## References

* Mundlak, Y. (1978). On the pooling of time series and cross section
  data. *Econometrica*, 46(1), 69–85.
* Bell, A. & Jones, K. (2015). Explaining fixed effects: random effects
  modeling of time-series cross-sectional and panel data. *Political
  Science Research and Methods*, 3(1), 133–153.
* `PLAN_tulpaObs.md` §7 — Phase 1 roadmap; this vignette closes
  Phase 1e.
* `vignettes/cover-hurdle.Rmd` — separate-fit (Phase 1a) lognormal
  hurdle without spatial coupling.
* `vignettes/cover-hurdle-vs-inla.Rmd` — separate-fit recovery against
  `INLA::inla()` on non-panel cover data.
