---
title: "Cover hurdle on resurvey panels (MOTIVATE reduction)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle on resurvey panels (MOTIVATE reduction)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

Vegetation cover is a zero-inflated quantity. A plot either holds a species or
it does not, and where the species is present, its cover lands somewhere in
$(0, 1]$. Two questions ride on the same data: where does the species occur,
and how much of it is there once it occurs. A naive Gaussian or beta regression
on cover mixes the two, because the zeros and the positive values answer
different questions. A hurdle model separates them into an occurrence arm and a
positive-cover arm.

The two arms are not independent. A patch of habitat that lifts the chance of
presence usually also lifts the cover where the species is present: the same
soil moisture, the same grazing history, the same nutrient gradient drives
both. `tulpaObs` couples the arms through one shared spatial field. The
occurrence arm sees the field at full amplitude; the cover arm sees a scalar
multiple of it, with the multiplier (the copy coefficient) fitted from the data
rather than assumed. Fitting the two arms separately throws this coupling away.
It also wastes the cover observations: a plot with high cover is evidence that
the latent field is high there, which sharpens the occurrence estimate at
neighbouring plots, and the separate fit never propagates that. The joint fit
borrows strength across arms and across space at once, and it reports the copy
coefficient with a credible interval so you can ask whether the two processes
really do track each other.

There is a third reason the joint model matters, specific to monitoring. When
the goal is to estimate change, not just a snapshot, the coupling decides how a
shift in occurrence translates into a shift in cover. If a warming climate
lifts the occurrence probability at a plot, the joint model says cover there
rises too, by an amount the copy coefficient sets; the separate model leaves
the two changes unlinked and can report a rising occurrence beside a flat cover
with no internal contradiction. Whether that linkage is real is an empirical
question, and the joint fit answers it with the posterior on the copy
coefficient rather than forcing it to zero (separate fits) or to one
(hard-copied fits).

This vignette builds the joint cover-hurdle end to end on a reduced generator
that mimics the structure of the **MOTIVATE** resurvey workflow: plots laid out
in space (BYM2 adjacency), visited at irregular intervals across decades, with
vegetation cover recorded on the natural $[0, 1]$ scale. It then switches to the
three-arm `occu_cover()` family, which adds an explicit detection layer and the
change-quantity prediction contract the resurvey literature wants. The two
families share an engine and a parameterization; the difference is whether you
trust your zeros. If a zero always means true absence, the two-arm `cover()`
family is enough. If a zero can mean the species was there and missed, the
three-arm `occu_cover()` family separates non-detection from absence and the
detection arm earns its parameters.

The reduction stays in synthetic territory so the vignette is reproducible and
CRAN-safe; the full MOTIVATE comparison runs as a local script under
`dev_notes/`.

## The model

Write $y_i$ for the cover at observation $i$, on the natural $[0, 1]$ scale,
with $y_i = 0$ recording absence. The hurdle splits into an occurrence
indicator and a positive part. Let $z_i = \mathbf{1}\{y_i > 0\}$ be the
occurrence indicator with probability $p_i$, and let the positive cover follow
a family $f$ (lognormal on $\log y_i$, or beta on $y_i$ directly):

$$
z_i \sim \mathrm{Bernoulli}(p_i), \qquad
\operatorname{logit} p_i = \mathbf{x}_i^\top \boldsymbol\beta^{\mathrm{occ}} + w_{s(i)},
$$
$$
y_i \mid z_i = 1 \sim f\!\left(\eta_i^{\mathrm{pos}}, \theta\right), \qquad
\eta_i^{\mathrm{pos}} = \mathbf{x}_i^\top \boldsymbol\beta^{\mathrm{pos}} + \alpha\, w_{s(i)} .
$$

Here $w_s$ is a BYM2 field over the spatial units $s$, with marginal amplitude
$\sigma$ and ICAR/IID mixing $\rho$, mapped to observation $i$ through
$s(i)$. The occurrence arm loads the field at amplitude one; the cover arm
loads the same field scaled by $\alpha$, the copy coefficient. So
$\alpha = 1$ means the two arms see the field identically, $\alpha = 0$
decouples them (the cover arm has no shared spatial structure), and a negative
$\alpha$ means the two processes run in opposition. The dispersion $\theta$ is
the beta precision $\phi$ or the lognormal residual SD $\sigma_{\mathrm{pos}}$,
depending on the positive family. The symbol $w_{s(i)}$ is the single object
that turns two regressions into a joint model: it is shared, and only its
amplitude on the cover arm is free.

The copy coefficient is what makes this a joint model rather than two
regressions stapled together. With $\alpha$ free, the cover observations
inform the field, and the field informs occurrence: a plot with high cover
pushes $w_s$ up there, which raises $\psi$ at that plot and, through the ICAR
penalty, at its neighbours. Information flows both ways across the arms and
across space. A separate fit severs the cross-arm path entirely, and a
hard-copied fit ($\alpha$ fixed at one) assumes the flow is at full strength
without checking. The free copy coefficient sits between those two and lets the
data set the strength.

The latent $z_i$ marginalises out of the likelihood in closed form (two states
per observation), so the fit is a nested Laplace approximation, with the
hyperparameters $(\sigma, \alpha, \theta)$ integrated on an outer grid and the
field plus coefficients solved by an inner Newton step at each grid point. That
is the `joint` engine, the default for `method = "nested_laplace"` on
`occu_cover()`. Because the hyperparameters are integrated rather than
optimized, the SEs on $\alpha$ and $\sigma$ reflect the full posterior over the
grid, not a curvature approximation at a single mode, and the coefficient SEs
fold in the uncertainty about the field amplitude through the law of total
variance. The price is that the outer grid has to be wide enough to bracket the
posterior mass; a grid that clips the mode biases the integrated estimate, a
point the practical guidance returns to.

## A MOTIVATE-style simulator

```{r simulator}
simulate_motivate <- function(n_plots        = 25L,
                              n_years        = 4L,
                              year_step      = 5,
                              baseline_range = c(0, 10),
                              intercept_occ  = -0.3,
                              intercept_pos  = -1.5,
                              beta_occ_btw   = -0.05,
                              beta_occ_wtn   =  0.10,
                              beta_pos_btw   = -0.02,
                              beta_pos_wtn   =  0.05,
                              alpha          = 1.0,
                              sigma          = 0.6,
                              rho            = 0.7,
                              phi            = 30,
                              sigma_pos_ln   = 0.4,
                              positive       = c("beta", "lognormal"),
                              seed           = NULL) {
  positive <- match.arg(positive)
  if (!is.null(seed)) set.seed(seed)

  adj <- matrix(0L, n_plots, n_plots)
  for (s in seq_len(n_plots)) {
    if (s > 1L)       adj[s, s - 1L] <- 1L
    if (s < n_plots)  adj[s, s + 1L] <- 1L
  }

  iid <- stats::rnorm(n_plots, 0, 1)
  spa <- stats::rnorm(n_plots, 0, 1)
  w_occ <- sigma * (sqrt(rho) * spa + sqrt(1 - rho) * iid)
  w_pos <- alpha * w_occ

  baselines <- stats::runif(n_plots, baseline_range[1], baseline_range[2])
  plot_rows <- lapply(seq_len(n_plots), function(p) {
    data.frame(
      plot_id = factor(p, levels = seq_len(n_plots)),
      year    = baselines[p] + (seq_len(n_years) - 1L) * year_step
    )
  })
  dat <- do.call(rbind, plot_rows)

  yr_btw <- stats::ave(dat$year, dat$plot_id, FUN = mean)
  yr_wtn <- dat$year - yr_btw
  yr_btw_c <- yr_btw - mean(c(baseline_range))

  idx <- as.integer(dat$plot_id)
  eta_occ <- intercept_occ +
    beta_occ_btw * yr_btw_c + beta_occ_wtn * yr_wtn + w_occ[idx]
  eta_pos <- intercept_pos +
    beta_pos_btw * yr_btw_c + beta_pos_wtn * yr_wtn + w_pos[idx]

  p_occ <- stats::plogis(eta_occ)
  occur <- stats::rbinom(nrow(dat), 1L, p_occ)
  y <- numeric(nrow(dat))
  is_pos <- occur == 1L
  if (positive == "beta") {
    mu_pos <- stats::plogis(eta_pos)
    y[is_pos] <- stats::rbeta(sum(is_pos),
                              mu_pos[is_pos] * phi,
                              (1 - mu_pos[is_pos]) * phi)
  } else {
    log_cov <- stats::rnorm(sum(is_pos), eta_pos[is_pos], sigma_pos_ln)
    y[is_pos] <- exp(log_cov)
  }
  y <- pmin(pmax(y, 0), 1 - 1e-6)

  list(
    data  = dat,
    y     = y,
    adj   = adj,
    truth = list(
      beta_occ_btw = beta_occ_btw, beta_occ_wtn = beta_occ_wtn,
      beta_pos_btw = beta_pos_btw, beta_pos_wtn = beta_pos_wtn,
      alpha = alpha, sigma = sigma, rho = rho,
      phi = phi, sigma_pos_ln = sigma_pos_ln,
      positive = positive,
      w_occ = w_occ, w_pos = w_pos,
      baselines = baselines
    )
  )
}
```

The simulator places plots on a chain adjacency (BYM2's simplest geometry),
draws a baseline year per plot uniformly over the resurvey window, and visits
each plot at `n_years` evenly-spaced occasions starting from that baseline. The
occurrence and positive-cover linear predictors share a BYM2 plot-level field
scaled by `alpha` on the positive arm. The known truth is in `sim$truth`:
occurrence slopes (`beta_occ_btw`, `beta_occ_wtn`), positive-cover slopes
(`beta_pos_btw`, `beta_pos_wtn`), the copy coefficient `alpha`, the field
amplitude `sigma`, the BYM2 mixing `rho`, and the per-plot field draws `w_occ`
on the occurrence arm and `w_pos = alpha * w_occ` on the cover arm. Holding
those fixed across two seeds lets us isolate the effect of the positive family:
the beta and lognormal panels below share an identical occurrence layer and
differ only in how the positive cover is drawn.

The field construction deserves a second look because it is the part most
people get wrong. The occurrence-arm field `w_occ` is drawn once per plot as a
BYM2-style mix of a spatially-structured component and an IID component,
weighted by `rho`, and scaled by `sigma`. The cover-arm field is not redrawn:
it is `alpha * w_occ`, the same field at a different amplitude. That is the
copy structure made explicit in the generator. If you instead drew an
independent field for the cover arm, you would be simulating the separate-fit
world, and no joint model could recover an `alpha`, because there would be no
shared field to scale. The two arms also carry their own within-plot and
between-plot slopes, so the simulator can produce a species whose occurrence
trends one way over time while its conditional cover trends another, which is
the realistic case the Mundlak split is built to disentangle.

## The two signals in `year`

A column called `year` in a resurvey table carries two distinct quantities:

* **Between-plot.** Plots that were first surveyed in 1985 versus 2005 differ
  for reasons that have nothing to do with temporal change: selection
  criteria, study designs, or accessible regions shift over decades. Cross-plot
  baseline year is a *site-property* covariate.
* **Within-plot.** A single plot's cover at resurvey, controlling for that
  plot's baseline, is the *actual temporal trajectory*.

A single slope on raw `year` mixes the two. The Mundlak decomposition (Mundlak
1978; Bell & Jones 2015) splits `year` into a per-plot mean (between) and the
deviation from that mean (within), and the regression recovers a separate slope
for each. `tulpaObs::within_between()` implements the split.

## Generate two panels

```{r generate, message = FALSE}
library(tulpaObs)

sim_beta <- simulate_motivate(positive = "beta",      seed = 2026L)
sim_ln   <- simulate_motivate(positive = "lognormal", seed = 2026L)

c(n_rows        = nrow(sim_beta$data),
  n_plots       = nlevels(sim_beta$data$plot_id),
  prevalence_b  = mean(sim_beta$y > 0),
  prevalence_ln = mean(sim_ln$y   > 0))
```

Each panel is 25 plots x 4 visits = 100 rows. The two simulations use the same
seed so the occurrence layer is identical between them; only the positive-cover
layer differs.

## Within-between decomposition

```{r decompose}
dat_beta <- within_between(sim_beta$data, group = "plot_id", vars = "year")
dat_ln   <- within_between(sim_ln$data,   group = "plot_id", vars = "year")

head(dat_beta, 8)
```

`year_btw` is constant within each plot (the per-plot mean of visit years);
`year_wtn` is the visit-level deviation from that mean. By construction
`year == year_btw + year_wtn`, and `year_wtn` sums to zero within every plot.

## Fitting the model

The BYM2 plot-level field is written into the formula as a `bym2()` term. The
graph is over plots, so the term carries `group_var = "plot_id"` to map each
visit row to its plot. The positive family is chosen by `cover("beta")` or
`cover("lognormal")`, the cover values pass through `y`, and
`method = "nested_laplace"` selects the joint shared-field engine. Everything
else is a default: the field amplitude `sigma`, the copy coefficient `alpha`,
the BYM2 mixing `rho`, and the dispersion are integrated on the outer
hyperparameter grid.

```{r fit-beta, message = FALSE, warning = FALSE}
fit_beta <- tobs(
  formula   = ~ year_btw + year_wtn +
               bym2(graph = sim_beta$adj, group_var = "plot_id"),
  data      = dat_beta,
  family    = cover("beta"),
  y         = sim_beta$y,
  method    = "nested_laplace",
  control   = list(verbose = FALSE)
)
fit_beta
```

The print method lays out both arms. The occurrence block is the binomial-logit
fit on $z_i$; the cover block is the beta fit (logit link on the conditional
mean) over the positive subset. The header carries the positive family, the
total and positive sample sizes, the fitted dispersion (`phi_pos` for beta,
`sigma_pos` for lognormal), and the convergence flag. The same call with the
lognormal family fits the second panel:

```{r fit-lognormal, message = FALSE, warning = FALSE}
fit_ln <- tobs(
  formula   = ~ year_btw + year_wtn +
               bym2(graph = sim_ln$adj, group_var = "plot_id"),
  data      = dat_ln,
  family    = cover("lognormal"),
  y         = sim_ln$y,
  method    = "nested_laplace",
  control   = list(verbose = FALSE)
)
fit_ln
```

Each fit exposes the coefficients and SEs per arm in `beta_occ` / `se_occ` and
`beta_pos` / `se_pos`, the dispersion in `sigma_pos` / `phi_pos` (with the
posterior SD in `sigma_pos_sd` / `phi_pos_sd`), and the raw joint engine result
in `fit$joint`. The copy coefficient and donor amplitude come off the joint
engine's hyperparameter posterior, `fit$joint$theta_mean`, under the names
`alpha` and `sigma_occ`.

### Exact row reduction

The joint engine collapses each arm to its exact sufficient statistic before the
inner solve. Plots that share the occurrence design row and every per-observation
latent component (the spatial cell, the trend weight) are exchangeable Bernoulli
trials, so the occurrence arm reduces to one Binomial row per group carrying the
trial count and the success count. The beta positive arm has no single-row form,
so it reduces instead to one row per group carrying `(n, sum log y, sum log(1 - y))`,
the statistics the beta log-density depends on. Neither step changes the
log-likelihood, gradient, or Hessian; the reduced fit returns the same numbers as
the per-plot fit with fewer rows in the inner scatter.

Both reductions run by default, through `control$aggregate.occ` and
`control$aggregate.pos`. The fits above already use the collapsed arms. Setting
either to `FALSE` recovers the full per-plot arm, which lets us check that nothing
moved:

```{r aggregate-equiv}
fit_full <- tobs(
  formula = ~ year_btw + year_wtn +
             bym2(graph = sim_beta$adj, group_var = "plot_id"),
  data    = dat_beta, family = cover("beta"), y = sim_beta$y,
  method  = "nested_laplace",
  control = list(verbose = FALSE, aggregate.occ = FALSE, aggregate.pos = FALSE)
)
max(abs(fit_beta$beta_pos - fit_full$beta_pos))
```

The collapsed and full positive-arm coefficients agree to numerical tolerance.
`aggregate.pos` is specific to the beta arm; on a lognormal positive arm it stays
off, and an explicit `aggregate.pos = TRUE` there raises an error rather than
silently doing nothing. The reduction earns its keep on resurvey datasets where
many plots share a cell and design row, the regime these joint cover fits are
built for.

## Parameter recovery

Both panels were generated with the same occurrence layer and the same slopes
on `year_btw` and `year_wtn`, only the positive-part family differs. Slopes
should recover near truth on both fits; the intercept shows a finite-sample
offset because the BYM2 spatial field's realised plot-level mean is not exactly
zero with 25 plots and absorbs into the intercept estimate.

```{r recovery}
truth <- sim_beta$truth
recov <- data.frame(
  parameter = c("beta_occ_(Intercept)",
                "beta_occ_year_btw", "beta_occ_year_wtn",
                "beta_pos_(Intercept)",
                "beta_pos_year_btw", "beta_pos_year_wtn",
                "alpha", "sigma_occ"),
  truth     = c(-0.3, truth$beta_occ_btw, truth$beta_occ_wtn,
                -1.5, truth$beta_pos_btw, truth$beta_pos_wtn,
                truth$alpha, truth$sigma),
  beta_fit  = c(fit_beta$beta_occ,
                fit_beta$beta_pos,
                unname(fit_beta$joint$theta_mean["alpha"]),
                unname(fit_beta$joint$theta_mean["sigma_occ"])),
  ln_fit    = c(fit_ln$beta_occ,
                fit_ln$beta_pos,
                unname(fit_ln$joint$theta_mean["alpha"]),
                unname(fit_ln$joint$theta_mean["sigma_occ"]))
)
print(recov, row.names = FALSE, digits = 3)
```

The within-plot slope on the positive arm, the quantity the resurvey
literature actually wants, recovers tightly under both positive families
(truth +0.05). The between-plot slopes are noisier because 25 plot-level means
is a small effective sample for that quantity, and the realised BYM2 field
absorbs into both the intercept and the between-slope estimates.

## Within-plot trajectories

The within slope is what the resurvey-monitoring literature actually wants:
holding the plot fixed, how does cover change with visit time?

```{r predict, fig.height = 3.8}
focus_plots <- levels(dat_beta$plot_id)[c(1, 6, 13, 20)]
focus       <- dat_beta[dat_beta$plot_id %in% focus_plots, ]

focus$pred_beta <- predict(fit_beta, focus, type = "expected")$mean
focus$pred_ln   <- predict(fit_ln,   focus, type = "expected")$mean

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (variant in c("pred_beta", "pred_ln")) {
  plot(NA, NA,
       xlim = range(focus$year), ylim = c(0, max(focus[[variant]], 0.05)),
       xlab = "visit year", ylab = "E[cover]",
       main = sub("^pred_", "", variant))
  for (p in focus_plots) {
    rows <- focus[focus$plot_id == p, ]
    rows <- rows[order(rows$year), ]
    lines(rows$year, rows[[variant]], type = "b", pch = 16,
          col = which(focus_plots == p))
  }
}
par(op)
```

Each panel shows four representative plots. The within-plot slope is positive in
the simulator (`beta_*_wtn > 0`), so each plot's predicted cover rises across
its four visits regardless of where its baseline sits along the x-axis. The
cross-plot offset between trajectories reflects the joint spatial field at that
plot.

## The three-arm model: `occu_cover()`

The `cover()` family above treats every visit as a clean observation: if a
plot reads zero, the species is absent. Resurvey data is rarely that clean. A
species can be present and missed, so a zero is a mixture of true absence and
non-detection. The `occu_cover()` family adds the missing layer: latent
occupancy $z_i \sim \mathrm{Bernoulli}(\psi_i)$ per cell, per-visit detection
$y_{ij} \mid z_i = 1 \sim \mathrm{Bernoulli}(p_{ij})$, and per-visit cover
$y^{\mathrm{pos}}_{ij} \mid y_{ij} = 1 \sim f$ on a third linear predictor. The
shared field rides on the occupancy state $\psi$ and is copied onto the cover
arm by $\alpha$, exactly as in the two-arm model. It is the analogue of
N-mixture, but for cover instead of counts.

The data shape is the `unmarked`-style site-by-visit matrix. Cell-level
covariates and the spatial graph live in one data frame; the detection history
`y` and the cover history `y_pos` are `n_sites x n_visits` matrices;
visit-level covariates pass through `visits`. The detection formula is its own
argument, the positive-cover formula goes through `positive`, and the
occupancy formula is the main `formula` and carries the `bym2()` term. We build
a fixture with `simulate_occu_cover()`, the generator the recovery tests use,
adding a cell-level `year` covariate on the occupancy arm so we can predict
change over time later.

```{r oc-build, message = FALSE, warning = FALSE}
N <- 40L; J <- 5L
adj <- matrix(0L, N, N)
for (s in seq_len(N)) {
  if (s > 1L) adj[s, s - 1L] <- 1L
  if (s < N)  adj[s, s + 1L] <- 1L
}

sim_oc <- simulate_occu_cover(
  N = N, J = J, positive = "lognormal",
  adj = adj, sigma = 0.8, alpha = 0.6, seed = 202L
)
```

The generating parameters here are worth naming, because the checks below test
recovery against them. The field amplitude `sigma = 0.8` sets how strongly the
shared spatial field swings occupancy; `alpha = 0.6` is the copy coefficient,
so the cover arm sees the field at 60% of the occupancy arm's amplitude. The
chain adjacency makes neighbouring plots correlated, which is what lets the
field be identified at all: with no spatial structure the field would be N free
parameters fit to N cells. The simulator draws latent occupancy per cell,
detection per visit given occupancy, and cover per detected visit, so a cell
that is occupied but never detected contributes only zeros to `y` and `NA` to
`y_pos`, exactly the non-detection case the detection arm exists to handle.

`sim_oc` returns `y` (detection history), `y_pos` (cover history, `NA` where not
detected), `data` (cell-level covariates: `occ_cov1`), `visit_data` (visit-level
covariates: `det_cov1`, `pos_cov1`), and `truth` (every generating parameter
plus the field `f`). We reshape to long form for `tobs_data()`, add the `year`
covariate at the cell level, and zero-fill the non-detected cover entries the
binder expects. The `year` covariate is a standardized cell index, a stand-in
for a baseline-year or climate gradient that varies across plots; putting it on
the occupancy arm means moving it changes presence, and the change prediction
later traces how that propagates to cover.

```{r oc-shape}
long <- data.frame(
  site_id  = rep(seq_len(N), each = J),
  visit    = rep(seq_len(J), times = N),
  y        = as.vector(t(sim_oc$y)),
  det_cov1 = sim_oc$visit_data$det_cov1,
  pos_cov1 = sim_oc$visit_data$pos_cov1
)
od <- tobs_data(long, y = "y", site = "site_id", visit = "visit",
                det.covs = c("det_cov1", "pos_cov1"))

year     <- as.numeric(scale(seq_len(N)))
cell_dat <- cbind(data.frame(site_id = seq_len(N)), sim_oc$data, year = year)
y_pos    <- sim_oc$y_pos; y_pos[is.na(y_pos)] <- 0
```

The fit threads the three arms through one call. The occupancy arm gets the
spatial field and the covariates; detection and cover get their own formulas.

```{r oc-fit, message = FALSE, warning = FALSE}
fit_oc <- suppressWarnings(tobs(
  formula   = ~ occ_cov1 + year + bym2(graph = adj),
  data      = cell_dat,
  family    = occu_cover("lognormal"),
  detection = ~ det_cov1,
  positive  = ~ pos_cov1,
  y         = od$y,
  y_pos     = y_pos,
  visits    = od$det.covs,
  method    = "nested_laplace",
  control   = list(verbose = FALSE, max.iter = 300L,
                   sigma.grid = c(0.5, 1.0),
                   alpha.grid = c(0, 0.5, 1.0))
))
```

The fitted object carries the three arms' coefficients in `fit_oc$means`, named
`psi_*` (occupancy), `p_*` (detection), and `pos_*` (cover), with SEs in
`fit_oc$sds`. The copy coefficient and donor amplitude are surfaced as the
hyperparameters `alpha` and `sigma`, and the per-cell field posterior is in
`fit_oc$spatial_field` plus the tidy `fit_oc$field_table`. The raw joint engine
result is `fit_oc$joint_fit`, the object `predict()` samples from.

```{r oc-extract}
coef_tbl <- data.frame(
  parameter = c("psi_occ_cov1", "psi_year", "p_det_cov1", "pos_pos_cov1",
                "alpha", "sigma"),
  estimate  = fit_oc$means[c("psi_occ_cov1", "psi_year", "p_det_cov1",
                             "pos_pos_cov1", "alpha", "sigma")],
  sd        = fit_oc$sds[c("psi_occ_cov1", "psi_year", "p_det_cov1",
                           "pos_pos_cov1", "alpha", "sigma")]
)
print(coef_tbl, row.names = FALSE, digits = 3)
```

The copy coefficient `alpha` was generated at 0.6: the cover arm sees the
spatial field at a little over half the amplitude the occupancy arm does. The
fitted `alpha` carries its own SD, so the question "do the two processes share
spatial structure" gets a credible interval rather than a yes/no.

## Checking the fit

A fitted field is only useful if it tracks the truth and if its uncertainty is
honest. Two checks cover this. First, the recovered field should correlate with
the latent field that generated the data, which the simulator hands back as
`sim_oc$truth$f`. Second, the in-sample occurrence probabilities the joint
posterior produces should agree with a plug-in predictor built from the
posterior-mean coefficients and field, which confirms the field amplitude is on
the right scale rather than collapsed or inflated.

```{r oc-check-field}
field_cor <- cor(fit_oc$spatial_field, sim_oc$truth$f)
field_cor
```

```{r oc-check-plugin}
pr_occ <- predict(fit_oc, type = "occurrence", nsim = 2000L)

X     <- model.matrix(~ occ_cov1 + year, cell_dat)
b     <- fit_oc$means[c("psi_(Intercept)", "psi_occ_cov1", "psi_year")]
eta   <- as.numeric(X %*% b) + fit_oc$spatial$sigma_mean * fit_oc$spatial_field
p_plug <- plogis(eta)

c(correlation = cor(pr_occ$mean, p_plug),
  max_abs_gap = max(abs(pr_occ$mean - p_plug)))
```

The field correlation runs high (a shared chain field is well-identified at
N = 40), and the plug-in agreement is tight: the small residual gap is Jensen's
inequality acting through the `plogis` link, not a scaling error. A broken
amplitude factor would crater both numbers. The plug-in check is the more
diagnostic of the two, because it compares the marginalized posterior mean
(integrated over the grid and the field) against a point reconstruction at the
posterior-mean field and coefficients. If the engine had the field amplitude
wrong, the two would diverge sharply; if it were collapsing the field to zero,
the correlation with the truth would fall apart. Both holding is evidence the
nested integration is on scale. The field posterior itself, with its per-cell
uncertainty, plots straight from `fit_oc$field_table`, whose columns are
`cell`, `z_mean`, `z_sd`, and the 95% bounds `z_lower` / `z_upper`.

```{r oc-field-plot, message = FALSE, fig.height = 3.5}
library(ggplot2)
ft <- fit_oc$field_table
ggplot(ft, aes(cell, z_mean)) +
  geom_ribbon(aes(ymin = z_lower, ymax = z_upper), alpha = 0.2) +
  geom_line() + geom_point(size = 0.8) +
  labs(x = "plot (cell)", y = "shared field z (posterior)") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The ribbon is the 95% band on the standardized field per cell. Cells where the
band excludes zero are where the spatial signal is real; cells where it
straddles zero are where presence is driven by the covariates alone. The field
varies smoothly along the chain because adjacent plots share the ICAR penalty.

## Interpreting results

The fitted coefficients are credible-interval objects, not point estimates. The
joint engine returns a posterior mean and SD per parameter, so a Wald interval
on each arm reads straight off `means` and `sds`. The occupancy slope on `year`
is the change in latent occupancy logit per standardized year; the detection
slope on `det_cov1` is the change in per-visit detectability; the cover slope on
`pos_cov1` is the change in conditional log-cover.

```{r oc-ci}
pars <- c("psi_year", "p_det_cov1", "pos_pos_cov1", "alpha")
ci <- data.frame(
  parameter = pars,
  mean      = fit_oc$means[pars],
  lwr       = fit_oc$means[pars] - 1.96 * fit_oc$sds[pars],
  upr       = fit_oc$means[pars] + 1.96 * fit_oc$sds[pars]
)
print(ci, row.names = FALSE, digits = 3)
```

The map of interest is usually expected cover per cell, which combines all three
arms: $E[y] = \psi \cdot \mu$ where $\mu$ is the conditional cover. `predict()`
returns this as `type = "cover_exp"`, marginalized over the joint posterior so
the uncertainty carries the correlation between arms rather than plugging in
means. The returned table is one row per cell with `mean`, `sd`, `lwr`, `upr`,
ready to join to spatial geometry.

```{r oc-exp-map, fig.height = 3.5}
pr_exp <- predict(fit_oc, newdata = cell_dat, type = "cover_exp", nsim = 1000L)

ggplot(pr_exp, aes(cell, mean)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line() +
  labs(x = "plot (cell)", y = "expected cover E[y]") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The per-cell band is wider where occupancy is uncertain and where the field is
poorly pinned. Reading the map without the band would overstate confidence in
exactly the cells where the data is thinnest. For a real region you would join
`pr_exp` to an `sf` polygon layer on the `cell` key and draw it with
`geom_sf(aes(fill = mean))`, with a companion panel on `sd` or interval width.

Two things make this map honest where a plug-in map would not be. First, every
draw runs the full product $\psi\mu$ before averaging, so the expected-cover
mean is the mean of a product, not the product of means; with a skewed cover
distribution those differ, and the per-draw route is the correct one. Second,
the occupancy and cover arms are correlated through the shared field, and the
draws carry that correlation, so the interval on $\psi\mu$ is neither the naive
sum nor the independent combination of the two arms' intervals. The same logic
is why the change prediction below differences draws rather than differencing
summaries: the credible interval on a change is only right if it is built from
paired draws that share the field realization at both time points.

## Prediction

The resurvey question is not "what is cover now" but "how does cover change". A
covariate that moves over time, here `year` on the occupancy arm, drives a
change in occupancy, and through the shared field and the hurdle structure, a
change in expected cover. `predict(type = "change")` evaluates the model at two
values of the time covariate and differences the draws per cell, so every
change quantity carries a credible interval.

```{r oc-change}
pr_chg <- predict(fit_oc, newdata = cell_dat, type = "change",
                  times = c(0, 1), time_col = "year", nsim = 1000L)
names(pr_chg)
```

The contract is fixed. Three headline deltas come with `.lwr` / `.upr` interval
columns: `delta_p` (change in occupancy probability), `delta_cover_cond`
(change in conditional cover given presence), and `delta_cover_exp` (change in
expected cover $\psi\mu$). Two further columns decompose the expected-cover
change into the part driven by occupancy moving and the part driven by
conditional cover moving, `delta_cover_from_occ` and `delta_cover_from_ab`, and
these sum to `delta_cover_exp` exactly, per draw. The states at each time point
are reported too (`p_T1`, `p_T2`, `cover_exp_T1`, and so on).

```{r oc-change-decomp}
with(pr_chg, head(data.frame(
  cell, delta_p,
  delta_cover_exp,
  from_occ = delta_cover_from_occ,
  from_ab  = delta_cover_from_ab,
  check    = delta_cover_from_occ + delta_cover_from_ab
), 6))
```

The `check` column reproduces `delta_cover_exp` because the additive
decomposition $\Delta(\psi\mu) = (\Delta\psi)\mu_1 + \psi_2(\Delta\mu)$ holds
exactly on every draw. The map of `delta_cover_exp` with its band is the
deliverable: where cover is rising, where it is falling, and where the interval
crosses zero so the change is not credibly distinguishable from none.

```{r oc-change-plot, fig.height = 3.5}
ggplot(pr_chg, aes(cell, delta_cover_exp)) +
  geom_ribbon(aes(ymin = delta_cover_exp.lwr, ymax = delta_cover_exp.upr),
              alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line() +
  labs(x = "plot (cell)", y = "change in expected cover") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

Cells whose band sits entirely above the dashed zero line are credibly
gaining; cells whose band straddles it have no detectable trend. The change is
not uniform across the region because `year` enters through occupancy and the
shared field modulates how much that occupancy change translates into cover.

The decomposition is where the joint model pays off for interpretation. A
rising expected cover can come from the species spreading into new plots
(occupancy up, conditional cover flat) or from the species thickening where it
already is (occupancy flat, conditional cover up), and the two have different
ecological meanings and different management responses. `delta_cover_from_occ`
is the first mechanism and `delta_cover_from_ab` the second, and because they
sum exactly to `delta_cover_exp` on every draw, their ratio is interpretable
cell by cell with its own uncertainty. In this fixture `year` enters only
through occupancy, so the change is dominated by `delta_cover_from_occ`; in a
real fit where a covariate sits on both arms, the split tells you which channel
is doing the work. Reporting only `delta_cover_exp` would collapse that
distinction.

## Model comparison

The coupling has to earn its keep, and the honest comparison is the joint
`occu_cover()` fit against fitting the arms apart. The two-arm `cover()` family
under `method = "laplace"` is the separate-fit baseline: a binomial GLM on
occurrence and a Gaussian (or beta) GLM on the positive cover, with no shared
field and no copy coefficient. We can score that baseline with WAIC because the
hurdle is fully observed and its pointwise log-likelihood is available; the
joint nested-Laplace fit integrates the field out and does not expose a
pointwise likelihood, so we compare it on the copy coefficient and the field
recovery instead.

```{r oc-compare-cover}
# Long form in site-major order so every cover value lines up with its visit's
# covariates (the same row layout `long` uses above).
cover_y <- as.vector(t(sim_oc$y_pos))
cover_y[is.na(cover_y)] <- 0
cover_df <- data.frame(
  occ_cov1 = rep(sim_oc$data$occ_cov1, each = J),
  pos_cov1 = long$pos_cov1
)

fit_sep <- tobs(
  formula = ~ occ_cov1 + pos_cov1,
  data    = cover_df,
  family  = cover("lognormal"),
  y       = pmin(cover_y, 1 - 1e-6),
  method  = "laplace"
)
sep_waic <- tobs_waic(fit_sep)
```

```{r oc-compare-table}
comparison <- data.frame(
  model          = c("joint occu_cover", "independent cover()"),
  shared_field   = c("yes (copy alpha)", "no"),
  detection_arm  = c("yes", "no"),
  alpha_estimate = c(round(fit_oc$means[["alpha"]], 3), NA),
  field_cor      = c(round(field_cor, 3), NA),
  waic           = c(NA, round(sep_waic$waic, 1))
)
print(comparison, row.names = FALSE)
```

The separate fit gives a defensible WAIC for the cover surface on its own terms,
but it cannot answer the question the joint model was built for: it has no
$\alpha$ to report, because it never couples the arms, and it has no detection
layer, so its zeros are taken at face value. The joint fit's `alpha` interval is
the formal test of coupling. If that interval excludes zero, the shared field is
real and the separate fit is discarding information; if it brackets zero, the
arms are effectively independent and the simpler model is adequate. Here `alpha`
sits near its generating value of 0.6 with the field recovered at high
correlation, so the coupling is doing work.

## Practical guidance

A few numbers and rules of thumb that hold up across the recovery suite.

* **At least 60 to 100 spatial units for the field.** The BYM2 amplitude `sigma`
  is a variance component, and Laplace underestimates variance components at
  small N. The recovery tests run at N = 100 and still allow `sigma` a
  tolerance of about 0.9 on a scale near 1.0; at N = 40 the field shape recovers
  (correlation above 0.8) but the amplitude is biased low. Below roughly 30
  units the field is barely identified and `alpha` rides on noise.
* **At least 4 to 6 visits per site for the detection arm.** Occupancy and
  detection are only separable when sites are revisited. The fixtures use J = 5
  or 6. With J = 1 the detection arm is confounded with occupancy and you should
  drop to the two-arm `cover()` family, which assumes perfect detection.
* **Roughly 30+ positive observations per cover-arm coefficient.** The positive
  arm only sees the detected-and-present visits. A cover formula with several
  covariates needs the positive subset to comfortably exceed the coefficient
  count; the encoder hard-errors below `ncol(X) + 1` and is unreliable well
  above that floor.

A decision table for picking the family and method:

| Situation | Family | Method |
|---|---|---|
| Cover in [0,1], detection assumed perfect, no spatial coupling needed | `cover()` | `laplace` |
| Cover with a shared spatial field across the two hurdle arms | `cover()` | `nested_laplace` |
| Repeat visits, imperfect detection, shared field, change-over-time prediction | `occu_cover()` | `nested_laplace` |
| Cover bounded strictly inside (0,1) | `*("beta")` | as above |
| Cover that can be very small but strictly positive, right-skewed | `*("lognormal")` | as above |

On the grid and priors: the outer hyperparameter grid is integrated, not
optimized, so a coarse `sigma.grid` biases `sigma` toward whatever points you
supply. The defaults are reasonable, but for a publication fit widen and refine
`control$sigma.grid` and `control$alpha.grid` and confirm the posterior is
interior to the grid rather than pinned at an edge. The copy coefficient is the
most prior-sensitive quantity at small N: with few positive observations the
data barely constrains `alpha`, and the posterior will track the grid you
specify. Run the fit at two grid resolutions and check that `alpha` and `sigma`
move by less than their SDs; if they swing, the grid is doing the inference and
you need more data or an informative prior on `alpha`.

One more sensitivity worth a sentence: the positive family is not innocuous.
Beta and lognormal place mass differently near zero, and a cover distribution
with many values just above zero is fit better by lognormal, while one
concentrated in the middle of $(0, 1)$ suits beta. Picking the wrong one biases
the conditional-cover slope and the dispersion, though it leaves the occupancy
arm and the copy coefficient largely alone. When in doubt, fit both and compare
the conditional-cover predictions against held-out positive observations.

When NOT to use the joint model: if detection is genuinely perfect (exhaustive
survey, sessile organism counted directly), the detection arm is unidentified
and the three-arm `occu_cover()` will spend parameters chasing a layer that is
not there, so use the two-arm `cover()` instead. If the copy coefficient's
interval brackets zero across grid resolutions, the arms are not coupled and the
shared field buys nothing; fit them independently and report that. And if you
have fewer than about 30 spatial units, the field and `alpha` are too weakly
identified for the joint machinery to pay off; a fixed-effects hurdle with the
spatial structure folded into covariates is more honest.

## What the reduction leaves out

A few features of the real MOTIVATE workflow are deliberately absent:

* **Irregular plot geometry.** Real plots are not chain-adjacent. The joint
  engine handles arbitrary `adj` matrices: swap the chain for the
  nearest-neighbour graph of plot coordinates and the rest of the pipeline is
  unchanged.
* **Observer / method effects.** Real resurveys carry a "visit protocol"
  covariate (taxonomic resolution, season, observer team). Add it as a
  visit-level covariate on the detection formula.
* **Missing visits.** This simulator gives every plot the same number of visits
  at the same spacing. The site-by-visit matrices accept `NA` for missing
  visits, and the binder masks them out of the likelihood.
* **Multiple species.** A single-species fit. Multi-species cover hurdle is a
  separate family.

## References

* Mundlak, Y. (1978). On the pooling of time series and cross section data.
  *Econometrica*, 46(1), 69-85.
* Bell, A. & Jones, K. (2015). Explaining fixed effects: random effects
  modeling of time-series cross-sectional and panel data. *Political Science
  Research and Methods*, 3(1), 133-153.
* `vignette("cover-hurdle")`: the independent-arm cover hurdle without spatial
  coupling.
* `vignette("cover-hurdle-vs-inla")`: separate-fit recovery against
  `INLA::inla()` on non-panel cover data.
