---
title: "Cover hurdle: tulpaObs vs INLA on simulated cover data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle: tulpaObs vs INLA on simulated cover data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
have_inla <- requireNamespace("INLA", quietly = TRUE)
```

`tulpaObs::cover(response = "lognormal")` fits a two-arm hurdle model: a
binomial GLM on the occurrence indicator $\mathbf{1}\{y_i > 0\}$, and a
Gaussian GLM on $\log y_i$ over the positive subset, with the lognormal
back-transform
$E[y_i \mid y_i > 0] = \exp(\eta_i^{\text{pos}} + \sigma^2 / 2)$ applied at
prediction time. The same generative model can be fit in `INLA::inla()` as
two independent calls (Michael Glaser's separate-hurdle workflow in
`example/MOT_abund_data.Rmd` §"INLA hurdles"). This vignette compares the
two on simulated data where the truth is known.

Michael's joint model couples the two arms through one shared spatial
field with INLA's `copy=` mechanism (the cover arm's field is a fitted
scalar multiple of the occurrence arm's). tulpaObs reproduces that
coupling in two places. The closest analogue is the spatial `cover()`
path: a `bym2()` / `icar()` / `car()` term on the latent-presence formula
under `method = "nested_laplace"`, which fits one shared areal field and a
copy coefficient `alpha` across the occurrence and cover arms.
`occu_cover()` is the further generalization that adds a replicated
detection arm on top of that coupling. This vignette starts on the
separate-fit case, where the two engines should agree closely, then walks
the function mapping, two side-by-side fits, the algorithmic differences,
and the cases where INLA remains the better tool.

## Function mapping

The table below maps the INLA constructs in Michael's two workflows to
their tulpaObs counterparts. The left block is the independent-arm path
(two `inla()` calls vs one `cover()` call); the right block is the joint
`copy=` path (one `inla()` list-model vs the spatial `cover()` / detection
`occu_cover()` engine).

| INLA construct | tulpaObs equivalent | Notes |
|---|---|---|
| `inla(occur ~ x, family = "binomial")` | `tobs(~ x, family = cover("lognormal"), y = cover)` (occurrence arm) | One `cover()` call fits both arms; no separate occurrence call. |
| `inla(log_cover ~ x, family = "gaussian", data = subset(occur == 1))` | the positive arm of the same `cover()` fit | tulpaObs subsets to `y > 0` internally inside `encode_cover_hurdle()`. |
| `family = "gaussian"` precision hyperparameter | `fit$sigma_pos` (residual SD, `1 / sqrt(precision)`) | tulpaObs reports the SD directly; INLA reports a precision posterior. |
| `family = c("binomial", "beta")` for cover in $(0, 1)$ | `cover("beta")` | Logit-link beta with precision `phi`; see `fit$phi_pos`. |
| `f(site, model = "bym2", graph = adj)` | `bym2(graph = adj, group_var = "site")` in the formula | Areal field on the latent-presence linear predictor. |
| `f(site, model = "iid")`, `"besag"`, `"besagproper2"` | `icar(graph = adj)`, `car(graph = adj)`, `car_proper(graph = adj)` | One areal term per latent-presence formula. |
| `copy = "site", fixed = FALSE` (shared field, fitted scale) | the joint `cover()` path: `method = "nested_laplace"` with a `bym2()` term, copy coefficient `alpha` | `alpha` is integrated on the outer grid, surfaced as `fit$joint$theta_mean["alpha"]`. |
| `inla.stack()` for the joint design | not needed | tulpaObs builds the two-arm design from one formula and one `y`. |
| `control.compute = list(waic = TRUE)` | `tobs_waic(fit)` | WAIC on the assembled pointwise log-likelihood. |
| `inla.posterior.sample()` + manual differencing | `predict(fit, newdata, type = "expected")` | Posterior draws are differenced per draw inside `predict()`. |
| `pc.prec` / `pc` hyperpriors via `hyper = list(...)` | `cover_priors()` (fixed effects); `control$prior.sigma`, `control$prior.alpha` (joint hyperparameters) | Fixed-effect priors on both arms; PC-style amplitude / copy priors on the joint grid. |
| replicated detection: `inla()` has no native single-species detection arm | `occu_cover()` with a `detection` formula and an `N x J` visit matrix | The generalization INLA cannot express directly (see `example/info.md`). |

## Simulate

```{r simulate}
library(tulpaObs)

sim <- simulate_cover(
  N         = 400,
  beta_occ  = c(-0.4, 0.8),
  beta_pos  = c(-1.0, 0.3),
  sigma_pos = 0.4,
  seed      = 11
)

dat <- sim$data
dat$occur     <- as.integer(dat$cover > 0)
dat$log_cover <- ifelse(dat$cover > 0, log(dat$cover), NA_real_)

truth <- sim$truth
c(sum(dat$occur), mean(dat$occur))
```

The simulator's `truth$beta_occ`, `truth$beta_pos`, and `truth$sigma_pos`
give the parameters used to generate `cover`. Site-level prevalence
`mean(occur)` lands near `mean(plogis(eta_occ))` once enough sites are
drawn. The generative model is exactly the one both engines target: a
Bernoulli occurrence layer with a logit link, and a Gaussian layer on
`log(cover)` for the occupied sites. Because the two layers are simulated
independently here (no shared field), the separate-fit and the
two-`inla()`-call workflows are fitting the same likelihood, and any gap in
the recovery table is approximation error rather than model misspecification.

## Fit with tulpaObs

```{r fit-tulpaobs}
fit_tulpa <- tobs(
  formula = ~ x,
  data    = dat,
  family  = cover(response = "lognormal"),
  y       = dat$cover
)
fit_tulpa
```

## Fit with INLA (Michael's separate-hurdle workflow)

```{r fit-inla, eval = have_inla}
library(INLA)

m_occ_inla <- INLA::inla(
  occur ~ x,
  family = "binomial",
  data   = dat,
  control.compute = list(waic = TRUE)
)

m_pos_inla <- INLA::inla(
  log_cover ~ x,
  family = "gaussian",
  data   = subset(dat, occur == 1),
  control.compute = list(waic = TRUE)
)

inla_beta_occ <- m_occ_inla$summary.fixed[, "mean"]
inla_beta_pos <- m_pos_inla$summary.fixed[, "mean"]

# Gaussian-arm residual SD from INLA's hyperpar (1 / sqrt(precision)).
inla_sigma_pos <- 1 / sqrt(
  m_pos_inla$summary.hyperpar["Precision for the Gaussian observations", "mean"]
)
```

```{r fit-inla-skip, eval = !have_inla, echo = !have_inla}
# INLA is not installed in this environment; the comparison below is
# skipped. To run the full vignette: install.packages("INLA",
# repos = "https://inla.r-inla-download.org/R/stable").
```

## Parameter recovery

```{r recover, eval = have_inla}
recovery <- data.frame(
  parameter = c("beta_occ_0", "beta_occ_x",
                "beta_pos_0", "beta_pos_x",
                "sigma_pos"),
  truth     = c(truth$beta_occ, truth$beta_pos, truth$sigma_pos),
  tulpaObs  = c(fit_tulpa$beta_occ, fit_tulpa$beta_pos, fit_tulpa$sigma_pos),
  INLA      = c(inla_beta_occ,    inla_beta_pos,    inla_sigma_pos)
)
recovery$delta_tulpa <- recovery$tulpaObs - recovery$truth
recovery$delta_inla  <- recovery$INLA     - recovery$truth
recovery
```

Both engines hit the same fixed-effects estimates to within the
Laplace/MAP vs. INLA-integrated marginal-mean tolerance. The two
discrepancies that matter:

* `sigma_pos` — tulpaObs computes the Gaussian residual SD post-hoc as
  `sqrt(SS_res / (n - p))` because `tulpa::tulpa_laplace(family = "gaussian")`
  treats `phi = 1` internally. INLA estimates the precision as a
  hyperparameter and reports its posterior mean. The two converge as
  `n_positive` grows.
* `beta_pos_0` — sensitive to the same Gaussian-precision detail; the
  intercept absorbs whatever residual scale the engine attributes to the
  data.

## Predictive equivalence

```{r predict, eval = have_inla}
newdata <- data.frame(x = seq(-2, 2, length.out = 50))

pred_tulpa <- data.frame(
  x       = newdata$x,
  E_cover = predict(fit_tulpa, newdata, type = "expected")
)

# INLA reconstruction: p * exp(eta_pos + sigma^2 / 2).
X <- model.matrix(~ x, newdata)
eta_occ_inla <- as.numeric(X %*% inla_beta_occ)
eta_pos_inla <- as.numeric(X %*% inla_beta_pos)
p_inla       <- plogis(eta_occ_inla)
mu_inla      <- exp(eta_pos_inla + inla_sigma_pos^2 / 2)
pred_inla    <- data.frame(x = newdata$x, E_cover = p_inla * mu_inla)

cor(pred_tulpa$E_cover, pred_inla$E_cover)
max(abs(pred_tulpa$E_cover - pred_inla$E_cover))
```

Both engines reconstruct expected cover the same way, `E[y] = p * mu` with
`mu = exp(eta_pos + sigma^2 / 2)`, so the curves track each other across the
covariate range. The correlation is effectively one, and the maximum
absolute gap traces back to the `sigma_pos` difference noted above: a small
shift in the residual SD feeds through the lognormal back-transform and
scales `mu`. On 400 sites with prevalence near 0.5 the gap stays a few
percent of the predicted mean; it shrinks as the positive subset grows.

```{r predict-plot, eval = have_inla, fig.height = 3.5}
plot(pred_tulpa$x, pred_tulpa$E_cover, type = "l", lwd = 2,
     xlab = "x", ylab = "E[cover]",
     main = "Expected cover: tulpaObs (black) vs INLA (red)")
lines(pred_inla$x, pred_inla$E_cover, col = 2, lwd = 2, lty = 2)
```

## The joint copy= model side by side

The separate fits above ignore any shared structure between the two arms.
Michael's `mod.joint` model assumes the occurrence and cover signals
respond to the same latent spatial field, and ties them with `copy=`. The
tulpaObs analogue is the spatial `cover()` path: add a `bym2()` term to the
formula and switch to `method = "nested_laplace"`. The engine identifies
one field `z` per region and writes `eta_occ = X beta_occ + sigma * z`,
`eta_pos = X beta_pos + alpha * sigma * z`, integrating the copy
coefficient `alpha` on the outer hyperparameter grid.

```{r joint-simulate}
n_s <- 25L
adj <- matrix(0L, n_s, n_s)
for (s in seq_len(n_s)) {
  if (s > 1L)   adj[s, s - 1L] <- 1L
  if (s < n_s)  adj[s, s + 1L] <- 1L
}

sim_joint <- simulate_cover_joint(
  N        = 300L,
  adj      = adj,
  beta_occ = c(-0.3, 0.7),
  beta_pos = c(-1.5, -0.5),
  sigma    = 0.6,
  rho      = 0.7,
  alpha    = 1.0,
  positive = "lognormal",
  seed     = 21L
)
```

The `simulate_cover_joint()` generator demeans both BYM2 sub-blocks before
scaling, matching the sum-to-zero constraint the engine applies, so the
recovered cover-arm intercept targets the population truth rather than a
field-mean offset. The cover-arm intercept sits well below zero so the
lognormal median stays under the cover ceiling of 1: the generator caps its
draws there, and an intercept near or above the ceiling censors most of the
positive responses and flattens the cover-arm slope. The truth carries
`sigma = 0.6`, `alpha = 1.0` (the two
arms see the field identically), and `sigma_pos = alpha * sigma`. The same
demeaning matters in INLA: a `bym2` field there is also sum-to-zero
constrained, so a simulator that skips the demean biases the intercept in
INLA the same way it does here. The constrained field causes it; the
behaviour is identical across both implementations.

```{r joint-fit-tulpa, message = FALSE, warning = FALSE}
fit_joint <- tobs(
  formula = ~ x + bym2(graph = adj, group_var = "region"),
  data    = sim_joint$data,
  family  = cover("lognormal"),
  y       = sim_joint$y,
  method  = "nested_laplace",
  control = list(
    sigma.grid = c(0.4, 0.8),
    rho.grid   = c(0.5, 0.9),
    alpha.grid = c(0.25, 0.75, 1.5, 2.5),
    verbose    = FALSE
  )
)

joint_recovery <- data.frame(
  parameter = c("beta_occ_x", "beta_pos_x", "alpha", "sigma"),
  truth     = c(sim_joint$truth$beta_occ[2], sim_joint$truth$beta_pos[2],
                sim_joint$truth$alpha,       sim_joint$truth$sigma),
  estimate  = c(fit_joint$beta_occ[2], fit_joint$beta_pos[2],
                unname(fit_joint$joint$theta_mean["alpha"]),
                unname(fit_joint$joint$theta_mean["sigma"]))
)
joint_recovery
```

The occurrence and cover slopes land on the right side of zero (truth +0.7
and -0.5), and `alpha` recovers near its true 1.0. The estimate of
`sigma` reads off the field amplitude on the occurrence arm; the cover arm
sees the same field at `alpha * sigma`. With only
25 regions the field-level quantities (`alpha`, `sigma`) carry visible
sampling noise; the fixed-effect slopes, which pool over all 300 sites,
recover more tightly. This is the same small-region behaviour Michael's
MOTIVATE fit shows, and it is intrinsic to the data, not to the engine.

The same model in INLA is a single `inla()` call over an `inla.stack()`
that lays the binomial and Gaussian responses on a shared index, with the
cover arm's field declared as a `copy` of the occurrence arm's. The copy
weight is the fitted scalar `Beta for ...` hyperparameter, the direct
counterpart of tulpaObs's `alpha`.

```{r joint-inla, eval = FALSE}
# Joint shared-field hurdle in INLA (Michael's mod.joint pattern). Not run
# here (INLA may be absent at build); shown for the structural mapping.
#
# n <- nrow(sim_joint$data)
# occur     <- as.integer(sim_joint$y > 0)
# log_cover <- ifelse(sim_joint$y > 0, log(sim_joint$y), NA_real_)
# site      <- as.integer(sim_joint$data$region)
#
# # Two-column response: occurrence in col 1, log-cover in col 2.
# Y <- matrix(NA_real_, 2 * n, 2)
# Y[1:n,        1] <- occur
# Y[(n + 1):(2 * n), 2] <- log_cover
#
# stk <- inla.stack(
#   data = list(Y = Y),
#   A    = list(1),
#   effects = list(list(
#     occ_int = c(rep(1, n), rep(NA, n)),
#     pos_int = c(rep(NA, n), rep(1, n)),
#     x_occ   = c(sim_joint$data$x, rep(NA, n)),
#     x_pos   = c(rep(NA, n), sim_joint$data$x),
#     s_occ   = c(site, rep(NA, n)),
#     s_pos   = c(rep(NA, n), site)
#   ))
# )
#
# joint <- inla(
#   Y ~ -1 + occ_int + pos_int + x_occ + x_pos +
#       f(s_occ, model = "bym2", graph = adj) +
#       f(s_pos, copy = "s_occ", fixed = FALSE),
#   family = c("binomial", "gaussian"),
#   data   = inla.stack.data(stk),
#   control.predictor = list(A = inla.stack.A(stk))
# )
# joint$summary.hyperpar["Beta for s_pos", "mean"]   # the copy coefficient
```

The structural difference is visible in the two code blocks. tulpaObs reads
one formula and one response vector; the two-arm design, the
subset-to-positive, and the shared-field bookkeeping happen inside the
dispatcher. INLA needs the response stacked into a two-column matrix, each
fixed effect split into an occurrence column and a cover column, and the
field declared once and copied. The `occu_cover()` family extends the
tulpaObs side one step further: it adds an `N x J` detection matrix and a
`detection` formula so each cell's presence is observed through replicated
visits, a structure the `copy=` stack does not express because INLA has no
native single-species detection arm.

## Key differences

**Algorithm.** INLA runs nested Laplace approximation throughout: a Laplace
approximation to the latent field conditional on hyperparameters, then a
grid or CCD integration over the hyperparameters, with no exact-MCMC step.
The separate `cover()` path runs two independent Laplace approximations
through `tulpa::tulpa_laplace()` (binomial on occurrence, Gaussian on
log-cover); the joint `cover()` path runs `tulpa::tulpa_nested_laplace_joint()`,
which is the same nested-Laplace recipe INLA uses, integrating `(sigma,
alpha, phi)` on an outer grid with an inner Newton solve per cell. The two
engines therefore agree closely on the means for these Gaussian-latent
blocks, which is what the recovery table shows.

**Speed.** For the separate-fit case both engines finish in well under a
second on 400 sites; the cost is dominated by the design-matrix algebra,
not the optimizer. For the joint shared-field case the cost is the outer
grid size: the example above pins a small one, which keeps the fit fast,
but a dense default grid with adaptive densification runs longer.
tulpaObs exposes the grid directly through `control$sigma.grid`,
`control$rho.grid`, and `control$alpha.grid`, so the time-accuracy
trade is in the user's hands.

**Output format.** INLA returns full posterior marginals for every fixed
effect and hyperparameter (`summary.fixed`, `summary.hyperpar`, plus the
marginal densities). tulpaObs returns a `cover_fit` object carrying
`beta_occ` / `beta_pos` with Hessian-based or grid-integrated SEs, the
dispersion summary (`sigma_pos` or `phi_pos`), and posterior draws reachable
through `predict()`. The joint fit additionally exposes the outer-grid
hyperparameter posterior under `fit$joint$theta_mean` and
`fit$joint$theta_grid`, where `alpha` and `sigma_occ` live. The grid is the
explicit object INLA hides behind its CCD integrator: each row of
`theta_grid` is one hyperparameter cell, and `fit$joint$weights` gives its
posterior mass. Reading `theta_mean` is a weighted average over those cells,
so a flat grid signals an under-resolved posterior the same way a
concentrated one signals a well-identified hyperparameter. When the team's
question is the copy coefficient itself (does the cover signal track the
occurrence signal, and by how much), `alpha` and its grid are the answer
the joint fit was built to give.

**Prior specification.** INLA places PC priors on the hyperparameters
through the `hyper = list(...)` argument of each `f()` term and Gaussian
priors on fixed effects through `control.fixed`. tulpaObs splits the same
two concerns: `cover_priors()` carries the fixed-effect Gaussian penalty for
both arms, and the joint hyperparameters take PC-style priors through
`control$prior.sigma` (field amplitude) and `control$prior.alpha` (the copy
coefficient). The `alpha` prior is the direct regularizer on the copy scalar
at small `n_positive`, the role INLA's `copy` hyperprior plays.

**Shared-field coupling.** This is the substantive modelling difference.
INLA's `copy=` declares the cover arm's field to be a scalar multiple of the
occurrence arm's: the field pattern is locked across arms and only the
magnitude `Beta for ...` is fitted. The tulpaObs joint `cover()` path
encodes the identical structure with `alpha`, `eta_pos = X beta_pos + alpha
* sigma * z`, and integrates `alpha` over the outer grid rather than
optimizing a single hyperparameter mode. The separate-fit `cover()` path
used in the recovery table does not couple the arms at all, which is why it
is the right baseline for the agreement check and the wrong tool when the
shared-field signal is real. `occu_cover()` keeps the same `alpha` coupling
and adds the detection arm on top.

## When NOT to switch

INLA is the better tool in several honest cases, and the migration is not
worth the effort when one of these holds.

**You need full posterior marginals, not summaries.** INLA hands back the
marginal density of every fixed effect and hyperparameter, ready for
`inla.tmarginal()` transforms, tail-probability queries, and exact credible
intervals on derived quantities. tulpaObs returns posterior draws and
moment summaries; if your downstream work differences or transforms
marginals heavily and you want INLA's smoothed densities rather than draws,
staying in INLA avoids re-deriving that layer.

**Your latent structure is one INLA already ships and tulpaObs does not.**
INLA's `f()` library covers SPDE continuous-space fields, `rw1` / `rw2`
smoothers, `ar1`, `seasonal`, `rgeneric` user blocks, and a large catalogue
of built-in models. The tulpaObs cover engine couples areal `bym2()` /
`icar()` / `car()` / `car_proper()` fields and the Mundlak within-between
temporal decomposition. If your model needs a structure outside that set,
INLA expresses it today and the tulpaObs port would be new engine code.

**You are coupling more than two arms, or arms tulpaObs does not model.**
The `copy=` mechanism generalizes to many linear predictors sharing many
fields with independent copy weights. The tulpaObs joint engine couples the
occurrence and cover arms (and, in `occu_cover()`, the detection arm)
through the shared field; a three-or-more-likelihood integrated model with
several copied fields is INLA territory.

**Your team's pipeline is already INLA.** Michael's MOTIVATE workflow
runs `inla.stack()`, `inla.posterior.sample()`, and the `copy=` joint fit
end to end. If the goal is to reproduce or extend that exact pipeline rather
than start fresh, the cost of rebuilding the stack-and-sample plumbing in
tulpaObs usually outweighs the gain. The migration pays off when you want
the replicated detection arm, exact-MCMC debiasing of the Laplace body, or
the single-formula interface, none of which the `copy=` stack provides.

## When the two diverge

The two engines diverge in three places that the separate-fit story
doesn't surface:

1. **Joint shared spatial field.** Michael's `mod.joint`/`copy=` model
   forces the cover linear predictor's spatial component to be a scalar
   multiple of the occurrence linear predictor's -- pattern locked,
   magnitude fitted. The separate-fit `cover()` path compared in the
   recovery table does not couple the two arms; the spatial `cover()` joint
   path shown above (a fitted copy coefficient `alpha` on a shared field) is
   the coupled analogue, and `occu_cover()` adds a detection arm on top.
2. **Beta positive part.** Michael's joint model uses
   `family = c("binomial", "beta")` for cover in $(0, 1)$ on the natural
   scale; `cover("beta")` and `occu_cover("beta")` are the tulpaObs
   equivalents.
3. **Uncertainty propagation.** INLA returns full posterior marginals on
   both arms; the separate-fit path here returns the Laplace mode and a
   Hessian-based SE per arm. The two agree on the means within sampling
   noise but diverge on tail behaviour for small `n_positive`.

## References

* `example/MOT_abund_data.Rmd` §"INLA hurdles" — Michael Glaser's
  separate-hurdle and joint-`copy=` workflows on the MOTIVATE resurvey
  data.
* `example/info.md` — Michael's note on why a proper single-species
  occupancy model is hard in INLA (the marginalization problem).
* `vignette("cover-hurdle-motivate")` — the joint shared-field cover model
  and the beta positive variant on a resurvey panel.
* `vignette("cover-hurdle")` — the independent-arm dispatcher in isolation.
