---
title: "Cover hurdle: tulpaObs vs INLA on simulated cover data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle: tulpaObs vs INLA on simulated cover data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
have_inla <- requireNamespace("INLA", quietly = TRUE)
```

`tulpaObs::cover(positive = "lognormal")` fits a two-arm hurdle model: a
binomial GLM on the occurrence indicator $\mathbf{1}\{y_i > 0\}$, and a
Gaussian GLM on $\log y_i$ over the positive subset, with the lognormal
back-transform
$E[y_i \mid y_i > 0] = \exp(\eta_i^{\text{pos}} + \sigma^2 / 2)$ applied at
prediction time. The same generative model can be fit in `INLA::inla()` as
two independent calls (Michael Glaser's separate-hurdle workflow in
`example/MOT_abund_data.Rmd` §"INLA hurdles"). This vignette compares the
two on simulated data where the truth is known.

The joint shared-spatial-field variant Michael uses in his
`mod.joint`/`copy=` model is not included here — that lands in Phase 1c
of the tulpaObs roadmap (`PLAN_tulpaObs.md` §7) when
`tulpa_nested_laplace()` gains multi-likelihood support. For the
separate-fit case the two approaches should agree closely; this vignette
documents what "closely" means.

## Simulate

```{r simulate}
library(tulpaObs)

sim <- simulate_cover_hurdle(
  N         = 400,
  beta_occ  = c(-0.4, 0.8),
  beta_pos  = c(-1.0, 0.3),
  sigma_pos = 0.4,
  seed      = 11
)

dat <- sim$data
dat$occur     <- as.integer(dat$cover > 0)
dat$log_cover <- ifelse(dat$cover > 0, log(dat$cover), NA_real_)

truth <- sim$truth
c(sum(dat$occur), mean(dat$occur))
```

The simulator's `truth$beta_occ`, `truth$beta_pos`, and `truth$sigma_pos`
give the parameters used to generate `cover`. Site-level prevalence
`mean(occur)` lands near `mean(plogis(eta_occ))` once enough sites are
drawn.

## Fit with tulpaObs

```{r fit-tulpaobs}
fit_tulpa <- tobs(
  formula = ~ x,
  data    = dat,
  family  = cover(positive = "lognormal"),
  y       = dat$cover
)
fit_tulpa
```

## Fit with INLA (Michael's separate-hurdle workflow)

```{r fit-inla, eval = have_inla}
library(INLA)

m_occ_inla <- INLA::inla(
  occur ~ x,
  family = "binomial",
  data   = dat,
  control.compute = list(waic = TRUE)
)

m_pos_inla <- INLA::inla(
  log_cover ~ x,
  family = "gaussian",
  data   = subset(dat, occur == 1),
  control.compute = list(waic = TRUE)
)

inla_beta_occ <- m_occ_inla$summary.fixed[, "mean"]
inla_beta_pos <- m_pos_inla$summary.fixed[, "mean"]

# Gaussian-arm residual SD from INLA's hyperpar (1 / sqrt(precision)).
inla_sigma_pos <- 1 / sqrt(
  m_pos_inla$summary.hyperpar["Precision for the Gaussian observations", "mean"]
)
```

```{r fit-inla-skip, eval = !have_inla, echo = !have_inla}
# INLA is not installed in this environment; the comparison below is
# skipped. To run the full vignette: install.packages("INLA",
# repos = "https://inla.r-inla-download.org/R/stable").
```

## Parameter recovery

```{r recover, eval = have_inla}
recovery <- data.frame(
  parameter = c("beta_occ_0", "beta_occ_x",
                "beta_pos_0", "beta_pos_x",
                "sigma_pos"),
  truth     = c(truth$beta_occ, truth$beta_pos, truth$sigma_pos),
  tulpaObs  = c(fit_tulpa$beta_occ, fit_tulpa$beta_pos, fit_tulpa$sigma_pos),
  INLA      = c(inla_beta_occ,    inla_beta_pos,    inla_sigma_pos)
)
recovery$delta_tulpa <- recovery$tulpaObs - recovery$truth
recovery$delta_inla  <- recovery$INLA     - recovery$truth
recovery
```

Both engines hit the same fixed-effects estimates to within the
Laplace/MAP vs. INLA-integrated marginal-mean tolerance. The two
discrepancies that matter:

* `sigma_pos` — tulpaObs computes the Gaussian residual SD post-hoc as
  `sqrt(SS_res / (n - p))` because `tulpa::tulpa_laplace(family = "gaussian")`
  treats `phi = 1` internally. INLA estimates the precision as a
  hyperparameter and reports its posterior mean. The two converge as
  `n_positive` grows.
* `beta_pos_0` — sensitive to the same Gaussian-precision detail; the
  intercept absorbs whatever residual scale the engine attributes to the
  data.

## Predictive equivalence

```{r predict, eval = have_inla}
newdata <- data.frame(x = seq(-2, 2, length.out = 50))

pred_tulpa <- data.frame(
  x       = newdata$x,
  E_cover = predict(fit_tulpa, newdata, type = "expected")
)

# INLA reconstruction: p * exp(eta_pos + sigma^2 / 2).
X <- model.matrix(~ x, newdata)
eta_occ_inla <- as.numeric(X %*% inla_beta_occ)
eta_pos_inla <- as.numeric(X %*% inla_beta_pos)
p_inla       <- plogis(eta_occ_inla)
mu_inla      <- exp(eta_pos_inla + inla_sigma_pos^2 / 2)
pred_inla    <- data.frame(x = newdata$x, E_cover = p_inla * mu_inla)

cor(pred_tulpa$E_cover, pred_inla$E_cover)
max(abs(pred_tulpa$E_cover - pred_inla$E_cover))
```

```{r predict-plot, eval = have_inla, fig.height = 3.5}
plot(pred_tulpa$x, pred_tulpa$E_cover, type = "l", lwd = 2,
     xlab = "x", ylab = "E[cover]",
     main = "Expected cover: tulpaObs (black) vs INLA (red)")
lines(pred_inla$x, pred_inla$E_cover, col = 2, lwd = 2, lty = 2)
```

## When the two diverge

The two engines diverge in three places that the separate-fit story
doesn't surface:

1. **Joint shared spatial field.** Michael's `mod.joint`/`copy=` model
   forces the cover linear predictor's spatial component to be a scalar
   multiple of the occurrence linear predictor's. Pattern is locked,
   magnitude is fitted. tulpaObs's Phase 1a cover-hurdle does not couple
   the two arms at all. The joint engine arrives in Phase 1c
   (`tulpa_nested_laplace()` multi-likelihood).
2. **Beta positive part.** Michael's earlier joint model uses
   `family = c("binomial", "beta")` for cover in $(0, 1)$ on the natural
   scale. tulpaObs's Phase 1d will add `cover(positive = "beta")` once
   `tulpa_laplace()` supports the beta likelihood (Phase 1b).
3. **Uncertainty propagation.** INLA returns full posterior marginals on
   both arms; tulpaObs returns the Laplace mode and a Hessian-based SE
   per arm. The two agree on the means within sampling noise but diverge
   on tail behaviour for small `n_positive`.

## References

* `example/MOT_abund_data.Rmd` §"INLA hurdles" — Michael Glaser's
  separate-hurdle and joint-`copy=` workflows on the MOTIVATE resurvey
  data.
* `example/info.md` — Michael's note on why a proper single-species
  occupancy model is hard in INLA (the marginalization problem).
* `PLAN_tulpaObs.md` §7 — Phase 1 roadmap for the joint engine and the
  beta positive variant.
