---
title: "Cover hurdle (independent arms)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cover hurdle (independent arms)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette shows the independent-arm cover-hurdle dispatcher on simulated
data. The fit uses two independent Laplace approximations through `tulpa`:

* a binomial GLM on the occurrence indicator $\mathbf{1}\{y_i > 0\}$;
* a Gaussian GLM on $\log y_i$ for the sites where $y_i > 0$.

The shared-field joint model -- occurrence, detection, and cover arms coupled
through one spatial field -- is the `occu_cover()` family;
`vignette("cover-hurdle-motivate")` builds it end to end, including the beta
positive variant. The `INLA::inla()` comparison on the real MOTIVATE data is
sketched below (marked `eval = FALSE`).

```{r simulate}
library(tulpaObs)
sim <- simulate_cover(
  N         = 400,
  beta_occ  = c(-0.4, 0.7),
  beta_pos  = c(-1.0, 0.3),
  sigma_pos = 0.4,
  seed      = 7
)
head(sim$data)
```

```{r fit}
fit <- tobs(
  formula = ~ x,
  data    = sim$data,
  family  = cover(positive = "lognormal"),
  y       = sim$y
)
fit
```

```{r predict}
newdata <- data.frame(x = seq(-2, 2, length.out = 50))
preds <- data.frame(
  x         = newdata$x,
  occupancy = predict(fit, newdata, type = "occupancy"),
  expected  = predict(fit, newdata, type = "expected")
)
head(preds)
```

```{r compare-inla, eval = FALSE}
# Phase 1e: compare against the joint INLA hurdle fit on the MOTIVATE
# resurvey data. Skeleton retained here so the comparison drops in
# without restructuring the vignette.
#
# library(INLA)
# joint <- inla(
#   list(
#     occur ~ x + f(site, model = "bym2", graph = adj),
#     log_cover ~ x + f(site, model = "bym2", graph = adj, copy = "site",
#                       fixed = FALSE)
#   ),
#   family = c("binomial", "gaussian"),
#   data   = inla_stack
# )
```

## Scope

This dispatcher predicts from fixed effects only. Spatial fields fit through
the `spatial` argument, and spatial projection at new coordinates -- together
with the within/between (Mundlak) temporal decomposition -- comes with the
joint `occu_cover()` engine; see `vignette("cover-hurdle-motivate")`.
