---
title: "Shaping data for tobs()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shaping data for tobs()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## Two pieces: the response and the site table

Every `tobs()` call takes a detection response `y` and a per-site `data`
frame. The formula on the left of the call names columns in `data`; the
`detection` formula names columns there too. The shape of `y` is what
changes between families, and getting it right is most of the work.

For single-season occupancy, `y` is an `N x J` matrix: one row per site,
one column per visit, holding `0` (surveyed, not detected), `1`
(detected), or `NA` (not surveyed). The `data` frame has exactly `N`
rows, one per site, aligned to the rows of `y`. A small simulated dataset
shows the layout.

```{r sim-occu}
sim <- simulate_occu(
  N = 120, J = 5,
  n_occ_covs = 1, n_det_covs = 1,
  beta_occ = c(0.2, 0.9),
  beta_det = c(0.5, 0.4),
  seed = 1
)
y   <- sim$y
dat <- sim$data
dim(y)
y[1:4, ]
head(dat, 4)
```

The matrix has 120 rows to match the 120 rows in `dat`. Row `i` of `y` is
the detection history for the site described by row `i` of `dat`. That
alignment is the contract `tobs()` checks first, and it is the one to get
right before anything else.

The other families reshape the same idea:

| Family | `y` shape |
|---|---|
| Single-season occupancy, N-mixture | `N x J` matrix |
| Community (`ms_occu`, `ms_abun`) | `N x J x species` array |
| Dynamic (`dyn_occu`) | `N x J x season` array |
| Integrated (`int_occu`) | list of matrices, one per source |

In every case the leading dimension is the site, the second is the visit,
and `data` carries one row per site.

## Site-level versus visit-level covariates

Occupancy covariates describe a site and do not change between visits, so
they live as plain columns in `data`. Detection covariates can vary by
visit (effort, observer, weather). When `data` has one row per site, a
detection covariate named in the `detection` formula is read from that
single row and broadcast across all `J` visits. That is the common case,
and it is what the simulators produce.

```{r fit-occu}
fit <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
coef(fit)
```

`occ_cov1` and `det_cov1` both came out of `dat`. The occupancy slope used
the value once per site; the detection slope used the same value for each
of the five visits at that site. When a detection covariate genuinely
changes between visits you supply a per-visit design instead of a single
site value, which the `tobs_data()` helper below builds for you from a
long table.

## Building `y` from a long detection table

Field data rarely arrives as a wide matrix. The usual starting point is a
long table with one row per site-visit: a site id, a visit number, and a
0/1 detection. Reshaping that into the `N x J` matrix is a `reshape()` or
a matrix-indexing step.

```{r long-table}
long <- data.frame(
  site     = rep(paste0("S", 1:6), each = 3),
  visit    = rep(1:3, times = 6),
  detected = rbinom(18, 1, 0.4),
  elev     = rep(rnorm(6), each = 3)   # a site-level covariate
)
head(long, 6)
```

The direct route fills a matrix by indexing on the site and visit
positions. Build the empty matrix at the right size, then scatter each
row of the long table into its `[site, visit]` cell.

```{r reshape-manual}
sites  <- sort(unique(long$site))
visits <- sort(unique(long$visit))
ymat   <- matrix(NA_integer_, length(sites), length(visits),
                 dimnames = list(sites, paste0("V", visits)))
si <- match(long$site,  sites)
vi <- match(long$visit, visits)
ymat[cbind(si, vi)] <- as.integer(long$detected)
ymat
```

The site-level table then takes one row per site, in the same site order
as the matrix rows.

```{r reshape-sitetable}
site_dat <- long[match(sites, long$site), "elev", drop = FALSE]
rownames(site_dat) <- NULL
site_dat
```

`tobs_data()` packages both steps. Hand it the long frame and the column
names; it returns a `tobs_data` object carrying the detection matrix, the
site-level covariates, and any visit-level detection covariates as their
own `N x J` matrices.

```{r tobs-data}
obs <- tobs_data(
  long,
  y     = "detected",
  site  = "site",
  visit = "visit",
  occ.covs = "elev"
)
obs
str(obs$y)
```

`obs$y` is the same `6 x 3` matrix, with site ids on the rows. `obs$occ.covs`
holds the per-site `elev` column. A detection covariate passed through
`det.covs` would come back as its own `N x J` matrix rather than a single
column, so a per-visit value lands in the right visit slot.

The `type` argument controls what the response column means. The default,
`"occurrence"`, builds an integer 0/1 matrix; `"abundance"` keeps integer
counts; `"cover"` keeps a continuous proportion in `[0, 1]` without coercing
it to integer. Picking the wrong one is a common quiet error: a cover value of
`0.6` run through the integer path truncates to `0`. The same long frame can
feed an occurrence and a cover model side by side as long as each call names
the matching column and `type`.

```{r tobs-data-cover}
cov_long <- data.frame(
  site  = rep(paste0("S", 1:4), each = 2),
  visit = rep(1:2, times = 4),
  cover = round(runif(8, 0, 0.8), 2)
)
cov_obs <- tobs_data(cov_long, y = "cover", site = "site", visit = "visit",
                     type = "cover")
cov_obs$y
```

## Building a `tobs_data` object by hand

When the matrix and covariates already sit in the right shape, skip the long
frame and assemble the object directly. `tobs_format()` is the low-level
constructor: it wraps a detection matrix, a site-covariate frame, and a named
list of detection-covariate matrices into the `tobs_data` class with no
reshaping. The named elements it expects are `y` (the `N x J` matrix),
`occ.covs` (a per-site data frame), and `det.covs` (a named list of `N x J`
matrices, one per visit-varying covariate).

```{r build-scratch}
N <- 5; J <- 3
y_hand   <- matrix(rbinom(N * J, 1, 0.4), N, J,
                   dimnames = list(paste0("S", 1:N), paste0("V", 1:J)))
occ_hand <- data.frame(elev = round(rnorm(N), 2))
eff_hand <- matrix(round(runif(N * J, 1, 4)), N, J)   # per-visit effort
obs_hand <- tobs_format(y = y_hand, occ.covs = occ_hand,
                        det.covs = list(effort = eff_hand))
obs_hand
str(obs_hand$y)
```

`obs_hand$det.covs$effort` carries one effort value per site-visit cell, so the
detection formula `~ effort` reads a genuinely per-visit covariate rather than a
broadcast site value. The site covariate `elev` stays a single column in
`occ.covs`. A scalar detection covariate that does not vary by visit can still
go through `det.covs` as an `N x J` matrix with the same value down each row, or
through `occ.covs` if it is constant within a site, since `tobs_format()` stores
whatever layout you hand it.

For the array families the named elements are the same except `y` carries an
extra dimension and `tobs_format_ms()` is the constructor. It accepts either a
3D `N x J x species` array or a named list of per-species matrices, and records
`species_names` and `n_species` alongside `y`, `occ.covs`, and `det.covs`.

```{r build-ms-scratch}
sp_list <- lapply(1:3, function(s) matrix(rbinom(N * J, 1, 0.3), N, J))
names(sp_list) <- c("warbler", "finch", "wren")
ms_hand <- tobs_format_ms(sp_list, occ.covs = occ_hand)
dim(ms_hand$y)          # sites x visits x species
ms_hand$species_names
```

The list-of-matrices form is convenient when species come from separate files:
each matrix is one species, the list names become the species dimension. Passing
a pre-built `N x J x species` array works the same way, with `species_names`
supplied separately or defaulted to `sp1`, `sp2`, and so on.

Which constructor to reach for follows from where the data already sits.
`tobs_data()` takes a long site-visit table and does the reshape for you, which
suits raw field exports. `tobs_format()` takes matrices you have already shaped,
which suits data that arrives wide or that you built in an earlier step.
`tobs_format_ms()` is the multi-species counterpart of `tobs_format()`. All
three return the same `tobs_data` class, so the `print()`, `summary()`, and
`plot()` methods and the `tobs()` fitter treat their output identically.

## Inspecting a `tobs_data` object

`tobs_format()` is the lower-level constructor: it wraps an existing
matrix, a site-covariate frame, and a named list of detection-covariate
matrices into the same `tobs_data` class without any reshaping. Reach for
it when your `y` is already a matrix.

```{r tobs-format}
obs2 <- tobs_format(
  y        = y,
  occ.covs = dat["occ_cov1"],
  det.covs = list(det_cov1 = matrix(dat$det_cov1, nrow(y), ncol(y)))
)
obs2
```

The `summary()` method reports the naive occupancy and detection rates,
the distribution of detections per site, and the per-visit detection rate.
Naive occupancy is the fraction of sites with any detection, which sits
below true occupancy because an occupied site can be missed on every visit.

```{r summary-data}
summary(obs2)
```

The `plot()` method draws the same information: detections per site,
detection rate by visit, and visit completeness (the share of sites
actually surveyed on each visit).

```{r plot-data, fig.alt = "Three panels showing detections per site, detection rate by visit, and survey completeness by visit"}
plot(obs2)
```

These summaries are a quick check that the reshape did what you meant. A
visit column that is mostly empty, or a naive occupancy of zero, usually
points to a site or visit id that did not line up during the reshape.

## Missing visits

Sites are not always visited the same number of times. A site surveyed
three times in a five-visit design has two `NA` entries in its row. The
fitter handles this directly: each `NA` becomes a missing marker
internally, and the likelihood sums only over the visits that were
actually observed. No imputation, no dropped sites.

```{r missing}
y_miss <- y
y_miss[1, 4:5] <- NA   # site 1 surveyed 3 times
y_miss[2, 5]   <- NA   # site 2 surveyed 4 times
y_miss[1:3, ]
```

The fit runs unchanged on the gappy matrix, and the estimates stay close
to the truth because the present visits still carry their information.

```{r missing-fit}
fit_miss <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y_miss,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
coef(fit_miss)
```

A site with every visit missing is a special case: with no observations
it contributes nothing to the detection likelihood, and under the spatial
nested-Laplace engine it becomes a held-out site whose occupancy is
interpolated from the latent field. The single-season Laplace fit simply
carries it with no detection information.

## Validation

`tobs()` checks the response shape before it builds anything, so a
malformed input fails with a named error rather than a confusing fit
failure several steps later. The two slips that catch people are a wrong
`y` type and a row-count mismatch between `y` and `data`.

A single-season `y` must be a matrix. Passing a data frame or a bare
vector stops the build with the expected shape spelled out.

```{r err-type, error = TRUE}
tobs(~ occ_cov1, data = dat, family = occu(),
     detection = ~ det_cov1, y = as.data.frame(y),
     method = "laplace", control = list(verbose = FALSE))
```

If the matrix has a different number of rows than `data`, the error quotes
both counts so you can see which side is wrong.

```{r err-rows, error = TRUE}
tobs(~ occ_cov1, data = dat[1:100, , drop = FALSE], family = occu(),
     detection = ~ det_cov1, y = y,
     method = "laplace", control = list(verbose = FALSE))
```

`tobs_data()` validates earlier, while it is still reshaping the long frame.
Naming a covariate column that is not in the frame stops with the missing name
quoted, before any matrix is built.

```{r err-col, error = TRUE}
tobs_data(long, y = "detected", site = "site", visit = "visit",
          occ.covs = "elevation")   # column is named 'elev', not 'elevation'
```

The `type` argument is checked against the response values. An occurrence
response carrying a `2` is rejected, since occurrence data is 0/1; an abundance
response with a negative count and a cover response outside `[0, 1]` fail the
same way against their own domains.

```{r err-type-data, error = TRUE}
bad <- long
bad$detected[1] <- 2
tobs_data(bad, y = "detected", site = "site", visit = "visit",
          type = "occurrence")
```

The community and dynamic builders apply the matching check on their own
shapes. A `dyn_occu()` fit wants a 3D `[site x visit x season]` array, and
handing it a 2D matrix reports the array requirement directly.

```{r err-array, error = TRUE}
tobs(~ 1, data = dat, family = dyn_occu(), detection = ~ 1, y = y,
     colonization = ~ 1, extinction = ~ 1,
     method = "laplace", control = list(verbose = FALSE))
```

The pattern is the same across families: declare the site table and the
response, let the builder reconcile them, and read the error when they do
not match. When a fit refuses to start, checking `dim(y)` against
`nrow(data)` resolves most of it before any inference runs.

## Response shapes per family

The wide-matrix logic extends to the array and list families, and the named
elements stay constant: `y` holds the response, `occ.covs` the per-site
covariate frame, `det.covs` the per-visit covariate matrices. Only the shape of
`y` moves. A single-season `occu()` or `abun()` fit reads an `N x J` matrix; a
community `ms_occu()` or `ms_abun()` fit reads an `N x J x species` array; a
dynamic `dyn_occu()` fit reads an `N x J x season` array; an integrated
`int_occu()` fit reads a list of matrices, one per source, that may differ in
their visit counts. The simulators are the fastest way to see each shape, since
they return a `y` already in the form the matching family expects.

A community model stacks species on a third dimension. `simulate_ms_occu()`
returns an `N x J x species` array, and `species` names that dimension.

```{r ms-shape}
sm <- simulate_ms_occu(N = 80, J = 4, n_species = 6, seed = 3)
dim(sm$y)   # sites x visits x species
```

An integrated model has one detection matrix per data source, with a
shared occupancy state across sources. `simulate_int_occu()` returns `y`
as a list of matrices that differ in their visit counts.

```{r int-shape}
si <- simulate_int_occu(N_total = 100, n_data = 2, J = c(4, 3),
                        n_shared = 20, seed = 42)
vapply(si$y, dim, integer(2))   # each source: sites x its own visit count
```

A dynamic model adds the season as a third dimension.
`simulate_dyn_occu()` returns an `N x J x season` array, and the fit needs
`colonization` and `extinction` for the colonisation and extinction
processes.

```{r dyn-shape}
sd <- simulate_dyn_occu(N = 60, J = 3, n_seasons = 4, seed = 1)
dim(sd$y)   # sites x visits x seasons
```

Each of those families has its own vignette that fits the model end to
end. This one stops at the data shape, which is the part that is the same
whether you go on to a Laplace fit or a full NUTS posterior.

## See also

- `?tobs_data`, `?tobs_format`, `?tobs_format_ms` for the constructors.
- The getting-started vignette for the fit, extract, and predict workflow.
- The family vignettes (`vignette("community-models")`,
  `vignette("integrated-occupancy")`, `vignette("dynamic-occupancy")`) for
  the array and list responses fit end to end.
