---
title: "Diagnostics: checking, comparing, and residuals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics: checking, comparing, and residuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## A fit to check

A coefficient table tells you what the model thinks. It does not tell you
whether the model is any good. This vignette covers the second question:
once `tobs()` returns a fit, how do you decide whether to trust it, how do
you choose between two candidates, and where do you look when something is
off. The tools split into three groups. WAIC ranks models against each
other. Posterior-predictive checks and the simulation-based tests ask
whether the fit reproduces features of the data it saw. Residuals localise
the trouble to particular sites or visits.

The three groups answer different questions, and confusing them is the most
common mistake. Model comparison is relative: it picks the best of the
candidates you supply and says nothing about whether the winner is right in
any absolute sense. Goodness-of-fit checks are absolute, or aim to be: they
compare the fit against the data directly, with no rival model in the
frame. Residual analysis is diagnostic: it does not return a verdict but
points at the sites and visits driving any misfit the other two flagged.
A thorough check uses all three, in that order, because a model can win a
comparison, fail a goodness-of-fit check, and have the failure traced to a
handful of sites by its residuals.

Occupancy adds a wrinkle that single-process models do not have. The data
you observe are detections, but the quantity of interest is occupancy, and
the link between them runs through a latent state you never see directly.
The posterior-predictive check and the PIT residuals both handle this by
conditioning on the latent state given the detection history, rather than
redrawing it from the fitted occupancy probability.

Everything below runs on one occupancy dataset, so the diagnostics stay
comparable. `simulate_occu()` gives us a model with known truth: occupancy
rises with `occ_cov1` (intercept 0.3, slope 1.0), detection depends on
`det_cov1` (intercept 0.7, so detection near 0.67), and six visits per
site keep the two processes apart.

```{r sim}
sim <- simulate_occu(
  N = 300, J = 6,
  n_occ_covs = 1, n_det_covs = 1,
  beta_occ = c(0.3, 1.0),
  beta_det = c(0.7, 0.6),
  seed = 1
)
dat <- sim$data
y   <- sim$y
```

Fit the model the data were generated from. The Laplace engine is the
default; `verbose = FALSE` keeps the EM from printing its iterations.

```{r fit}
fit <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
coef(fit)
```

The coefficients land near the truth. That is a good start and nothing
more. A model can recover its parameters and still fit badly somewhere the
point estimates do not reveal: it might miss a cluster of unusual sites,
underpredict the spread of detections, or leave spatial structure in its
errors, none of which a coefficient table shows. The job of the diagnostics
below is to surface exactly those failures, on a fit whose coefficients
already look right, so that the checks are doing real work rather than
confirming an obviously broken model.

## WAIC and model comparison

`tobs_waic()` computes the widely applicable information criterion from the
pointwise log-likelihood, marginalised over the latent occupancy state. It
returns four numbers.

```{r waic-shape}
w <- tobs_waic(fit)
str(w)
```

`lppd` is the log pointwise predictive density, the in-sample fit summed
over sites. For each site it averages the likelihood across posterior
draws, takes the log, and sums; a model that assigns high likelihood to the
observed history at every site scores high. `p_waic` is the effective
number of parameters, an estimate of model flexibility read off the
variance of the pointwise log-likelihood across draws. The intuition is
that a flexible model has a log-likelihood that swings widely from draw to
draw as the parameters move, so the variance of that swing measures how
much the model can bend to the data. `elpd` is `lppd - p_waic`, the
flexibility-penalised fit, an estimate of the expected log predictive
density on new data drawn from the same process. `waic` is `-2 * elpd` so
that it sits on the deviance scale where lower is better and where a
difference reads like a likelihood-ratio drop.

WAIC trades fit against complexity the way AIC does, with one practical
advantage. It reads the penalty from the posterior rather than counting
parameters by hand, so it handles models where the parameter count is
ambiguous, such as random effects or shrinkage priors that sit between zero
and one effective parameter per group. For the fixed-effect occupancy
models here the two penalties agree closely, but the posterior-based
penalty is what lets the same criterion carry over to the structured
models in the spatial and random-effect vignettes without a special rule
for counting their parameters.

The number means nothing alone. WAIC compares; it does not validate. Build
a small ladder of nested models and read the differences. Here the rungs
are an intercept-only occupancy model, the true single-covariate model,
and an over-specified model that adds a spurious quadratic term the data
were not generated with.

```{r waic-ladder}
dat$occ_cov1_sq <- dat$occ_cov1^2

fit0 <- tobs(~ 1, data = dat, family = occu(),
             detection = ~ det_cov1, y = y, method = "laplace",
             control = list(verbose = FALSE))
fit2 <- tobs(~ occ_cov1 + occ_cov1_sq, data = dat, family = occu(),
             detection = ~ det_cov1, y = y, method = "laplace",
             control = list(verbose = FALSE))

waic_tab <- data.frame(
  model  = c("intercept", "linear", "quadratic"),
  waic   = c(tobs_waic(fit0)$waic, tobs_waic(fit)$waic, tobs_waic(fit2)$waic),
  p_waic = c(tobs_waic(fit0)$p_waic, tobs_waic(fit)$p_waic, tobs_waic(fit2)$p_waic)
)
waic_tab$delta <- round(waic_tab$waic - min(waic_tab$waic), 2)
waic_tab[order(waic_tab$waic), ]
```

The linear model wins, which matches the recipe. Dropping the covariate
costs many WAIC units: the intercept-only model cannot represent the
occupancy slope, so it predicts the data worse at sites with extreme
covariate values. Adding the quadratic term barely moves the criterion.
Its `p_waic` ticks up because the extra coefficient adds flexibility, but
the data do not reward that flexibility, so the penalty roughly cancels the
tiny gain in fit. This is the behaviour you want from a comparison
criterion. A term that mirrors the generating process earns a large drop in
WAIC; a term that merely decorates earns nothing. A `delta` of a few units
or less is a near-tie; prefer the simpler model, since the richer one buys
no predictive accuracy with its extra parameter.

How large a `delta` counts as real depends on the scale. WAIC sits on the
deviance scale, so a drop of two corresponds to the rough threshold AIC
uses for preferring a model, and a drop of ten or more is decisive. The
gaps here run to many tens of units between the intercept model and the
others, which leaves no doubt about the covariate. Narrow margins, a unit
or two, are the cases that warrant caution, especially when the contested
term strains the Laplace approximation that the pointwise log-likelihood is
read from.

One caveat matters for the structured models. The pointwise log-likelihood
behind `tobs_waic()` is evaluated from the process fixed-effect
coefficients. Spatial fields, temporal terms, and random effects are not
folded into the linear predictor for the score, so for a model carrying
those components the WAIC is conditional on the fixed-effect part of the
predictor. For the fixed-effect models compared here that is the whole
model and the score is exact; for a spatial or random-effect model treat
the WAIC as a fixed-effect comparison rather than a full-model one.

What WAIC does not tell you, in any case, is whether the winning model is
calibrated. It ranks the candidates you handed it. If every candidate is
misspecified in the same way, WAIC happily picks the least-bad one and
reports nothing about the shared flaw. The checks in the rest of this
vignette catch that.

## Posterior-predictive checks

A posterior-predictive check simulates fresh detection histories from the
fitted model, computes a discrepancy statistic on the real data and on each
replicate, and asks how often the replicates are more extreme than the
data. The gap between the observed and replicate statistics is the reading;
a smaller gap means the fit reproduces the data more closely.

`tobs_ppc()` offers two discrepancy statistics. The Freeman-Tukey statistic
sums $(\sqrt{\text{obs}} - \sqrt{\text{exp}})^2$ over cells, which tames the
mean-variance link of binary counts. The chi-squared statistic sums
$(\text{obs} - \text{exp})^2 / \text{exp}$ in the usual way. The expected
cell value is $z_i \, p_i$, where $z_i$ is drawn from its full conditional
given the detection history: a site with any detection is occupied, an
all-zero site is occupied with probability
$\psi_i (1 - p_i)^J / [\psi_i (1 - p_i)^J + (1 - \psi_i)]$.

```{r ppc-shape}
ppc <- tobs_ppc(fit, fit.stat = "freeman-tukey")
str(ppc)
```

`fit.y` is the statistic on the real data across draws, `fit.y.rep` the
same on the replicated data, and `bayesian.p` the fraction of draws where
the replicate statistic exceeds the observed one. Plotting the two
distributions against each other shows where the data sit relative to a
draw from the fit.

```{r ppc-plot, fig.alt = "Distributions of the observed and replicate Freeman-Tukey fit statistics, with the observed values sitting above the replicate values"}
plot(density(ppc$fit.y), col = "firebrick", lwd = 2,
     xlim = range(ppc$fit.y, ppc$fit.y.rep),
     xlab = "Freeman-Tukey statistic", main = "Observed vs replicate")
lines(density(ppc$fit.y.rep), col = "steelblue", lwd = 2)
legend("topright", c("observed", "replicate"), lwd = 2,
       col = c("firebrick", "steelblue"), bty = "n")
```

The observed and replicate clouds overlap here, and the Bayesian p-value
sits near the middle of its range, which is what a correctly specified model
produces. The chi-squared statistic reads the same way.

```{r ppc-chisq}
tobs_ppc(fit, fit.stat = "chi-squared")$bayesian.p
```

Conditioning the latent state on the detection history is what keeps the
p-value calibrated. A site with at least one detection is occupied with
probability 1; a site with an all-zero history is occupied with the
posterior probability $\psi_i (1 - p_i)^J / [\psi_i (1 - p_i)^J +
(1 - \psi_i)]$. Sampling $z_i$ from the prior $\psi_i$ instead would force a
truly occupied but $z_i = 0$ draw to register a large observed discrepancy
with no matching replicate term, pulling the p-value toward 0 even when the
model is correct. A value far from 0.5 now flags a real misfit: a p-value
near 0 means the data are more dispersed than the fit predicts, near 1 means
less.

## PIT residuals and uniformity

The probability integral transform offers a sharper, per-site view than a
single discrepancy statistic collapses into one number. The idea is general.
If you evaluate the
cumulative distribution function of a correct predictive model at the value
it actually observed, the result is uniform on $[0, 1]$, because that is
what running a random variable through its own CDF does. Departures from
uniformity are departures of the predictive distribution from the truth,
and the shape of the departure names the flaw.

For each site, `tobs_pit_residuals()` evaluates the fitted cumulative
probability of the observed detection history, averaged over posterior
draws. A site with at least one detection is occupied with certainty, so
its predictive CDF is pinned at the top; a site with all-zero visits gets
the probability mass the model places on the all-zero outcome, which is the
occupied-but-undetected path plus the unoccupied path. A small jitter is
added so that the discreteness of the binary response does not pile the
values onto a few points and break the uniformity test. If the model is
correct, the jittered values are uniform on $[0, 1]$.

```{r pit-shape}
pit <- tobs_pit_residuals(fit)
length(pit); summary(pit)
```

`tobs_test_uniformity()` runs a Kolmogorov-Smirnov test of the PIT values
against the uniform. It returns the standard `htest` object.

```{r pit-test}
tobs_test_uniformity(pit)
```

A large p-value, as here, says the PIT values are consistent with
uniformity, so the model's predictive distribution matches the data across
the range. The histogram makes the shape concrete.

```{r pit-hist, fig.alt = "Histogram of PIT residuals lying roughly flat between zero and one, consistent with the uniform reference"}
hist(pit, breaks = 12, col = "grey80", border = "white",
     xlab = "PIT residual", main = "PIT residuals vs uniform")
abline(h = length(pit) / 12, col = "steelblue", lwd = 2, lty = 2)
```

The shape of a non-uniform PIT histogram tells you what went wrong, and the
mapping is worth committing to memory because it generalises across
families. A U-shape, with mass piling up at both ends, means the predictive
distribution is too narrow: the data fall in the tails more often than the
model expects, the signature of underdispersion. A hump in the middle, the
mirror image, means the predictive distribution is too wide and the data
cluster nearer the centre than the model allows, the signature of
overdispersion. A slope, with mass drifting toward one end, means the
predictive mean is biased that way and a covariate or an offset is missing.
With a flat histogram, as here, the predictive distribution matches the
data across its whole range and there is nothing to chase.

The KS p-value formalises the eye test. It can be conservative when the
response is discrete and the jitter leaves ties, which the function handles
by suppressing the ties warning while keeping the asymptotic p-value, but
the histogram is the part to read first. A test that passes with a histogram
that looks wrong is a sign of low power at the current sample size, not a
clean bill of health.

## Dispersion, zero-inflation, and outliers

Three simulation-based tests target specific failure modes. Each draws
replicate datasets from the fit with `simulate()`, computes a feature on
the real data and on the replicates, and reports where the data sit. They
share a shape: an observed value, an expected value averaged over
replicates, a ratio of the two, and a p-value that is the fraction of
replicates whose feature meets or exceeds the observed one. Because they
simulate from the model rather than redrawing the latent state in
isolation, these tests sidestep the calibration issue that affects the
omnibus posterior-predictive check, and their p-values are closer to honest
tail probabilities. Each one isolates a single named feature, which makes
the verdict easy to act on.

`tobs_test_dispersion()` compares the variance of the per-site detection
totals against what the model predicts. The ratio is the observed variance
over the expected.

```{r dispersion}
tobs_test_dispersion(fit)
```

A ratio near 1 and a p-value away from the extremes mean the spread of
detections matches the fit. A ratio well above 1 is overdispersion: the
detection totals vary more than the model allows, usually because sites
differ in occupancy or detection in ways the covariates do not capture, so
some sites pile up detections while others stay empty beyond what the fit
predicts. The cure is a richer mean structure, a random effect on site, or
a spatial field. A ratio well below 1 is the rarer underdispersion, where
the data are smoother than the model expects. For counts rather than
detections the same diagnostic drives the choice between Poisson and
negative binomial abundance: an overdispersion flag on a Poisson N-mixture
fit is the standard reason to switch to the negative binomial, covered in
the abundance vignette.

`tobs_test_zero_inflation()` counts sites with no detections at all and asks
whether the model produces that many all-zero sites.

```{r zero-inflation}
tobs_test_zero_inflation(fit)
```

The observed count of empty sites matches the expected count, so there is
no excess of zeros. This test matters for occupancy because the model
already produces zeros two ways, through unoccupied sites and through
occupied sites that went undetected, and the question is whether those two
sources together account for all the empty sites in the data. An observed
value far above the expected, with a p-value near 0, would say they do not:
the data have more empty sites than occupancy plus detection can explain.
The usual culprits are a third zero-generating process the model omits,
such as habitat that is structurally unsuitable, or detection that collapses
to near zero at a subset of sites. Either points toward a zero-inflated or
hurdle structure, or toward a detection covariate that separates the dead
sites from the live ones.

`tobs_test_outliers()` flags sites whose detection total falls outside the
central 95% of the replicate distribution, then checks whether the number
of such sites is itself surprising.

```{r outliers}
tobs_test_outliers(fit)
```

A handful of outliers is expected by chance; 5% of sites sit outside a 95%
interval by construction, so the raw count is not alarming on its own. The
p-value does the calibration: it asks whether the count of outliers in the
data exceeds what the model produces on its own replicates, which centres
the comparison on the model's own expectation rather than the nominal 5%.
A small p-value points to a few sites the model cannot accommodate. Those
are worth inspecting individually, because a population-level fit can look
fine while a small number of sites with unusual detection histories distort
the estimates. Cross-referencing the flagged sites with the largest Pearson
residuals from the next section usually identifies the same handful, which
is the cue to check them for data-entry errors, unusual survey effort, or a
covariate value the model did not see.

## Residuals

`residuals()` returns a list with two elements, because an occupancy model
has two processes. `$occ` is one residual per site, comparing the observed
occupancy indicator (any detection) against the fitted occupancy
probability. `$det` is a site-by-visit matrix of detection residuals, with
`NA` where a visit was not made.

```{r resid-shape}
r <- residuals(fit, type = "pearson")
length(r$occ)
dim(r$det)
```

The `type` argument follows the usual convention. `"response"` is the raw
difference between observed and fitted, on the scale of the data.
`"pearson"` divides that difference by the fitted standard deviation, which
standardises sites with different fitted probabilities onto a common scale
so that a residual of 2 means the same thing at a rare site and a common
one. `"deviance"` is the signed root of the per-observation deviance
contribution, the residual whose sum of squares is the model deviance, and
the one whose distribution is closest to normal for binary data. Pearson
residuals are the natural default for spotting sites the model handles
badly, since the common scale makes large values comparable; deviance
residuals are the better choice when you want the residual distribution
itself to look normal in a QQ plot.

The two-part return reflects that an occupancy fit predicts two things and
can be wrong about either. The occupancy residual asks whether the model
got the site's occupancy probability right given whether it was ever
detected. The detection residuals ask, visit by visit, whether the model
got the per-visit detection probability right at sites it believes are
occupied. A model can fit occupancy well and detection badly, or the
reverse, and keeping the two residual streams separate is what lets you
tell which.

A residuals-against-fitted plot is the first thing to draw. Structure in
the cloud, such as a trend or a fan, signals a misspecified mean or
variance. Here the occupancy residuals scatter without pattern.

```{r resid-plot, fig.alt = "Pearson occupancy residuals plotted against fitted occupancy probability, scattered around zero with no trend"}
fv <- fitted(fit)
plot(fv$psi, r$occ, pch = 16, col = rgb(0, 0, 0, 0.35),
     xlab = "fitted occupancy probability", ylab = "Pearson residual (occupancy)")
abline(h = 0, col = "steelblue", lwd = 2)
```

The detection residuals live in the matrix `r$det`, one row per site and
one column per visit, with `NA` in the cells for visits that were not made.
Collapsing them to a per-site mean gives a quick read on which sites
contribute the most detection misfit, again with no structure expected
under a good model.

```{r resid-det}
det_site <- rowMeans(r$det, na.rm = TRUE)
summary(det_site)
```

A per-site mean near zero across the board, as here, says the detection
model is not systematically over- or under-predicting at any site. If you
suspect a detection covariate is missing, plotting these per-site means
against the candidate covariate is the direct check: a trend in that plot
is the covariate announcing itself. The same matrix supports a per-visit
view, by taking column means instead, which catches a detection process
that drifts across the survey season.

## Spatial residual diagnostics

When sites carry coordinates, leftover residual structure in space is a
sign of a missing spatial field. The reasoning is the same as for a missing
covariate. If nearby sites share an unmodelled influence, such as a climate
gradient or a dispersal corridor, their residuals will share sign and
magnitude, and that shared pattern is detectable even though the influence
itself is unobserved. A correctly specified model leaves residuals that are
spatially independent, scattered without regard to which sites are
neighbours.

The `tulpa` engine that powers tulpaObs exports two helpers for this,
inherited through the fit's class. `moran_i()` reduces the spatial pattern
to a single autocorrelation statistic and tests it; `tulpa_variogram()`
shows how residual similarity changes with distance, which carries more
detail about the range of the structure. Both accept a residual vector
directly, so feed them the `$occ` element from the residual list rather
than the whole list.

To show the diagnostic working, simulate occupancy on a grid where the
occupancy intercept follows a smooth spatial surface. A model with no
spatial term will leave that surface in its residuals.

```{r spatial-sim}
side   <- 12L
coords <- expand.grid(lon = seq_len(side), lat = seq_len(side))
N      <- nrow(coords)

adj <- matrix(0L, N, N)                     # rook adjacency on the grid
for (i in seq_len(N)) for (j in seq_len(N)) {
  if (i < j &&
      abs(coords$lon[i] - coords$lon[j]) +
      abs(coords$lat[i] - coords$lat[j]) == 1) {
    adj[i, j] <- 1L; adj[j, i] <- 1L
  }
}

field <- 2.2 * sin(coords$lon / 2.2) + 2.2 * cos(coords$lat / 2.2)
xg  <- rnorm(N)
z   <- rbinom(N, 1, plogis(-0.2 + 0.5 * xg + field))
yg  <- matrix(0L, N, 7L)
for (i in seq_len(N)) if (z[i] == 1L) yg[i, ] <- rbinom(7L, 1, 0.6)
datg <- data.frame(x = xg)
```

Fit two models: a non-spatial Laplace fit that ignores the surface, and a
nested-Laplace fit with an intrinsic CAR (`icar()`) field on the same
adjacency graph.

```{r spatial-fits}
fit_ns <- tobs(~ x, data = datg, family = occu(),
               detection = ~ 1, y = yg, method = "laplace",
               control = list(verbose = FALSE))
fit_sp <- tobs(~ x + icar(graph = adj), data = datg, family = occu(),
               detection = ~ 1, y = yg, method = "nested_laplace",
               control = list(max.iter = 10L, verbose = FALSE))
```

Moran's I on the occupancy residuals measures how strongly neighbouring
sites share residual sign and size. The function returns an `htest` with
the statistic and a p-value under the randomisation null.

```{r moran}
mi_ns <- tulpa::moran_i(residuals(fit_ns)$occ, coords = coords,
                        weights = "knn", k = 8)
mi_sp <- tulpa::moran_i(residuals(fit_sp)$occ, coords = coords,
                        weights = "knn", k = 8)

data.frame(
  model    = c("non-spatial", "icar field"),
  moran_I  = round(c(mi_ns$statistic, mi_sp$statistic), 3),
  p_value  = signif(c(mi_ns$p.value, mi_sp$p.value), 2)
)
```

Moran's I runs from roughly $-1$ to $+1$, with values near its small
negative expected value indicating no spatial pattern. The non-spatial fit
leaves strong positive autocorrelation in its residuals: neighbouring sites
are wrong in the same direction because the shared spatial surface is
unmodelled, and the statistic sits far above its expected value with a
p-value near zero. Adding the CAR field absorbs most of that structure, and
Moran's I drops sharply toward its null. The field is doing the work the
residuals show is needed. The `weights` argument controls how neighbours
are defined: `"knn"` with a chosen `k` uses the k nearest sites, while
`"inverse"` weights every pair by inverse distance. The k-nearest choice is
the steadier one on a regular grid, since it does not let a few close pairs
dominate the statistic.

The variogram tells the same story as a curve, with the advantage that it
shows the range over which the structure operates rather than collapsing it
to one number. It bins pairs of sites by distance and plots half the mean
squared residual difference in each bin, the semivariance. Under spatial
structure, nearby sites have similar residuals and the semivariance is
small at short distances, rising as distance grows until pairs are far
enough apart to be effectively independent, where the curve levels off. The
distance at which it levels is the range of the residual structure, which
tells you roughly how far a spatial field would need to reach. A flat curve
across all distances means no structure is left.

```{r variogram, fig.alt = "Empirical semivariogram of non-spatial occupancy residuals rising with distance before levelling, indicating residual spatial structure"}
vg <- tulpa::tulpa_variogram(residuals(fit_ns)$occ, coords = coords)
plot(vg)
```

The rise in the non-spatial residual variogram confirms the Moran's I
verdict. If you compute the same curve on the spatial fit's residuals, it
flattens, which is the visual form of the autocorrelation shrinking. Both
helpers also accept a fitted model directly and call `residuals()` for you,
but the tobs residual list means passing the `$occ` vector is the safe
route.

## A diagnostic workflow

The checks fall into an order that catches the cheap problems first and
spends simulation effort only once the cheap checks pass. The first two
steps cost nothing beyond the fit you already have. The middle steps
simulate from the model, which takes a few seconds. The spatial step needs
coordinates and a second fit. Running them in this order means you rarely
pay for the expensive checks on a model that a cheap check would have
rejected.

1. **Read the coefficients against their intervals.** A coefficient whose
   credible interval spans zero is not earning its place, whatever its
   point estimate. Start with `summary()` and `confint()`.
2. **Compare against nested alternatives with `tobs_waic()`.** Drop the
   covariate, add a plausible one, and read the `delta`. A change of a few
   units or less is a tie; prefer the simpler model. WAIC ranks, it does
   not validate.
3. **Run a posterior-predictive check with `tobs_ppc()`.** The Bayesian
   p-value sits near 0.5 for a fit that reproduces the data. A value near 0
   means the data are more dispersed than the fit predicts, near 1 means
   less. The check conditions the latent state on the detection history, so
   the value is calibrated rather than purely relative.
4. **Inspect the PIT histogram and `tobs_test_uniformity()`.** A flat
   histogram is the target. A U-shape means underdispersion, a central hump
   overdispersion, a slope a biased mean.
5. **Target specific failures.** `tobs_test_dispersion()` for spread,
   `tobs_test_zero_inflation()` for excess empty sites,
   `tobs_test_outliers()` for individual sites the model cannot fit.
6. **Plot residuals against fitted values.** Structure in the cloud points
   to a misspecified mean or variance. Use Pearson residuals for a common
   scale.
7. **If sites have coordinates, test for residual spatial structure** with
   `moran_i()` and `tulpa_variogram()` on `residuals(fit)$occ`. Significant
   autocorrelation is the cue to add a spatial field and refit, then
   confirm the autocorrelation shrinks.

The point throughout is that a clean coefficient table is necessary and not
sufficient. The fit earns trust when it survives the checks, and each check
that fails names the next term to add: a dispersion flag asks for a random
effect, a zero-inflation flag for a hurdle or a detection covariate, a
spatial-autocorrelation flag for a field. Diagnostics are not a final exam
the model either passes or fails. They are the loop that turns a first fit
into a defensible one, and the most useful outcome of a failed check is the
specific term it tells you to add before the next pass.

## Where to go next

- The accessors behind these checks, `coef()`, `summary()`, `confint()`,
  `fitted()`: `vignette("occupancy")`.
- Model comparison by marginal likelihood at the engine level, and what
  WAIC and LOO need: the `tulpa` model-comparison vignette.
- Adding spatial fields once the residual diagnostics call for them:
  `vignette("occupancy")` and the spatial model vignettes.
- Dispersion in counts, where the negative binomial answers an
  overdispersion flag: `vignette("abundance")`.
```
