---
title: "Dynamic (multi-season) occupancy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic (multi-season) occupancy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## When occupancy changes over time

A single-season occupancy model treats a site as either occupied or empty and
asks only whether you detected the species. That picture holds for a snapshot.
Run the same survey for several years and the picture moves: a pond that was
empty fills with frogs, a meadow loses its butterflies, a forest patch is
colonised and then goes quiet again. The state itself is no longer fixed. It is
a sequence, and the interesting questions are about the changes between the
entries in that sequence.

Two rates govern those changes. Colonisation is the chance that an empty site
becomes occupied by the next season. Local extinction is the chance that an
occupied site becomes empty. A site's trajectory is a chain of occupied and
empty states linked by these two rates, and the species is detected only when
the site is occupied and you happen to catch it on a visit. The dynamic
occupancy model of MacKenzie and colleagues separates all three pieces:
where the species starts, how it moves between states, and how often you see it
when it is there.

The reason to separate them is that the raw data hide all three behind the same
zeros. A site that records no detections in a season might be empty, or occupied
but missed. A site occupied this year but not last might have been colonised, or
might have been occupied all along and undetected the first year. Detection
error and state change both turn a present species into a string of zeros, and
only a model that names the two processes can tell which zeros are which. The
dynamic model layers a detection process over a Markov chain on the state and
estimates the rates of each from the pattern of zeros and ones across visits and
seasons.

`dyn_occu()` fits that model in tulpaObs. You hand it a detection history with
a season dimension, a formula for first-season occupancy, a formula for
colonisation, a formula for extinction, and a detection formula. This vignette
simulates a known trajectory, fits it, checks that the rates come back, derives
turnover and the long-run occupancy the rates imply, and works through when the
rates are identifiable and when they are not.

## The model

Let $z_{i,t}$ be the latent occupancy state of site $i$ in season $t$, equal to
$1$ if the species is present and $0$ if not. The first season is drawn from an
initial occupancy probability,

$$
z_{i,1} \;\sim\; \mathrm{Bernoulli}(\psi_1),
$$

and every later season follows from the one before it through colonisation
$\gamma$ and extinction $\epsilon$,

$$
z_{i,t} \mid z_{i,t-1} \;\sim\;
\begin{cases}
\mathrm{Bernoulli}(1 - \epsilon) & z_{i,t-1} = 1 \\[2pt]
\mathrm{Bernoulli}(\gamma)       & z_{i,t-1} = 0.
\end{cases}
$$

An occupied site stays occupied with probability $1 - \epsilon$ and goes empty
with probability $\epsilon$. An empty site is colonised with probability
$\gamma$ and stays empty with probability $1 - \gamma$. Those four numbers are a
$2 \times 2$ transition matrix on the state, with rows indexed by last season's
state and columns by this season's,

$$
T \;=\;
\begin{pmatrix}
1 - \gamma & \gamma \\
\epsilon   & 1 - \epsilon
\end{pmatrix},
$$

where the first row starts empty and the second row starts occupied. The state
sequence $z_{i,1}, z_{i,2}, \dots$ is a two-state Markov chain: the next state
depends on the present one and nothing earlier. This is the hidden part of a
hidden Markov model, hidden because you never see $z$ directly.

Reading the matrix row by row recovers the two rates. The empty row, $(1 -
\gamma,\ \gamma)$, says an empty site stays empty unless it is colonised. The
occupied row, $(\epsilon,\ 1 - \epsilon)$, says an occupied site stays occupied
unless it goes extinct. The columns of any one row sum to one because the site
must land in one of the two states. Multiplying a row vector of occupancy
probabilities by $T$ advances it one season, which is the recipe the trajectory
section uses to draw the modelled curve forward.

Detection is the observation layer on top of the chain. On visit $j$ of season
$t$ you record $y_{i,j,t} = 1$ if you saw the species and $0$ if not, and a
sighting is possible only when the site is occupied,

$$
y_{i,j,t} \mid z_{i,t} \;\sim\; \mathrm{Bernoulli}(z_{i,t}\, p),
$$

so an empty site produces only zeros and an occupied site produces a $1$ with
probability $p$ on each visit. A season of all-zero visits is therefore
ambiguous: the site was empty, or it was occupied and you missed the species
every time. The model resolves that ambiguity in two directions at once. Within
a season the repeat visits separate a true absence from a run of misses, since an
occupied site with detection $p$ produces at least one sighting in $J$ visits
with probability $1 - (1 - p)^J$. Across seasons the transition rates tie a
site's silent year to its neighbours in time, so a site detected before and
after a blank season is judged more likely to have been occupied throughout than
one blank at the end of its record.

tulpaObs sums the latent chain out with the forward algorithm, the standard
recursion for a hidden Markov model. It carries forward the probability of the
detection history up to season $t$ jointly with each possible current state,
multiplies by the transition matrix to step to season $t + 1$, and folds in that
season's detections. Summing over the two states at the end gives the likelihood
of the whole history with $z$ integrated out, so the latent states never appear
as parameters and the likelihood is a function of $\psi_1, \gamma, \epsilon, p$
alone. This is why the fit returns four rates rather than a state for every
site and season.

Each of those four can carry its own linear predictor. By default they are
intercept-only, fitted on the logit scale and reported back as probabilities,
but `gamma` and `epsilon` accept covariate formulas the same way occupancy and
detection do, which is the subject of a later section.

## Simulating a trajectory

`simulate_dyn_occu()` builds a dataset with a known initial occupancy, known
colonisation and extinction rates, and a known detection probability, so the fit
has a truth to recover. The arguments `gamma` and `epsilon` are the
colonisation and extinction probabilities directly; `beta_occ` and `beta_det`
are logit-scale intercepts, so `beta_occ = c(0)` means $\psi_1 = 0.5$ and
`beta_det = c(0)` means $p = 0.5$.

```{r sim}
sim <- simulate_dyn_occu(
  N = 200, J = 4, n_seasons = 6,
  beta_occ = c(0),    # psi1 = plogis(0) = 0.5
  beta_det = c(0),    # p    = plogis(0) = 0.5
  gamma    = 0.3,     # colonisation: empty -> occupied
  epsilon  = 0.2,     # extinction:   occupied -> empty
  seed     = 2026
)
str(sim, max.level = 2)
```

The response `sim$y` is a three-dimensional array indexed by site, visit, and
season. The third dimension is what makes the model dynamic; a two-dimensional
matrix would be a single-season detection history and `dyn_occu()` rejects it.

```{r y-dim}
dim(sim$y)             # 200 sites x 4 visits x 6 seasons
sim$y[1, , ]           # one site: rows are visits, columns are seasons
```

A missing visit is encoded as `NA` and dropped from that season's detection
count, so seasons can carry different numbers of visits and a site can be unvisited
in some seasons without breaking the chain. The state still transitions across a
season with no data; the model simply has nothing to update detection with for
that cell, and the transition rates carry the occupancy forward. This is what
lets the model handle the ragged survey effort real monitoring programmes
produce, where not every site is visited every year.

The truth list carries the latent state matrix `sim$truth$z`, one row per site
and one column per season, which the fit never sees. Averaging it down the
columns shows the realised occupancy in each season. With colonisation above
extinction, occupancy climbs from its starting value toward a higher level and
then settles.

```{r z-traj, fig.alt = "Realised occupancy fraction across six seasons, rising from one half toward about six tenths"}
occ_by_season <- colMeans(sim$truth$z)
round(occ_by_season, 3)
plot(seq_along(occ_by_season), occ_by_season, type = "b", pch = 16,
     ylim = c(0, 1), xlab = "season", ylab = "occupied fraction")
abline(h = 0.3 / (0.3 + 0.2), col = "steelblue", lwd = 2, lty = 2)
```

The dashed line marks $\gamma / (\gamma + \epsilon) = 0.6$, the occupancy the
two rates pull the system toward in the long run. The realised fraction wobbles
around it because two hundred sites is a finite sample of a stochastic chain, not
because the rates are drifting. A site's own path is a string of occupied and
empty seasons; the population fraction is the average over those strings.

The ecological reading of the two rates is worth stating before fitting. A high
$\gamma$ describes a species that spreads readily into empty habitat: good
dispersal, abundant propagules, vacant patches that are easy to reach. A high
$\epsilon$ describes a species that blinks out of occupied patches easily: small
local populations, harsh between-season conditions, habitat that degrades. The
balance between them sets whether a metapopulation grows, shrinks, or holds
steady, and the ratio sets the level it holds at.

The two rates also separate processes that a single occupancy number blurs
together. Suppose occupancy holds at sixty percent across a decade. That same
stable level can come from a placid system where the same sites stay occupied
year after year, or from a churning one where many sites wink out each season and
just as many are colonised. The static occupancy is identical; the dynamics are
opposite, and they imply different conservation responses. Estimating $\gamma$
and $\epsilon$ rather than occupancy alone is what tells the two apart, and it is
the reason a multi-season survey is worth the extra effort over repeating a
single-season one.

## Fitting

The call mirrors the single-season one with two extra formulas. The first
argument is the first-season occupancy formula $\psi_1$, `detection` is the
per-visit detection formula $p$, and `col_formula` and `ext_formula` are the
colonisation $\gamma$ and extinction $\epsilon$ formulas. All four are
intercept-only here. The response is the three-dimensional array, and
`method = "laplace"` is the default.

```{r fit}
fit <- tobs(
  ~ 1,                                 # psi1: first-season occupancy
  data        = sim$data,
  family      = dyn_occu(),
  detection   = ~ 1,                   # p: per-visit detection
  y           = sim$y,
  col_formula = ~ 1,                   # gamma: colonisation
  ext_formula = ~ 1,                   # epsilon: extinction
  method      = "laplace",
  control     = list(verbose = FALSE)  # the EM prints its iterations otherwise
)
fit
```

The response array must be three-dimensional. Passing a flat matrix, the shape a
single-season model expects, raises an error that names the requirement rather
than fitting the wrong model silently.

```{r dim-error, error = TRUE}
tobs(~ 1, data = sim$data, family = dyn_occu(), detection = ~ 1,
     y = sim$y[, , 1],                 # a single season: 2D matrix
     col_formula = ~ 1, ext_formula = ~ 1,
     control = list(verbose = FALSE))
```

The fitted object holds the four rates back-transformed to the probability scale
in `$intercepts`. This is the quickest read of the result: where occupancy
starts, how fast empty sites fill, how fast occupied sites empty, and how often a
present species is seen.

```{r intercepts}
fit$intercepts
```

The values land near the simulated truth of $\psi_1 = 0.5$, $\gamma = 0.3$,
$\epsilon = 0.2$, and $p = 0.5$. `coef()` returns the same four processes on the
logit scale, the scale the linear predictors live on, with one element per
process.

```{r coef}
coef(fit)
```

`summary()` adds the posterior standard deviation and credible bounds for every
coefficient, with the process prefix on each row: `psi1_` for first-season
occupancy, `p_` for detection, `gamma_` for colonisation, and `epsilon_` for
extinction.

```{r summary}
summary(fit)
```

## Recovering the rates

A model passing its plumbing checks is not the same as a model returning the
right numbers. The honest test is to compare the back-transformed rates against
the truth they were simulated from. Putting them side by side shows each
estimate within a few hundredths of its target.

```{r recovery}
truth <- c(psi1 = sim$truth$psi1, gamma = sim$truth$gamma,
           epsilon = sim$truth$epsilon, p = sim$truth$p)
est   <- c(psi1 = fit$intercepts$psi1, gamma = fit$intercepts$gamma,
           epsilon = fit$intercepts$epsilon, p = fit$intercepts$p)

data.frame(
  parameter = names(truth),
  truth     = round(truth, 3),
  estimate  = round(est, 3),
  abs_error = round(abs(est - truth), 3),
  row.names = NULL
)
```

All four absolute errors sit below the tolerance the package test asserts for
this regime. Recovery this clean rests on the design: two hundred sites give the
chain room to express both transitions, six seasons give five transition steps
to estimate the rates from, and four visits per season with detection near one
half keep $p$ separated from the occupancy states. Thin any of those and the
rates start trading off against one another, the subject of the practical
guidance below.

The recovery check is worth running as a habit, not just once. Simulating from
the rates you intend to report, fitting, and confirming the estimates return is
the cheapest way to learn whether a given survey design can estimate the
quantities you care about before you collect the data. If the rates come back
biased or with intervals too wide to be useful in the simulation, no analysis
of the real data will rescue them; the information was never in the design. The
gap between an estimate and its truth here is a few hundredths because the design
is generous. A design with three seasons and two visits would show a wider gap
and warn you off before fieldwork.

## Turnover and equilibrium occupancy

The two transition rates imply quantities that are often more interesting than
the rates themselves. The first is the equilibrium occupancy, the level the
metapopulation settles at once the initial condition washes out. Setting the
season-to-season change to zero in the transition matrix gives

$$
\psi_{\mathrm{eq}} \;=\; \frac{\gamma}{\gamma + \epsilon},
$$

the fraction of sites occupied when colonisation and extinction balance. Reading
it off the fitted rates gives a number close to the dashed line in the
trajectory plot.

```{r psi-eq}
g <- fit$intercepts$gamma
e <- fit$intercepts$epsilon
psi_eq <- g / (g + e)
c(psi_eq = round(psi_eq, 3), truth = round(0.3 / (0.3 + 0.2), 3))
```

The second is turnover, the share of the occupied set that is made up of
newly colonised sites rather than holdovers from the previous season. At
equilibrium it works out to

$$
\tau \;=\; \frac{\gamma\,(1 - \psi_{\mathrm{eq}})}{\psi_{\mathrm{eq}}}
       \;=\; \frac{\gamma\,\epsilon}{\gamma + \epsilon}\Big/\psi_{\mathrm{eq}},
$$

the colonisation flow into empty sites divided by the equilibrium occupied
fraction. A high turnover marks a restless metapopulation whose occupied set is
constantly being rebuilt; a low turnover marks a stable one where the same sites
stay occupied year after year.

```{r turnover}
turnover <- (g * (1 - psi_eq)) / psi_eq
round(turnover, 3)
```

A turnover of this size means a meaningful share of the occupied sites in any
season are recent arrivals rather than long-term residents. Two metapopulations
can sit at the same equilibrium occupancy with very different turnover: a
low-$\gamma$, low-$\epsilon$ system holds the same sites for years, while a
high-$\gamma$, high-$\epsilon$ system reshuffles its occupied set constantly even
as the headline occupancy stays put. Turnover is the quantity that distinguishes
them, and it is invisible to a single-season survey. Reporting it alongside the
equilibrium gives a reader both the level the system holds at and the rate at
which the membership of that level churns.

The expected occupancy trajectory follows from iterating the transition once per
season starting from $\psi_1$. Each season's occupancy is last season's pushed
through the two rates,

$$
\psi_{t} \;=\; \psi_{t-1}(1 - \epsilon) + (1 - \psi_{t-1})\,\gamma,
$$

so the modelled curve can be drawn forward from the fitted $\psi_1$, $\gamma$,
and $\epsilon$ and laid over the realised fractions from the simulation.

```{r traj-plot, fig.alt = "Modelled occupancy trajectory from the fitted rates overlaid on the realised occupancy fractions across six seasons"}
n_seasons <- dim(sim$y)[3]
psi_t <- numeric(n_seasons)
psi_t[1] <- fit$intercepts$psi1
for (t in 2:n_seasons) {
  psi_t[t] <- psi_t[t - 1] * (1 - e) + (1 - psi_t[t - 1]) * g
}

plot(seq_len(n_seasons), colMeans(sim$truth$z), type = "b", pch = 16,
     ylim = c(0, 1), xlab = "season", ylab = "occupied fraction")
lines(seq_len(n_seasons), psi_t, col = "darkorange", lwd = 2)
abline(h = psi_eq, col = "steelblue", lwd = 2, lty = 2)
legend("bottomright",
       c("realised (truth)", "modelled trajectory", "equilibrium"),
       col = c("black", "darkorange", "steelblue"),
       lwd = c(1, 2, 2), pch = c(16, NA, NA), lty = c(1, 1, 2), bty = "n")
```

The modelled curve rises from the fitted starting occupancy toward the
equilibrium line and tracks the realised fractions across the six seasons. The
gap between the orange curve and the black points is sampling noise in two
hundred finite chains, not a failure of fit; the curve is the expectation those
chains scatter around.

The speed of approach is itself informative. The distance to equilibrium shrinks
by a factor of $1 - \gamma - \epsilon$ each season, so when the two rates are
small the system creeps toward its level over many seasons, and when they are
large it snaps there in one or two. Here $\gamma + \epsilon = 0.5$, so the curve
covers half its remaining distance to the line each season and arrives quickly.
A survey that opens far from equilibrium and runs only a few seasons will see
occupancy still in transit, which is exactly the case where reporting a single
average occupancy across seasons would mislead and the rates carry the real
story.

## Covariates on colonisation and extinction

Colonisation and extinction need not be constant across sites. A patch near an
occupied source colonises faster; a small or exposed patch goes extinct more
often. `col_formula` and `ext_formula` take covariate formulas exactly as the
occupancy and detection formulas do, so a site-level predictor enters the rate on
its logit scale. Add a covariate to the data and put it on the colonisation
formula.

```{r cov-fit}
dat_cov <- sim$data
dat_cov$connectivity <- rnorm(nrow(dat_cov))   # a site-level predictor

fit_cov <- tobs(
  ~ 1,
  data        = dat_cov,
  family      = dyn_occu(),
  detection   = ~ 1,
  y           = sim$y,
  col_formula = ~ connectivity,                # colonisation varies by site
  ext_formula = ~ 1,
  method      = "laplace",
  control     = list(verbose = FALSE)
)
coef(fit_cov)$gamma
```

The colonisation process now has an intercept and a `connectivity` slope on the
logit scale. The slope is near zero here because `connectivity` was drawn
independently of the simulated trajectory; in a real dataset it would carry the
effect of the predictor on the colonisation rate. Extinction takes a formula the
same way through `ext_formula`, and the two can carry different covariates. To
read a fitted rate at a given covariate value, push the logit-scale linear
predictor through `plogis()`; the intercept-only `$intercepts` shortcut applies
only when the formula is `~ 1`.

Covariates on the transitions answer questions the constant-rate model cannot.
A connectivity covariate on colonisation tests whether patches near occupied
sources fill faster, the signature of dispersal limitation. A patch-area
covariate with a negative slope on extinction tests whether larger patches hold
their populations longer, the signature of demographic rescue. Because each rate
carries its own formula, you can let colonisation depend on the landscape between
patches and extinction depend on conditions within them, which matches how the
two processes actually differ: one is about reaching empty habitat, the other
about persisting in occupied habitat. The first-season occupancy formula and the
detection formula accept covariates in the same way, so every layer of the model
can be made a function of site or visit attributes without changing the call
shape.

## Full posterior with NUTS

The Laplace fit finds the mode of the four rates and the curvature there, which
is fast and accurate when the likelihood is well behaved. When you want draws
from the full posterior, or when few seasons and low detection make the mode an
incomplete summary, `method = "nuts"` runs the gradient sampler over the same
model. The control names are dotted: `n.iter`, `n.warmup`, and `seed`.

```{r nuts}
fit_nuts <- tobs(
  ~ 1,
  data        = sim$data,
  family      = dyn_occu(),
  detection   = ~ 1,
  y           = sim$y,
  col_formula = ~ 1,
  ext_formula = ~ 1,
  method      = "nuts",
  control     = list(n.iter = 600, n.warmup = 300, seed = 1, verbose = FALSE)
)
fit_nuts$intercepts
```

The sampler returns the same four rates near their simulated values. Its
`$intercepts` are summaries of the posterior draws rather than a single mode, so
the two backends agree on the rates while the sampler additionally reports the
spread the Laplace approximation collapses to its curvature. Use Laplace for the
first look and for speed; reach for NUTS when the rates are weakly identified or
when you need the posterior shape rather than a Gaussian summary at the mode.

The difference between the two backends matters most for the extinction rate
when occupancy is high or for colonisation when occupancy is low, because the
relevant transitions are then rare and the likelihood for that rate is skewed
rather than bell-shaped. A Gaussian summary at the mode reports a symmetric
interval that can stray below zero or above one on the probability scale once
back-transformed, while the sampler traces the true skew and keeps the interval
where it belongs. When the equilibrium occupancy you derive depends sensitively
on a rate near its boundary, the sampler's draws also let you propagate the rate
uncertainty into the derived quantity directly, by computing the equilibrium and
turnover on every draw and summarising the result, rather than plugging point
estimates into the formula.

## Practical guidance

Dynamic occupancy asks more of a dataset than a single-season model does,
because it estimates change from repeated snapshots. A few rules of thumb keep
the rates identifiable.

- **Use at least three seasons, and prefer five or more.** Two seasons give a
  single transition step, which estimates one blended change but cannot separate
  colonisation from extinction cleanly. Three seasons give two steps and make the
  two rates jointly estimable; five or more seasons let the rates and the
  equilibrium be read with usable precision. The simulation here uses six.

- **Keep at least three visits per season, more when detection is low.** Within a
  season the model separates a true absence from a missed detection using the
  repeat visits, exactly as in the single-season case. With one visit per season
  detection and occupancy are confounded; with three or four visits and detection
  above about 0.3 the season-level state is well resolved. Below that detection
  level, add visits or move to NUTS.

- **Watch the sites-times-seasons budget.** The transition rates are estimated
  from the count of sites that change state between consecutive seasons. A few
  dozen sites over three seasons leaves few such transitions and wide intervals on
  $\gamma$ and $\epsilon$; a couple of hundred sites over five or more seasons, as
  here, gives hundreds of transition opportunities and tight rates. If the
  realised occupancy is near zero or near one, the relevant transitions become
  rare even at large site counts, so a mid-range occupancy is the easiest regime.

- **Read the derived quantities, not just the rates.** Equilibrium occupancy
  $\gamma / (\gamma + \epsilon)$ and turnover are usually the ecological targets,
  and they are smooth functions of the two rates that summarise the metapopulation
  better than either rate alone. A small change in $\epsilon$ near zero moves the
  equilibrium a long way, so report the equilibrium with the uncertainty that
  flows from the rate intervals rather than from a point plug-in.

- **When not to fit it.** If you have a single season, fit `occu()`; there is no
  transition to estimate. If you have many seasons but the species never changes
  state, the rates collapse to zero or one and the dynamic structure adds
  parameters that the data cannot inform, so a static model per season is more
  honest. And if seasons are so far apart that the Markov assumption fails, with
  several unobserved transitions folded into each observed step, the colonisation
  and extinction you estimate describe the observation interval rather than the
  biological process.

## Where to go next

- Single-season occupancy, detection, and the latent state: `vignette("occupancy")`.
- Many species surveyed at once: `vignette("community-models")`.
- Combining detection histories from several sources: `vignette("integrated-occupancy")`.
- Posterior-predictive checks, WAIC, and residual diagnostics: `vignette("diagnostics")`.
