---
title: "Integrated occupancy: combining data sources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated occupancy: combining data sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## Two surveys, one species

A structured survey is expensive. A field crew visits a fixed set of sites a
fixed number of times, records presence and absence under a known protocol,
and the resulting detection histories support a clean occupancy model. The
catch is reach: a few hundred sites, often clustered near roads or stations,
covering a fraction of the range you care about.

Citizen-science records cover the rest. Birders, naturalists, and app users
report sightings across the whole landscape, far beyond where any crew can go.
The records are plentiful and broad, but the detection process behind them is
murky. Nobody logged the visits that found nothing, effort varies from one
observer to the next, and the protocol is whatever each person happened to do
that day.

An integrated occupancy model uses both. The two sources disagree about how
species are detected, yet they agree about one thing: whether a site is
occupied. Occupancy is a property of the place and the species, not of who is
looking. So you let the sources share a single latent occupancy state and give
each its own detection process. The structured survey anchors the detection
calibration; the opportunistic records extend the spatial and covariate reach.
`int_occu()` fits exactly this.

This vignette simulates two sources with a known shared truth, fits the
integrated model, reads the per-source detection, and shows where integration
sharpens the occupancy estimate over a single source fit alone. It closes with
rules for when sources may share a state and when they may not.

## The model

Let $z_i \in \{0, 1\}$ be the occupancy state of site $i$, shared across every
source. It depends on site covariates $x_i$ through one occupancy regression:

$$
z_i \sim \mathrm{Bernoulli}(\psi_i), \qquad
\mathrm{logit}(\psi_i) = x_i^\top \beta_{\mathrm{occ}}.
$$

Each source $s$ then sees that same state through its own detection process. On
a visit $j$ to site $i$ by source $s$, the observation $y_{sij}$ is a detection
only when the site is occupied and the source happens to catch it:

$$
y_{sij} \mid z_i \sim \mathrm{Bernoulli}\!\left(z_i \, p_{sij}\right),
\qquad
\mathrm{logit}(p_{sij}) = w_{sij}^\top \beta_{\mathrm{det},\,s}.
$$

The occupancy coefficient vector $\beta_{\mathrm{occ}}$ is one object,
estimated from all sources at once. The detection coefficients
$\beta_{\mathrm{det},\,s}$ are a separate vector per source, so a precise
survey and a noisy citizen-science stream get their own detection
probabilities. A site visited by both sources contributes two detection
histories that both inform the one $z_i$. A site seen by only one source
contributes through that source alone. The shared $\psi$ is where the pooling
happens: every detection history, from every source, is evidence about the
same occupancy surface.

This is the structure that makes integration pay. A single sparse source has to
estimate $\psi$ and its own $p$ from a thin set of visits, and the two trade off
against each other. Adding a second source that observes overlapping sites
brings more visits to bear on $z$ without forcing the detection processes to
match. The occupancy slope sharpens; each $p_s$ stays free.

The likelihood spells out how the pooling works. For one site, the latent
state $z_i$ is unobserved, so the model sums over its two values. With every
source's visits to that site collected into the histories
$y_{1i}, y_{2i}, \dots$, the site's contribution is

$$
\psi_i \prod_s \Pr(y_{si} \mid z_i = 1)
\;+\; (1 - \psi_i)\, \mathbf{1}\!\left[\text{no detection in any source}\right].
$$

The first term is the occupied case, where every source's detection history
enters through its own $p_s$. The second is the unoccupied case, possible only
when no source detected anything. A single detection in any one source rules out
$z_i = 0$ for that site, which is why a deeper source can resolve the state of a
site that a shallow source left ambiguous. The product over sources inside the
occupied term is the channel through which the structured survey's clean
detection sharpens the state estimate that the casual records also depend on.

## Simulating two sources

`simulate_int_occu()` draws a shared occupancy state and then observes it
through several sources with their own visit counts and detection
probabilities. Here source 1 is a shallow survey with two visits per site;
source 2 is a deeper opportunistic stream with five visits. Both observe a
block of shared sites, and each also covers sites the other does not.

```{r sim}
si <- simulate_int_occu(
  N_total  = 500,                       # unique sites across both sources
  n_data   = 2,                         # two sources
  J        = c(2, 5),                   # visits per source: shallow vs deep
  n_shared = 250,                       # sites observed by both
  beta_occ = c(0.3, 0.9),               # occupancy intercept + slope on x
  beta_det = list(c(0.5, 0), c(0.7, 0)),# per-source detection intercept
  seed     = 11
)
```

The response `si$y` is a list with one detection-history matrix per source.
They have different visit counts and, in general, different numbers of rows,
because each source covers its own set of sites.

```{r y-structure}
str(si$y, max.level = 1)
c(survey_sites = nrow(si$y[[1]]), casual_sites = nrow(si$y[[2]]))
```

The site covariates live in a single data frame with one row per unique site.
`si$site_maps` records which rows of that frame each source observed, and
`si$truth` carries the occupancy coefficients and the latent state for checking
the fit.

```{r truth}
str(si$data)
si$truth$beta_occ
```

## Fitting the integrated model

The call mirrors a single-season occupancy fit. The occupancy formula is the
first argument and applies to the shared state. `detection` is the detection
formula. `y` is the list of source matrices, and naming the list elements names
the detection processes in the output. `family = int_occu()` selects the
integrated likelihood.

```{r fit}
fit <- tobs(
  ~ x,
  data      = si$data,
  family    = int_occu(),
  detection = ~ 1,
  y         = list(survey = si$y[[1]], casual = si$y[[2]]),
  method    = "laplace",
  control   = list(verbose = FALSE)
)
```

`coef()` returns the shared occupancy block plus one detection block per
source, keyed by the names you gave `y`.

```{r coef}
coef(fit)
```

The occupancy slope on `x` lands near the simulated `r si$truth$beta_occ[2]`.
The two detection intercepts come back separately, one labelled `survey` and
one `casual`, because each source carries its own detection process.
`summary()` adds the posterior standard deviation and credible bounds; the
`psi_` rows are the shared occupancy predictor, the `survey_` and `casual_` rows
the two detection predictors.

```{r summary}
summary(fit)
```

## Why integration helps

The payoff is in the occupancy slope. To see it, fit the shallow survey on its
own and compare. Source 1 alone is a single-season occupancy model on the
sites that source observed, pulled out through `site_maps`.

```{r survey-only}
rows1 <- si$site_maps[[1]]
d1    <- si$data[rows1, , drop = FALSE]

fit_survey <- tobs(
  ~ x, data = d1, family = occu(),
  detection = ~ 1, y = si$y[[1]],
  method = "laplace", control = list(verbose = FALSE)
)
coef(fit_survey)$psi
```

The point estimates are in the same neighbourhood, so the comparison that
matters is the uncertainty, and that needs an honest interval. The default
Laplace approximation runs its occupancy intervals optimistically narrow on the
integrated fit, because the penalised EM weights the pooled data heavily at the
mode. For comparing uncertainty across the two fits, run both under NUTS, which
samples the full posterior and reports a calibrated spread.

```{r nuts-payoff}
fit_int_nuts <- tobs(
  ~ x, data = si$data, family = int_occu(),
  detection = ~ 1, y = list(survey = si$y[[1]], casual = si$y[[2]]),
  method = "nuts",
  control = list(n.iter = 1000, n.warmup = 500, seed = 1, verbose = FALSE)
)

fit_survey_nuts <- tobs(
  ~ x, data = d1, family = occu(),
  detection = ~ 1, y = si$y[[1]],
  method = "nuts",
  control = list(n.iter = 1000, n.warmup = 500, seed = 1, verbose = FALSE)
)
```

Read the occupancy slope from each posterior. The integrated fit borrows the
deeper source's visits and the extra sites it covers, so its slope sits closer
to the truth of `r si$truth$beta_occ[2]` and carries a smaller standard
deviation and a narrower credible interval.

```{r payoff-table}
row_x <- function(f) {
  s <- summary(f)["psi_x", c("mean", "sd", "q2.5", "q97.5")]
  s <- as.numeric(s)
  c(estimate = s[1], sd = s[2], lower = s[3], upper = s[4],
    width = s[4] - s[3])
}

comparison <- rbind(
  `survey only` = row_x(fit_survey_nuts),
  integrated    = row_x(fit_int_nuts)
)
round(comparison, 3)
```

The integrated credible interval is narrower than the survey-only one, and its
centre is closer to the true slope. The shallow survey, with two visits per
site, struggles to separate occupancy from detection on its own. Folding in the
deeper source tightens the shared state without making the two sources pretend
to share a detection process.

The gain has a shape worth naming. Integration does not manufacture
information; it pools what the sources already carry onto the one quantity they
agree about. When the second source is itself thin, or observes few of the same
sites, the interval barely moves. The benefit scales with how much real
occupancy signal the added source brings to the shared state.

It is worth being precise about what gets sharper and what does not. The
occupancy slope tightens because more sites and more visits inform the shared
state surface. The detection parameters do not pool across sources, by design,
so each source's detection precision still depends only on that source's own
visits. A casual stream with few visits per site keeps a loose detection
estimate even inside a large integrated fit. That separation is the feature, not
a shortcoming: the survey's detection stays calibrated by survey data, and the
occupancy estimate borrows strength from both. If you find the casual detection
intercept wandering, the fix is more casual visits, not more survey sites.

## Source-specific detection

Each source detects the species at its own rate, and the model estimates those
rates separately. Make the difference stark: a careful survey with high
detection and a casual stream with low detection, observing the same occupancy
surface.

```{r sim-detection}
sd2 <- simulate_int_occu(
  N_total  = 500, n_data = 2, J = c(5, 4), n_shared = 200,
  beta_occ = c(0.3, 0.9),
  beta_det = list(c(1.2, 0),   # survey: logit p = 1.2, so p about 0.77
                  c(-0.8, 0)), # casual: logit p = -0.8, so p about 0.31
  seed = 21
)
```

Fit it and read the two detection intercepts back on the probability scale.
NUTS keeps the detection blocks honest here.

```{r fit-detection}
fit2 <- tobs(
  ~ x, data = sd2$data, family = int_occu(),
  detection = ~ 1, y = list(survey = sd2$y[[1]], casual = sd2$y[[2]]),
  method = "nuts",
  control = list(n.iter = 800, n.warmup = 400, seed = 1, verbose = FALSE)
)

data.frame(
  source       = c("survey", "casual"),
  true_p       = round(plogis(c(1.2, -0.8)), 2),
  fitted_p     = round(plogis(c(coef(fit2)$survey, coef(fit2)$casual)), 2),
  row.names    = NULL
)
```

The casual stream comes back near its true detection of 0.31 and the survey
near its higher rate, exactly the contrast the data carry. This is the point of
the separate detection blocks: a single pooled detection probability would
average the two and misrepresent both. The survey arm can attenuate a little
when occupancy and detection trade off along their shared ridge, so read the
detection intercepts as a calibration of each source's reporting rate rather
than an exact constant.

The detection formula can also differ in structure across sources, beyond
differing in its fitted value. Pass a list of formulas, one per source, to give a source its
own detection covariates. A structured survey often records effort or observer,
while opportunistic records carry no such fields, so the survey arm can take a
covariate the casual arm omits.

```{r per-source-formula}
sd2$data$effort <- rnorm(nrow(sd2$data))   # a survey-side detection covariate

fit3 <- tobs(
  ~ x, data = sd2$data, family = int_occu(),
  detection = list(survey = ~ effort, casual = ~ 1),
  y = list(survey = sd2$y[[1]], casual = sd2$y[[2]]),
  method = "laplace", control = list(verbose = FALSE)
)
names(coef(fit3))
```

The `survey` block now carries an `effort` slope alongside its intercept, while
the `casual` block stays intercept-only. Detection covariates must be columns of
the site data frame, since the integrated builder reads each source's detection
design from that frame; place any spatial, temporal, or random-effect term on
the occupancy formula, where the shared field lives, not on a per-source
detection formula.

## Practical guidance

Integration is a modelling commitment, not a free lunch. It rests on the
assumption that the sources observe the same latent state, and that assumption
does real work. A few rules keep it honest.

- **Sources must share the same definition of occupied, over the same closure
  period.** The shared $z_i$ is one event: the species is present at site $i$
  during the season both sources sampled. If the structured survey ran in
  spring and the citizen records span the whole year, they are not measuring
  the same state, and pooling them biases $\psi$ toward whichever source sampled
  the wider window. Align the temporal window before you align the model.

- **Overlap is the currency. Aim for a real block of shared sites, not a
  handful.** The deeper source sharpens the shared slope only through sites it
  observes in common with the others, since those are where two detection
  histories pin one $z_i$. As a working target, want at least 20 to 30 shared
  sites, and more when detection is low, so the cross-source calibration has
  something to stand on. With near-zero overlap the model reduces to fitting
  each source independently and stacking them, and the integration buys little.

- **Lean on integration when the primary source is thin.** The gain is largest
  when the structured survey is small or shallow, two or three visits per site,
  a few hundred sites, and a broad opportunistic source can extend its reach.
  When the structured survey is already deep and wide, a single-source occupancy
  fit may be cleaner, and the second source mostly adds detection nuisance
  parameters.

- **Give each source its own detection process, always.** The reason to
  integrate rather than concatenate is that the sources detect differently.
  Forcing one shared $p$ across a calibrated survey and a casual stream throws
  away the calibration the survey provides. Keep the detection blocks separate
  and let the survey anchor the scale.

- **Read the occupancy slopes, treat the intercepts with care.** Occupancy and
  detection trade off along a ridge, more so with few visits, and the
  occupancy intercept absorbs part of that trade. The slopes are the
  inferential payload and recover cleanly; the intercept is a level to report
  with its interval, not a precise baseline rate. NUTS gives the honest spread
  when the default Laplace intervals on the integrated fit run narrow.

**When not to integrate.** The model fails quietly when the sources measure
different states. A breeding-season survey and an autumn-migration dataset both
record the same species, but presence in one is not presence in the other, and
a shared $z$ forces a contradiction the fit will resolve by splitting the
difference. The same applies to a survey that scores a $5 \times 5$ km cell as
occupied and an app that pins a point: the spatial grain differs, so the
occupancy units differ, and they cannot share one state without first being
brought to a common resolution. When in doubt, ask whether a single coin flip,
the one $z_i$, could have generated both sources' data at that site. If not, fit
them separately.

## Where to go next

- Single-season occupancy, identifiability, and model comparison:
  `vignette("occupancy")`.
- Many species at once, with community hyperpriors: `vignette("community-models")`.
- Posterior-predictive checks and residual diagnostics on any fit:
  `vignette("diagnostics")`.
- The shared engine, tiers, and how the Laplace and NUTS backends differ:
  `vignette("quickstart")`.
