---
title: "Occupancy with a continuous SPDE spatial field"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Occupancy with a continuous SPDE spatial field}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

Occupancy data are usually spatially structured: nearby sites share habitat
and species pressures that the row-wise covariates can't fully capture. The
Matern field via the SPDE approach (Lindgren, Rue & Lindstrom, 2011) gives
a continuous spatial random effect with an interpretable range and marginal
SD, and an `spde()` term in the formula wires it through tulpaObs's
Laplace EM engine.

This vignette simulates an occupancy dataset with a known smooth spatial
signal, fits a non-spatial model and an SPDE model, and compares them on
parameter recovery and the latent state.

## Simulation

```{r simulate}
library(tulpaObs)

n_sites <- 400
J <- 8
coords <- cbind(runif(n_sites), runif(n_sites))

# A smooth spatial signal: amplitude ~0.8 on the logit scale
u_true <- 0.8 * cos(3 * coords[, 1]) * sin(3 * coords[, 2])

x_cov   <- rnorm(n_sites)
det_cov <- rnorm(n_sites)
beta_occ_true <- c(-0.5, 0.7)
beta_det_true <- c(-0.3, 0.4)

eta_occ <- beta_occ_true[1] + beta_occ_true[2] * x_cov + u_true
psi <- plogis(eta_occ)
z <- rbinom(n_sites, 1, psi)

eta_det <- beta_det_true[1] + beta_det_true[2] * det_cov
p <- plogis(eta_det)

y <- matrix(0L, n_sites, J)
for (i in seq_len(n_sites)) {
  if (z[i] == 1L) y[i, ] <- rbinom(J, 1, p[i])
}

dat <- data.frame(occ_cov = x_cov, det_cov = det_cov,
                  lon = coords[, 1], lat = coords[, 2])
c(true_occupancy_rate = mean(z),
  naive_any_detection = mean(rowSums(y) > 0))
```

The naive any-detection rate is biased low by `(1 - p)^J` at each occupied
site; a proper occupancy fit should correct that bias, and a spatial fit
should additionally recover the smooth field that drives the
site-to-site variation in `psi`.

## Non-spatial fit

```{r fit-non-spatial}
fit_ns <- tobs(
  formula   = ~ occ_cov,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov,
  y         = y,
  engine    = "laplace",
  control   = list(verbose = FALSE)
)
round(fit_ns$means, 3)
```

## SPDE fit

The spatial field is written directly into the occupancy formula as an
`spde()` term, the way `mgcv` writes `s(lon, lat)`. `spde()` builds a
triangular mesh over the observation coordinates (via tulpaMesh) and
constructs the SPDE precision at the prior modes of `prior_range` and
`prior_sigma`. The mesh covers the data extent plus a small buffer;
`max_edge = c(inner, outer)` controls the spacing inside and outside the
data. The coordinate columns (`lon`, `lat`) are taken from `data`.

```{r fit-spde}
fit_sp <- tobs(
  formula   = ~ occ_cov + spde(lon, lat,
                               max_edge    = c(0.3, 0.6),
                               nu          = 1,
                               prior_range = c(0.3, 0.5),
                               prior_sigma = c(0.7, 0.5)),
  data      = dat,
  family    = occu(),
  detection = ~ det_cov,
  y         = y,
  engine    = "laplace",
  control   = list(verbose = FALSE)
)
round(fit_sp$means, 3)
```

## Coefficient recovery

```{r compare-betas}
compare <- data.frame(
  parameter = c("psi_(Intercept)", "psi_occ_cov",
                "p_(Intercept)",   "p_det_cov"),
  truth     = c(beta_occ_true, beta_det_true),
  non_spat  = unname(round(fit_ns$means, 3)),
  spde      = unname(round(fit_sp$means, 3))
)
compare
```

The detection coefficients are essentially unchanged: detection covariates
don't carry spatial structure in this simulation, so the detection
submodel sees the same data both times. Occupancy coefficients shift
because the SPDE absorbs site-to-site logit deviations that the
non-spatial fit had to attribute to the global intercept.

## Recovered spatial field

The fitted SPDE field is stored on the fit as `spatial_field`, a length-
`n_units` numeric vector at the mesh nodes. Project it to observation
locations through the spec's `A` matrix to compare against the truth.

```{r field-recovery, fig.height = 4}
u_hat <- fit_sp$spatial_field
field_at_sites <- as.numeric(fit_sp$spatial$tulpa_spec$A %*% u_hat)

corr <- cor(field_at_sites, u_true)
cat(sprintf("Correlation field-at-sites vs truth: %.3f\n", corr))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(u_true, field_at_sites, pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "truth u(s)", ylab = "fitted u_hat(s)",
     main = "Field recovery at sites")
abline(0, 1, col = 2, lwd = 2)

plot(coords[, 1], coords[, 2], pch = 16, cex = 0.6,
     col = ifelse(field_at_sites > 0, "red", "blue"),
     xlab = "x", ylab = "y",
     main = "Fitted field sign")
par(op)
```

The fitted field is regularized by the SPDE prior, so it is smoother than
the truth and slightly attenuated in amplitude. The site-level
correlation with the truth is well above what the non-spatial fit could
achieve (where there is no field at all).

## When to reach for SPDE

The SPDE field earns its keep when:

- spatial autocorrelation in `psi` is real and not captured by available
  covariates, and ignoring it would bias the regression coefficients
- you want continuous prediction across the study area (the field is
  defined on a mesh that can be extended beyond the sites)
- you care about uncertainty: the field's prior carries through to the
  posterior, unlike point covariates added post-hoc

It is not a free lunch:

- a mesh adds latent variables — usually a few hundred to a few thousand;
  fits are slower than non-spatial Laplace
- the prior at the prior modes (range, sigma) is what the engine currently
  uses; nested-Laplace integration over hyperparameters lands in a later
  release
- detection-side SPDE is not plumbed for the Laplace engine yet; keep the
  `spde()` term in the occupancy formula, not in `detection`

## References

* Lindgren, F., Rue, H., & Lindstrom, J. (2011). An explicit link between
  Gaussian fields and Gaussian Markov random fields: the stochastic
  partial differential equation approach. *Journal of the Royal
  Statistical Society B*, 73, 423-498.
* `?tobs` for the unified fitter and the in-formula `spde()` term.
* `vignette("occupancy-vs-inla", package = "tulpaObs")` for the
  non-spatial occupancy story.
