---
title: "Occupancy: tulpaObs vs INLA aggregated-binomial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Occupancy: tulpaObs vs INLA aggregated-binomial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
have_inla <- requireNamespace("INLA", quietly = TRUE)
set.seed(1)
```

This is a migration guide for anyone who fits occupancy or related
detection models in `INLA::inla()` and wants to know what changes when
the same model moves to tulpaObs. It maps the INLA constructs onto their
tulpaObs equivalents, shows the same model fit both ways on simulated
data, lays out where the two engines differ, and closes with the cases
where INLA is the better tool and you should stay put.

The short version: INLA fits a latent Gaussian model with a row-wise
likelihood, and a single-season occupancy model is neither. Its
likelihood is a two-component mixture at the site level, not a GLM family
INLA recognises, so the standard route is a row-stacked binomial that
fits the wrong quantity. tulpaObs writes the marginalised occupancy
likelihood directly and integrates the hyperparameters with the same
nested-Laplace recipe INLA uses, plus an exact-MCMC option INLA does not
have. Where your model *is* a latent Gaussian model with a Gaussian or
standard-exponential-family likelihood, INLA stays the faster, more
mature choice, and the closing section says so plainly.

A proper single-season occupancy model has a latent occupancy state
$z_i \sim \text{Bernoulli}(\psi_i)$ and a conditional detection process
$y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \, p_{ij})$. The site-level
marginal likelihood is

$$
L_i
\;=\;
\psi_i \prod_{j=1}^{J} p_{ij}^{y_{ij}} (1 - p_{ij})^{1 - y_{ij}}
\;+\;
\mathbf{1}\!\left\{\sum_{j} y_{ij} = 0\right\}
(1 - \psi_i).
$$

That mixture cannot be written as a row-wise GLM likelihood, so
`INLA::inla()` cannot fit it directly without `inla.rgeneric`. Michael
Glaser's `example/info.md` is the same observation: in INLA, occupancy at
the site level and detection at the visit level don't share a row
structure, and the marginalization over $z_i$ has to be programmed
explicitly.

The practical workaround in `MOT_abund_data.Rmd` §"VISIT-LEVEL BERNOULLI
MODEL" is to fit a single binomial on the row-stacked detection
histories. This vignette shows what's lost compared to the true
marginalized fit.

## Function mapping

The table below maps the INLA constructs you would reach for against the
tulpaObs call that does the same job. The grouping mirrors how an INLA
workflow is built: format the data, name the likelihood, write the latent
structure, choose the inference, and pull summaries off the result.

| Step | `INLA` construct | tulpaObs equivalent | Notes |
|---|---|---|---|
| Format data | `inla.stack()`, hand-built long frame | `tobs_data()`, `tobs_format()`, or pass `data` + `y` directly to `tobs()` | tulpaObs keeps the N x J detection matrix as `y`; no row-stacking |
| Single-season occupancy | none (needs `inla.rgeneric` for the mixture) | `tobs(family = occu())` | the mixture likelihood is built in, not user-coded |
| Multi-season / dynamic | none direct | `tobs(family = dyn_occu())` with `colonization` / `extinction` | colonisation + extinction HMM |
| Multispecies / community | per-species `f(species, model = "iid")` hyperpriors | `tobs(family = ms_occu(), species = ...)` | community random effects over species coefficients |
| Integrated (multi-source) | shared latent field across stacks | `tobs(family = int_occu())` | one occupancy state, per-source detection |
| Joint species distribution | latent-factor `f(..., model = "iid")` | `tobs(family = jsdm(), species = ...)` | shared latent factors, probit link |
| N-mixture abundance | none direct | `tobs(family = abun())` | Royle binomial-N marginal, Poisson or `mixture = "negbin"` |
| Likelihood family | `family = "binomial"` / `"poisson"` / ... | the `family` argument: `occu()`, `abun()`, `cover()`, ... | a tulpaObs family encodes the *whole* latent-state model, not just the response distribution |
| Areal spatial field | `f(region, model = "bym2", graph = adj)` | `bym2(graph = adj)` term in the `formula` | also `icar()`, `car()`, `car_proper()` |
| Continuous spatial field | `inla.spde2.pcmatern()` + `f(field, model = spde)` | `spde(lon, lat)` term in the `formula` | SPDE mesh built via tulpaMesh |
| Gaussian process | none (SPDE is the GP) | `gp(lon, lat)` term | exact GP, separate from the SPDE term |
| IID random effect | `f(group, model = "iid")` | `re(group)` or `(1 | group)` bar syntax | `lme4`-style bars are sugar for `re()` |
| Random slope | `f(group, x, model = "iid")` | `(x | group)` (correlated) / `(x \|\| group)` | correlation toggled by single vs double bar |
| Temporal field | `f(time, model = "ar1" / "rw1")` | `temporal(time, type = ...)` | AR1 / RW kernels |
| Shared field across arms | `copy = ...` in a second `f()` | `id = "u"` tag + `share("u")` in the other formula | the occupancy / detection arms share one realisation |
| Prior on fixed effect | `control.fixed = list(...)` | `priors =` (`occu_priors()`) or `control$sigma.beta` | weakly-informative quadratic prior by default |
| Hyperparameter prior | `hyper = list(prec = list(prior = "pc.prec"))` | PC priors built per block by the engine | default PC + LKJ on covariance blocks |
| Inference | (only) nested Laplace | `method = "laplace"` (default) | nested approximation, no debias |
| | | `method = "nested_laplace"` | outer grid over hyperparameters, INLA's recipe |
| | | `method = "nuts"` | exact HMC, no INLA analogue |
| | | `method = "laplace_gibbs"` / `"laplace_mi"` | Laplace body + MCMC bias correction |
| Coefficient summary | `m$summary.fixed` | `coef()`, `summary()`, `confint()` | per-process lists, `psi_` / `p_` prefixes |
| Fitted / predicted | `m$summary.fitted.values` | `fitted()` (`psi`, `p`, `z`), `predict(X.0 = )` | `z` is posterior occupancy given the history |
| Marginal effect | manual `inla.tmarginal()` | `tobs_marginal_effect()` | inverse-link curve with credible band |
| Model comparison | `control.compute = list(waic = TRUE)` | `waic()` | WAIC from the pointwise marginal log-lik |
| Held-out prediction | `NA` response rows | `NA` rows in `y` under `method = "nested_laplace"` | same convention as INLA |

The one structural difference to keep in mind from the start: in INLA the
`family` argument names a response distribution, while a tulpaObs family
names a complete latent-state generative model. `occu()` is not "the
binomial family"; it is the occupancy-state-plus-detection model whose
likelihood is the mixture below. That is why `occu()` has no INLA
one-liner equivalent and the next sections fit it side by side.

## Simulate detection histories with known $\psi$ and $p$

```{r simulate}
library(tulpaObs)

sim <- simulate_occu(
  N          = 600,
  J          = 6,
  n_occ_covs = 1,
  n_det_covs = 1,
  beta_occ   = c(0.5, 1.2),    # intercept + slope on occ_cov1
  beta_det   = c(0.0, 0.8),    # intercept + slope on det_cov1
  seed       = 42
)

dat   <- sim$data
y     <- sim$y                 # N x J detection-history matrix
truth <- sim$truth

c(naive_occ      = mean(rowSums(y) > 0),
  true_psi_mean  = mean(truth$psi),
  true_p_mean    = mean(truth$p))
```

The naive proportion of sites with any detection is biased low by
$(1 - p)^J$ at each occupied site — exactly the bias an occupancy model
is supposed to correct.

## tulpaObs: true occupancy fit via `tobs(family = occu())`

```{r fit-tulpa-laplace}
fit_tulpa_lap <- tobs(
  formula   = ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
```

For the parameter-recovery story below the EM-Laplace point estimates
underestimate detection in this small-J setting (the joint MAP picks one
mode of the $\psi$-$p$ ambiguity rather than integrating over both).
NUTS samples the full posterior and is unbiased:

```{r fit-tulpa-nuts}
fit_tulpa_nuts <- tobs(
  formula   = ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "nuts",
  control   = list(n.iter = 1500, n.warmup = 750, seed = 42, verbose = FALSE)
)
fit_tulpa_nuts
```

## INLA workaround: row-stack the detection histories, fit a single binomial

This is the structural shape Michael's `mod.visit` uses. Each site
contributes `J` Bernoulli rows; site-level occupancy covariates broadcast
across the J visits.

```{r reshape, eval = have_inla}
N <- nrow(y)
J <- ncol(y)

long <- data.frame(
  y        = as.integer(t(y)),
  site     = rep(seq_len(N), each = J),
  visit    = rep(seq_len(J), times = N),
  occ_cov1 = rep(dat$occ_cov1, each = J),
  det_cov1 = rep(dat$det_cov1, each = J)
)

head(long)
```

```{r fit-inla, eval = have_inla}
library(INLA)

m_inla <- INLA::inla(
  y ~ occ_cov1 + det_cov1,
  family = "binomial",
  data   = long,
  control.compute = list(waic = TRUE)
)

inla_fixed <- m_inla$summary.fixed[, c("mean", "0.025quant", "0.975quant")]
inla_fixed
```

## What the workaround estimates

Site $i$ contributes $J$ Bernoulli rows with success probability
$\psi_i\, p_{ij}$. Because the response is binary and the model is
row-wise, the linear predictor $\eta_{ij}$ fits

$$
P(y_{ij} = 1)
\;=\;
\psi_i \, p_{ij}.
$$

There is no factor that separates the $\psi_i$ contribution from the
$p_{ij}$ contribution: the workaround can identify the product, never the
two factors. The `occ_cov1` coefficient is pulled toward zero by the
detection probability, and the `det_cov1` coefficient soaks up whatever
signal is left.

```{r compare, eval = have_inla}
compare <- data.frame(
  parameter        = c("(occ Intercept)", "occ_cov1",
                       "(det Intercept)", "det_cov1"),
  truth            = c(truth$beta_occ, truth$beta_det),
  tulpaObs_NUTS    = c(
    fit_tulpa_nuts$means["psi_(Intercept)"],
    fit_tulpa_nuts$means["psi_occ_cov1"],
    fit_tulpa_nuts$means["p_(Intercept)"],
    fit_tulpa_nuts$means["p_det_cov1"]
  ),
  tulpaObs_Laplace = c(
    fit_tulpa_lap$means["psi_(Intercept)"],
    fit_tulpa_lap$means["psi_occ_cov1"],
    fit_tulpa_lap$means["p_(Intercept)"],
    fit_tulpa_lap$means["p_det_cov1"]
  ),
  INLA_workaround  = c(
    inla_fixed["(Intercept)", "mean"],
    inla_fixed["occ_cov1",    "mean"],
    NA_real_,    # no separate detection intercept exists
    inla_fixed["det_cov1",    "mean"]
  ),
  row.names = NULL
)
compare
```

NUTS lands on the truth on both processes. Laplace gives biased point
estimates with this $J = 6$ setting — the joint MAP underestimates
detection probability when sites are weakly identified. INLA's
workaround returns *one* intercept that mixes occupancy and detection,
and the slopes on occupancy-side and detection-side covariates pull
toward each other.

## Predictive divergence at the site level

Despite the Laplace point-estimate bias, the predicted *per-site*
occupancy probability still correlates strongly with the truth — the
bias mostly cancels in the linear predictor. The INLA workaround cannot
recover site-level $\psi$ even in principle, because what it estimates is
not $\psi$.

```{r predict, eval = have_inla}
# tulpaObs returns per-site psi_hat. fitted() also returns $p (per-site
# detection probability) and $z (posterior occupancy probability given
# the detection history); $psi is the right comparator for true psi.
psi_hat_nuts <- fitted(fit_tulpa_nuts)$psi
psi_hat_lap  <- fitted(fit_tulpa_lap)$psi

# INLA workaround: predicted "any-visit detection" probability at the
# site, back-projected to a single row per site. This is the product
# psi * p, not psi.
X_site       <- model.matrix(~ occ_cov1 + det_cov1, data = dat)
eta_site     <- as.numeric(X_site %*% inla_fixed[, "mean"])
psi_hat_inla <- plogis(eta_site)

cor_with_truth <- c(
  tulpaObs_NUTS    = cor(psi_hat_nuts, truth$psi),
  tulpaObs_Laplace = cor(psi_hat_lap,  truth$psi),
  INLA_workaround  = cor(psi_hat_inla, truth$psi),
  naive_any        = cor(as.numeric(rowSums(y) > 0), truth$psi)
)
round(cor_with_truth, 3)
```

For population-trend questions ("did $\psi$ change?") the workaround
answers a different question ("did $\psi \cdot p$ change?"), and its
trend can be driven by detection improvements that have nothing to do
with the species' actual occupancy.

```{r predict-plot, eval = have_inla, fig.height = 4}
op <- par(mfrow = c(1, 2))
plot(truth$psi, psi_hat_nuts, pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "true psi", ylab = "tulpaObs psi_hat (NUTS)",
     main = "occu(): recovers psi")
abline(0, 1, col = 2, lwd = 2)
plot(truth$psi, psi_hat_inla, pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "true psi", ylab = "INLA aggregated p(detection)",
     main = "binomial workaround: confounds psi and p")
abline(0, 1, col = 2, lwd = 2)
par(op)
```

## What about the "detectability proxies" trick?

Michael's `mod.visit` adds plot-level fixed effects (`area.sc`,
`richness.sc`, `sample.density.sc`, `sample.extent.sc`) to the same
linear predictor that drives the occupancy-side covariates. Those proxies
*can* improve fit if they correlate with $p$, but their coefficients are
still confounded with the occupancy coefficients on the same linear
predictor — there is no hierarchical structure in the model that
attributes one slope to $\psi$ and another to $p$. The workaround does
not become an occupancy model by adding more covariates.

The only way to recover $\psi$ and $p$ separately from a row-wise INLA
likelihood is `inla.rgeneric`: a custom likelihood that implements the
mixture above. That is the same level of work as building tulpaObs's
EM/NUTS engine; if you're going to write the marginalized likelihood,
you may as well use the package that already has it.

## A second example: an IID site random effect both ways

The first comparison was a model INLA cannot fit at all. This one is a
model both packages fit, so it shows the migration on equal footing: a
single-season occupancy model with an extra site-level random effect on
occupancy. In INLA that random effect is `f(site, model = "iid")` on the
linear predictor; in tulpaObs it is an `re(site)` term, or the `lme4`-style
`(1 | site)` bar, written straight into the occupancy formula. The
detection process keeps a covariate so there are two arms to separate.

```{r sim2}
sim2 <- simulate_occu(
  N          = 400,
  J          = 5,
  n_occ_covs = 1,
  n_det_covs = 1,
  beta_occ   = c(0.2, 0.9),
  beta_det   = c(0.4, 0.5),
  seed       = 7
)

dat2 <- sim2$data
y2   <- sim2$y
dat2$site <- factor(seq_len(nrow(dat2)))   # one group per site
```

The tulpaObs side reads the random effect as a formula term. When a
random effect carries no slope, the Laplace path debiases the variance
component with adaptive Gauss-Hermite quadrature by default; `re(site)`
and `(1 | site)` are the same term written two ways.

```{r fit-tulpa-re}
fit_re <- tobs(
  formula   = ~ occ_cov1 + (1 | site),
  data      = dat2,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y2,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
coef(fit_re)
```

The same model in INLA needs the mixture likelihood it does not have, so
the honest INLA-side comparison is again the row-stacked binomial, now
with the site random effect added as `f(site, model = "iid")`. The
random effect absorbs the same site-level heterogeneity, but it rides on
the confounded $\psi \cdot p$ linear predictor rather than on the
occupancy process alone.

```{r fit-inla-re, eval = have_inla}
N2 <- nrow(y2); J2 <- ncol(y2)
long2 <- data.frame(
  y        = as.integer(t(y2)),
  site     = rep(seq_len(N2), each = J2),
  occ_cov1 = rep(dat2$occ_cov1, each = J2),
  det_cov1 = rep(dat2$det_cov1, each = J2)
)

m_inla_re <- INLA::inla(
  y ~ occ_cov1 + det_cov1 + f(site, model = "iid"),
  family = "binomial",
  data   = long2,
  control.compute = list(waic = TRUE)
)
m_inla_re$summary.fixed[, c("mean", "0.025quant", "0.975quant")]
```

Two practical points make the migration easier here. First, the grouping
factor is a plain factor column in the site-level `data`, the same way
INLA wants an integer or factor index for `f(site, ...)`, so the data
preparation is identical on both sides. Second, the term placement
carries meaning: an `re(site)` written in the occupancy `formula` is a
site effect on occupancy, while the same term written in the `detection`
formula would be a site effect on detection, and the two are different
models. INLA makes the same distinction by which linear predictor the
`f()` enters, so the habit transfers directly. If you want one shared
site effect across both arms, tag it with `id` and `share()` it into the
second formula, which is the tulpaObs spelling of INLA's `copy` feature.

The migration is mechanical for the structured terms: every `f(...)` in
the INLA formula becomes a term inside the tulpaObs `formula`, and the
term enters whichever linear predictor it is written in. What does not
migrate is the likelihood. The INLA fit still estimates the product
$\psi \cdot p$ on a single linear predictor, so its `occ_cov1` slope is
attenuated toward zero for the reason the first example showed, and its
random-effect variance mixes site-level occupancy heterogeneity with
site-level detection heterogeneity. The tulpaObs fit keeps the two
processes apart and the random effect lives on occupancy alone.

## Key differences

**Algorithm.** INLA is a nested Laplace approximation with no debiasing
step: it approximates each latent marginal by a (skew-corrected) Gaussian
at the conditional mode and integrates the hyperparameters on an outer
grid. That is exact when the latent field is Gaussian and the
approximation degrades smoothly as the conditional posterior departs from
Gaussian. tulpaObs offers the same nested-Laplace path
(`method = "nested_laplace"`), and its default `method = "laplace"` is an
EM-Laplace loop that marginalises the latent occupancy state in closed
form. On top of that it adds two things INLA does not have: an exact HMC
sampler (`method = "nuts"`) that integrates the full posterior rather than
sitting at a mode, and Laplace-plus-correction routes
(`method = "laplace_gibbs"` / `"laplace_mi"`) that run a short MCMC or
multiple-imputation correction on the residual non-Gaussian directions.
The design is the nested approximation INLA pioneered, with an optional
debias layer bolted on for the directions where the approximation is
biased.

**Speed.** The Laplace and nested-Laplace paths are deterministic and
return in a fraction of a second on a few hundred sites, the same order
of cost as an INLA call on a comparable latent Gaussian model; both pay
for the outer hyperparameter grid the same way. NUTS costs more because
it samples, and the `occupancy.Rmd` vignette puts a short NUTS run on a
few hundred sites at seconds rather than the sub-second Laplace fit. The
package ships timing scripts under `tests/benchmark*.R`, but they are
live-timing harnesses rather than recorded results, so this guide does
not quote a head-to-head number; run them on your own hardware if you
need the figure for your data size. The honest summary is that for the
deterministic paths the two engines are in the same ballpark, and the
extra cost in tulpaObs buys the exact-MCMC option when you want it.

**Output format.** INLA returns marginals: `summary.fixed`,
`summary.hyperpar`, and `summary.fitted.values`, each a data frame of
posterior summaries, with the full marginal densities available through
`inla.tmarginal()` and friends. tulpaObs returns a fitted object that
answers the standard R generics. `coef()` gives per-process coefficient
lists with `psi_` and `p_` prefixes so the two arms never run together;
`summary()` adds SDs and credible bounds (plus Rhat and ESS under NUTS);
`confint()` and `vcov()` read the fixed-effect block; `fitted()` returns
per-site `psi`, `p`, and the posterior occupancy `z`; `predict(X.0 = )`
evaluates occupancy at a chosen design; `tobs_marginal_effect()` returns
an inverse-link response curve with a credible band; and `waic()`
gives WAIC from the pointwise marginal log-likelihood. The mental shift
is from "pull a marginal off a list" to "call the accessor the model
type provides".

**Prior specification.** INLA priors are set through `control.fixed` for
the fixed effects and `hyper` lists per `f()` term, with PC priors the
modern default. tulpaObs uses weakly-informative quadratic priors on the
fixed effects by default (they break the $\psi$-$p$ identifiability ridge
at small $J$, which is why the detection intercept is pulled toward
$p = 0.5$); you override them with a `priors` argument built from
`occu_priors()`, or disable them with `priors = FALSE` to recover the
unpenalised MAP, or rescale the Gaussian width with `control$sigma.beta`.
Hyperparameter priors on the covariance and spatial blocks are
PC-and-LKJ by construction, built per block by the engine, the same
family of priors INLA recommends. The difference is mostly where you
write them: INLA threads priors through `control.*` lists, tulpaObs
threads them through the `priors` argument and a small number of
`control` knobs.

## When NOT to switch

INLA is the better tool in several honest cases, and this guide would be
dishonest to hide them.

**Your model is a genuine latent Gaussian model with a standard
likelihood.** If you are fitting a spatial GLM, a disease-mapping BYM2
model, a geostatistical SPDE, or any model whose likelihood is one of
INLA's built-in families on a Gaussian latent field, INLA does exactly
that, fast, and with a decade of validation behind it. tulpaObs is built
for latent-state observation models where the latent state is discrete
(occupancy, abundance) and marginalises out, which is precisely the case
INLA handles awkwardly. Outside that niche the comparison runs the other
way.

**You need INLA's breadth of latent structures.** INLA's `f()` zoo,
besag, bym2, rw1, rw2, ar1, ar(p), seasonal, Matern SPDE, copies,
replicates, and the rest, is broad and battle-tested. tulpaObs covers the
spatial, temporal, and random-effect structures listed in the mapping
table, but it does not match INLA term for term, and a model that leans
on a structure tulpaObs does not implement belongs in INLA.

**You depend on INLA's maturity and ecosystem.** INLA has years of
published applications, a large user base, `inlabru` for point-process
and nonlinear-predictor work, and extensive diagnostics. tulpaObs is a
young package. For production work where a reviewer expects INLA, or
where you need a feature only the INLA ecosystem provides, the
established tool wins.

**Your occupancy design has no replication.** This is not an
INLA-versus-tulpaObs question at all. A single visit per site cannot
separate occupancy from detection in any engine, and switching packages
will not rescue an unidentified design. If your data lack repeat visits,
neither tool fits a real occupancy model, and the honest move is the one
the occupancy vignette describes: add visits, borrow strength from a
spatial structure, or report a detection-uncorrected map.

The case for moving to tulpaObs is narrow and specific: you have a
latent-state observation model with replication, the latent state
marginalises out, and you want either the exact marginalised likelihood
INLA cannot write directly or the exact-MCMC check INLA does not provide.
For everything else, INLA is a fine place to stay.

## References

* `example/MOT_abund_data.Rmd` §"VISIT-LEVEL BERNOULLI MODEL" — Michael
  Glaser's binomial workaround, with detectability proxies.
* `example/info.md` — Michael's note on why a proper single-species
  occupancy model is hard in INLA (the marginalization problem).
* `?tobs` and `?occu` — the tulpaObs entry points used here.
* `?simulate_occu` — the detection-history simulator.
