---
title: "Occupancy with a continuous SPDE spatial field"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Occupancy with a continuous SPDE spatial field}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

```{r load, message = FALSE}
library(tulpaObs)
library(ggplot2)
```

## Introduction

Occupancy data are usually spatially structured: nearby sites share habitat
and species pressures that the row-wise covariates can't fully capture. The
Matern field via the SPDE approach (Lindgren, Rue & Lindstrom, 2011) gives
a continuous spatial random effect with an interpretable range and marginal
SD, and an `spde()` term in the formula wires it through tulpaObs's
Laplace EM engine.

The choice between a continuous field and an areal field follows from what
your sites are. Areal data arrive pre-divided into regions joined by borders:
census tracts, grid cells, administrative units. The neighbour graph is given,
and an ICAR or BYM2 field rides on "who borders whom". Point data arrive as raw
coordinates, transects walked, traps set, plots placed on a continuous
landscape, with no natural partition into cells. Forcing those points into an
adjacency means binning them into arbitrary squares first, and the answer then
depends on the grid you drew. A continuous field skips that step. It places a
Matern process directly on the coordinates, so two sites 200 m apart are
correlated more strongly than two sites 2 km apart, and the strength is set by
a range parameter rather than a hand-drawn graph. That is the case this
vignette covers: occupancy surveyed at scattered locations, with a smooth
spatial signal in `psi` that no covariate measures.

Ignoring that signal costs twice. The occupancy residuals come out spatially
correlated, which means the effective sample size is smaller than the site
count and the standard errors run too narrow. And when an occupancy covariate
is itself patterned across space, it competes with the unmeasured spatial
signal for the same variance, so the fitted slope absorbs part of that signal
and drifts away from its true value. A spatial field gives the location-driven
part of occupancy a place to live, separate from the covariate effects, so the
slope is read against a background that already accounts for where a site sits.
The field also turns the missing-data problem around: a location with no survey
borrows occupancy from its neighbours through the smooth surface, which a
covariate-only model cannot do at all.

This vignette simulates an occupancy dataset with a known smooth spatial
signal, fits a non-spatial model and an SPDE model, and compares them on
parameter recovery and the latent state. It then predicts the field at
unsurveyed coordinates, checks the residuals for leftover spatial structure,
and closes with the rules of thumb that decide when the field is worth fitting.

## The model

The single-season occupancy model has two layers. A latent presence state
$z_i \in \{0, 1\}$ at site $i$ is Bernoulli with occupancy probability
$\psi_i$, and the detection history $y_{ij}$ over $J$ visits is Bernoulli with
detection probability $p_i$ when $z_i = 1$ and zero otherwise. The occupancy
logit carries the spatial field:

$$
\operatorname{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol\beta + u(\mathbf{s}_i),
\qquad
\operatorname{logit}(p_i) = \mathbf{w}_i^\top \boldsymbol\alpha .
$$

Here $\mathbf{x}_i$ is the occupancy covariate row at site $i$,
$\boldsymbol\beta$ its coefficients, $\mathbf{w}_i$ and $\boldsymbol\alpha$ the
detection covariates and their coefficients, and $\mathbf{s}_i$ the spatial
coordinates of site $i$. The term $u(\mathbf{s})$ is a continuous Gaussian
random field: a smooth surface defined everywhere on the study area, evaluated
at each site's coordinates.

The field is a Matern process. Its covariance between two points depends only
on the distance $d$ between them,

$$
\operatorname{Cov}\bigl(u(\mathbf{s}), u(\mathbf{s}')\bigr)
   = \sigma^2 \, \frac{2^{1-\nu}}{\Gamma(\nu)} (\kappa d)^\nu K_\nu(\kappa d),
$$

with marginal standard deviation $\sigma$, smoothness $\nu$, and inverse range
$\kappa$; $K_\nu$ is the modified Bessel function of the second kind. The range
$\rho = \sqrt{8\nu}/\kappa$ is the distance at which correlation drops to about
0.1, the practical reach of the field. Lindgren, Rue & Lindstrom (2011) show
this Matern field is the stationary solution of a stochastic partial
differential equation, and that solving the SPDE on a triangular mesh gives a
Gaussian Markov random field: a discrete approximation $u(\mathbf{s}) \approx
\sum_k A_{ik} u_k$ where $u_k$ are values at the mesh nodes and $A$ is a sparse
projection matrix that interpolates them to the site. The node values $u_k$
have a sparse precision matrix built from the mesh geometry and
$(\kappa, \sigma)$, which is what keeps the field cheap even on thousands of
sites. The `spde()` term builds the mesh, the projection $A$, and the precision
at the prior modes of $\rho$ and $\sigma$.

The smoothness $\nu$ is fixed rather than estimated, and the engine defaults to
$\nu = 1$, which makes the field once-differentiable and is the common choice
for ecological surfaces; it is rarely identifiable from data the way $\rho$ and
$\sigma$ are, so fixing it costs little. The two hyperparameters that do carry
information, the range $\rho$ and the marginal SD $\sigma$, are set through PC
priors (Simpson et al., 2017). A PC prior is given as a tail statement: for the
range, $P(\rho < \rho_0) = p$, naming a small reference range $\rho_0$ and the
probability the true range falls below it; for the SD, $P(\sigma > \sigma_0) =
p$. That form penalises the field toward the simpler no-field model (infinite
range, zero SD) and pulls back only as far as the data demand, which is why a
range or SD prior set at a sensible scale rarely over-fits. The field enters the
occupancy logit with mean zero, so the intercept $\beta_0$ carries the overall
occupancy level and the field carries only the contrasts between locations,
matching the way the data identify the two.

## Simulating data

A spatial fit is only honest against a known truth, because a recovered map
that merely looks plausible proves nothing. We set a smooth field on the unit
square by hand, draw occupancy from it, simulate a detection history with
imperfect detection, and ask the fit to find the field again. The field is
$0.8\cos(3x)\sin(3y)$ on the logit scale: a smooth ripple with an amplitude of
about 0.8, peaks and troughs at opposite corners, and real structure that no
single intercept can fake. The amplitude matters. At 0.8 on the logit scale the
field moves occupancy from roughly 0.3 to 0.7 across the square, enough signal
for the model to separate from binomial noise, and small enough that detection
still has teeth.

```{r simulate}
n_sites <- 400
J <- 8
coords <- cbind(runif(n_sites), runif(n_sites))

# A smooth spatial signal: amplitude ~0.8 on the logit scale
u_true <- 0.8 * cos(3 * coords[, 1]) * sin(3 * coords[, 2])

x_cov   <- rnorm(n_sites)
det_cov <- rnorm(n_sites)
beta_occ_true <- c(-0.5, 0.7)
beta_det_true <- c(-0.3, 0.4)

eta_occ <- beta_occ_true[1] + beta_occ_true[2] * x_cov + u_true
psi <- plogis(eta_occ)
z <- rbinom(n_sites, 1, psi)

eta_det <- beta_det_true[1] + beta_det_true[2] * det_cov
p <- plogis(eta_det)

y <- matrix(0L, n_sites, J)
for (i in seq_len(n_sites)) {
  if (z[i] == 1L) y[i, ] <- rbinom(J, 1, p[i])
}

dat <- data.frame(occ_cov = x_cov, det_cov = det_cov,
                  lon = coords[, 1], lat = coords[, 2])
c(true_occupancy_rate = mean(z),
  naive_any_detection = mean(rowSums(y) > 0))
```

The naive any-detection rate is biased low by `(1 - p)^J` at each occupied
site; a proper occupancy fit should correct that bias, and a spatial fit
should additionally recover the smooth field that drives the
site-to-site variation in `psi`. The `data.frame` keeps the coordinate columns
`lon` and `lat` alongside the covariates, because the `spde()` term resolves
those columns from `data` the way `mgcv` resolves `s(lon, lat)`. There is no
separate spatial-field simulator in the package; `simulate_occu()` draws
covariate-only occupancy, so a known continuous field is set by hand as above.
A map of the truth shows what the fit has to recover.

```{r truth-map, fig.height = 4, fig.alt = "Scatter of the 400 site coordinates coloured by the true latent spatial field, a smooth ripple across the unit square"}
truth_df <- data.frame(lon = coords[, 1], lat = coords[, 2], u = u_true)
ggplot(truth_df, aes(lon, lat, colour = u)) +
  geom_point(size = 2) +
  scale_colour_viridis_c() +
  labs(title = "True latent field u(s)", colour = "u") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The ripple is the part we want back: a smooth surface that the non-spatial fit
has no place to put, so it will leak into the intercept and the residuals.

Two design choices in the simulation make the test fair. Eight visits per site
at moderate detection give the model enough replication to pull occupancy apart
from detection; with one or two visits the latent state is barely identified and
any field would inherit that ambiguity rather than the spatial signal. And the
sites are scattered uniformly over the square rather than clustered, so every
part of the field has data nearby and the recovery measures the field, not the
sampling design. A clustered design would leave large gaps where the field is
pure prior, which is a real situation but a different question from whether the
estimator works. The covariate `occ_cov` is independent of the coordinates by
construction, so the simulation isolates the field: any shift in the occupancy
slope between the non-spatial and SPDE fits comes from the field absorbing
residual variance, not from confounding between the covariate and the spatial
signal. On real data those two are usually entangled, which is the harder case
the field is built for, but the clean simulation is the right place to confirm
the machinery first.

## Fitting the model

The spatial field is written directly into the occupancy formula as an
`spde()` term, the way `mgcv` writes `s(lon, lat)`. `spde()` builds a
triangular mesh over the observation coordinates (via tulpaMesh) and
constructs the SPDE precision at the prior modes of `prior_range` and
`prior_sigma`. The mesh covers the data extent plus a small buffer;
`max_edge = c(inner, outer)` controls the spacing inside and outside the
data. The coordinate columns (`lon`, `lat`) are taken from `data`. We fit a
non-spatial model first as the baseline, then the SPDE model, so the two are
directly comparable.

```{r fit-non-spatial}
fit_ns <- tobs(
  formula   = ~ occ_cov,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
round(fit_ns$means, 3)
```

The SPDE call adds one term to the occupancy formula and changes nothing else.
The `prior_range = c(0.3, 0.5)` is a PC prior on the range: the first number is
the median range guess (0.3 of the unit square), the second the probability the
range is below it. `prior_sigma = c(0.7, 0.5)` is the matching prior on the
marginal SD. `max_edge = c(0.3, 0.6)` sets the mesh spacing inside and outside
the data; `nu = 1` fixes the Matern smoothness.

```{r fit-spde}
fit_sp <- tobs(
  formula   = ~ occ_cov + spde(lon, lat,
                               max_edge    = c(0.3, 0.6),
                               nu          = 1,
                               prior_range = c(0.3, 0.5),
                               prior_sigma = c(0.7, 0.5)),
  data      = dat,
  family    = occu(),
  detection = ~ det_cov,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)
)
round(fit_sp$means, 3)
```

The fixed-effect coefficients come back through `coef()`, split by process:
`psi` for occupancy, `p` for detection. The standard errors and credible
intervals are in `summary()`, which reports the estimate, standard error, and
the lower and upper credible bounds for every fixed effect.

```{r coef-summary}
coef(fit_sp)
round(summary(fit_sp), 3)
```

The fitted SPDE field itself is stored on the fit as `spatial_field`, a
length-`n_mesh` numeric vector at the mesh nodes, the posterior mode of the
node values $u_k$. The mesh that carries it lives under
`fit_sp$spatial$tulpa_spec`, including the projection matrix `A` and the node
count `n_mesh`.

```{r field-extract}
u_hat <- fit_sp$spatial_field
mesh  <- fit_sp$spatial$tulpa_spec$mesh
c(n_mesh_nodes  = fit_sp$spatial$tulpa_spec$n_mesh,
  field_length  = length(u_hat),
  field_sd      = round(sd(u_hat), 3))
```

The 400 sites expand to a few hundred mesh nodes, more nodes than sites because
the mesh adds a buffer of nodes around the data so the field does not distort at
the edges. The field SD around 0.4 is the fitted analogue of the simulated
amplitude, damped by the SPDE prior's smoothing.

## Checking the fit

The first check is recovery of the field at the sites. Project the node values
to the observation locations through the spec's `A` matrix and line them up
against the truth. A correlation well above zero says the field found the
signal; the non-spatial fit has no field at all, so its only comparison is the
flat population mean.

```{r field-recovery, fig.height = 4, fig.alt = "Two panels: fitted field at sites against the true field near the identity line, and the site coordinates coloured by fitted field sign"}
field_at_sites <- as.numeric(fit_sp$spatial$tulpa_spec$A %*% u_hat)
corr <- cor(field_at_sites, u_true)
cat(sprintf("Correlation field-at-sites vs truth: %.3f\n", corr))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(u_true, field_at_sites, pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "truth u(s)", ylab = "fitted u_hat(s)",
     main = "Field recovery at sites")
abline(0, 1, col = 2, lwd = 2)
plot(coords[, 1], coords[, 2], pch = 16, cex = 0.6,
     col = ifelse(field_at_sites > 0, "red", "blue"),
     xlab = "x", ylab = "y", main = "Fitted field sign")
par(op)
```

The fitted field is regularized by the SPDE prior, so it is smoother than the
truth and slightly attenuated in amplitude. The site-level correlation with the
truth is well above what the non-spatial fit could achieve, where there is no
field at all.

The formal check is whether any spatial structure is left in the residuals. If
the field has absorbed the smooth signal, the occupancy residuals should look
spatially random; leftover autocorrelation means the field missed something or
the range was set wrong. `check_model()` runs the model diagnostics and, given
the coordinates, computes Moran's I on the occupancy residuals, the standard
test for spatial autocorrelation. It also reports WAIC and a posterior
predictive Bayesian p-value. Run it on both fits.

```{r moran-check}
chk_ns <- check_model(fit_ns, coords = coords, n.samples = 200, plot = FALSE)
chk_sp <- check_model(fit_sp, coords = coords, n.samples = 200, plot = FALSE)
```

A significant Moran's I on the non-spatial residuals (small p-value) is the
signature of a spatial signal the covariates missed. After the field is added,
the Moran's I should move toward zero and its p-value above 0.05: the field
soaked up the structure the non-spatial residuals carried. The Bayesian
p-value near 0.5 says the model reproduces the detection histories; values
below 0.05 or above 0.95 would flag a misfit the field did not cure.

Moran's I is computed on the occupancy residuals, the difference between the
realised ever-detected indicator and the smoothed occupancy state, weighted by
inverse distance between sites. The test asks whether residuals at nearby sites
are more alike than residuals at distant sites, which is exactly the structure a
spatial field is meant to remove. Reading the two `check_model()` outputs side by
side is the cleanest single-number summary of whether the field earned its
place: the non-spatial residuals should carry autocorrelation the SPDE residuals
do not. If the SPDE residuals still test significant, the field is
under-fitting, usually because the prior range is too long for the signal, and
shortening it (a smaller `prior_range` median) lets the field track finer
variation. If neither fit shows autocorrelation, there was no spatial signal to
capture and the field is not needed; the comparison section makes that call
formal.

A second diagnostic worth a glance is the recovered field amplitude against the
truth. The field SD reported in the fitting section sits around 0.4, below the
simulated 0.8, the expected attenuation from a smoothing prior: the SPDE trades
a little amplitude for a smoother surface, which is the bias-variance bargain
the prior strikes. A recovered SD near zero would say the field collapsed and
found no signal; a recovered SD matching or exceeding the truth would suggest
the prior range is so short the field is chasing noise. The value here, damped
but well clear of zero, is the healthy middle.

## Interpreting results

The occupancy coefficients shifted between the two fits, and the reason is the
field. In the non-spatial model the smooth signal had nowhere to go, so part of
it leaked into the intercept and inflated the residual variance. The SPDE
absorbs those site-to-site logit deviations, which frees the intercept to sit
at the true level and tightens the slope. Detection sits on its own predictor,
so the detection coefficients barely move.

```{r compare-betas}
compare <- data.frame(
  parameter = c("psi_(Intercept)", "psi_occ_cov",
                "p_(Intercept)",   "p_det_cov"),
  truth     = c(beta_occ_true, beta_det_true),
  non_spat  = unname(round(fit_ns$means, 3)),
  spde      = unname(round(fit_sp$means, 3))
)
compare
```

The detection coefficients are essentially unchanged: detection covariates
don't carry spatial structure in this simulation, so the detection submodel
sees the same data both times. Occupancy coefficients shift because the SPDE
absorbs site-to-site logit deviations that the non-spatial fit had to attribute
to the global intercept.

Uncertainty on the occupancy slope reads off `confint()`, which returns the
Wald credible interval from the Laplace posterior. The interval should bracket
the true slope of 0.7. Reporting the SPDE interval beside the non-spatial one
shows what the field does to the error: by accounting for the spatial signal,
it gives the slope its honest standard error rather than the optimistic one a
model that ignores correlation reports.

```{r confint}
ci_ns <- confint(fit_ns)["psi_occ_cov", ]
ci_sp <- confint(fit_sp)["psi_occ_cov", ]
rbind(non_spatial = round(ci_ns, 3), spde = round(ci_sp, 3))
```

A map of the fitted field with its sign and magnitude shows where occupancy
runs high and low after the covariate is removed. Colour the sites by the
projected field value: the gradient should trace the simulated ripple, high in
the corners where $\cos(3x)\sin(3y)$ peaks, low where it troughs.

```{r field-map, fig.height = 4, fig.alt = "Site coordinates coloured by the fitted SPDE field, tracing the simulated ripple across the unit square"}
fitted_df <- data.frame(lon = coords[, 1], lat = coords[, 2],
                        u_hat = field_at_sites)
ggplot(fitted_df, aes(lon, lat, colour = u_hat)) +
  geom_point(size = 2) +
  scale_colour_viridis_c() +
  labs(title = "Fitted SPDE field at sites", colour = "u_hat") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The fitted surface is a damped, smoothed version of the truth map from the
simulation section: the SPDE prior pulls neighbouring sites together, so the
recovered ripple has lower amplitude and softer edges than the field that
generated the data, while keeping the location of the peaks and troughs.

## Prediction

The continuous field buys something an areal model on these points could not:
occupancy at coordinates that were never surveyed. The field is defined
everywhere on the mesh, so a new location inside the mesh extent has a field
value, interpolated from the node values through the same FEM basis that built
the fit. Build a regular grid of prediction coordinates over the unit square,
project the fitted node field onto it with a fresh projection matrix from
`tulpaMesh::fem_matrices()`, and add it to the fixed-effect predictor.

```{r predict-grid}
grid <- expand.grid(lon = seq(0.05, 0.95, length.out = 30),
                    lat = seq(0.05, 0.95, length.out = 30))
A_new <- tulpaMesh::fem_matrices(mesh, obs_coords = as.matrix(grid))$A
field_new <- as.numeric(A_new %*% u_hat)
str(field_new)
```

The fixed-effect part of the prediction comes from `predict()` with a design
matrix at the new locations. Setting `occ_cov` to its mean isolates the spatial
surface: the prediction is occupancy at an average-covariate site, varying only
through the field. `predict()` returns the posterior mean, SD, and quantiles
of `psi` from the fixed-effect draws, so the SD column carries the fixed-effect
uncertainty, and the field is added on the logit scale before the inverse-link.

```{r predict-psi}
X0 <- cbind(1, rep(mean(dat$occ_cov), nrow(grid)))
pred_fixed <- predict(fit_sp, X.0 = X0)
eta_grid <- qlogis(pred_fixed$mean) + field_new
psi_grid <- plogis(eta_grid)
grid$psi <- psi_grid
summary(grid$psi)
```

Mapping the predicted occupancy over the grid gives the continuous surface,
filled in between the surveyed points. This is the deliverable a continuous
field exists for: a smooth occupancy map at every coordinate, not just at the
sites that happened to be visited.

```{r predict-map, fig.height = 4, fig.alt = "Predicted occupancy probability over a 30 by 30 grid of new coordinates, a smooth filled surface with the surveyed sites overlaid"}
ggplot(grid, aes(lon, lat, fill = psi)) +
  geom_raster() +
  geom_point(data = dat, aes(lon, lat), inherit.aes = FALSE,
             size = 0.4, colour = "white", alpha = 0.5) +
  scale_fill_viridis_c() +
  labs(title = "Predicted occupancy (mean covariate)", fill = "psi") +
  theme(panel.background = element_rect(fill = "transparent"),
        plot.background  = element_rect(fill = "transparent"))
```

The fixed-effect SD in `pred_fixed$sd` propagates the coefficient uncertainty
into the map, and a wider band appears where the data are sparse and the
intercept is less pinned. The field's own posterior uncertainty is not in this
band: the Laplace path stores the field mode, not a per-node SD, so the
interval here is conditional on the fitted field. For a fully marginalised
field band on coordinates, the areal `method = "nested_laplace"` path computes
a per-cell `psi_lower` / `psi_upper` directly; the continuous-field counterpart
reports the mode.

## Model comparison

In simulation a high correlation with a known truth says the field works. On
real data there is no truth, so the question becomes whether the field improves
prediction enough to justify the latent variables it adds. WAIC estimates
out-of-sample predictive accuracy from the pointwise log-likelihood, lower is
better, and it docks the model for the effective parameters it spends. One
caveat applies here: the tulpaObs WAIC scores the fixed-effect predictor and
does not fold the SPDE field contribution into the per-site likelihood, so on a
spatial fit it is conditional on the fixed effects. It still answers whether the
covariate-and-detection structure predicts the histories, and it is the
apples-to-apples comparison against the non-spatial fit, which has no field
either.

```{r waic}
w_ns <- waic(fit_ns)
w_sp <- waic(fit_sp)
data.frame(
  model  = c("non_spatial", "spde"),
  waic   = round(c(w_ns$waic, w_sp$waic), 1),
  p_waic = round(c(w_ns$p_waic, w_sp$p_waic), 2),
  elpd   = round(c(w_ns$elpd, w_sp$elpd), 1)
)
```

The cleaner test for the field on these data is the residual diagnostic from
the checking section: a non-spatial fit leaves spatial autocorrelation in its
occupancy residuals, the SPDE fit removes it. That is the direct evidence the
field is doing work, where WAIC on the fixed-effect predictor is the conditional
view. Pairing the two, residual Moran's I for whether the field is needed and
WAIC for whether the mean structure predicts, is the honest read.

The comparison is asymmetric by design. WAIC and the residual test do not
reward the field for flexibility alone; they reward it only for flexibility the
data use. Had the simulated occupancy been spatially flat, the field would have
found nothing to fit, the residual Moran's I would have been non-significant in
both fits, and the WAIC gap would have closed or reversed as the field's extra
latent variables bought no predictive gain. A model selection that punishes an
unused field as readily as it credits a used one is the kind that keeps you
honest on real data, where the truth map is unavailable and the only evidence is
out-of-sample fit and leftover structure.

A second comparison is the field correlation itself, which only a simulation
affords. The SPDE recovers the smooth surface at a correlation of roughly 0.4
to 0.5 with the truth at this sample size and mesh, where the non-spatial model
has no surface to recover. Put the two evidence lines in one table.

```{r compare-table}
data.frame(
  metric      = c("field cor (vs truth)", "WAIC (fixed-effect)"),
  non_spatial = c(NA_real_,    round(w_ns$waic, 1)),
  spde        = c(round(corr, 3), round(w_sp$waic, 1))
)
```

The field correlation is the simulation-only number, available only for the
SPDE fit since the non-spatial model has no surface to recover, and the WAIC is
conditional on the fixed effects. Read alongside the residual Moran's I from the
`check_model()` output above, they say the field captured a real spatial signal
that the non-spatial model could not, which is exactly how the data were built.

## Practical guidance

A few rules of thumb for continuous occupancy fields, with the numbers that
make them usable.

- **Set the range prior to roughly a third of the study extent, then check.**
  `prior_range = c(0.3, 0.5)` on a unit square says the median range is 0.3 of
  the domain, the distance over which sites stay correlated, with even odds it
  is smaller. A range much shorter than the typical inter-site spacing leaves
  the field unable to borrow strength (every site is effectively isolated); a
  range near the full extent makes the field nearly flat and indistinguishable
  from the intercept. Start at a third of the extent, fit, and look at the
  recovered field SD: if it collapses toward zero the prior range is fighting
  the data.

- **Tie `max_edge` to the range, inner edge well below it.** The mesh
  resolution has to be fine enough to represent the field's wiggles. A useful
  default is an inner `max_edge` of one-fifth to one-third of the prior range,
  with the outer edge two to three times the inner so the buffer stays cheap.
  Here `max_edge = c(0.3, 0.6)` against a range of 0.3 is at the coarse end and
  gives a few hundred nodes; halving the inner edge to 0.15 roughly doubles the
  node count and sharpens the field, at a proportional cost in fit time. The
  N = 1500 recovery tier in the package's own tests uses `c(0.18, 0.45)` for
  this reason.

- **At least a few hundred sites, and detection that is not too sparse.** The
  field's range and marginal SD are learned from how occupancy varies across
  space, and a few dozen scattered points leave both barely informed. Below
  roughly 200 sites a continuous field is mostly prior, and a covariate-only
  fit is the more honest choice. The recovery here uses 400 sites with 8 visits
  at moderate detection; with very few visits or very low detection the model
  struggles to tell an unoccupied site from an occupied-but-undetected one, and
  the field inherits that ambiguity.

The decision between a continuous field and an areal field follows the data
geometry:

| Your sites | Field to use | Term | Engine |
|---|---|---|---|
| Raw coordinates on a continuous landscape | Matern SPDE | `spde(lon, lat)` | `method = "laplace"` |
| Regions joined by borders, graph given | ICAR / BYM2 | `icar(graph)` / `bym2(graph)` | `method = "nested_laplace"` |
| Regions, autocorrelation strength of interest | proper CAR | `car_proper(graph)` | `method = "nested_laplace"` |
| No spatial signal in residuals | none | covariates only | `method = "laplace"` |

On priors, the PC prior on `(range, sigma)` is weakly informative and the
default usually suffices, but the range is the sensitive one. Refitting with
the median range halved and doubled is the cheap sensitivity check: if the
recovered field and the occupancy slope move little, the prior is not driving
the answer; if the field SD swings with the prior range, the data are not
informing the range and you should report that the field is prior-dominated.
The fixed-effect prior is separate and threads through `priors = occu_priors()`
if you want to tighten the detection intercept or break the psi-p
identifiability ridge at small `J`; `priors = FALSE` recovers the unpenalised
MAP.

Three cases call for a different tool. When a non-spatial fit leaves no spatial
autocorrelation in its residuals (Moran's I p-value above 0.05), the field adds
latent variables for nothing and the WAIC will not improve; do not fit it. When
the sites are genuinely areal, regions with shared borders rather than points,
the adjacency graph carries information the SPDE throws away by treating
centroids as bare coordinates, so use `icar()` or `bym2()` on the
`nested_laplace` path instead. And when you need a fully marginalised credible
band on the field at new coordinates rather than a point surface, the
continuous Laplace path stores the field mode only; the areal nested path is
the one that returns calibrated per-cell intervals today, so route through it if
the field's own uncertainty is the deliverable.

Two limitations of the current SPDE-Laplace path are worth stating plainly. The
prior at the prior modes of `(range, sigma)` is what the engine uses; nested
integration over those hyperparameters, which would propagate their uncertainty
into the field band, lands in a later release. And detection-side SPDE is not
plumbed for the Laplace engine, so keep the `spde()` term in the occupancy
formula, not in `detection`.

## References

* Lindgren, F., Rue, H., & Lindstrom, J. (2011). An explicit link between
  Gaussian fields and Gaussian Markov random fields: the stochastic
  partial differential equation approach. *Journal of the Royal
  Statistical Society B*, 73, 423-498.
* Simpson, D., Rue, H., Riebler, A., Martins, T. G., & Sorbye, S. H. (2017).
  Penalising model component complexity: a principled, practical approach to
  constructing priors. *Statistical Science*, 32, 1-28.
* `?tobs` for the unified fitter and the in-formula `spde()` term.
* `vignette("spatial-occupancy", package = "tulpaObs")` for the areal
  ICAR / BYM2 / CAR fields on a neighbour graph.
* `vignette("occupancy-vs-inla", package = "tulpaObs")` for the
  non-spatial occupancy story.
* `vignette("diagnostics", package = "tulpaObs")` for posterior-predictive
  checks and spatial residual diagnostics.
