---
title: "Occupancy: TulpaObs vs INLA aggregated-binomial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Occupancy: TulpaObs vs INLA aggregated-binomial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
have_inla <- requireNamespace("INLA", quietly = TRUE)
set.seed(1)
```

A proper single-season occupancy model has a latent occupancy state
$z_i \sim \text{Bernoulli}(\psi_i)$ and a conditional detection process
$y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \, p_{ij})$. The site-level
marginal likelihood is

$$
L_i
\;=\;
\psi_i \prod_{j=1}^{J} p_{ij}^{y_{ij}} (1 - p_{ij})^{1 - y_{ij}}
\;+\;
\mathbf{1}\!\left\{\sum_{j} y_{ij} = 0\right\}
(1 - \psi_i).
$$

That mixture cannot be written as a row-wise GLM likelihood, so
`INLA::inla()` cannot fit it directly without `inla.rgeneric`. Michael
Glaser's `example/info.md` is the same observation: in INLA, occupancy at
the site level and detection at the visit level don't share a row
structure, and the marginalization over $z_i$ has to be programmed
explicitly.

The practical workaround in `MOT_abund_data.Rmd` §"VISIT-LEVEL BERNOULLI
MODEL" is to fit a single binomial on the row-stacked detection
histories. This vignette shows what's lost compared to the true
marginalized fit.

## Simulate detection histories with known $\psi$ and $p$

```{r simulate}
library(TulpaObs)

sim <- simulate_occu(
  N          = 600,
  J          = 6,
  n_occ_covs = 1,
  n_det_covs = 1,
  beta_occ   = c(0.5, 1.2),    # intercept + slope on occ_cov1
  beta_det   = c(0.0, 0.8),    # intercept + slope on det_cov1
  seed       = 42
)

dat   <- sim$data
y     <- sim$y                 # N x J detection-history matrix
truth <- sim$truth

c(naive_occ      = mean(rowSums(y) > 0),
  true_psi_mean  = mean(truth$psi),
  true_p_mean    = mean(truth$p))
```

The naive proportion of sites with any detection is biased low by
$(1 - p)^J$ at each occupied site — exactly the bias an occupancy model
is supposed to correct.

## TulpaObs: true occupancy fit via `occu()`

```{r fit-tulpa-laplace}
mod_tulpa <- occu(
  occ_formula = ~ occ_cov1,
  det_formula = ~ det_cov1,
  data        = dat,
  y           = y
)
fit_tulpa_lap <- occu_fit(mod_tulpa, method = "laplace", verbose = FALSE)
```

For the parameter-recovery story below the EM-Laplace point estimates
underestimate detection in this small-J setting (the joint MAP picks one
mode of the $\psi$-$p$ ambiguity rather than integrating over both).
NUTS samples the full posterior and is unbiased:

```{r fit-tulpa-nuts}
fit_tulpa_nuts <- occu_fit(mod_tulpa, method = "nuts",
                           iter = 1500, warmup = 750,
                           seed = 42, verbose = FALSE)
fit_tulpa_nuts
```

## INLA workaround: row-stack the detection histories, fit a single binomial

This is the structural shape Michael's `mod.visit` uses. Each site
contributes `J` Bernoulli rows; site-level occupancy covariates broadcast
across the J visits.

```{r reshape, eval = have_inla}
N <- nrow(y)
J <- ncol(y)

long <- data.frame(
  y        = as.integer(t(y)),
  site     = rep(seq_len(N), each = J),
  visit    = rep(seq_len(J), times = N),
  occ_cov1 = rep(dat$occ_cov1, each = J),
  det_cov1 = rep(dat$det_cov1, each = J)
)

head(long)
```

```{r fit-inla, eval = have_inla}
library(INLA)

m_inla <- INLA::inla(
  y ~ occ_cov1 + det_cov1,
  family = "binomial",
  data   = long,
  control.compute = list(waic = TRUE)
)

inla_fixed <- m_inla$summary.fixed[, c("mean", "0.025quant", "0.975quant")]
inla_fixed
```

## What the workaround estimates

Site $i$ contributes $J$ Bernoulli rows with success probability
$\psi_i\, p_{ij}$. Because the response is binary and the model is
row-wise, the linear predictor $\eta_{ij}$ fits

$$
P(y_{ij} = 1)
\;=\;
\psi_i \, p_{ij}.
$$

There is no factor that separates the $\psi_i$ contribution from the
$p_{ij}$ contribution: the workaround can identify the product, never the
two factors. The `occ_cov1` coefficient is pulled toward zero by the
detection probability, and the `det_cov1` coefficient soaks up whatever
signal is left.

```{r compare, eval = have_inla}
compare <- data.frame(
  parameter        = c("(occ Intercept)", "occ_cov1",
                       "(det Intercept)", "det_cov1"),
  truth            = c(truth$beta_occ, truth$beta_det),
  TulpaObs_NUTS    = c(
    fit_tulpa_nuts$means["psi_(Intercept)"],
    fit_tulpa_nuts$means["psi_occ_cov1"],
    fit_tulpa_nuts$means["p_(Intercept)"],
    fit_tulpa_nuts$means["p_det_cov1"]
  ),
  TulpaObs_Laplace = c(
    fit_tulpa_lap$means["psi_(Intercept)"],
    fit_tulpa_lap$means["psi_occ_cov1"],
    fit_tulpa_lap$means["p_(Intercept)"],
    fit_tulpa_lap$means["p_det_cov1"]
  ),
  INLA_workaround  = c(
    inla_fixed["(Intercept)", "mean"],
    inla_fixed["occ_cov1",    "mean"],
    NA_real_,    # no separate detection intercept exists
    inla_fixed["det_cov1",    "mean"]
  ),
  row.names = NULL
)
compare
```

NUTS lands on the truth on both processes. Laplace gives biased point
estimates with this $J = 6$ setting — the joint MAP underestimates
detection probability when sites are weakly identified. INLA's
workaround returns *one* intercept that mixes occupancy and detection,
and the slopes on occupancy-side and detection-side covariates pull
toward each other.

## Predictive divergence at the site level

Despite the Laplace point-estimate bias, the predicted *per-site*
occupancy probability still correlates strongly with the truth — the
bias mostly cancels in the linear predictor. The INLA workaround cannot
recover site-level $\psi$ even in principle, because what it estimates is
not $\psi$.

```{r predict, eval = have_inla}
# TulpaObs returns per-site psi_hat. fitted() also returns $p (per-site
# detection probability) and $z (posterior occupancy probability given
# the detection history); $psi is the right comparator for true psi.
psi_hat_nuts <- fitted(fit_tulpa_nuts)$psi
psi_hat_lap  <- fitted(fit_tulpa_lap)$psi

# INLA workaround: predicted "any-visit detection" probability at the
# site, back-projected to a single row per site. This is the product
# psi * p, not psi.
X_site       <- model.matrix(~ occ_cov1 + det_cov1, data = dat)
eta_site     <- as.numeric(X_site %*% inla_fixed[, "mean"])
psi_hat_inla <- plogis(eta_site)

cor_with_truth <- c(
  TulpaObs_NUTS    = cor(psi_hat_nuts, truth$psi),
  TulpaObs_Laplace = cor(psi_hat_lap,  truth$psi),
  INLA_workaround  = cor(psi_hat_inla, truth$psi),
  naive_any        = cor(as.numeric(rowSums(y) > 0), truth$psi)
)
round(cor_with_truth, 3)
```

For population-trend questions ("did $\psi$ change?") the workaround
answers a different question ("did $\psi \cdot p$ change?"), and its
trend can be driven by detection improvements that have nothing to do
with the species' actual occupancy.

```{r predict-plot, eval = have_inla, fig.height = 4}
op <- par(mfrow = c(1, 2))
plot(truth$psi, psi_hat_nuts, pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "true psi", ylab = "TulpaObs psi_hat (NUTS)",
     main = "occu(): recovers psi")
abline(0, 1, col = 2, lwd = 2)
plot(truth$psi, psi_hat_inla, pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "true psi", ylab = "INLA aggregated p(detection)",
     main = "binomial workaround: confounds psi and p")
abline(0, 1, col = 2, lwd = 2)
par(op)
```

## What about the "detectability proxies" trick?

Michael's `mod.visit` adds plot-level fixed effects (`area.sc`,
`richness.sc`, `sample.density.sc`, `sample.extent.sc`) to the same
linear predictor that drives the occupancy-side covariates. Those proxies
*can* improve fit if they correlate with $p$, but their coefficients are
still confounded with the occupancy coefficients on the same linear
predictor — there is no hierarchical structure in the model that
attributes one slope to $\psi$ and another to $p$. The workaround does
not become an occupancy model by adding more covariates.

The only way to recover $\psi$ and $p$ separately from a row-wise INLA
likelihood is `inla.rgeneric`: a custom likelihood that implements the
mixture above. That is the same level of work as building TulpaObs's
EM/NUTS engine; if you're going to write the marginalized likelihood,
you may as well use the package that already has it.

## References

* `example/MOT_abund_data.Rmd` §"VISIT-LEVEL BERNOULLI MODEL" — Michael
  Glaser's binomial workaround, with detectability proxies.
* `example/info.md` — Michael's note on why a proper single-species
  occupancy model is hard in INLA (the marginalization problem).
* `?occu` and `?occu_fit` — the TulpaObs entry points used here.
* `?simulate_occu` — the detection-history simulator.
