---
title: "Getting started with tulpaObs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tulpaObs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## What tulpaObs fits

A survey that visits a site a handful of times and records whether a
species was seen carries two things at once: whether the species is there
at all, and whether you happened to detect it on a given visit. tulpaObs
fits the hierarchical models that separate those two. You pass the latent
process formula, a `detection` formula, the detection-history matrix `y`,
and a `family`; `tobs()` builds the design, plugs the family likelihood
into the [`tulpa`](https://github.com/gcol33/tulpa) engine, and returns a
fitted object you can summarise, predict from, and compare.

The package covers occupancy (single-season, dynamic, community,
integrated, joint species distribution), N-mixture abundance (single
species and community, Poisson or negative binomial), and a vegetation
cover hurdle. Spatial fields, temporal structure, and random effects
attach to any of them through terms inside the formula. The default
engine is a fast Laplace approximation; the same model fits under NUTS
when you want the full posterior.

This vignette stays on the spine: simulate a dataset with known truth,
fit it, read the estimates, predict, compare. The dedicated vignettes go
deeper on each family.

## A first occupancy model

`simulate_occu()` draws a single-season dataset with parameters you set,
so the fit has a truth to recover. Occupancy depends on one site
covariate (`occ_cov1`) with intercept 0.3 and slope 1.0; detection
depends on `det_cov1` with intercept 0.7 (so detection probability is
about 0.67) and slope 0.6. Six visits per site keep the two processes
identifiable.

```{r sim}
sim <- simulate_occu(
  N = 300, J = 6,
  n_occ_covs = 1, n_det_covs = 1,
  beta_occ = c(0.3, 1.0),    # occupancy intercept + slope on occ_cov1
  beta_det = c(0.7, 0.6),    # detection intercept + slope on det_cov1
  seed = 1
)
dat <- sim$data              # site covariates, one row per site
y   <- sim$y                 # 300 x 6 detection-history matrix
y[1:3, ]
```

The raw proportion of sites with any detection understates true
occupancy, because an occupied site can go undetected on every visit.

```{r naive}
c(naive_occupied = mean(rowSums(y) > 0),
  true_occupied  = mean(sim$truth$z))
```

Fit the proper model with `tobs()`. The latent occupancy formula is the
first argument, detection is its own formula, and `occu()` selects the
single-season family. `method = "laplace"` is the default: it finds the
posterior mode and the curvature there and returns in a fraction of a
second.

```{r fit}
fit <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)   # the EM prints its iterations otherwise
)
```

`coef()` returns the estimates as a list with one element per process.
The occupancy and detection coefficients both land near the simulated
truth.

```{r coef}
coef(fit)
```

`summary()` adds the posterior standard deviation and the 2.5%/50%/97.5%
credible bounds for every coefficient. The `psi_` rows are the occupancy
linear predictor, the `p_` rows the detection one.

```{r summary}
summary(fit)
```

`confint()` returns the same bounds as a matrix, reading the fixed-effect
block of the Hessian the Laplace fit already computed, so there is no
extra sampling step.

```{r confint}
confint(fit)
```

## Occupancy, detection, and the latent state

`fitted()` returns three per-site quantities: `psi` (the modelled
occupancy probability), `p` (detection probability), and `z` (the
posterior probability the site is occupied given its detection history).
A site with at least one detection has `z = 1`; a site with all-zero
visits gets the partial credit the model assigns from `psi` and `p`.

```{r fitted}
fv <- fitted(fit)
str(lapply(fv, head, 4))
```

Plotting the fitted `psi` against the true occupancy probability shows
the model recovers the site-level signal, not just the population mean.

```{r fitted-plot, fig.alt = "Fitted occupancy probability against the true simulated probability, scattered around the identity line"}
plot(sim$truth$psi, fv$psi, pch = 16, col = rgb(0, 0, 0, 0.35),
     xlab = "true psi", ylab = "fitted psi")
abline(0, 1, col = "steelblue", lwd = 2)
```

## Laplace and NUTS

The Laplace fit conditions on a Gaussian posterior shape. That is exact
for the fixed-effect curvature here and immediate to compute. When you
want draws from the full posterior, `method = "nuts"` runs the
gradient sampler; it reports `Rhat` and effective sample size alongside
the estimates.

```{r nuts}
fit_nuts <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "nuts",
  control   = list(n.iter = 800, n.warmup = 400, seed = 1, verbose = FALSE)
)
summary(fit_nuts)
```

The two backends agree on the coefficients to within their uncertainty.
`Rhat` near 1 and effective sample sizes in the hundreds say the chains
mixed. With six visits and detection near 0.67 the Laplace mode and the
posterior mean coincide; at fewer visits or lower detection the joint
mode can drift, and NUTS is the honest choice.

```{r compare-backends}
data.frame(
  term    = names(unlist(coef(fit))),
  laplace = round(unlist(coef(fit)), 3),
  nuts    = round(unlist(coef(fit_nuts)), 3),
  row.names = NULL
)
```

## Prediction

`tobs_marginal_effect()` varies one covariate across its range, holds the
others at their mean, and returns the predicted probability with a
credible band. Here it traces how occupancy responds to `occ_cov1`.

```{r marginal, fig.alt = "Predicted occupancy probability rising with occ_cov1, with a shaded 95% credible band"}
me <- tobs_marginal_effect(fit_nuts, "occ_cov1", process = "occupancy")
plot(me)
```

For prediction at an explicit design, pass a matrix to `X.0` with one
column per occupancy coefficient. The result carries the posterior mean,
SD, and credible bounds of `psi` at each new row.

```{r predict-design}
X0 <- cbind(`(Intercept)` = 1, occ_cov1 = c(-1, 0, 1))
predict(fit_nuts, X.0 = X0)
```

## Comparing models

`tobs_waic()` computes the widely applicable information criterion from
the pointwise log-likelihood. Lower is better. Compare the covariate
model against an intercept-only occupancy model to check the covariate
earns its place.

```{r waic}
fit0 <- tobs(~ 1, data = dat, family = occu(),
             detection = ~ det_cov1, y = y, method = "laplace",
             control = list(verbose = FALSE))

c(intercept_only = tobs_waic(fit0)$waic,
  with_covariate = tobs_waic(fit)$waic)
```

The covariate model has the lower WAIC, matching how the data were
generated. The drop reflects the occupancy slope the intercept-only model
cannot represent.

## Other families

The same call shape fits counts and cover; only `family` and the response
change. Each gets its own vignette, so these are tastes, not tours.

### N-mixture abundance

`abun()` fits Royle's N-mixture model: latent abundance per site, binomial
detection per visit. The abundance arm uses a log link, so its intercept
back-transforms with `exp()`.

```{r abun}
sa  <- simulate_abun(N = 200, J = 5, n_abund_covs = 1, n_det_covs = 1,
                     beta_lambda = c(log(3), 0.5), beta_p = c(0.4, 0.3),
                     seed = 2)
fa  <- tobs(~ abund_cov1, data = sa$data, family = abun(),
            detection = ~ det_cov1, y = sa$y, method = "laplace",
            control = list(verbose = FALSE))
coef(fa)
```

`abun(mixture = "negbin")` adds an overdispersion parameter for counts
with more spread than Poisson. See `vignette("abundance")`.

### Community occupancy

`ms_occu()` fits many species at once, with community-level hyperpriors
tying the species coefficients together. `y` is a sites-by-visits-by-species
array and `species` names the third dimension.

```{r ms}
sm <- simulate_ms_occu(N = 120, J = 5, n_species = 8, seed = 3)
fm <- tobs(~ x, data = sm$data, family = ms_occu(), detection = ~ 1,
           y = sm$y, species = paste0("sp", 1:8), method = "laplace",
           control = list(verbose = FALSE))
coef(fm)
```

`tobs_richness()` turns the per-species occupancy posteriors into a
site-level species-richness estimate with uncertainty. See
`vignette("community-models")`.

### Vegetation cover

`cover()` fits a hurdle for continuous cover: a presence process plus a
positive-cover process (beta or lognormal) for where the species occurs.
See `vignette("cover-hurdle")`.

```{r cover}
sc <- simulate_cover(N = 200, beta_occ = c(-0.3, 0.8),
                     beta_pos = c(-1, 0.3), sigma_pos = 0.4, seed = 4)
str(sc$data, max.level = 1)
```

## Extracting results

The fitted object answers the standard accessors. They behave the same on
a Laplace fit and a NUTS fit; the Laplace tier reports the Gaussian
approximation, the sampler tier reports empirical posterior summaries.

| Accessor | Returns |
|---|---|
| `coef()` | per-process fixed-effect estimates |
| `summary()` | estimate, SD, credible bounds (plus Rhat/ESS under NUTS) |
| `confint()` | credible intervals (matrix) |
| `vcov()` | fixed-effect covariance |
| `ranef()` | group-level random effects (when present) |
| `fitted()` | per-site `psi`, `p`, `z` |
| `predict()` | occupancy at a new design or varied term |
| `tobs_waic()` | WAIC, elpd, effective parameters |
| `residuals()` | deviance / Pearson / response residuals |
| `simulate()` | replicate detection histories from the fit |

## Practical guidance

- **Start with `method = "laplace"`.** It is deterministic and returns
  in well under a second on a few hundred sites. Use it for the first
  look and for WAIC comparison. Move to `method = "nuts"` when detection
  is low, visits are few, or you need the full posterior shape rather than
  a Gaussian approximation at the mode.
- **Visits drive identifiability.** Occupancy and detection trade off
  against each other. With two or three visits and detection below about
  0.3, the two processes are weakly separated and the Laplace mode can
  drift; NUTS or more visits resolve it. The
  [identifiability discussion in `vignette("occupancy")`](occupancy.html)
  works through where recovery fails.
- **Detection covariates broadcast across visits** when the data frame
  has one row per site, as in `simulate_occu()`. For covariates that
  change between visits (effort, observer, weather), pass a per-visit
  design; see `vignette("data-formatting")`.
- **Read uncertainty, not just point estimates.** `confint()` and the
  `summary()` quantile columns carry the credible bounds. A coefficient
  whose interval spans zero is not pulling its weight regardless of the
  point estimate.
- **WAIC compares, it does not validate.** A lower WAIC says one model
  predicts the in-sample data better; it does not say the model is
  calibrated. Pair it with a posterior-predictive check
  (`tobs_ppc()`, see `vignette("diagnostics")`).

## Where to go next

- Single-season occupancy end to end, with model comparison and
  identifiability: `vignette("occupancy")`.
- N-mixture abundance, negative binomial, and areal spatial counts:
  `vignette("abundance")`.
- Many species at once: `vignette("community-models")`.
- Colonisation and extinction across seasons:
  `vignette("dynamic-occupancy")`.
- Random effects with `lme4` bar syntax: `vignette("random-effects")`.
- Posterior-predictive checks, WAIC, and spatial residual diagnostics:
  `vignette("diagnostics")`.
```
