---
title: "Getting started with tulpaObs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tulpaObs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

```{r ggtheme, include = FALSE}
have_gg <- requireNamespace("ggplot2", quietly = TRUE)
if (have_gg) {
  library(ggplot2)
  transparent <- theme(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
}
```

## What tulpaObs fits

A survey that visits a site a handful of times and records whether a
species was seen carries two things at once: whether the species is there
at all, and whether you happened to detect it on a given visit. tulpaObs
fits the hierarchical models that separate those two. You pass the latent
process formula, a `detection` formula, the detection-history matrix `y`,
and a `family`; `tobs()` builds the design, plugs the family likelihood
into the [`tulpa`](https://github.com/gcol33/tulpa) engine, and returns a
fitted object you can summarise, predict from, and compare.

The package covers occupancy (single-season, dynamic, community,
integrated, joint species distribution), N-mixture abundance (single
species and community, Poisson or negative binomial), and a vegetation
cover hurdle. Spatial fields, temporal structure, and random effects
attach to any of them through terms inside the formula. The default
engine is a fast Laplace approximation; the same model fits under NUTS
when you want the full posterior.

The split that defines the package is between the state and the
observation. The state is the thing you care about: occupied or not,
how many individuals, what fraction of ground a plant covers. The
observation is what the field crew wrote down, and it is the state seen
through a detection process that misses some of what is there. A naive
analysis reads the observation as the state and undercounts everywhere
detection is below one. Every family in tulpaObs writes the two levels
separately, integrates the latent state out of the likelihood in closed
form where it can, and reports both the state coefficients and the
detection coefficients. You read occupancy against a background that
already accounts for the visits where the species was present and missed.

Where this differs from the tools next to it is the engine underneath.
tulpaObs is a thin observation layer; the inference lives in `tulpa`,
which composes a fast deterministic approximation with an exact sampler.
The Laplace path finds the posterior mode and the curvature there and
returns in a fraction of a second, which is the right tool for the first
look and for comparing candidate models. The NUTS path samples the full
posterior and reports `Rhat` and effective sample size, which is the
right tool when detection is low, visits are few, or the posterior is
skewed enough that a Gaussian summary at the mode would mislead. The same
`tobs()` call switches between them with one argument, and the fitted
object answers the same accessors either way. Against `unmarked`, the
draw is a Bayesian posterior with credible intervals rather than a
maximum-likelihood point fit; against a hand-written Stan model, the draw
is the speed of the deterministic path and a formula interface that reads
like `lme4`. You write the model once and choose the inference per run.

This vignette stays on the spine: simulate a dataset with known truth,
fit it, read the estimates, predict, compare. It then takes one short
taste of each model family, fits a random effect, and walks the standard
S3 accessors. The dedicated vignettes go deeper on each family, and the
taste tests link out to them rather than repeating their depth here.

## A first occupancy model

`simulate_occu()` draws a single-season dataset with parameters you set,
so the fit has a truth to recover. Occupancy depends on one site
covariate (`occ_cov1`) with intercept 0.3 and slope 1.0; detection
depends on `det_cov1` with intercept 0.7 (so detection probability is
about 0.67) and slope 0.6. Six visits per site keep the two processes
identifiable.

```{r sim}
sim <- simulate_occu(
  N = 300, J = 6,
  n_occ_covs = 1, n_det_covs = 1,
  beta_occ = c(0.3, 1.0),    # occupancy intercept + slope on occ_cov1
  beta_det = c(0.7, 0.6),    # detection intercept + slope on det_cov1
  seed = 1
)
dat <- sim$data              # site covariates, one row per site
y   <- sim$y                 # 300 x 6 detection-history matrix
y[1:3, ]
```

The simulated object carries three pieces. `sim$data` is the site frame,
one row per site, holding the covariates the occupancy and detection
formulas will read. `sim$y` is the detection history, a matrix with one
row per site and one column per visit, each entry a 0 or 1. `sim$truth`
holds the parameters and the latent occupancy state, which the fit never
sees and which you keep only to check recovery. Real data arrive without
that last piece, so the recovery checks here are a luxury of working from
a simulation; on a field dataset you fit, predict, and compare without a
truth to measure against.

The raw proportion of sites with any detection understates true
occupancy, because an occupied site can go undetected on every visit.

```{r naive}
c(naive_occupied = mean(rowSums(y) > 0),
  true_occupied  = mean(sim$truth$z))
```

The gap between those two numbers is the whole reason an occupancy model
exists. The naive rate counts a site as occupied only if the species was
seen at least once, and with detection near two thirds across six visits
a present species is missed on every visit about one time in a thousand,
so the gap is small here. Cut the visits to two or the detection to 0.3
and the gap widens fast, because a present species then slips through
undetected often enough to drag the observed rate well below the truth.
The model recovers the difference by reading how often a species seen on
one visit was missed on the next.

The mechanism is worth one sentence before the fit. Across six visits an
occupied site at detection $p$ produces at least one sighting with
probability $1 - (1 - p)^6$, so the fraction of occupied sites that never
register a single detection is $(1 - p)^6$. Invert that and the observed
detection rate tells you both how many occupied sites you missed and what
$p$ must have been to miss them, provided you have enough visits for the
two to separate. Six visits at detection 0.67 leave almost no occupied
site silent, which is why the naive and true rates nearly agree here and
why this design recovers cleanly. The same model run on a two-visit
survey would carry a wide interval on $p$ and a noticeably higher
estimated occupancy than the naive count, because more occupied sites
would have gone unseen.

Fit the proper model with `tobs()`. The latent occupancy formula is the
first argument, detection is its own formula, and `occu()` selects the
single-season family. `method = "laplace"` is the default: it finds the
posterior mode and the curvature there and returns in a fraction of a
second.

```{r fit}
fit <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "laplace",
  control   = list(verbose = FALSE)   # the EM prints its iterations otherwise
)
```

Five arguments carry the whole model, and they map one to one onto the
two-level structure. The first formula is the occupancy linear predictor
on the logit of $\psi$. The `detection` formula is the per-visit
detection predictor on the logit of $p$. The `family` names the
likelihood, here single-season occupancy. The `y` matrix is the
observation. The `method` picks the inference. Everything that is not a
statistical choice (verbosity, iteration caps, tolerances, thread counts)
lives inside `control`, so the call signature stays statistical and the
knobs stay in one place. Swapping `occu()` for `abun()` and handing it a
count matrix turns the same call into an N-mixture fit; the shape does
not change.

`coef()` returns the estimates as a list with one element per process.
The occupancy and detection coefficients both land near the simulated
truth.

```{r coef}
coef(fit)
```

The list has two elements because the model has two processes. `coef(fit)$psi`
holds the occupancy intercept and the slope on `occ_cov1`, both on the
logit scale, and `coef(fit)$p` holds the detection coefficients. Reading
the occupancy intercept of about 0.3 through `plogis()` gives a baseline
occupancy near 0.57 at the covariate mean; the slope of about 1.0 says
occupancy rises steeply with `occ_cov1`. Keep the link in mind whenever
you read a coefficient: occupancy and detection are both logit-scale, so
a coefficient is a change in log-odds, not in probability, and the
mapping to probability is steepest near the middle of the range and flat
at the ends.

The per-process list is the shape every family in the package returns, so
the same line of code reads the right block whatever the model. A
single-season occupancy fit has `psi` and `p`; an N-mixture fit has
`lambda` and `p`; an integrated fit has one occupancy block and a
detection block per source. Reaching into the named element rather than a
flat coefficient vector keeps the two arms distinct, which matters because
they live on different links and mean different things. A common slip is
to back-transform the whole vector with one function; the occupancy
intercept needs `plogis()` to read as a probability, while an abundance
intercept needs `exp()` to read as a count, and mixing them produces
numbers that look plausible and are wrong.

`summary()` adds the posterior standard deviation and the 2.5%/50%/97.5%
credible bounds for every coefficient. The `psi_` rows are the occupancy
linear predictor, the `p_` rows the detection one.

```{r summary}
summary(fit)
```

Read the summary as four numbers per coefficient: the posterior mean, its
standard deviation, and the lower and upper credible bounds. The bounds
are the honest report. A slope whose interval clears zero is doing work;
one whose interval straddles zero is consistent with no effect regardless
of how large the point estimate looks. Here the occupancy slope on
`occ_cov1` sits well clear of zero, which is the signal the simulation
built in. The standard deviation and the bounds come straight from the
curvature of the log-posterior at the mode, so the Laplace fit reports
them without a sampling step.

`confint()` returns the same bounds as a matrix, reading the fixed-effect
block of the Hessian the Laplace fit already computed, so there is no
extra sampling step.

```{r confint}
confint(fit)
```

## Occupancy, detection, and the latent state

`fitted()` returns three per-site quantities: `psi` (the modelled
occupancy probability), `p` (detection probability), and `z` (the
posterior probability the site is occupied given its detection history).
A site with at least one detection has `z = 1`; a site with all-zero
visits gets the partial credit the model assigns from `psi` and `p`.

```{r fitted}
fv <- fitted(fit)
str(lapply(fv, head, 4))
```

The three quantities answer three different questions about a site. `psi`
is what the model expects before seeing the visits: the occupancy
probability the covariates imply. `p` is the per-visit detection
probability the covariates imply. `z` is the posterior after the visits:
given what was actually recorded, how sure are we the site is occupied. A
site detected once has `z = 1` because detection rules out absence. A site
with six blank visits has a `z` below one, pulled down from `psi` by the
chance that six visits at this detection would all miss a present species.
That posterior occupancy is the quantity richness sums over in a community
model, and it is the right thing to map when you want occupancy rather
than detections.

The three numbers also pull apart the two ways a model can be wrong about
a site. If `psi` is high but the visits are all blank and `z` stays high,
the model is saying the covariates expect occupancy and the few visits
were not enough to overturn that expectation. If `psi` is low and `z` is
also low, the model and the data agree the site is probably empty. The
disagreements are the interesting cases: a low `psi` with a detection
pulling `z` up means the species turned up somewhere the covariates did
not predict, which is the kind of residual a map of `z` minus `psi` makes
visible. Reading the three together, rather than any one alone, is how you
tell a confident prediction from a guess the covariates had to make.

Plotting the fitted `psi` against the true occupancy probability shows
the model recovers the site-level signal, not just the population mean.

```{r fitted-plot, fig.alt = "Fitted occupancy probability against the true simulated probability, scattered around the identity line"}
plot(sim$truth$psi, fv$psi, pch = 16, col = rgb(0, 0, 0, 0.35),
     xlab = "true psi", ylab = "fitted psi")
abline(0, 1, col = "steelblue", lwd = 2)
```

The points scatter along the identity line rather than collapsing onto a
single horizontal band, which is the test that matters. A model that
recovered only the population mean would predict the same `psi` for every
site and the cloud would be flat. This one tracks the covariate-driven
spread, so a site with high `occ_cov1` is predicted high and a low one
low. The scatter around the line is the irreducible uncertainty six binary
visits leave behind, not bias.

## Laplace and NUTS

The Laplace fit conditions on a Gaussian posterior shape. That is exact
for the fixed-effect curvature here and immediate to compute. When you
want draws from the full posterior, `method = "nuts"` runs the
gradient sampler; it reports `Rhat` and effective sample size alongside
the estimates.

```{r nuts}
fit_nuts <- tobs(
  ~ occ_cov1,
  data      = dat,
  family    = occu(),
  detection = ~ det_cov1,
  y         = y,
  method    = "nuts",
  control   = list(n.iter = 800, n.warmup = 400, seed = 1, verbose = FALSE)
)
summary(fit_nuts)
```

The two backends agree on the coefficients to within their uncertainty.
`Rhat` near 1 and effective sample sizes in the hundreds say the chains
mixed. With six visits and detection near 0.67 the Laplace mode and the
posterior mean coincide; at fewer visits or lower detection the joint
mode can drift, and NUTS is the honest choice.

The two diagnostics report different things, and a fit is trustworthy only
when both pass. `Rhat` compares the variance within each chain to the
variance across chains; a value near 1 means the chains explored the same
region and agree on it, while a value above about 1.01 says they have not
converged on a common answer and the summaries are premature. Effective
sample size discounts the autocorrelation in the chain: a thousand draws
that wander slowly might carry the information of only a few dozen
independent ones, and a small effective sample size means the reported
standard error is itself noisy. The Laplace fit reports neither, because
it has no chains; its summaries come from the curvature at the mode, which
is exact for the fixed-effect block and immediate. The decision between
the two is not Laplace-versus-NUTS in general but per run: fast and
deterministic when the posterior is well behaved, sampled when it is
skewed or weakly identified.

```{r compare-backends}
data.frame(
  term    = names(unlist(coef(fit))),
  laplace = round(unlist(coef(fit)), 3),
  nuts    = round(unlist(coef(fit_nuts)), 3),
  row.names = NULL
)
```

## Prediction

`tobs_marginal_effect()` varies one covariate across its range, holds the
others at their mean, and returns the predicted probability with a
credible band. Here it traces how occupancy responds to `occ_cov1`.

```{r marginal, fig.alt = "Predicted occupancy probability rising with occ_cov1, with a shaded 95% credible band"}
me <- tobs_marginal_effect(fit_nuts, "occ_cov1", process = "occupancy")
plot(me)
```

The curve rises across the covariate range because the occupancy slope is
positive, and the band widens at the ends where the data thin out. The
`process` argument switches the same call to the detection arm; pass
`process = "detection"` and the same machinery traces $p$ against one of
its covariates instead. Either way the result is a tidy table you can plot
with the built-in method or feed to your own ggplot.

For prediction at an explicit design, pass a matrix to `X.0` with one
column per occupancy coefficient. The result carries the posterior mean,
SD, and credible bounds of `psi` at each new row.

```{r predict-design}
X0 <- cbind(`(Intercept)` = 1, occ_cov1 = c(-1, 0, 1))
predict(fit_nuts, X.0 = X0)
```

The three rows ask for occupancy at `occ_cov1` of -1, 0, and 1 with the
intercept held in. The returned mean climbs across the rows and each row
carries its own credible bounds, which is the prediction you would map
across a landscape: build `X.0` from the covariate raster, predict, and
plot the posterior mean with its band. A ggplot of the marginal effect
draws the same picture as a smooth curve, with the band carried straight
from the prediction table.

```{r predict-plot, eval = have_gg, fig.alt = "Predicted occupancy probability against occ_cov1 as a line with a shaded 95 percent credible band"}
ggplot(me, aes(x = x, y = estimate)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(colour = "steelblue", linewidth = 1) +
  labs(x = "occ_cov1", y = "occupancy probability") +
  transparent
```

## Random effects

Surveys carry structure that is not a covariate of interest: sites fall
into clusters, observers differ in effort, regions sit under their own
unmeasured baseline. A grouping factor with many exchangeable levels
belongs in a random effect, which assumes the level effects are draws
from a common distribution $N(0, \sigma^2)$ and estimates the spread
rather than one free intercept per level. A level with little data is
pulled toward the population mean; a level with plenty keeps its own
signal. tulpaObs writes random effects the `lme4` way, with a bar inside
the formula, and `(1 | g)` enters whichever linear predictor it sits in.

Simulate occupancy that varies by a site cluster: thirty clusters of
twenty-five sites, each with its own intercept drawn from $N(0, 0.9^2)$,
plus one site covariate.

```{r re-sim}
set.seed(101)
ng <- 30L; per <- 25L; N <- ng * per
g  <- rep(seq_len(ng), each = per)
x  <- rnorm(N)
b_true <- rnorm(ng, 0, 0.9)                       # per-cluster intercepts
z  <- rbinom(N, 1, plogis(0.3 - 0.6 * x + b_true[g]))
y_re <- matrix(0L, N, 6L)
for (i in seq_len(N)) y_re[i, ] <- rbinom(6L, 1, z[i] * 0.45)
d_re <- data.frame(g = factor(g), x = x)
```

Put `(1 | g)` in the occupancy formula. The default Laplace path fits the
variance component through a variance-component EM with an adaptive
Gauss-Hermite debias for small clusters; `coef()` returns the fixed
effects and `ranef()` the per-group BLUPs.

```{r re-fit}
fit_re <- tobs(~ x + (1 | g), data = d_re, y = y_re, detection = ~ 1,
               family = occu(), method = "laplace",
               control = list(verbose = FALSE))
coef(fit_re)$psi
```

The variance component lands in `fit$means`, named for the group it
belongs to, and the per-cluster intercepts come back from `ranef()` as a
table with one row per group. Plotting the recovered BLUPs against the
simulated cluster effects shows the random effect tracks the cluster-level
signal, with the shrinkage that pulls thin clusters toward zero.

```{r re-blup, eval = have_gg, fig.alt = "Recovered per-cluster random intercepts against the simulated truth, scattered along the identity line"}
re <- ranef(fit_re)
bl <- data.frame(truth = b_true, blup = re$estimate)
ggplot(bl, aes(x = truth, y = blup)) +
  geom_abline(slope = 1, intercept = 0, colour = "steelblue", linewidth = 1) +
  geom_point(alpha = 0.6) +
  labs(x = "true cluster effect", y = "estimated BLUP") +
  transparent
```

The estimated intercepts line up along the identity with a slight pull
toward zero, the shrinkage the random effect applies where a cluster's own
data are thin. Correlated random slopes `(1 + x | g)`, the uncorrelated
form `(x || g)`, crossed `(1 | g:h)`, and nested `(1 | g/h)` grouping all
fit the same way, and a bar in the `detection` formula moves the random
effect onto the detection arm instead. The dedicated treatment, including
the small-cluster bias and when to reach for NUTS, is in
`vignette("random-effects")`.

## Comparing models

`tobs_waic()` computes the widely applicable information criterion from
the pointwise log-likelihood. Lower is better. Compare the covariate
model against an intercept-only occupancy model to check the covariate
earns its place.

```{r waic}
fit0 <- tobs(~ 1, data = dat, family = occu(),
             detection = ~ det_cov1, y = y, method = "laplace",
             control = list(verbose = FALSE))

c(intercept_only = tobs_waic(fit0)$waic,
  with_covariate = tobs_waic(fit)$waic)
```

The covariate model has the lower WAIC, matching how the data were
generated. The drop reflects the occupancy slope the intercept-only model
cannot represent.

The same comparison weighs the random effect. Fit the clustered data with
and without the grouping term, then read the two WAICs side by side. The
random-intercept model should win when the clusters genuinely differ,
because the flat model has nowhere to put the between-cluster variance and
predicts the clustered sites worse.

```{r waic-re}
fit_re0 <- tobs(~ x, data = d_re, y = y_re, detection = ~ 1,
                family = occu(), method = "laplace",
                control = list(verbose = FALSE))

c(no_random_effect = tobs_waic(fit_re0)$waic,
  with_random_int  = tobs_waic(fit_re)$waic)
```

The clustered model carries the lower WAIC, the random intercept paying
for its one variance parameter by predicting the between-cluster spread
the flat model averages away. `tobs_waic()` returns more than the headline
number: `$waic` is the criterion, `$elpd` the expected log pointwise
predictive density it is built from, and `$p_waic` the effective number of
parameters, which is what docks a flexible model for its complexity. Read
WAIC as a comparison, not a verdict: a lower value says one model predicts
the in-sample data better, not that its intervals are calibrated. Pair it
with a posterior-predictive check (`tobs_ppc()`) before trusting a fit.
The full model-comparison and diagnostic workflow is in
`vignette("diagnostics")`.

## Taste tests

The same call shape fits counts, communities, cover, and the spatial,
dynamic, and integrated extensions; only `family`, the response, and a
formula term or two change. These are tastes, one short fit each, with a
plot and a pointer to the vignette that goes deep.

### Single-season occupancy

The model from the worked example is the base case, and it is worth seeing
its output as a print rather than a coefficient list. The `print` method
lays out both processes with their estimates and the family that produced
them, which is the first thing you read off a fresh fit.

```{r taste-occu}
fit
```

A bar plot of the two occupancy coefficients against zero reads the model
at a glance: the slope clears zero, the intercept sits where the baseline
occupancy puts it. The same picture for the detection arm comes from
`coef(fit)$p`.

```{r taste-occu-plot, eval = have_gg, fig.alt = "Occupancy coefficients as bars against a zero reference line"}
co <- data.frame(term = names(coef(fit)$psi), est = coef(fit)$psi)
ggplot(co, aes(x = term, y = est)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 0, colour = "grey40") +
  labs(x = NULL, y = "estimate (logit scale)") +
  transparent
```

Single-season occupancy is the spine the other families extend, and the
identifiability tension between occupancy and detection runs through all
of them. The tension has a simple source: a string of zeros at a site is
consistent with absence and with a present species missed on every visit,
and only the repeat visits and the covariates pull the two apart. When the
visits are few or detection is low, the likelihood flattens along the
ridge where occupancy and detection trade off, the mode becomes a poor
summary, and the Laplace standard errors run optimistic. The same fit
under NUTS traces the ridge and reports the honest, wider interval. Where
recovery fails, how visits and detection trade off, and how to read a
model that has too few visits to separate the two are worked through in
`vignette("occupancy")`.

### N-mixture abundance

`abun()` fits Royle's N-mixture model: a latent count per site drawn from
a Poisson, thinned by binomial detection on each visit. The abundance arm
uses a log link, so its intercept back-transforms with `exp()`. The fit
reads the spread of counts across visits to separate abundance from
detection.

```{r taste-abun}
sa <- simulate_abun(N = 200, J = 5, n_abund_covs = 1, n_det_covs = 1,
                    beta_lambda = c(log(3), 0.5), beta_p = c(0.4, 0.3),
                    seed = 2)
fa <- tobs(~ abund_cov1, data = sa$data, family = abun(),
           detection = ~ det_cov1, y = sa$y, method = "laplace",
           control = list(verbose = FALSE))
exp(coef(fa)$lambda[["(Intercept)"]])    # baseline expected abundance
```

The back-transformed intercept lands near the simulated baseline of three
animals per site. Predicting abundance across the covariate range and
plotting it shows the log-linear rise on the response scale.

```{r taste-abun-plot, eval = have_gg, fig.alt = "Predicted expected abundance rising with abund_cov1 across its range"}
pr <- predict(fa, terms = "abund_cov1")
ggplot(pr, aes(x = x, y = estimate)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(colour = "darkorange", linewidth = 1) +
  labs(x = "abund_cov1", y = "expected abundance") +
  transparent
```

The curve climbs because the abundance slope is positive, and the band
carries the count uncertainty through the exponential. What lets the model
see eight animals through counts that never reach eight is the spread of
the counts across visits. A site that averages two animals per visit could
hold twenty seen one time in ten or three seen two times in three, and the
two readings produce counts with very different variance: many animals at
low detection jump around relative to their mean, few at high detection
barely move. The N-mixture reads that spread and infers the abundance
behind it, which is why replicate visits are not optional here. The model
also leans on closure, that the same animals stay put across the survey
window, so a season long enough to let animals move violates it and reads
the movement as detection noise. `abun(mixture = "negbin")` adds an
overdispersion parameter for counts with more spread than a Poisson
allows, common when animals cluster, and the same `icar()` / `bym2()`
terms that smooth occupancy smooth a count surface. The closure
assumption, the count ceiling $K$, the negative-binomial state, and areal
spatial counts are in `vignette("abundance")`.

### Community models

`ms_occu()` fits many species at once, with community-level hyperpriors
tying the species coefficients together so a rare species borrows the
shape of the assemblage. `y` is a sites-by-visits-by-species array and
`species` names the third dimension. The community mean is the fixed
effect `coef()` returns.

```{r taste-ms}
sm <- simulate_ms_occu(N = 120, J = 5, n_species = 8, seed = 3)
fm <- tobs(~ x, data = sm$data, family = ms_occu(), detection = ~ 1,
           y = sm$y, species = paste0("sp", 1:8), method = "laplace",
           control = list(verbose = FALSE))
coef(fm)$psi                                # community occupancy mean
```

`tobs_richness()` turns the per-species occupancy posteriors into a
site-level species-richness estimate with uncertainty, summing occupancy
probabilities across species rather than counting detections. Plotting
richness against site, ordered, shows the spread across the survey with
its band.

```{r taste-ms-plot, eval = have_gg, fig.alt = "Estimated species richness per site, sorted, with a shaded 95 percent credible band"}
rich <- tobs_richness(fm)
rich <- rich[order(rich$mean), ]
rich$site <- seq_len(nrow(rich))
ggplot(rich, aes(x = site, y = mean)) +
  geom_ribbon(aes(ymin = q2.5, ymax = q97.5), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 1) +
  labs(x = "site (ordered by richness)", y = "expected richness") +
  transparent
```

The modelled richness exceeds the naive detected count because it adds the
species probably present but undetected, and the band carries every
species' occupancy uncertainty into the sum. The reason to fit the
assemblage jointly rather than species by species is the rare species. A
bird seen at three sites out of a hundred carries almost no information on
its own, and its single-species slope would run from deeply negative to
deeply positive. The community prior assumes the species coefficients are
draws from a shared distribution and pulls a thin species toward the
assemblage mean, hardest where its own data are weakest. The pull strength
is the fitted community covariance, learned from how the species scatter
rather than set by hand, so a homogeneous assemblage pools hard and a
heterogeneous one barely pools at all. The count analogue is `ms_abun()`,
the community N-mixture, which pools abundance and detection across the
assemblage the same way and exposes per-species deviations through
`ranef()`. When pooling helps, when it hurts, and how to read the
community covariance are in `vignette("community-models")`.

### Cover hurdle

`cover()` fits a hurdle for continuous vegetation cover: a binomial
occurrence process on whether the species is present at all, plus a
positive-cover process (lognormal or beta) for how much ground it covers
where it occurs. The two arms fit as independent Laplace approximations,
and `predict()` reads occupancy and expected cover at new covariate
values.

```{r taste-cover}
sc <- simulate_cover(N = 400, beta_occ = c(-0.4, 0.7),
                     beta_pos = c(-1.0, 0.3), sigma_pos = 0.4, seed = 7)
fc <- tobs(~ x, data = sc$data, family = cover(positive = "lognormal"),
           y = sc$y)
nd <- data.frame(x = seq(-2, 2, length.out = 50))
nd$expected <- predict(fc, nd, type = "expected")
head(nd)
```

The expected-cover curve combines both arms: the chance the species occurs
times the cover it carries where it does. Plotting it across the covariate
range traces the unconditional expected cover.

```{r taste-cover-plot, eval = have_gg, fig.alt = "Expected vegetation cover rising with the covariate across its range"}
nd$occupancy <- predict(fc, nd, type = "occupancy")
ggplot(nd, aes(x = x, y = expected)) +
  geom_line(colour = "forestgreen", linewidth = 1) +
  labs(x = "x", y = "expected cover") +
  transparent
```

The rise reflects both a higher occurrence probability and more cover
where present as the covariate climbs. Cover data have a shape that breaks
an ordinary regression: a pile of exact zeros where the species is absent,
and a continuous positive distribution where it occurs. Modelling the two
together with one Gaussian forces a single mean to explain both the zeros
and the spread, and it predicts impossible negative cover at the low end.
The hurdle splits them, a binomial arm on presence and a positive arm on
the cover where present, so the zeros inform occurrence and the positive
values inform the cover distribution without contaminating each other. The
expected cover the plot shows is the product of the two arms, which is why
it can rise both because the species occurs more often and because it
covers more ground where it does. This dispatcher predicts from fixed
effects only; the shared-field joint model, where occurrence, detection,
and cover couple through one spatial field, is the `occu_cover()` family.
The independent-arm hurdle is in `vignette("cover-hurdle")`, and the joint
engine in `vignette("cover-hurdle-motivate")`.

### Spatial occupancy

A neighbourhood graph lets adjacent regions share what the covariates
miss. An areal spatial field adds one smoothed random effect per cell,
ties neighbours together, and soaks up the part of occupancy that depends
on where a cell sits. The fields are formula terms: `icar()`, `bym2()`,
and `car_proper()`, each taking a dense 0/1 adjacency matrix. They route
through `method = "nested_laplace"`, which integrates the smoothing
strength over a grid.

```{r taste-spatial}
gx <- 5; n <- gx * gx
coord <- expand.grid(cx = seq_len(gx), cy = seq_len(gx))
adj <- matrix(0, n, n)
for (i in seq_len(gx)) for (j in seq_len(gx)) {
  a <- (j - 1) * gx + i
  if (i < gx) { b <- (j - 1) * gx + i + 1; adj[a, b] <- adj[b, a] <- 1 }
  if (j < gx) { b <- j * gx + i;           adj[a, b] <- adj[b, a] <- 1 }
}
field <- 1.4 * exp(-((coord$cx - 3)^2 + (coord$cy - 3)^2) / 4)
field <- field - mean(field)
z <- rbinom(n, 1, plogis(field))
ys <- matrix(0L, n, 8L)
for (i in seq_len(n)) if (z[i]) ys[i, ] <- rbinom(8L, 1, 0.5)
fit_sp <- tobs(~ bym2(graph = adj), data = data.frame(site = seq_len(n)),
               family = occu(), detection = ~ 1, y = ys,
               method = "nested_laplace",
               control = list(max.iter = 30L, tol = 1e-5, verbose = FALSE))
```

`predict(type = "state")` returns the smoothed per-cell occupancy
posterior. Laying it back on the grid coordinates and shading each cell
maps the field, which peaks where the simulated bump put it.

```{r taste-spatial-plot, eval = have_gg, fig.alt = "Grid map of fitted occupancy shaded by cell, peaking in the grid centre"}
sp <- predict(fit_sp, type = "state")
mp <- cbind(coord, psi = sp$psi)
ggplot(mp, aes(x = cx, y = cy, fill = psi)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = NULL, y = NULL, fill = "psi") +
  transparent
```

The high-occupancy ridge runs through the grid centre, smoothed from the
binary noise an isolated per-cell estimate would carry. The field earns
its keep two ways. It widens the slope standard errors to their honest
size, because adjacent cells share unmeasured drivers and a non-spatial
fit treats them as independent and runs its errors too narrow. And it
keeps a patterned covariate from absorbing the regional signal it competes
with, so the fitted slope reads against a background that already accounts
for location. The smoothing strength is itself a parameter, and the nested
path integrates it over a grid rather than fixing it, so the slope errors
inherit the field's uncertainty instead of pretending the smoothing was
known. ICAR is the plain smoother, BYM2 splits structured from
unstructured variation, and the proper CAR estimates the autocorrelation
strength. Areal fields, the floating intercept under ICAR, and
held-out-cell interpolation are in `vignette("spatial-occupancy")`;
continuous fields on point coordinates are in
`vignette("occupancy-spatial-spde")`.

### Dynamic occupancy

`dyn_occu()` fits the multi-season model of MacKenzie and colleagues: a
first-season occupancy, a colonisation rate for empty sites, an extinction
rate for occupied sites, and a detection process over the latent state.
The response carries a season dimension, and the model sums the latent
Markov chain out with the forward algorithm.

```{r taste-dyn}
sd_dyn <- simulate_dyn_occu(N = 200, J = 4, n_seasons = 6,
                            beta_occ = c(0), beta_det = c(0),
                            gamma = 0.3, epsilon = 0.2, seed = 2026)
fit_dyn <- tobs(~ 1, data = sd_dyn$data, family = dyn_occu(),
                detection = ~ 1, y = sd_dyn$y,
                col_formula = ~ 1, ext_formula = ~ 1,
                method = "laplace", control = list(verbose = FALSE))
fit_dyn$intercepts                          # psi1, gamma, epsilon, p
```

The four rates land near their simulated values. The two transition rates
imply an equilibrium occupancy $\gamma / (\gamma + \epsilon)$, and
iterating the transition forward from $\psi_1$ draws the modelled
trajectory across seasons against the realised occupancy fractions.

```{r taste-dyn-plot, eval = have_gg, fig.alt = "Modelled occupancy trajectory across six seasons against the realised fractions"}
g <- fit_dyn$intercepts$gamma; e <- fit_dyn$intercepts$epsilon
ns <- dim(sd_dyn$y)[3]; psi_t <- numeric(ns); psi_t[1] <- fit_dyn$intercepts$psi1
for (t in 2:ns) psi_t[t] <- psi_t[t - 1] * (1 - e) + (1 - psi_t[t - 1]) * g
tj <- data.frame(season = seq_len(ns), realised = colMeans(sd_dyn$truth$z),
                 modelled = psi_t)
ggplot(tj, aes(x = season)) +
  geom_line(aes(y = modelled), colour = "darkorange", linewidth = 1) +
  geom_point(aes(y = realised)) +
  ylim(0, 1) + labs(y = "occupied fraction") + transparent
```

The modelled curve rises from the fitted starting occupancy toward the
equilibrium the rates imply, tracking the realised fractions. The reason
to estimate the two rates rather than occupancy alone is that the same
standing occupancy can come from opposite dynamics. A level holding at
sixty percent could be a placid system where the same sites stay occupied
for years, or a churning one where many sites wink out each season and as
many are colonised. The static occupancy is identical; the rates tell them
apart, and they imply different conservation responses. The model reads
the rates from the pattern of which sites change state between consecutive
seasons, which is why a multi-season survey is worth the extra effort over
repeating a single-season one. Covariates on colonisation and extinction,
turnover, and the count analogue `dyn_abun()` are in
`vignette("dynamic-occupancy")`.

### Integrated occupancy

`int_occu()` combines several data sources that disagree about detection
but agree about occupancy. The sources share one latent occupancy state
and each carries its own detection process, so a structured survey anchors
the calibration while opportunistic records extend the reach. `y` is a
list of detection-history matrices, one per source.

```{r taste-int}
si <- simulate_int_occu(N_total = 500, n_data = 2, J = c(2, 5),
                        n_shared = 250, beta_occ = c(0.3, 0.9),
                        beta_det = list(c(0.5, 0), c(0.7, 0)), seed = 11)
fit_int <- tobs(~ x, data = si$data, family = int_occu(), detection = ~ 1,
                y = list(survey = si$y[[1]], casual = si$y[[2]]),
                method = "laplace", control = list(verbose = FALSE))
coef(fit_int)$psi                           # shared occupancy block
```

The fit returns one shared occupancy block plus a detection block per
source, keyed by the names you gave `y`. Plotting the two detection
intercepts on the probability scale shows each source's reporting rate,
estimated separately.

```{r taste-int-plot, eval = have_gg, fig.alt = "Per-source detection probability as bars, one per data source"}
dp <- data.frame(source = c("survey", "casual"),
                 p = plogis(c(coef(fit_int)$survey, coef(fit_int)$casual)))
ggplot(dp, aes(x = source, y = p)) +
  geom_col(fill = "steelblue", width = 0.5) +
  ylim(0, 1) + labs(x = NULL, y = "detection probability") +
  transparent
```

The two bars sit at different heights because the sources detect at
different rates, which is the point of the separate detection blocks. The
shared state borrows strength from both; the detection arms stay
independent. The logic that makes integration pay is that occupancy is a
property of the place and the species, not of who is looking, so a
structured survey and a casual stream can share one $z$ while keeping
their own $p$. A single thin source has to estimate occupancy and its own
detection from a few visits, and the two trade off; a second source that
observes overlapping sites brings more visits to bear on the shared state
without forcing the detection processes to match. The gain scales with the
overlap and with how much real occupancy signal the added source carries.
Where integration sharpens the occupancy slope, the overlap that drives
the gain, and source-specific detection covariates are in
`vignette("integrated-occupancy")`. The joint species distribution family
`jsdm()` and the detection-protocol families `removal()`, `distance()`,
and `fp_occu()` extend the same observation layer.

## S3 methods

The fitted object answers the standard R accessors, and they behave the
same on a Laplace fit and a NUTS fit; the Laplace tier reports the
Gaussian approximation at the mode, the sampler tier reports empirical
posterior summaries. `print()` lays out both processes and the family.
`summary()` adds the standard deviation and credible bounds, plus `Rhat`
and effective sample size under NUTS. `coef()` returns the per-process
estimates as a list, `confint()` the credible intervals as a matrix, and
`vcov()` the fixed-effect covariance.

```{r s3-methods}
co  <- coef(fit)            # list, one element per process
ci  <- confint(fit)         # matrix of credible bounds
vc  <- vcov(fit)            # fixed-effect covariance
list(coef = names(co), confint_rows = rownames(ci), vcov_dim = dim(vc))
```

Beyond the four above, `fitted()` returns the per-site `psi`, `p`, and
`z`; `ranef()` returns the group-level random effects when a model carries
them; `residuals()` returns deviance, Pearson, or response residuals;
`predict()` returns occupancy at a new design or a varied term; and
`simulate()` draws replicate detection histories from the fit. The
`plot()` method on a prediction object draws the marginal-effect curve
with its band, and on a data object summarises the detection histories.
The accessor surface is deliberately the base-R one, so a `tobs_fit`
drops into code written for `lm` or `glmer` outputs.

The accessors that involve uncertainty read it from whichever inference
produced the fit, without you having to ask. On a Laplace fit, `confint()`
and `vcov()` come from the curvature of the log-posterior at the mode, so
they are the Gaussian approximation and they return instantly. On a NUTS
fit, the same calls summarise the posterior draws, so the interval is the
empirical quantile of the chain and `summary()` carries the `Rhat` and
effective-sample-size columns alongside. The code you write does not
change between the two; only the numbers and their provenance do. That is
the practical payoff of keeping the observation layer thin and the
inference in the engine: you choose the backend per run and read the
result through one set of methods.

| Accessor | Returns |
|---|---|
| `coef()` | per-process fixed-effect estimates |
| `summary()` | estimate, SD, credible bounds (plus Rhat/ESS under NUTS) |
| `confint()` | credible intervals (matrix) |
| `vcov()` | fixed-effect covariance |
| `ranef()` | group-level random effects (when present) |
| `fitted()` | per-site `psi`, `p`, `z` |
| `predict()` | occupancy at a new design or varied term |
| `tobs_waic()` | WAIC, elpd, effective parameters |
| `residuals()` | deviance / Pearson / response residuals |
| `simulate()` | replicate detection histories from the fit |

## Practical guidance

- **Start with `method = "laplace"`.** It is deterministic and returns
  in well under a second on a few hundred sites. Use it for the first
  look and for WAIC comparison. Move to `method = "nuts"` when detection
  is low, visits are few, or you need the full posterior shape rather than
  a Gaussian approximation at the mode.
- **Visits drive identifiability.** Occupancy and detection trade off
  against each other. With two or three visits and detection below about
  0.3, the two processes are weakly separated and the Laplace mode can
  drift; NUTS or more visits resolve it. The
  [identifiability discussion in `vignette("occupancy")`](occupancy.html)
  works through where recovery fails.
- **Detection covariates broadcast across visits** when the data frame
  has one row per site, as in `simulate_occu()`. For covariates that
  change between visits (effort, observer, weather), pass a per-visit
  design; see `vignette("data-formatting")`.
- **Read uncertainty, not just point estimates.** `confint()` and the
  `summary()` quantile columns carry the credible bounds. A coefficient
  whose interval spans zero is not pulling its weight regardless of the
  point estimate.
- **WAIC compares, it does not validate.** A lower WAIC says one model
  predicts the in-sample data better; it does not say the model is
  calibrated. Pair it with a posterior-predictive check
  (`tobs_ppc()`, see `vignette("diagnostics")`).

## Where to go next

- Single-season occupancy end to end, with model comparison and
  identifiability: `vignette("occupancy")`.
- N-mixture abundance, negative binomial, and areal spatial counts:
  `vignette("abundance")`.
- Many species at once: `vignette("community-models")`.
- Colonisation and extinction across seasons:
  `vignette("dynamic-occupancy")`.
- Combining detection histories from several sources:
  `vignette("integrated-occupancy")`.
- The vegetation cover hurdle, independent and joint:
  `vignette("cover-hurdle")` and `vignette("cover-hurdle-motivate")`.
- Areal and continuous spatial fields:
  `vignette("spatial-occupancy")` and `vignette("occupancy-spatial-spde")`.
- Random effects with `lme4` bar syntax: `vignette("random-effects")`.
- Posterior-predictive checks, WAIC, and spatial residual diagnostics:
  `vignette("diagnostics")`.
```
