---
title: "Random effects with lme4 bar syntax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random effects with lme4 bar syntax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## When a grouping factor needs a random effect

Surveys come with structure that is not a covariate of interest. Sites
fall into clusters, observers differ in how hard they look, regions sit
under their own unmeasured baseline. You could enter the grouping factor
as a fixed effect, one coefficient per level, and for a handful of levels
that is fine. With thirty observers or two hundred site clusters it stops
being fine: each level gets its own intercept estimated from its own
sparse data, the levels with little data get wild estimates, and nothing
ties them together.

A random effect ties them together. Instead of estimating each level
freely, you assume the level effects are draws from a common distribution,
$b_g \sim N(0, \sigma^2)$, and estimate the spread $\sigma$. A level with
plenty of data keeps most of its own signal; a level with little data is
pulled toward the population mean. The model spends one parameter on the
spread rather than one per level, and the per-level estimates borrow
strength from each other. This is the difference between asking "what is
this observer's intercept" and "how much do observers vary, and where does
this one sit".

The honest question to ask first is which one you want. If the grouping
factor has a few levels and each is interesting in its own right (three
treatment arms, two field protocols), a fixed effect names them and is the
right tool. If it has many exchangeable levels and you care about the
variance more than any single level, a random effect is the right tool.
tulpaObs writes random effects the way `lme4` does, inside the formula,
with a bar.

The pull toward the mean is called shrinkage, and it is the feature, not a
side effect. A cluster surveyed once contributes one noisy observation; on
its own that observation would give a wild intercept. The random effect
treats it as a draw from the population and reports an estimate close to
the population mean, because the data give little reason to move it. A
cluster surveyed two hundred times anchors its own estimate, because its
data dominate the prior. The amount of shrinkage is set by the data, not
chosen: it falls out of the ratio between within-group noise and the
estimated between-group variance $\sigma^2$. This is also why a random
effect reports honest uncertainty. A fixed effect with one observation per
level claims a precise intercept it has no right to; the random effect
admits it is mostly guessing the mean.

## Bar syntax

A random-effect term is `(<effects> | <grouping factor>)`. The left side
says which coefficients vary by group; the right side names the grouping
factor (a `factor` in your data frame). The double bar `||` drops the
correlation between a group's intercept and its slopes. The supported
forms:

| Form | Meaning |
|---|---|
| `(1 | g)` | random intercept per level of `g` |
| `(x | g)` | random intercept and slope on `x`, correlated |
| `(1 + x | g)` | the same, written explicitly |
| `(x || g)` | random intercept and slope, uncorrelated |
| `(1 + x || g)` | the same, written explicitly |
| `(0 + x | g)` | random slope on `x`, no group intercept |
| `(1 | g:h)` | random intercept for each crossed combination of `g` and `h` |
| `(1 | g/h)` | nested: a `g` intercept plus an `h`-within-`g` intercept |

A term enters whichever linear predictor its formula sits in. Put
`(1 | g)` in the occupancy formula and it varies occupancy by group; put
it in the `detection` formula and it varies detection by group. The
process membership comes from the formula, not from a separate argument.

Under the hood `tulpaObs` rewrites the bars into `re()` calls on the
formula tree before parsing, so there is one parser and one term type. You
never call `re()` yourself; the bar is the surface you type.

## A random intercept under Laplace

Start with the simplest case: occupancy varies by a site cluster. We
simulate thirty clusters of twenty-five sites each, give each cluster its
own intercept drawn from $N(0, 0.9^2)$, and add one site covariate `x`.
Detection is constant at about 0.45 across six visits.

```{r sim-intercept}
sim_re <- function(seed = 101, ng = 30L, per = 25L, J = 6L,
                   b0 = 0.3, b1 = -0.6, sigma = 0.9, p = 0.45) {
  set.seed(seed)
  N <- ng * per
  g <- rep(seq_len(ng), each = per)
  x <- rnorm(N)
  b_true <- rnorm(ng, 0, sigma)               # the per-cluster intercepts
  z <- rbinom(N, 1, plogis(b0 + b1 * x + b_true[g]))
  y <- matrix(0L, N, J)
  for (i in seq_len(N)) y[i, ] <- rbinom(J, 1, z[i] * p)
  list(y = y, d = data.frame(g = factor(g), x = x), b_true = b_true)
}

s <- sim_re()
str(s$d)
```

The grouping factor `g` is a `factor`. Fit occupancy with a fixed slope on
`x` and a random intercept for `g`. The default `method = "laplace"` fits
the random effect through a variance-component EM; pass
`verbose = FALSE` so the EM does not print its iterations.

```{r fit-intercept}
fit <- tobs(
  ~ x + (1 | g),
  data      = s$d,
  y         = s$y,
  detection = ~ 1,
  family    = occu(),
  method    = "laplace",
  control   = list(verbose = FALSE)
)
```

The fixed effects come back from `coef()` as usual. The occupancy
intercept and slope land near the simulated 0.3 and -0.6.

```{r coef-intercept}
coef(fit)$psi
```

The variance component is a hyperparameter in `fit$means`, named for the
group it belongs to. There is one `sigma_` entry for the random intercept.

```{r sigma-intercept}
sig_name <- grep("^sigma_", names(fit$means), value = TRUE)
sig_name
c(estimate = fit$means[[sig_name]], truth = 0.9)
```

`ranef()` returns the per-group effects (the BLUPs, best linear unbiased
predictors) as a data frame with one row per group and term: which
grouping factor, which level, which coefficient, the estimate, and its
standard error.

```{r ranef-intercept}
re <- ranef(fit)
head(re)
nrow(re)
```

The point of the random effect is that these per-cluster intercepts track
the truth. Plotting the recovered BLUPs against the simulated cluster
effects shows the model recovers the cluster-level signal, with the
shrinkage that pulls small-data clusters toward zero.

```{r blup-plot, fig.alt = "Recovered per-cluster random intercepts against the simulated truth, scattered along the identity line with slight shrinkage toward zero"}
plot(s$b_true, re$estimate, pch = 16, col = rgb(0, 0, 0, 0.5),
     xlab = "true cluster effect", ylab = "estimated BLUP")
abline(0, 1, col = "steelblue", lwd = 2)
cor(s$b_true, re$estimate)
```

## The EM, the small-cluster bias, and the AGHQ debias

The default Laplace path integrates the random-effect block with a
Gaussian approximation at its mode. For binary occupancy data that is the
`glmer` `nAGQ = 1` regime: the per-group integral is approximated by a
single Gaussian, and at small per-group sample sizes that approximation
pulls the estimated variance toward zero. The point estimates and the
fixed-effect standard errors are fine; the variance component itself comes
out attenuated.

tulpaObs corrects this by default with an adaptive Gauss-Hermite
quadrature pass. After the EM converges it replaces the single-Gaussian
per-group integral with a quadrature grid, which captures the curvature
the Laplace approximation misses, and re-estimates the variance component.
The control is `re.aghq` (on by default) and `n.quad` (the number of
quadrature nodes, default 9). The fit records what happened in `fit$aghq`.

```{r aghq-status}
fit$aghq$applied
fit$aghq$n_quad
```

The effect is largest when groups are small. With per-group $n = 8$ the
measured sigma bias runs about 18% under the raw EM and drops to about 4%
after the AGHQ pass, matching a NUTS fit. We can see the direction on a
single small-cluster dataset by turning the pass off and on.

```{r aghq-compare}
s8 <- sim_re(seed = 21, ng = 30L, per = 8L, sigma = 0.9)

fit_em <- tobs(~ x + (1 | g), data = s8$d, y = s8$y, detection = ~ 1,
               family = occu(), method = "laplace",
               control = list(re.aghq = FALSE, verbose = FALSE))
fit_aghq <- tobs(~ x + (1 | g), data = s8$d, y = s8$y, detection = ~ 1,
                 family = occu(), method = "laplace",
                 control = list(re.aghq = TRUE, verbose = FALSE))

get_sigma <- function(f) f$means[[grep("^sigma_", names(f$means), value = TRUE)]]
c(truth = 0.9,
  EM    = get_sigma(fit_em),
  AGHQ  = get_sigma(fit_aghq))
```

The EM estimate sits below the truth; the AGHQ pass corrects it upward.
The mechanism is worth one sentence: a single Gaussian centred at the
per-group mode underweights the tails of the integral over the group
effect, and for binary data that loss falls on the variance, so the
estimate shrinks. Adding quadrature nodes samples those tails and recovers
the spread. You rarely need to think about this. The default does the
right thing, and `fit$aghq$applied` tells you it ran. The pass applies to one grouping
factor on one predictor with random-effect dimension up to 3; outside that
(a random effect split across both occupancy and detection, crossed or
nested groupings) the fit keeps the EM estimate, and NUTS is available for
the full treatment.

## Correlated random slopes

A random slope lets the covariate effect vary by group, not just the
intercept. `(1 + x | g)` gives each group its own intercept and its own
slope on `x`, with a correlation between the two estimated as well. We
simulate from a known $2 \times 2$ covariance with a positive
intercept-slope correlation.

```{r sim-corr}
sim_corr <- function(seed = 402, ng = 40L, per = 25L, J = 6L,
                     b0 = 0.2, b1 = -0.4, p = 0.5,
                     Sigma = matrix(c(0.6, 0.3, 0.3, 0.4), 2, 2)) {
  set.seed(seed)
  N <- ng * per
  g <- rep(seq_len(ng), each = per)
  x <- rnorm(N)
  U <- matrix(rnorm(ng * 2L), ng, 2L) %*% chol(Sigma)   # rows ~ N(0, Sigma)
  eta <- b0 + U[g, 1] + (b1 + U[g, 2]) * x
  z <- rbinom(N, 1, plogis(eta))
  y <- matrix(0L, N, J)
  for (i in seq_len(N)) y[i, ] <- rbinom(J, 1, z[i] * p)
  list(y = y, d = data.frame(g = factor(g), x = x), U = U, Sigma = Sigma)
}

sc <- sim_corr()
fit_c <- tobs(~ x + (1 + x | g), data = sc$d, y = sc$y, detection = ~ 1,
              family = occu(), method = "laplace",
              control = list(verbose = FALSE))
```

The block reports two standard deviations (one for the intercept, one for
the slope) and the off-diagonal as a correlation named `cor_<g>_...`. The
true correlation here is $0.3 / \sqrt{0.6 \times 0.4} \approx 0.61$.

```{r corr-params}
fit_c$means[grep("^sigma_|^cor_", names(fit_c$means))]
```

The Laplace correlation is regularized by a default LKJ($\eta = 1.5$)
penalty (`re.lkj`) on each correlated block. A random-effect correlation
is often weakly identified, and unregularized maximum likelihood can pin
it against $\pm 1$. The LKJ penalty pulls a weakly-identified correlation
off the boundary toward zero without touching the marginal standard
deviations. Set `re.lkj = 1` to disable it; for a full posterior treatment
of the correlation, use NUTS.

The per-group slopes recover too. The `ranef()` table now carries both an
`(Intercept)` term and an `x` term per group.

```{r corr-ranef}
re_c <- ranef(fit_c)
unique(re_c$term)
b1_hat <- re_c$estimate[re_c$term == "x"]
cor(b1_hat, sc$U[, 2])
```

The uncorrelated forms `(x || g)`, `(1 + x || g)`, and the slope-only
`(0 + x | g)` fit the same way; they drop the off-diagonal, so you get the
standard deviations without a `cor_` entry. Use `||` when you have no
reason to expect the intercept and slope deviations to move together, or
when the correlation is too weakly informed to estimate.

## Crossed and nested grouping

Two grouping factors can apply at once. They are crossed when every level
of one can pair with every level of the other (sites visited by any of
several observers), and nested when each level of the inner factor belongs
to exactly one level of the outer (transects within plots). The bar syntax
follows `lme4`: `(1 | g:h)` is crossed, `(1 | g/h)` is nested.

Crossed grouping builds one random intercept for each observed combination
of the two factors. We cross eight site groups with six survey rounds and
give each combination its own occupancy intercept.

```{r sim-crossed}
set.seed(55)
ng <- 8L; nh <- 6L; per <- 12L
combos <- expand.grid(g = seq_len(ng), h = seq_len(nh))
N <- nrow(combos) * per
g <- rep(combos$g, each = per); h <- rep(combos$h, each = per)
gh <- interaction(factor(g), factor(h), drop = TRUE)
b_gh <- rnorm(nlevels(gh), 0, 0.8)
x <- rnorm(N)
z <- rbinom(N, 1, plogis(0.2 - 0.4 * x + b_gh[gh]))
y <- matrix(0L, N, 6L)
for (i in seq_len(N)) y[i, ] <- rbinom(6L, 1, z[i] * 0.5)
d_cr <- data.frame(g = factor(g), h = factor(h), x = x)

fit_cr <- tobs(~ x + (1 | g:h), data = d_cr, y = y, detection = ~ 1,
               family = occu(), method = "laplace",
               control = list(verbose = FALSE))
nrow(ranef(fit_cr))      # one BLUP per observed g:h combination
coef(fit_cr)$psi
```

The crossed term `(1 | g:h)` collapses to a single interaction factor, so
the Laplace path treats it as one grouping factor and the AGHQ debias
applies. The nested form `(1 | g/h)` expands to two terms, an outer `g`
intercept and an inner `h`-within-`g` intercept; the EM fits both, and the
AGHQ debias steps aside because there is more than one grouping factor.
Read the nested variance components as the attenuated EM estimates, or
refit under NUTS when the variance partition matters.

## A random effect on detection

Detection effort varies by observer. Put the bar in the `detection`
formula and the random effect moves the detection probability instead of
occupancy. Here observers each have their own detection intercept, drawn
from $N(0, 0.8^2)$, while occupancy depends on a site covariate with no
random part.

```{r sim-det}
sim_det <- function(seed = 1, N = 400L, J = 6L, ng = 40L,
                    b0 = 0.4, b1 = -0.7, d0 = 0.2, sigma = 0.8) {
  set.seed(seed)
  occ_cov  <- rnorm(N)
  observer <- sample.int(ng, N, replace = TRUE)
  b_obs    <- rnorm(ng, 0, sigma)
  z <- rbinom(N, 1, plogis(b0 + b1 * occ_cov))
  p <- plogis(d0 + b_obs[observer])
  y <- matrix(0L, N, J)
  for (i in seq_len(N)) if (z[i] == 1L) y[i, ] <- rbinom(J, 1L, p[i])
  list(y = y, d = data.frame(occ_cov = occ_cov, observer = factor(observer)),
       b_obs = b_obs)
}

sd_re <- sim_det()
fit_d <- tobs(~ occ_cov, detection = ~ (1 | observer), family = occu(),
              data = sd_re$d, y = sd_re$y, method = "laplace",
              control = list(verbose = FALSE))
```

The variance component now belongs to the detection arm, so it is named
for the detection process: `sigma_p<t>_` rather than the occupancy
`sigma_g<t>_`. The AGHQ pass records that it ran on the detection arm.

```{r det-params}
grep("^sigma_", names(fit_d$means), value = TRUE)
fit_d$aghq$arm
```

The detection arm is harder than the occupancy arm because only occupied
sites carry information about detection. A site that is empty tells you
nothing about how well the observer detects. The raw EM attenuates the
detection-RE variance heavily for that reason (about 70% in the measured
sims), and the AGHQ pass on the exact detection marginal restores it to
within about 1% with 88-96% fixed-effect interval coverage. The per-group
detection BLUPs track the simulated observer effects.

```{r det-ranef}
re_d <- ranef(fit_d)
rp <- re_d[re_d$group == "p1", ]
cor(rp$estimate, sd_re$b_obs)
```

## Laplace versus NUTS for random effects

The Laplace path with the AGHQ debias covers the common single-grouping
cases: a random intercept or slope on one of the two predictors, with the
variance component corrected for the small-cluster bias. NUTS fits every
random-effect form and returns the full posterior, including the full
posterior of the correlation rather than a regularized point estimate. It
is the tool for the forms Laplace declines and for any case where you want
draws rather than a Gaussian approximation.

Here is the same random intercept under NUTS. The variance component comes
back on the log scale as `log_sigma_`, and `ranef()` reconstructs the
per-group BLUPs by marginalising over the draws.

```{r nuts}
fit_n <- tobs(~ x + (1 | g), data = s$d, y = s$y, detection = ~ 1,
              family = occu(), method = "nuts",
              control = list(n.iter = 400, n.warmup = 200, seed = 1,
                             verbose = FALSE))

sig_l <- fit$means[[grep("^sigma_", names(fit$means), value = TRUE)]]
sig_n <- exp(fit_n$means[[grep("^log_sigma_", names(fit_n$means), value = TRUE)]])
c(laplace = sig_l, nuts = sig_n, truth = 0.9)
```

The two backends agree on the variance, and the BLUPs agree across
engines.

```{r nuts-blups}
re_n <- ranef(fit_n)
cor(re$estimate, re_n$estimate)
```

A few random-effect specifications are NUTS-only. The Laplace path errors,
with a pointer to NUTS, for: a single random effect shared across both the
occupancy and detection predictors (`share()` semantics); a random effect
combined with a spatial field in the same fit; and a random effect
combined with visit-level detection covariates. In each case the
deterministic path would have to fit something it cannot represent
exactly, so it stops rather than silently fitting a different model. NUTS
handles all three.

The reason for the boundary is in how the Laplace path works. The
variance-component EM splits the random effects into an occupancy arm and
a detection arm by which predictor they sit in, then fits each arm's block
separately while feeding its mode back into the other process. A single
effect that has to be the same realization in both predictors breaks that
split: there is no way to fit it once on the occupancy arm and once on the
detection arm and call the two the same. Likewise the AGHQ debias works on
a per-site marginal with the latent occupancy state integrated out in
closed form; a spatial field or a per-visit detection design changes that
marginal in a way the closed form does not cover. Rather than approximate
silently, the validator stops and names NUTS, which samples the joint
posterior directly and needs none of these closed forms. The cost is
speed: a NUTS fit takes seconds where Laplace takes a fraction of one, and
you read convergence from `Rhat` and effective sample size rather than
trusting a mode.

## Practical guidance

- **Want at least five to eight levels before the variance is informed.**
  A random effect estimates the spread across levels. With two or three
  levels there is almost no information about that spread, the variance is
  badly determined, and a fixed effect that names each level is the honest
  choice. Five to eight levels is the rough floor where the variance starts
  to mean something; the recovery here used thirty and forty.
- **Use a fixed effect when the levels are few and each is interesting.**
  Treatment arms, field protocols, a handful of named regions: these are
  fixed effects. You want their individual values, not their variance, and
  there are too few to estimate a variance well.
- **Keep at least ten or so observations per group for slopes.** A random
  slope asks each group's data to inform its own slope before shrinkage.
  Groups with two or three observations contribute almost nothing and the
  block leans entirely on the population estimate. Random intercepts
  tolerate sparser groups than random slopes do.
- **Trust the AGHQ default on small clusters.** When per-group counts drop
  below about ten, the raw EM variance is attenuated; the default
  `re.aghq = TRUE` corrects it. Check `fit$aghq$applied` to confirm it ran.
  When it does not apply (split arms, crossed or nested groupings, RE
  dimension above 3), read the variance as a lower bound and refit under
  NUTS if the variance matters to your conclusion.
- **Reach for `||` when the correlation is not the point.** The correlated
  block `(1 + x | g)` spends a parameter on the intercept-slope
  correlation, which is often weakly identified. If you have no use for
  that correlation, `(x || g)` drops it and estimates the two standard
  deviations more stably.
- **When not to use a random effect at all.** If the grouping factor is the
  exposure you care about (the thing whose effect is your research
  question), it belongs in the fixed part of the model. A random effect is
  for nuisance structure you want to account for, not for the signal you
  came to measure.

## Where to go next

- The single-season occupancy model the random effects attach to:
  `vignette("occupancy")`.
- Areal spatial fields, which use the same in-formula term machinery:
  `vignette("spatial-occupancy")`.
- Community models, where species are the grouping factor and the
  community covariances are fit internally: `vignette("community-models")`.
