---
title: "Areal spatial occupancy with ICAR, BYM2, and CAR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Areal spatial occupancy with ICAR, BYM2, and CAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpaObs)
```

## Adjacent regions share what the covariates miss

Two survey cells that sit next to each other tend to agree. They share
soil, climate, land use, and the dispersal of whatever you are surveying,
so a species present in one is more likely present in the next. A
plain occupancy regression treats every cell as an independent draw from
its covariates. That assumption costs you twice. The residuals come out
spatially correlated, which means the effective sample size is smaller
than the cell count and the standard errors run too narrow. And when a
covariate is itself patterned across space, it competes with the
unmeasured regional signal for the same variance, so the fitted slope
absorbs part of that signal and drifts.

Areal data carry a fix for free: a neighbourhood graph. The cells come
pre-divided into discrete units joined by borders, and "who borders whom"
is the structure a spatial field rides on. You add one random effect per
cell, tie neighbouring effects together so the field stays smooth, and
let it soak up the part of occupancy that depends on *where* a cell sits
rather than on its covariates. The slope is then read against a
background that already accounts for location, and its standard error
widens to its honest size.

The smoothing is the part that takes care. A free random effect per cell
with no ties between neighbours would absorb the spatial signal, but with
one parameter per cell it would also absorb the covariate effect you came
to estimate, because that many free parameters can mimic almost anything.
The neighbour ties are what keep the surface smooth and leave the
covariate its own variance. How strong those ties should be is a
parameter in its own right, the smoothing precision, and a fit that fixes
it by hand makes a choice the data could make better. The nested-Laplace
engine lays a grid over that precision, solves the field at each grid
point, and pools the answers weighted by their evidence, so the slope
errors inherit the field's uncertainty instead of pretending the
smoothing strength was known all along.

This vignette stays on the **areal** path: regions are discrete units
joined by an adjacency matrix, and the field is built from that graph. If
your sites are raw points on a map, coordinates rather than regions with
borders, you want a continuous field instead. That is the
`vignette("occupancy-spatial-spde")` story, and the closing section here
says when each one is appropriate. The engine throughout is
`method = "nested_laplace"`, which integrates the field's smoothing
strength over a grid rather than fixing it by hand.

## The three fields

tulpaObs builds areal fields through formula terms. Three matter here:
`icar()`, `bym2()`, and the proper CAR `car_proper()`. They differ in
what they assume about the spatial signal.

**ICAR** (intrinsic conditional autoregressive) is the plain smoother.
Each cell's effect is centred on the average of its neighbours. Write the
field $\phi$ over the regions; its joint distribution is a Gaussian
Markov random field with precision

$$
Q \;=\; \tau\,(D - W),
$$

where $W$ is the adjacency matrix, $D$ the diagonal of neighbour counts,
and $\tau$ a precision that sets the smoothing strength. The conditional
form is the one to carry in your head: each cell's effect is normally
distributed around the mean of its neighbours, with variance
$1/(\tau\,n_r)$ where $n_r$ is the neighbour count, so a cell with many
neighbours is held more tightly to their average than a cell with few. A
large $\tau$ forces neighbours to agree and flattens the field; a small
$\tau$ relaxes the penalty and lets each cell drift toward an independent
effect.

The off-diagonal entries of $Q$ are nonzero only between neighbours,
which is the Markov property of the field written as a matrix: a cell is
conditionally independent of every non-neighbour given its own
neighbours. That sparsity is what keeps the model cheap even on thousands
of cells. The matrix $D - W$ is the graph Laplacian, rank-deficient by
one with the constant vector in its null space, so the ICAR pins the
field only up to an additive level. Shifting every $\phi_r$ by the same
amount leaves the joint density unchanged. The intercept identifies that
level, and the field carries only the contrasts between cells, the part
that actually describes spatial pattern. ICAR assumes the signal is
entirely smooth: all regional variation is spatially structured, none of
it independent noise.

**BYM2** relaxes that assumption. It splits the field into a structured
ICAR part and an unstructured per-cell part, mixed by a fraction
$\lambda$:

$$
\phi_r \;=\; \frac{1}{\sqrt{\tau}}\left(\sqrt{\lambda}\,u_r^{\star}
  + \sqrt{1-\lambda}\,v_r\right),
$$

where $u^{\star}$ is a scaled ICAR field, $v$ an IID per-cell effect, and
$\lambda \in [0,1]$ the share of variance that is spatially structured.
At $\lambda = 1$ the field is pure ICAR; at $\lambda = 0$ it is pure
noise. The ICAR component is scaled following Riebler et al. (2016) so
that $\lambda$ reads on a comparable scale from one graph to the next,
which a raw ICAR variance does not. Without that scaling the same
$\lambda$ would mean different things on a sparse graph and a dense one,
because the marginal variance of an unscaled ICAR field depends on the
graph's connectivity; the Riebler scaling normalises the structured
component to unit generalised variance first, so $\lambda$ is a clean
fraction. BYM2 is the common default in disease mapping for this reason:
it separates "smooth regional driver" from "cell-specific quirk" and
reports how the variance splits between them. It also tames the intercept
confounding the plain ICAR carries, because the IID component gives the
field somewhere to put cell-level variation that is not on the flat
ridge.

**CAR proper** (`car_proper()`) keeps the single field of the ICAR but
replaces the fixed unit autocorrelation with an estimated $\rho$, so the
precision becomes $Q = \tau\,(D - \rho W)$. With $\rho$ free in its valid
range the field can run from near-independence to near-ICAR, and the
precision is proper rather than rank-deficient, which tames the intercept
confounding the ICAR carries. The cost is one more parameter to integrate.

For most occupancy smoothing tasks ICAR or BYM2 is the starting point.
Reach for the proper CAR when the autocorrelation strength is itself a
quantity you want to report.

## Building the adjacency

An areal field needs one thing from you: a dense 0/1 adjacency matrix,
one row and one column per site, with a 1 where two sites are neighbours
and 0 elsewhere. The matrix is symmetric with a zero diagonal. The site
order has to match the row order of your data and your detection-history
matrix `y`, because the field's cell `i` is the model's site `i`.

Here is a 5-by-5 rook-adjacency grid: each cell borders the cells
directly north, south, east, and west of it.

```{r adjacency}
gx <- 5; gy <- 5
n  <- gx * gy
coord <- expand.grid(cx = seq_len(gx), cy = seq_len(gy))
idx_of <- function(i, j) (j - 1) * gx + i

adj <- matrix(0, n, n)
for (i in seq_len(gx)) for (j in seq_len(gy)) {
  a <- idx_of(i, j)
  if (i < gx) { b <- idx_of(i + 1, j); adj[a, b] <- adj[b, a] <- 1 }
  if (j < gy) { b <- idx_of(i, j + 1); adj[a, b] <- adj[b, a] <- 1 }
}
dim(adj)
rowSums(adj)[1:6]   # neighbour count per cell: corners 2, edges 3, interior 4
```

A chain graph is even simpler: site $i$ borders $i-1$ and $i+1$. The same
matrix shape works, and a chain is the areal analogue of a random walk
along a transect.

```{r chain}
K <- 12
chain <- matrix(0, K, K)
for (i in seq_len(K - 1)) chain[i, i + 1] <- chain[i + 1, i] <- 1
rowSums(chain)   # endpoints have 1 neighbour, interior 2
```

Whatever the source (a shapefile's `poly2nb`, a regular lattice, a
hand-drawn graph), the field only ever sees this dense matrix. Real
shapefile workflows usually produce a neighbour list; convert it to the
0/1 matrix before passing it in.

## Simulating data with a known field

A spatial fit is only honest to check against a known truth. Real data
never reveal the field, so a recovered map that merely looks plausible
proves nothing. We set a smooth field on the grid by hand, draw occupancy
from it, simulate a detection history, and ask the fit to find the field
again. The response is binary and detection is imperfect, so this has
teeth: the model has to separate a smooth occupancy surface from the
binomial noise of eight coin-flip visits per site.

The true field is a smooth bump centred on the grid plus a gentle
gradient. A bump-plus-gradient is a fair test on two counts. It is
smooth, so an ICAR prior that rewards agreement between neighbours tracks
it rather than fights it, and it has real structure, a peak in the middle
and a rising corner, that no single intercept can fake. A field of pure
noise would be the opposite test, one the prior is designed to smooth
away, and recovering it would say little. We centre the field to mean
zero so the intercept carries the overall level and the field carries the
contrasts, matching the way the ICAR identifies them. Occupancy is
`plogis(field)`; eight visits per site at detection 0.5 give the model
enough to pull occupancy apart from detection.

```{r simulate}
field_true <- 0.9 * scale(coord$cx)[, 1] + 0.7 * scale(coord$cy)[, 1] +
              1.4 * exp(-((coord$cx - 3)^2 + (coord$cy - 3)^2) / 4)
field_true <- field_true - mean(field_true)

J     <- 8
p_det <- 0.5
psi_true <- plogis(field_true)
z <- rbinom(n, 1, psi_true)
y <- matrix(0L, n, J)
for (i in seq_len(n)) if (z[i]) y[i, ] <- rbinom(J, 1, p_det)

dat <- data.frame(site = seq_len(n))
c(occupied_cells = sum(z), naive_rate = round(mean(rowSums(y) > 0), 3),
  true_rate = round(mean(z), 3))
```

The naive detection rate undercounts occupancy, the usual occupancy
problem. The field is what we want back: a smooth surface that peaks in
the grid centre and rises toward the high-coordinate corner.

## Fitting an areal field

The spatial term lives inside the occupancy formula. `bym2(graph = adj)`
adds a BYM2 field over the sites named by the adjacency matrix. The
`method = "nested_laplace"` engine runs the inner Laplace solve at each
point of a grid over the smoothing hyperparameters and pools the results,
so the coefficient errors inherit the field's uncertainty rather than
pretending the smoothing strength is known.

```{r fit-bym2}
fit_bym2 <- tobs(
  ~ bym2(graph = adj),
  data      = dat,
  family    = occu(),
  detection = ~ 1,
  y         = y,
  method    = "nested_laplace",
  control   = list(max.iter = 30L, tol = 1e-5, verbose = FALSE)
)
```

The fixed effects come back from `coef()`. With an intercept-only mean
structure the only occupancy coefficient is the floating level the field
sits against.

```{r coef-bym2}
coef(fit_bym2)
```

The block that did the spatial work is recorded on the fit. The
nested-Laplace path stores its prior blocks under
`fit$nested_laplace$multi_prior`, one entry per latent term, and the
first block's `type` confirms BYM2 was the field used.

```{r prior-type}
fit_bym2$nested_laplace$multi_prior[[1]]$type
```

Swapping the field is a one-token change. Fit the same data with an ICAR
field and read its block type.

```{r fit-icar}
fit_icar <- tobs(
  ~ icar(graph = adj),
  data      = dat,
  family    = occu(),
  detection = ~ 1,
  y         = y,
  method    = "nested_laplace",
  control   = list(max.iter = 30L, tol = 1e-5, verbose = FALSE)
)
fit_icar$nested_laplace$multi_prior[[1]]$type
```

ICAR has one fewer parameter than BYM2 (no mixing fraction $\lambda$),
assumes the whole field is smooth, and is the lighter call for a quick
smoothing pass. BYM2 buys the structured-versus-noise split at the cost
of that extra parameter. On this simulation the field is smooth by
construction, so the two should track each other closely; the next
section checks that.

Both fits route through the same nested-Laplace machinery and return the
same `tobs_fit` shape. The only line that changed between the two calls
is the field term inside the formula. That is the composition the package
is built around: the family, the detection model, and the spatial field
combine orthogonally, so switching the smoother never touches the rest of
the model.

The proper CAR term, `car_proper(graph = adj)`, estimates the
autocorrelation strength $\rho$ rather than fixing it at one. That makes
the precision proper and the intercept identifiable, at the cost of
integrating one extra parameter, and it is the field to reach for when
$\rho$ is a number you want to report rather than a nuisance. On the
current backend the proper CAR is wired for the cover-hurdle
nested-Laplace path rather than the occupancy engine, so for occupancy
the choice is between `icar()` and `bym2()`; the practical guidance below
covers when each fits.

## Recovering and mapping the field

The smoothed surface is usually the point of the exercise. The clean way
to read it off a nested fit is `predict(type = "state")`, which returns
the marginalised per-site occupancy posterior: the smoothed `psi` for
every cell, with a calibrated credible band, integrated over the
smoothing grid.

```{r state-bym2}
sp_bym2 <- predict(fit_bym2, type = "state")
head(sp_bym2)
```

The `psi` column is the field on the probability scale. Plotting it
against the true occupancy probability shows the fit recovered the
cell-by-cell surface, well past the population mean a flat model would
give.

```{r recover-plot, fig.alt = "Fitted occupancy from the BYM2 field against the true simulated occupancy, scattered around the identity line"}
plot(psi_true, sp_bym2$psi, pch = 19, col = "steelblue",
     xlab = "true occupancy", ylab = "fitted occupancy (BYM2)")
abline(0, 1, lwd = 2, col = "grey40")
cor(psi_true, sp_bym2$psi)
```

The ICAR fit recovers the same surface. Lining up the two fitted fields
shows the choice of field barely moves the smoothed map on data this
smooth, which is what the simulation predicts.

```{r icar-vs-bym2, fig.alt = "Fitted occupancy from the ICAR field against the BYM2 field, near the identity line"}
sp_icar <- predict(fit_icar, type = "state")
plot(sp_icar$psi, sp_bym2$psi, pch = 19, col = "darkorange",
     xlab = "fitted occupancy (ICAR)", ylab = "fitted occupancy (BYM2)")
abline(0, 1, lwd = 2, col = "grey40")
```

A map makes the smoothing visible. Lay the fitted `psi` back on the grid
coordinates and shade each cell. The high-occupancy ridge runs through
the grid centre and the upper corner, exactly where the simulated bump
and gradient put it.

```{r field-map, fig.width = 7, fig.height = 3.4, fig.alt = "Two grid maps side by side, true occupancy and BYM2-fitted occupancy, both peaking in the grid centre"}
shade <- function(p) grey(1 - (p - min(p)) / (max(p) - min(p)) * 0.85)
op <- par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
plot(coord$cx, coord$cy, pch = 22, cex = 4, bg = shade(psi_true),
     xlab = "", ylab = "", main = "true occupancy")
plot(coord$cx, coord$cy, pch = 22, cex = 4, bg = shade(sp_bym2$psi),
     xlab = "", ylab = "", main = "fitted (BYM2)")
par(op)
```

Darker cells are higher occupancy. The fitted map smooths the binary
noise that an isolated per-cell estimate would carry: a cell with few
detections is pulled toward the surface its neighbours trace.

### The floating intercept

One detail of an ICAR fit surprises people the first time. The intercept
comes back with a very wide standard error, often two or three orders of
magnitude larger than any slope error would be. This is by design, not a
fitting failure, and it follows straight from the rank-deficiency of the
ICAR precision. The overall level of the field and the intercept are
confounded: raise every $\phi_r$ by a constant and lower the intercept by
the same constant, and the linear predictor does not move, so the
likelihood cannot tell the two apart. They trade off along a flat ridge,
and no amount of data pins them both. tulpaObs lets the intercept float
against a near-flat prior, which is what produces the wide error.

The practical rule is short. Under an intrinsic field, read the slopes,
which are identified, and treat the intercept as a floating reference
rather than a baseline rate. A reader who quotes the intercept as an
absolute occupancy level will be misled. If you need a clean intercept,
`bym2()` and `car_proper()` both break the exact confounding (the BYM2
through its IID component, the CAR through its proper precision), at the
cost of one more estimated parameter. The intercept-only mean structure
here makes the point starkly, since the floating level is the only
occupancy coefficient there is; with real covariates the slopes sit
beside it, unaffected by the ridge.

## Predicting unsurveyed cells

Here is what the areal field buys that a non-spatial model cannot. A cell
whose detection history is entirely missing carries no direct
information, yet its neighbours do. A non-spatial model has nothing to say
about such a cell beyond the population mean, because it learns occupancy
only from the cell's own detections. The areal model learns it from the
neighbourhood. The nested path keeps the all-missing cell in the latent
field, drops it from the likelihood (its number of trials is zero, so it
contributes no term), and interpolates its occupancy from the smoothed
surface the surrounding cells define. This is the INLA-style NA-response
prediction. It works for any of the areal fields, and for a mixed-scale
field such as BYM2 the prediction reads the engine's per-cell fitted
linear predictor directly rather than reconstructing it from the modes,
so the interpolation is exact for every prior.

Hold out a scattered set of cells by setting their detection histories to
`NA`. Scattering them (rather than blanking a contiguous block) keeps each
held-out cell with observed neighbours, so the field interpolates rather
than extrapolates across a gap.

```{r holdout}
heldout <- seq(2, n, by = 4)        # every fourth cell, scattered across the grid
y_ho <- y
y_ho[heldout, ] <- NA
length(heldout)
```

Fit the field on the punctured data. `predict(type = "state")` flags the
held-out rows and returns their interpolated occupancy with a band.

```{r holdout-fit}
fit_ho <- tobs(
  ~ icar(graph = adj),
  data      = dat,
  family    = occu(),
  detection = ~ 1,
  y         = y_ho,
  method    = "nested_laplace",
  control   = list(max.iter = 30L, tol = 1e-5, verbose = FALSE)
)
sp_ho <- predict(fit_ho, type = "state")
table(sp_ho$heldout)
```

The `heldout = TRUE` rows are the cells the model never saw a detection
for. Their `psi` is the field's best guess, and `psi_lower` / `psi_upper`
carry the 95% band. Check the guess against the truth the held-out cells
actually had.

```{r holdout-recover, fig.alt = "Interpolated occupancy for held-out cells against their true occupancy, with vertical 95 percent bands"}
ho <- sp_ho[sp_ho$heldout, ]
truth_ho <- psi_true[heldout]

ord <- order(truth_ho)
plot(truth_ho[ord], ho$psi[ord], pch = 19, col = "steelblue", ylim = c(0, 1),
     xlab = "true occupancy (held out)", ylab = "interpolated occupancy")
arrows(truth_ho[ord], ho$psi_lower[ord], truth_ho[ord], ho$psi_upper[ord],
       length = 0.02, angle = 90, code = 3, col = adjustcolor("steelblue", 0.5))
abline(0, 1, lwd = 2, col = "grey40")
c(cor = round(cor(ho$psi, truth_ho), 3),
  mae = round(mean(abs(ho$psi - truth_ho)), 3))
```

The interpolated points track the truth and the bands cover it, all from
neighbour borrowing alone. The band is the calibrated mixture-CDF
interval the nested path computes from the exact-marginal field pass, so
it reflects both the field's uncertainty and the smoothing grid. A held-
out cell in a high-occupancy neighbourhood predicts high, one in a sparse
neighbourhood predicts low, by exactly the field contrast its neighbours
trace.

## Does the field earn its place?

In simulation a high correlation with a known truth says the field
works. On real data there is no truth, so the question becomes whether
the field improves prediction enough to justify its parameters. A field
adds one effective parameter per cell, give or take, tempered by the
smoothing, so it is never free. WAIC answers whether it pays for itself:
it estimates out-of-sample predictive accuracy from the pointwise
log-likelihood, lower is better, and it docks the model for the effective
parameters the field spends. Compare the BYM2 fit against a non-spatial
occupancy fit on the same data.

```{r waic}
fit_flat <- tobs(
  ~ 1, data = dat, family = occu(), detection = ~ 1, y = y,
  method = "laplace", control = list(verbose = FALSE)
)

c(non_spatial = tobs_waic(fit_flat)$waic,
  bym2        = tobs_waic(fit_bym2)$waic)
```

The spatial fit carries the lower WAIC, matching how the data were built:
occupancy here is driven entirely by a smooth field, so a model with no
place to put it predicts worse. The gap is the field paying for its
parameters and still coming out ahead. Had the data been spatially flat,
the comparison would have reported equal or higher WAIC for the spatial
model, the field's honest signal that it was not needed. That symmetry is
the value of the comparison: it does not reward the field for flexibility
alone, only for flexibility the data actually use.

WAIC compares predictive accuracy; it does not certify that the model is
right. A lower WAIC says the spatial model predicts the in-sample cells
better, not that its credible intervals are calibrated or its
assumptions hold. Pair it with a residual check for leftover spatial
autocorrelation (a Moran's I test on the residuals) before trusting the
field. The diagnostics vignette walks through those checks.

## Practical guidance

A few rules of thumb for areal occupancy fields, with the numbers that
make them usable.

- **At least 8 to 10 regions, more is better.** The smoothing precision
  $\tau$ is learned from how much neighbouring cells differ, and a handful
  of units leaves it barely informed. Below roughly eight regions a fixed
  effect per region is usually the more honest choice, since there is
  little to gain from borrowing and the smoothing is mostly prior. The
  field repays its machinery when you have dozens to thousands of cells,
  many of them thinly surveyed.

- **Keep the joint grid small, watch the soft warning past about 50
  cells.** The nested path lays a grid over the smoothing hyperparameters
  and solves the field at each point. Stacking several latent terms (a
  spatial field plus a temporal block plus a random effect) multiplies
  those grids, and past roughly 50 joint cells the engine raises a soft
  warning that integration is getting expensive. One spatial field on a
  few hundred sites is comfortable; if you stack terms and trip the
  warning, simplify the latent structure or accept the slower fit.

- **Read slopes, not the intercept, under ICAR.** The intrinsic field
  confounds with the overall level: shift every $\phi_r$ up and the
  intercept down by the same amount and the likelihood does not move. The
  intercept's standard error runs to the prior scale as a result, far
  wider than any slope error. That gap is the signature of the intrinsic
  ridge, not a convergence problem. If you need a clean intercept, move to
  `bym2()`, whose IID component breaks the exact confounding.

- **ICAR versus BYM2 for occupancy.** Reach for `icar()` when you want the
  simplest smooth field and do not need to separate structured from
  unstructured variation; one fewer parameter, clean interpretation as
  pure smoothing. Reach for `bym2()` when some cell variation is plausibly
  independent noise rather than smooth signal, when you want the
  interpretable spatial fraction $\lambda$, or when the Riebler scaling
  matters for comparing field magnitude across maps with different graphs.
  The proper CAR (`car_proper()`), which estimates the autocorrelation
  $\rho$ directly, is wired for the cover-hurdle nested-Laplace path and
  the areal N-mixture rather than the occupancy engine, so it is not a
  choice for `occu()` on the current backend.

- **When areal is the wrong tool.** Three cases call for a different
  approach. If a non-spatial fit leaves no neighbour correlation in its
  residuals, a field adds parameters for nothing and WAIC will say so. If
  you have fewer than about eight regions, the smoothing parameter is too
  weakly informed to help. And if the sites are not truly areal, raw
  points on a map rather than regions joined by borders, a continuous
  field fits the geometry better than an adjacency you would have to
  invent by binning points into arbitrary cells. That last case is the
  `vignette("occupancy-spatial-spde")` story: a Matern field on
  coordinates, with a range parameter that says how far correlation
  reaches, rather than a graph that says who borders whom.

## Where to go next

- Continuous spatial fields on point coordinates (SPDE, Matern):
  `vignette("occupancy-spatial-spde")`.
- The engine internals, the integration grid, and the law-of-total-
  variance error correction: the `tulpa` package's spatial-models
  vignette.
- Areal spatial counts with the same `icar()` / `bym2()` / `car_proper()`
  terms on an N-mixture: `vignette("abundance")`.
- Posterior-predictive checks and spatial residual diagnostics (Moran's
  I, variograms): `vignette("diagnostics")`.
```
